---
title: TK
subtitle: TK
date: TK
author:
  - name: TK
    orcid: TK
    email: griffin@switch.box
    affiliations:
      - Switchbox

keywords: [line extension allowances, natural gas, future of gas, Illinois]

bibliography: ../references.bib
license: "CC BY-NC"

toc: true
notebook-links: false
reference-location: margin
fig-cap: true
fig-cap-location: margin
tbl-cap-location: margin

appendix-style: default
citation-location: document
citation:
  container-title: Switchbox
---

# Introduction

In order to connect new customers to the gas distribution system, gas utilities
subsidize the extension of their pipelines by raising rates for existing
customers. The size of these subsidies are set by regulators under rules called
**Line Extension Allowances** (LEAs).

But how much are Illinois ratepayers paying to subsidize the expansion of the
natural gas system under LEAs?
Could they be paying too much?
If so, what alternatives are available?

Regardless, any expansion of the gas system conflicts with Illinois's
decarbonization goals. In September 2021, Governor Pritzker signed into law the
**Climate and Equitable Jobs Act** (CEJA) -- one of the nation's most ambitious
state climate laws. CEJA commits Illinois to a 100% clean energy future by
2050. That's simply incompatible with a growing gas system: never mind the fact
that such growth risks increasing inequitable energy burdens and the risk of
stranded assets.

This memo presents the most complete data to date on LEA spending by Illinois's
three major gas utilities: Nicor Gas, Peoples Gas, and Ameren Illinois Gas.
While these companies either cannot or choose not to make the full
information on their LEA-subsidized spending available to the public, we
present the most comprehensive and complete data to date for each of them.

# Executive Summary

Illinois's three major gas utilities spent an estimated **$113.8 million** in
2022 subsidizing new gas connections under Line Extension Allowances.

- **Utilities do not consistently track or disclose LEA spending.** No utility
voluntarily published this data. Instead, we extracted it from discovery
responses in rate case proceedings before the Illinois Commerce Commission
(ICC). **No utility reports a single, standardized LEA spending figure**,
and some have explicitly objected to requests for LEA data.

- **Nicor Gas accounts for the majority of LEA spending.** Nicor's ratepayers
bore approximately **$65.7 million** in LEA costs in 2022, covering both main
extensions ($31.1 million) and service lines ($34.6 million).

- **Nicor's per-connection costs are rising rapidly.** Nicor's average LEA
subsidy per new main extension connection more than doubled from **$6,412 in 2019**
to **$15,202 in 2023**.

- **Peoples Gas ratepayers paid $26.7 million in 2022** to extend mains and
build service lines for new customers, at an average cost of **$11,784 per new single-family connection**.

- **Ameren's LEA spending is difficult to isolate but still risky for ratepayers.**
Ameren does not track line extension costs as a separate category. We use the
closest available proxy to show that **Ameren ratepayers paid an estimated $21.3 million in 2022**.
And that's in spite of a shrinking customer base. The company has lost
roughly 1,000–2,000 service connections per year since 2017 -- creating a real
stranded asset risk for Ameren's ratepayers.

- **The Illinois Commerce Commission can act to effectively eliminate gas &
service LEAs statewide.** The rules governing LEAs are regulatory, not
legislative. This means the ICC can act on its own initiative without the need
for legislation to directly address this issue.

- **Eight states have already acted.**
California, Colorado, Washington, Oregon, Minnesota, Maryland, Massachusetts,
and New York have taken substantive action to reduce or eliminate gas LEAs, with
estimated aggregate savings of **$750 million per year**.


# Background

## What are Line Extension Allowances?

Each year, Illinois gas utilities connect thousands of new customers to the
state's gas distribution system.

Between 2018 and 2025, these utilities built between 27,500 and 51,000 new
**service** connections per year, and hundreds of miles of new **mains**. While
some of these connections are replacements for existing pipes, we calculate that
**net service connections ranged between 3,000 and 6,000 per year**.
^["Main" refers to the main gas distribution line that runs through streets and
neighborhoods. "Service line" refers to the smaller pipe that branches off the
main to connect an individual building.]

Some of the customers serviced by these new connections are constructing new
buildings on existing sites already served by the gas network. Others are
newly-built or existing buildings that are being connected to the network for
the first time. In either case, the connections effectively grow the network's
pipeline network (and thus, its carbon footprint).

The new customers don't usually pay for this expansion work themselves. Instead,
most of the cost is covered by subsidies set by regulators through rules called
Line Extension Allowances (LEAs). Those subsidies are paid for by the existing
gas customers by raising their monthly bills.

Specifically, in Illinois, new customers are not required to pay for the first
100 to 400 feet of a main extension, or for the first 60 feet of a new service
line to connect their building to the main. The costs of that construction are
passed through to the existing ratepayers.
^[These thresholds are codified in Title 83, Part 501.600 of the Illinois
Administrative Code (for mains) as well as in individual ICC-approved tariff
schedules (for service lines). A tariff is a utility's ICC-approved schedule of
rates and service terms. Unlike main extensions, which are governed by a
statewide administrative rule, service line allowances are set utility-by-utility
in each company's tariff.]

In effect, LEAs hide the true cost of heating a building with natural gas
from the developers and building owners who benefit from the new connections.
Because existing customers pay for the connection, gas heating seems cheaper and
more attractive than it actually is. That creates a market distortion which
makes greener alternatives to gas heating (such as heat pumps) seem less
competitive.

The problem is statewide. While Illinois law only requires gas utilities to
provide LEA subsidies to urban customers, most utilities also offer them to
rural customers, and the ICC currently approves this practice.
^[Title 83, Part 501.600 of the Illinois Administrative Code mandates free main
extension footage for urban customers. The companion provision, Section 501.610,
governs rural extensions through a deposit-based mechanism rather than a free
footage mandate. However, Ameren Illinois (which services much of rural
Illinois) voluntarily provides 400 feet of free main to both urban and rural
new customers under an agreement with ICC Staff (ICC Docket No. 03-0767,
Appendix to Final Order, Jan. 25, 2006). The ICC currently approves this
practice.]

## How do LEAs affect current customers' bills?

When a utility company builds a new main extension or service line, the cost of
that infrastructure is added to the utility's **rate base**. The rate base is
the total value of the company's assets, which regulators use to set the amount
of profit that the utility can earn for its shareholders.

So when these companies build a new pipe under an LEA, existing customers pay
for the cost of the construction plus the utility's guaranteed profit margin on
that investment. And those payments are recouped over the entire life of the
pipe. Gas infrastructure remains in the company's rate base for its full
depreciation life (typically 40–60 years in Illinois). That means customers keep
paying for the new connection, plus company profits, for decades after the
infrastructure is built.

## Why are states revisiting LEA policies?

Historically, LEAs were established to grow the gas distribution system,
allowing utilities to capitalize on economies of scale. More customers meant
cheaper service for everyone. LEAs made it cheaper to connect new customers to
the gas network, and that made sense when customers could be assumed to stay in
the system indefinitely. A growing customer base also meant the costs associated
with LEAs would be recouped through increased revenue over time: making them
even more attractive.

But that's changing.

Today, rising gas bills and broader awareness of the environmental damage
associated with natural gas emissions are leading regulators to reevaluate the
role of LEAs.

As states progress towards their climate goals, they are increasingly looking to
electrify buildings and transportation systems in order to reduce emissions.
Continued expansion of the natural gas system not only runs counter to those
reduction targets, but also encourages additional investment in infrastructure
that may no longer be needed as alternatives to gas heating become more widely
available.

Unfortunately, the growing adoption of cost-effective electric alternatives to
gas can create **stranded assets**. These are pieces of gas infrastructure
(such as pipeline networks) that are rendered obsolete or uneconomic before the
end of their predicted useful life. So as some customers shift away from natural
gas, a shrinking pool of ratepayers would be responsible for paying off
investments in infrastructure they no longer use.

Worse still, low-income ratepayers are often those who are least able to
electrify their homes and businesses. Renters have little say in how their
landlord chooses to heat their home, and residents of much older buildings face
increased costs for retrofitting their homes with electric alternatives.

## Eight states lead the way on LEA reform

California, Colorado, Washington, Oregon, Minnesota, Maryland, Massachusetts,
and New York have all taken measures to reduce or completely end gas LEAs. Each
of the three states (California, Colorado, and New York) that eliminated LEAs
entirely have saved their ratepayers an estimated **$100–$200 million per year**.

Washington's and Oregon's utility commissions have required that their states'
largest gas utilities eliminate all LEAs, with some utilities meeting deadlines
as early as January 2025 and the remainder scheduled to follow by 2027.

It's true that reform efforts in Minnesota, Maryland, and Massachusetts have
been more modest. So far they've been limited to reducing the size of LEAs,
draft regulations and proposals, and studies of current LEA policies. But in all
three cases, state regulators have made it clear that they have serious concerns
about the climate effects and long-term viability of LEAs.

There's also early evidence that eliminating LEAs can shift the market away from
gas. In 2023, California's largest combined gas and electric utility reported that
**72 percent of new residential and 91 percent of new non-residential extension
requests** were for electric service only. Prior to the state's LEA reform,
**only 8 to 9 percent** of existing residential customers opted for electric
service only.

## What can the ICC do about LEAs in Illinois?

The ICC has already begun a major review of the state's LEA
policies in order to determine what role, if any, these subsidies should play in
Illinois's future energy system.

In November 2023, the ICC's Final Orders in the Nicor, Ameren, Peoples Gas, and
North Shore Gas rate cases directed the commencement of a Future of Gas
proceeding to "explore issues tied to decarbonization of the gas system." The
Commission formally launched that proceeding in March 2024 (Docket No. 24-0158).
LEAs and their alternatives were explicitly identified as issues of concern. ICC
staff, working in conjunction with an independent facilitator, are due to produce
a report on their findings along with recommendations for action in 2026.

If the ICC decided to act, it could do so swiftly, effectively, and
without legislative deliberation. Specifically, it could use two main
regulatory mechanisms to address LEAs:

**Main extensions** are governed by Title 83, Part 501.600 of the Illinois
Administrative Code. That regulation mandates that gas utilities provide free
main extension footage to new customers. The ICC can repeal this rule through a
**rulemaking proceeding** that doesn't require legislation.
^["Rulemaking proceeding" refers to the formal process by which the ICC
adopts, amends, or repeals administrative rules. This process involves public
notice, comment periods, and commission votes, but does not require action by
the General Assembly.]

**Service line extensions** are governed differently. Utilities recover service
line costs by including them in their rate base during **rate cases**. The ICC
can simply disallow this inclusion in any future rate cases.
^["Rate cases" are the periodic proceedings where utilities ask the ICC to let
them adjust the rates they charge customers.]

# Findings

Illinois's three major gas utilities -- Nicor Gas, Peoples Gas, and Ameren
Illinois Gas -- maintain and/or provide limited information on what they spend
on LEA projects. Nevertheless, we collected a mix of historical and forecasted
spending data from ICC rate cases. We combined these with our best estimates for
missing data in order to give as full a picture as possible of just how much
ratepayers are subsidizing new gas connections.

Understanding just how much money Illinois residents are spending on new gas
connections is essential. It's the first step in any honest evaluation of
whether or not gas heating is the best choice for Illinois.
^[For more details on the data presented, please see the data table in the Appendix.]

## How much are LEAs costing Illinois ratepayers?

In 2022, the last year for which data is available across all three utilities,
existing gas customers in Illinois paid an estimated total of
**$113.8 million** on LEA subsidies. That amount combines main extension and
service line spending, net of customer contributions, for Nicor Gas,
Peoples Gas, and Ameren Illinois Gas.

| Utility | Mains (net of customer contributions) | Service lines | Total |
|---|---|---|---|
| Nicor Gas | $31,080,414 | $34,624,181 | $65,704,595 |
| Peoples Gas | $6,401,127 | $20,348,036 | $26,749,163 |
| Ameren Illinois Gas | $21,349,683 (combined) | — | $21,349,683 |
| **Total** | | | **$113,803,441** |
^[Peoples Gas's rate cases do not indicate whether or not service line spending
is net of CIAC. The company explicitly lists that its main line figures are net
of these contributions in the same filing. Because Peoples does not provide
any figures for CIAC, we here assume that the same is true for service line
spending -- that is, it is also net of customer contributions.]

Nicor Gas accounts for more than half of total statewide LEA spending, with
Peoples Gas and Ameren Illinois Gas making up the remainder — a split visible
both by utility and by pipeline type (@fig-statewide-lea-2022).

:::{.column-page-inset-right}
{{< embed notebooks/analysis.qmd#fig-statewide-lea-2022 >}}
:::

Once more, this estimate is an approximation. In particular, we're limited by
the unavailability of two pieces of information.

1. Ameren does not report LEA spending as a separate line item. The
figure given is for infrastructure work which may include some costs beyond
traditional line extensions. Conversely, it may also undercount LEA activity
because that number combines main and service costs into a single category.

2. Peoples Gas's rate cases do not indicate whether or not service line
spending is net of **Customer Contributions in Aid of Construction** (CIAC).
These are payments new customers make when LEA subsidies will not fully
cover the cost of a new connection. The out-of-pocket payments offset
between 3 and 8 percent of gross LEA spending.

## How much does Nicor Gas spend on LEAs?

From 2018 to 2022, existing Nicor customers paid an average of **$23.2 million**
per year to extend gas mains, and **$25.3 million** per year to build service
lines for new customers (net of customer contributions).

And Nicor's customer base is only getting bigger. US Department of
Transportation data confirms Nicor is the only major Illinois gas utility which
is consistently growing, adding roughly 5,000 to 7,000 net new service
connections per year (from approximately 2.03 million in 2017 to 2.07 million in
2025). A bigger customer base means more new connections, and thus even bigger
LEA spending.

Both main extension and service line spending have grown in step with Nicor's
expanding customer base, and the utility's own forecasts show that trajectory
continuing through 2025 (@fig-nicor-lea-trend).

:::{.column-page-inset-right}
{{< embed notebooks/analysis.qmd#fig-nicor-lea-trend >}}
:::

### What does each Nicor connection cost?

Nicor's language in its rate case filings make it unclear how much it costs to
connect each new customer to the gas network. What we can say is this:

- In 2023, Nicor spent an average of **$15,202** in LEA subsidies per new main
extension connection. The filings (and Nicor) do not clarify whether a
"connection" refers to a building, a dwelling unit, or a gas meter. For
instance, in a multi-family building, it is unclear whether each unit counts as
a separate connection. We use the term as reported.

- The average LEA subsidy per service line "project" was **$3,716** in 2022.
Nicor does not actually report a cost per service line connection, so we
calculated the figure based on total service line extension cost divided by the
number of service line projects. Again, we interpret one project as one
connection, as Nicor does not specifically define the term.

These costs are growing. The average cost per connection for a main extension
more than **doubled from $6,412 in 2019 to the $15,202 it cost in 2023**. Main
line extensions also got more expensive per foot, rising more than 100%, from
**$32.06** to **$76.01**, over the same period.

The chart below traces this escalation for both main extensions and service
lines, showing how rapidly the cost of each new Nicor connection has grown
since 2019 (@fig-nicor-per-connection-cost).

:::{.column-page-inset-right}
{{< embed notebooks/analysis.qmd#fig-nicor-per-connection-cost >}}
:::

## How much does Peoples Gas spend on LEAs?

From 2018 to 2022, Peoples Gas customers paid an average of **$5.6 million**
per year to extend gas mains and **$15.5 million** per year to build service
lines for new customers. That's **$21.1 million** per year.

And while Peoples forecasted that it would decrease its spending on main
extensions from $5.2 million in 2023 to $4.0 million in 2024, it simultaneously
forecasted that it would increase spending on service line extensions from $17.2
million in 2023 to $18.5 million in 2024.

Peoples Gas's LEA spending on mains and service lines from 2018 through 2024,
with forecasted figures for 2023–2024, shows service line spending sustained at
over $15 million per year while main extension spending declines (@fig-peoples-lea-trend).

:::{.column-page-inset-right}
{{< embed notebooks/analysis.qmd#fig-peoples-lea-trend >}}
:::

### What does each Peoples Gas connection cost?

Peoples Gas also does not report a cost per connection for either main or
service line extensions. However, information in its rate cases before the ICC
tells us that in 2022, Peoples Gas customers paid an average of **$11,784** per
new single-family connection to both extend gas mains and build new service
lines. This figure represents the average subsidized cost of connecting a single
new residential single-family home to the distribution system, whether that's
done by a main or a service line.
^[Multi-family connection costs were not provided by the utility.]

The chart below shows the combined per-connection cost — covering both main
extensions and service lines — for new single-family residential connections to
Peoples Gas's system between 2020 and 2022 (@fig-peoples-per-connection-cost).

:::{.column-page-inset-right}
{{< embed notebooks/analysis.qmd#fig-peoples-per-connection-cost >}}
:::

## How much does Ameren Gas spend on LEAs?

Ameren has denied requests to report directly on LEA spending. As an
approximation, we have included information on a distribution
**standing work order** (SWO). The SWO is an internal standard
that describes the costs of gas mains, service lines, fittings, and other
materials for new service requests.

Using SWO data from 2021, we estimate that Ameren spent **$20.1 million** using
LEA subsidies that year. In 2022, that spending increased to **$21.3 million**.

This spending is particularly notable because unlike Nicor and Peoples Gas,
Ameren's customer base is shrinking. US Department of Transportation data shows
Ameren has lost approximately 1,000 to 2,000 service connections per year
consistently since 2017. The total number of connections has declined from
roughly 819,000 in 2017 to 809,000 in 2025.

In other words, Ameren's ratepayers are likely funding more than **$20 million**
annually of new infrastructure. That's for a pipeline network that serves fewer
and fewer customers each year. Fewer customers paying for more spending is a
clear and concrete instance of a stranded asset risk.

Ameren's annual LEA spending rose steadily from 2021 through 2024
(@fig-ameren-spending-trend), even as its total service connections declined.

:::{.column-page-inset-right}
{{< embed notebooks/analysis.qmd#fig-ameren-spending-trend >}}
:::

### What does each Ameren connection cost?

Ameren reports an **average standard cost equivalent** per new connection.
These figures cover main extensions only. Because they do not include service
line extensions they may also underestimate the true cost of connecting a new
customer to Ameren's network.

Our best estimate is that Ameren's per connection cost rose from
**$5,539** in 2021 to **$7,331** in 2025.

## Why is LEA spending data so difficult to obtain?

No utility voluntarily published any of the data in this memo. Instead,
we extracted it from ICC rate case filings. Specifically, we looked at discovery
responses that utilities provided to **Public Interest Organizations** (PIOs). A
PIO is a party that represents the public interest in an ICC rate case. They
may be consumer advocates, environmental groups, or other similar organizations.
PIOs have formal discovery rights, which means they can compel utilities to
produce data that isn't otherwise public.

Regrettably, even formal discovery procedures aren't sufficient to give the
public a complete picture of LEA spending. Utilities can object to requests for
data on the grounds that isolating the information would require too
much work. The legal term is "an objection on grounds of undue burden." As a
result, no utility reports a standardized LEA spending figure. Nicor provided
the most detailed data, including per-connection costs, but no other utility
replicated that level of disclosure. Peoples Gas does not track service line
extension costs as a separate category, while Ameren has denied requests to
report directly on LEA spending.

## The Bottom Line

In 2022, Illinois ratepayers spent roughly **$114 million** on extending gas
pipelines.

Did that money improve the quality and reliability of services they
need?

No. It went towards enticing new customers into choosing gas heating, even
though the state has committed to a 100% clean energy future in less than 30
years.

Was that money a safe investment in Illinois's energy future?

No. It built polluting fossil fuel infrastructure that stand directly in the
way of the state's clean energy goals.

Was that money responsibily tracked and the records made available to the
gas utilities ratepayers?

No. The companies made it harder for Illinois ratepayers to see the true costs
of gas heat and made greener alternatives seem more expensive by comparison.


# Appendix

## Acknowledgments

## Data and Methods

### Data collected from rate case filings

LEA spending figures in this memo come from three Illinois Commerce
Commission dockets. No utility voluntarily publishes this data.

**Nicor Gas** (Docket 23-0066, PIO Exhibit 12):

- *Main extension spending*: Reported as `extension_cost` by year
  (2018–2022 historical, 2023–2026 forecast), net of Contributions in Aid
  of Construction (CIAC) which Nicor reports separately as
  `customer_contribution`. Figures in this memo reflect the net amount
  (gross minus CIAC).
- *Service line spending*: Same structure as main extensions, with
  separate CIAC figures for service lines.
- *Per-connection main extension cost*: Reported directly by Nicor as
  `lea_cost_per_customer` (2019–2023). Nicor is the only utility in our
  sample that reports a standardized per-connection figure.
- *Service line per-project cost*: Not directly reported. We calculate it
  as total service line extension cost (net of CIAC) divided by the number
  of service line projects (`n_projects`). We interpret one project as
  one connection.
- *Per-foot main extension cost*: Reported directly by Nicor for
  high-pressure mains (2019–2023).

**Peoples Gas** (Docket 23-0068, PIO Exhibit 12; Docket 23-0069, PIO
Cross Exhibit 10.0):

- *Main extension spending*: Reported as `lea_cost` by year (2018–2022
  historical, 2023–2024 forecast). Peoples' filing explicitly states main
  figures are "net of customer contributions."
- *Service line spending*: Reported as `lea_cost` by year with the same
  structure. Peoples does not separately disclose CIAC amounts for service
  lines. We therefore assume these figures are also net of customer
  contributions, consistent with the treatment of main extensions in the
  same filing.
- *Per-connection combined cost*: Reported as `cost_per_new_service`
  (2020–2022), covering single-family residential connections only.
  Multi-family connection costs were not disclosed.
- *Service line project counts*: Reported for 2018–2022. Peoples does not
  forecast service line project counts, describing this work as
  customer-driven.

**Ameren Illinois Gas** (Docket 23-0067, PIO Exhibit 12; Docket 25-0084,
PIO Exhibit 1.2):

- *LEA spending*: Ameren does not track LEA spending as a separate
  category. We use the distribution standing work order (SWO, internal
  code C3231) as the closest available proxy. Per Ameren's filing, C3231
  covers "gas mains, service lines, fittings and other material associated
  with new service requests." Figures are reported as
  `distribution_extensions_total` (gross) and
  `distribution_extensions_customer_contributions` (CIAC). This memo uses
  net figures. A separate transmission SWO (C3160) is excluded as it
  covers large-scale equipment rather than new-customer connections.
- *Per-connection cost*: Reported as the standard cost equivalent for 400
  feet of main extension (`max_lea_cost_per_connect`, 2021–2025). This
  covers main extensions only. Service line costs are not reported
  separately.

### Data availability

The data available for each utility varies substantially in both
completeness and comparability. Nicor Gas provides the most detailed
disclosures, including per-connection costs for main extensions and
separate CIAC figures for both mains and service lines. Peoples Gas
provides total spending figures but does not isolate service line CIAC
amounts, and does not forecast service line project counts. Ameren
Illinois does not track LEA spending as a distinct category at all. The
SWO C3231 figures used in this memo are the closest available proxy.

This inconsistency is itself a finding. Illinois ratepayers cannot make
informed decisions about gas system expansion when the three utilities
serving the state use three different accounting methods, and when one of
them does not separately track LEA spending at all.

```{r}
#| label: setup-gt-table
#| include: false

suppressPackageStartupMessages({
  library(gt)
  library(tidyverse)
})

load("report_vars.RData")

# Post-load override: recolor D7 (Peoples Gas forecast, Services Total)
# from red to yellow, and update displayed value from "—" to "2023-2024".
# Evidence: Peoples Gas discovery response in ICC Docket 23-0068 reports
# estimated 2023 service line expenditures ($17.2M) and 2024 service line
# budget ($18.5M) net of customer contributions, though Peoples explicitly
# states it does not maintain forecasts for new service work.
gt_table$`_data`[6, "Services Total"] <- "2023-2024"
gt_table <- gt_table |>
  tab_style(
    style = cell_fill(color = "#fff3cd"),
    locations = cells_body(columns = `Services Total`, rows = 6)
  )
```

```{r}
#| label: tbl-data-availability
#| tbl-cap: "Data availability by utility, metric, and period. Green indicates complete data; yellow indicates partial or caveated data; red indicates data not available or not provided. Peoples Gas forecast service line spending (2023–2024) is reported net of customer contributions in ICC Docket 23-0068, though the utility does not maintain forecasts for this category."
#| echo: false

gt_table
```

### Estimated data

The following memo figures involve calculation or estimation beyond the
raw values in utility filings:

- **Nicor service line per-project cost ($3,716 in 2022)**. Calculated as
  net service line spending ($34,624,181) divided by the number of
  service line projects (9,318). Nicor does not directly report a
  cost-per-project figure for service lines.
- **Nicor main extension spending, net of CIAC**. Calculated as
  `extension_cost` + `customer_contribution`, where CIAC is stored as a
  negative value. Gross and net figures differ by approximately 3–8
  percent in each year.
- **Peoples Gas service line spending, CIAC treatment**. Peoples does not
  explicitly disclose whether its reported service line spending is gross
  or net of customer contributions. The same filing states that Peoples'
  main extension figures are net of customer contributions. We assume the
  same treatment applies to service lines. If service line figures are
  instead gross, the actual ratepayer burden would be approximately 3–8
  percent lower.
- **Ameren LEA spending**. Estimated using the distribution SWO (C3231)
  net of CIAC. The SWO covers gas mains, service lines, fittings, and
  other materials associated with new service requests. This proxy may
  overstate LEA spending to the extent that C3231 includes costs beyond
  traditional line extensions. Conversely, it may understate LEA spending
  to the extent that relevant costs are charged to other work orders.
  Because Ameren does not track LEA spending as a separate category, no
  direct comparison is possible.
- **Statewide 2022 total ($113.8 million)**. Calculated as the sum of
  each utility's 2022 net LEA spending: Nicor ($65.7M) + Peoples ($26.7M)
  + Ameren ($21.3M). All three figures are net of customer contributions.
- **PHMSA-derived connection counts**. Statewide net growth (3,000–6,000
  new connections per year), gross new installations (27,500–51,000 per
  year), and per-utility net change figures (Ameren: -1,000 to -2,000 per
  year & Nicor: +5,000 to +7,000 per year) are calculated from U.S.
  Department of Transportation Pipeline and Hazardous Materials Safety
  Administration Gas Distribution Annual Reports (Form F7100.1-1),
  2017–2025. PHMSA data tracks all physical service line installations
  regardless of funding mechanism. It does not distinguish LEA-subsidized
  connections from customer-paid connections or from replacements of
  existing service lines. Gross installation counts from PHMSA are
  therefore not directly comparable to LEA-funded project counts from
  utility rate case filings.

## References

::: {#refs}
::
