---
title: "Line Extension Allowances in Illinois"
subtitle: "How much are Illinois ratepayers paying to connect new customers to the gas network?"
date: 2026-04-22
author:
  - name: Alexandra Smith
    email: alex@switch.box
    affiliations:
      - Switchbox
  - name: Juan-Pablo Velez 
    email: jpv@switch.box
    affiliations:
      - Switchbox
  - name: Griffin Sharps
    email: griffin@switch.box
    affiliations:
      - Switchbox
keywords: line extension allowances, natural gas, future of gas, Illinois
bibliography: ../references.bib
license: "CC BY-NC"
toc: true
notebook-links: false
reference-location: margin
fig-cap: true
fig-cap-location: margin
tbl-cap-location: margin
appendix-style: default
citation-location: document
citation:
  container-title: Switchbox

# Uncomment this when PDF version has been added to local docs/
#other-links:
#  - text: PDF
#    icon: file-earmark-pdf
#    href: docs/switchbox_report.pdf
---

# Introduction

In order to connect new customers to the gas distribution system, gas utilities
subsidize the extension of their pipelines by raising rates for existing
customers. The size of these subsidies are set by regulators under rules called
**Line Extension Allowances** (LEAs).

But how much are Illinois ratepayers paying to subsidize the expansion of the
natural gas system under LEAs?
Could they be paying too much?
If so, what alternatives are available?

Any expansion of the gas system conflicts with Illinois's decarbonization goals. 
In September 2021, Governor Pritzker signed into law the 
**Climate and Equitable Jobs Act** (CEJA) -- one of the nation's most ambitious
state climate laws. CEJA commits Illinois to a 100% clean energy future by
2050. These emission targets are incompatible with a growing gas system, which
itself risks increasing inequitable energy burdens.

This memo presents the most complete data to date on how many ratepayer dollars 
Illinois’ three major utilities (Nicor, Peoples Gas, and Ameren) are spending on 
these gas system expansion subsidies. We present the most complete view of the 
data available, which is incomplete because these companies either cannot or 
choose not to make the full information on their LEA-subsidized spending 
available to the public.

# Executive Summary

Illinois's three major gas utilities spent an estimated **$113.8 million** in
2022 subsidizing new gas connections under Line Extension Allowances.

- **Nicor’s average subsidy per new main extension was $15,202 in 2023**. During 
the same period, 
**Peoples Gas ratepayers paid $11,784 per new single-family connection**.

- In 2022, **Nicor’s ratepayers bore approximately $65.7 million** in LEA costs. 
**Peoples Gas ratepayers paid $26.7 million** and we estimate that 
**Ameren ratepayers paid $21.3 million**.

- **Eight states have already acted.**
California, Colorado, Washington, Oregon, Minnesota, Maryland, Massachusetts,
and New York have taken substantive action to reduce or eliminate gas LEAs, with
estimated aggregate savings of **$750 million per year**.

- **The Illinois Commerce Commission can act to effectively eliminate gas &
service LEAs statewide.** The rules governing LEAs are regulatory, not
legislative. This means the ICC can act on its own initiative without the need
for legislation to directly address this issue.

- **Utilities do not consistently track or disclose LEA spending.** No utility
voluntarily published this data. Instead, we extracted it from discovery
responses in rate case proceedings before the Illinois Commerce Commission
(ICC). **No utility reports a single, standardized LEA spending figure**,
and some have explicitly objected to requests for LEA data.

# Background

## What are Line Extension Allowances?

Each year, Illinois gas utilities connect thousands of new customers to the
state's gas distribution system.

Between 2018 and 2025, these utilities built hundreds of miles of **mains** (the 
gas distribution lines that run through streets and neighborhoods). They also 
built between 27,500 and 51,000 new **service** lines (the smaller pipes that 
branch off the main to connect an individual building) each year. While some of 
these connections are replacements for existing pipes, we calculate that net 
service connections still ranged between **3,000 and 6,000 per year**.

Some of the customers serviced by these new connections are constructing new 
buildings on existing sites already served by the gas network. Others are 
newly-built or existing buildings that are being connected to the network for 
the first time. In either case, the connections effectively grow the utilities’ 
pipeline network (and thus, their emissions).

The customers receiving service lines don’t usually pay for this expansion work 
themselves. Instead, most of the cost is covered by subsidies set by regulators 
through rules called Line Extension Allowances (LEAs). Those subsidies are paid 
for by existing gas customers through increases to their monthly bills.

Specifically, in Illinois, new customers are not required to pay for the first 
100 to 400 feet of a main extension, or for the first 60 feet of a new service 
line to connect their building to the main. The costs of that construction are 
passed through to the existing ratepayers.
^[These thresholds are codified in [Title 83, Part 501.600](https://www.ilga.gov/agencies/JCAR/EntirePart?titlepart=08300501) of the Illinois
Administrative Code (for mains) as well as in individual ICC-approved tariff
schedules (for service lines). A tariff is a utility's ICC-approved schedule of
rates and service terms. Unlike main extensions, which are governed by a
statewide administrative rule, service line allowances are set utility-by-utility
in each company's tariff.]

In effect, LEAs hide the true cost of heating a building with natural gas from 
the developers and building owners who benefit from the new connections. Because 
existing customers pay for the connection, gas heating seems cheaper and more 
attractive than it actually is. That creates a market distortion which makes 
cleaner alternatives to gas heating (such as heat pumps) seem less competitive.
^[Bill SB2269, introduced in the Illinois General Assembly in 2025, would have 
required gas utilities to charge customers for the full incremental cost of new 
development and growth their connection creates. § 9-228.5 states: "Gas main and 
gas service extension policies shall be based on the principle that the full 
incremental cost associated with new development and growth shall be borne by 
the customers that cause those incremental costs." 
See [§ 9-228.5](https://www.documentcloud.org/documents/26212423-il-sb2269-villanueva/#document/p29/a2676215). [@villanueva_SB2269_2025]]

The problem is statewide. While Illinois law only requires gas utilities to
provide LEA subsidies to urban customers, most utilities also offer them to
rural customers, and the ICC currently approves this practice.
^[Title 83, Part 501.600 of the Illinois Administrative Code mandates free main
extension footage for urban customers. The companion provision, Section 501.610,
governs rural extensions through a deposit-based mechanism rather than a free
footage mandate. However, Ameren Illinois (which services much of rural
Illinois) voluntarily provides 400 feet of free main to both urban and rural
new customers under an agreement with ICC Staff (ICC Docket No. 03-0767,
Appendix to Final Order, Jan. 25, 2006). The ICC currently approves this
practice. [@iccTitle83Part2006]]

## How do LEAs affect current customers' bills?

When a utility company builds a new main extension or service line, the cost of 
that infrastructure is added to the utility’s rate base. The rate base is the 
total value of the company’s assets, which regulators use to set the amount of 
profit that the utility can earn for its shareholders.

So when these companies build a new pipe under an LEA, existing customers pay 
for the cost of the construction plus the utility’s guaranteed profit margin on 
that investment. And those payments are recouped from gas customers, through 
their bills, over the entire life of the pipe. Gas infrastructure remains in the 
company’s rate base for its full depreciation life (typically 40–60 years in 
Illinois). That means customers keep paying for the new connection, plus company 
profits, for decades after the infrastructure is built.

## Why are states revisiting LEA policies?

Historically, LEAs were established to grow the gas distribution system, 
allowing utilities to capitalize on economies of scale. More customers meant 
cheaper service for everyone. LEAs made it cheaper to connect new customers to 
the gas network, and that made sense when customers could be assumed to stay in 
the system indefinitely. A growing customer base also meant the costs associated 
with LEAs would be recouped through increased revenue over time, making them 
even more attractive.[^vu_historical]

[^vu_historical]: See Vu and Bagdanov, "The End of Gas System Subsidies," 
Building Decarbonization Coalition, August 2025. [@vu_EndGasSystem_2025]

But that's changing.

Today, rising gas bills and broader awareness of the environmental damage 
associated with natural gas emissions, from both leaks and combustion, are 
leading regulators to reevaluate the role of LEAs.

As states progress towards their climate goals, they are increasingly looking to 
electrify buildings and transportation systems in order to reduce emissions. 
Continued expansion of the natural gas system not only runs counter to those 
reduction targets, but also encourages additional investment in infrastructure 
that may no longer be needed as alternatives to gas heating become more widely 
available.

Unfortunately, the growing adoption of cost-effective electric alternatives to 
gas can create pieces of gas infrastructure (such as pipeline networks) that are 
rendered obsolete or uneconomic before the end of their predicted useful life. 
So as some customers shift away from natural gas, a shrinking pool of ratepayers 
would be responsible for paying off investments in infrastructure they no longer 
use.

Worse still, low-income ratepayers are often those who are least able to 
electrify their homes and businesses. Renters have little say in how their 
landlord chooses to heat their home, and residents of much older buildings face 
increased costs for retrofitting their homes with electric alternatives.

## Eight states lead the way on LEA reform

California, Colorado, Washington, Oregon, Minnesota, Maryland, Massachusetts,
and New York have all taken measures to reduce or completely end gas LEAs. Each
of the three states (California, Colorado, and New York) that eliminated LEAs
entirely have saved their ratepayers an estimated **$100–$200 million per year**.[^rmi_savings]

[^rmi_savings]: Rocky Mountain Institute (RMI) estimates that each of California, Colorado, and New York saves ratepayers $100–$200 million per year following LEA elimination. [@henchen_CuttingTheseSubsidies_2025a]

Washington’s and Oregon’s utility commissions have required that their states’ 
largest gas utilities eliminate all LEAs, with some utilities meeting deadlines 
as early as January 2025 and the remainder scheduled to follow by 2027.[^wa_or_deadlines]

[^wa_or_deadlines]: Washington: Avista and Puget Sound Energy eliminated gas LEAs effective January 2025; Cascade Natural Gas by March 2027. Oregon: Avista and NW Natural required to phase out LEAs by 2027. [@vu_EndGasSystem_2025]

Reform efforts in Minnesota, Maryland, and Massachusetts have been more modest. 
So far they’ve been limited to reducing the size of LEAs, draft regulations and 
proposals, and studies of current LEA policies. But in all three cases, state 
regulators have made it clear that they have serious concerns about the climate 
effects and long-term viability of LEAs.[^mn_md_ma]

[^mn_md_ma]: Minnesota's two largest gas utilities reduced per-connection allowances through rate case settlements in 2022 and 2023. Maryland PSC issued Order No. 91683 in June 2025 directing staff to draft LEA reform regulations. Massachusetts DPU issued an order in August 2025 signaling intent to eliminate allowances. [@vu_EndGasSystem_2025]

There’s also early evidence that eliminating LEAs can shift the market away from 
gas. In 2023, California’s largest combined gas and electric utility reported 
that **72 percent of new residential and 91 percent of new non-residential 
extension requests** were for electric service only. Prior to the state’s LEA 
reform, **only 8 to 9** percent of existing residential customers opted for 
electric service only.[^pge_data]

[^pge_data]: PG&E data reported in [@henchen_CuttingTheseSubsidies_2025a]. The 72%/91% figures reflect 2023 new extension requests following California's LEA reform. The 8–9% baseline reflects the share of existing residential customers on all-electric service prior to reform.

## What can the ICC do about LEAs in Illinois?

The ICC has already begun a major review of the state's LEA policies in order to 
determine what role, if any, these subsidies should play in Illinois's future 
energy system.

In November 2023, the ICC’s Final Orders in the Nicor, Ameren, and Peoples Gas 
rate cases directed the commencement of a Future of Gas proceeding to “explore 
issues tied to decarbonization of the gas system.” The Commission formally 
launched that proceeding in March 2024 (Docket No. 24-0158). LEAs and their 
alternatives were explicitly identified as issues of concern.[^icc_fog]

[^icc_fog]: Illinois Commerce Commission, Future of Gas Proceeding, Docket No. 24-0158, launched March 2024. [@iccFutureGasProceeding2024]

ICC staff, working in conjunction with an independent facilitator, are due to 
produce a report on their findings along with recommendations for action in 
2026. If the ICC decided to act, it could do so swiftly, effectively, and 
without legislative deliberation. Specifically, it could use two main regulatory 
mechanisms to address LEAs:

**Main extensions** are governed by Title 83, Part 501.600 of the Illinois
Administrative Code. That regulation mandates that gas utilities provide free
main extension footage to new customers. The ICC can repeal this rule through a
**rulemaking proceeding** that doesn't require legislation.
^["Rulemaking proceeding" refers to the formal process by which the ICC
adopts, amends, or repeals administrative rules. This process involves public
notice, comment periods, and commission votes, but does not require action by
the General Assembly.]

**Service line extensions** are governed differently. Utilities recover service
line costs by including them in their rate base during **rate cases**. The ICC
can simply disallow this inclusion in any future rate cases.
^["Rate cases" are the periodic proceedings where utilities ask the ICC to let
them adjust the rates they charge customers.]

# Findings

Illinois's three major gas utilities -- Nicor Gas, Peoples Gas, and Ameren
Illinois Gas -- maintain and/or provide limited information on what they spend
on LEA projects. Nevertheless, we collected a mix of historical and forecasted
spending data from ICC rate cases — specifically, Docket 23-0066 (Nicor Gas), 
Docket 23-0068 and 23-0069 (Peoples Gas), and Docket 23-0067 and 25-0084 
(Ameren Illinois Gas).[^findings_dockets]

We combined these with our best estimates for
missing data in order to give as full a picture as possible of just how much
ratepayers are subsidizing new gas connections.

[^findings_dockets]: See [ICC Docket 23-0066](https://www.documentcloud.org/documents/26188653-nicor-p2023-0066-pio-exhibit-12-30205654/) (Nicor Gas PIO Exhibit 12), [ICC Docket 23-0068](https://www.documentcloud.org/documents/26189740-peoplesgas-p2023-0068-pio-exhibit-12-64269107/) and [23-0069](https://www.documentcloud.org/documents/26189763-peoples-gas-p2023-0069-pio-cross-exhibit-100-95312469/) (Peoples Gas), and [ICC Docket 23-0067](https://www.documentcloud.org/documents/26188848-ameren-p2023-0067-pio-exhibit-12-96925852/) and [25-0084](https://www.documentcloud.org/documents/26193410-ameren-p2025-0084-pio-exhibit-12-57399815/) (Ameren Illinois Gas). No utility voluntarily publishes this data. [@northernillinoisgascompany_Docket230066Nicor_2023; @peoplesgas_Docket230068Peoples_2023; @kilhoffer_Docket230067Ameren_2023]

## How much are LEAs costing Illinois ratepayers?

In 2022, the last year for which data is available across all three utilities,
existing gas customers in Illinois paid an estimated total of
**$113.8 million** on LEA subsidies.[^service_costs]

| Utility | Mains (net of customer contributions) | Service lines | Total |
|---|---|---|---|
| Nicor Gas | $31,080,414 | $34,624,181 | $65,704,595 |
| Peoples Gas | $6,401,127 | $20,348,036 | $26,749,163 |
| Ameren Illinois Gas | $21,349,683 (combined) | — | $21,349,683 |
| **Total** | | | **$113,803,441**

That amount combines main extension and service line spending, net of customer 
contributions, for Nicor Gas, Peoples Gas, and Ameren Illinois Gas.

Nicor Gas accounts for more than half of total statewide LEA spending, with
Peoples Gas and Ameren Illinois Gas making up the remainder — a split visible
both by utility and by pipeline type (@fig-statewide-lea-2022).


:::{.column-page-inset-right}
{{< embed notebooks/analysis.qmd#fig-statewide-lea-2022 >}}
:::

[^service_costs]: Nicor main and service costs: ICC Docket 23-0066, PIO Exhibit 12, [p. 33](https://www.documentcloud.org/documents/26188653-nicor-p2023-0066-pio-exhibit-12-30205654/#document/p33/a2675139) (main gross), [p. 33](https://www.documentcloud.org/documents/26188653-nicor-p2023-0066-pio-exhibit-12-30205654/#document/p33/a2675140) (service gross), [p. 34](https://www.documentcloud.org/documents/26188653-nicor-p2023-0066-pio-exhibit-12-30205654/#document/p34/a2675143) (CIAC). Peoples Gas main and service costs: ICC Docket 23-0068, PIO Exhibit 12, [p. 36](https://www.documentcloud.org/documents/26189740-peoplesgas-p2023-0068-pio-exhibit-12-64269107/#document/p36/a2674688) (mains), [p. 36](https://www.documentcloud.org/documents/26189740-peoplesgas-p2023-0068-pio-exhibit-12-64269107/#document/p36/a2674690) (service). Peoples main figures explicitly net of customer contributions; same treatment assumed for service lines as no separate CIAC is disclosed. Ameren SWO C3231: ICC Docket 23-0067, PIO Exhibit 12, [p. 13](https://www.documentcloud.org/documents/26188848-ameren-p2023-0067-pio-exhibit-12-96925852/#document/p13/a2674478). [@northernillinoisgascompany_Docket230066Nicor_2023; @peoplesgas_Docket230068Peoples_2023; @kilhoffer_Docket230067Ameren_2023]


Once more, this estimate is an approximation. In particular, we're limited by
the unavailability of two pieces of information.[^data_caveats]

[^data_caveats]: In brief: Ameren does not report LEA spending as a separate line item, so its figure may include costs beyond traditional line extensions. Peoples Gas's rate cases do not indicate whether service line spending is net of customer contributions (CIAC); we assume the same net treatment as its main extension figures.

## How much does Nicor Gas spend on LEAs?

From 2018 to 2022, existing Nicor customers paid an average of **$23.2 million**
per year to extend gas mains, and **$25.3 million** per year to build service
lines for new customers (net of customer contributions).[^nicor_historical]

[^nicor_historical]: See ICC Docket 23-0066, PIO Exhibit 12. [@northernillinoisgascompany_Docket230066Nicor_2023]

And Nicor’s customer base is only getting bigger. US Department of 
Transportation data confirms Nicor is the only major Illinois gas utility which 
is consistently growing, adding roughly 5,000 to 7,000 net new service 
connections per year (from approximately 2.03 million in 2017 to 2.07 million in 
2025). A bigger customer base means more new connections, and thus even bigger 
LEA spending.[^phmsa_nicor]

[^phmsa_nicor]: U.S. Department of Transportation, Pipeline and Hazardous Materials Safety Administration, Gas Distribution Annual Report (Form F7100.1-1), 2017–2025. [@phmsa_GasDistributionAnnual_2025]

Both main extension and service line spending have grown in step with Nicor's
expanding customer base, and the utility's own forecasts show that trajectory
continuing through 2029 (@fig-nicor-lea-trend).^[Forecast data for 2023–2026 comes from ICC Docket 23-0066 PIO Exhibit 12. Forecast data for 2026–2029 (plotted as combined, without service/main breakdown) comes from ICC Docket 25-0055 PIO Cross Exhibit 2.0, [p. 46](https://www.documentcloud.org/documents/26193423-nicor_p2025-0055-pio-cross-exhibit-20-51703648/#document/p46/a2675138). 2026 appears in both filings; the chart stacks the 23-0066 service/main breakdown and shows the 25-0055 delta as combined.]

:::{.column-page-inset-right}
{{< embed notebooks/analysis.qmd#fig-nicor-lea-trend >}}
:::

### What does each Nicor connection cost?

Nicor's language in its rate case filings make it unclear how much it costs to
connect each new customer to the gas network. What we can say is this:

- In 2023, Nicor spent an average of **$15,202** in LEA subsidies per new main
extension connection.[^nicor_15202] The filings (and Nicor) do not clarify 
whether a "connection" refers to a building, a dwelling unit, or a gas meter. For
instance, in a multi-family building, it is unclear whether each unit counts as
a separate connection. We use the term as reported.

[^nicor_15202]: Reported as `lea_cost_per_customer` for high-pressure mains. ICC Docket 23-0066, PIO Exhibit 12, [p. 31](https://www.documentcloud.org/documents/26188653-nicor-p2023-0066-pio-exhibit-12-30205654/#document/p31/a2674817). [@northernillinoisgascompany_Docket230066Nicor_2023]

- The average LEA subsidy per service line "project" was **$3,716** in 2022.[^nicor_3716]
Nicor does not actually report a cost per service line connection, so we
calculated the figure based on total service line extension cost divided by the
number of service line projects. Again, we interpret one project as one
connection, as Nicor does not specifically define the term.

[^nicor_3716]: Calculated as net service line spending ($34,624,181) ÷ number of service line projects (9,318). ICC Docket 23-0066, PIO Exhibit 12, [p. 33](https://www.documentcloud.org/documents/26188653-nicor-p2023-0066-pio-exhibit-12-30205654/#document/p33/a2675140) (costs), [p. 35](https://www.documentcloud.org/documents/26188653-nicor-p2023-0066-pio-exhibit-12-30205654/#document/p35/a2675145) (project counts). [@northernillinoisgascompany_Docket230066Nicor_2023]

These costs are growing. The average cost per connection for a main extension
more than **doubled from $6,412 in 2019 to the $15,202 it cost in 2023**. Main
line extensions also got more expensive per foot, rising more than 100%, from
**$32.06** to **$76.01**, over the same period.[^nicor_perfoot]

[^nicor_perfoot]: Per-foot costs for high-pressure mains. ICC Docket 23-0066, PIO Exhibit 12, [p. 31](https://www.documentcloud.org/documents/26188653-nicor-p2023-0066-pio-exhibit-12-30205654/#document/p31/a2674817). [@northernillinoisgascompany_Docket230066Nicor_2023]

The chart below traces this escalation for both main extensions and service
lines, showing how rapidly the cost of each new Nicor connection has grown
since 2019 (@fig-nicor-per-connection-cost).

:::{.column-page-inset-right}
{{< embed notebooks/analysis.qmd#fig-nicor-per-connection-cost >}}
:::

## How much does Peoples Gas spend on LEAs?

From 2018 to 2022, Peoples Gas customers paid an average of **$5.6 million**
per year to extend gas mains and **$15.5 million** per year to build service
lines for new customers. That's **$21.1 million** per year.

And while Peoples forecasted that it would decrease its spending on main
extensions from $5.2 million in 2023 to $4.0 million in 2024, it simultaneously
forecasted that it would increase spending on service line extensions from $17.2
million in 2023 to $18.5 million in 2024.[^peoples_spending]

[^peoples_spending]: Peoples Gas main figures reported as "NonQIP main expenditures" net of customer contributions. Historical (2018–2022): ICC Docket 23-0068, PIO Exhibit 12, [p. 36](https://www.documentcloud.org/documents/26189740-peoplesgas-p2023-0068-pio-exhibit-12-64269107/#document/p36/a2674688). Forecast (2023–2024): [p. 39](https://www.documentcloud.org/documents/26189740-peoplesgas-p2023-0068-pio-exhibit-12-64269107/#document/p39/a2674694). Service line historical: [p. 36](https://www.documentcloud.org/documents/26189740-peoplesgas-p2023-0068-pio-exhibit-12-64269107/#document/p36/a2674690), forecast: [p. 40](https://www.documentcloud.org/documents/26189740-peoplesgas-p2023-0068-pio-exhibit-12-64269107/#document/p40/a2674695). [@peoplesgas_Docket230068Peoples_2023]

:::{.column-page-inset-right}
{{< embed notebooks/analysis.qmd#fig-peoples-lea-trend >}}
:::

Peoples Gas's LEA spending on mains and service lines from 2018 through 2024,
with forecasted figures for 2023–2024, shows service line spending sustained at
over $15 million per year while main extension spending declines (@fig-peoples-lea-trend).

### What does each Peoples Gas connection cost?

Peoples Gas also does not report a cost per connection for either main or
service line extensions. However, information in its rate cases before the ICC
tells us that in 2022, Peoples Gas customers paid an average of **$11,784** to
extend gas mains and build service lines for each new single-family connection.[^peoples_11784]

[^peoples_11784]: Combined main and service line cost per new single-family residential connection. Multi-family costs not disclosed. ICC Docket 23-0069, PIO Cross-Exhibit 10.0, [p. 1](https://www.documentcloud.org/documents/26189763-peoples-gas-p2023-0069-pio-cross-exhibit-100-95312469/#document/p1/a2674697). [@peoplesgas_Docket230069PIO_2023]

:::{.column-page-inset-right}
{{< embed notebooks/analysis.qmd#fig-peoples-per-connection-cost >}}
:::

The chart above shows the combined per-connection cost — covering both main
extensions and service lines — for new single-family residential connections to
Peoples Gas's system between 2020 and 2022 (@fig-peoples-per-connection-cost).

## How much does Ameren Gas spend on LEAs?

Ameren has denied requests to report directly on LEA spending. As an
approximation, we use Ameren's distribution **standing work order** (SWO) 
spending as a proxy for LEA costs.[^ameren_swo]

[^ameren_swo]: The SWO (internal code C3231) covers gas mains, service lines, fittings, and other materials associated with new service requests. A separate transmission SWO (C3160) is excluded as it covers large-scale equipment rather than new-customer connections. [@kilhoffer_Docket230067Ameren_2023]

Using SWO data from 2021, we estimate that Ameren spent **$20.1 million** using
LEA subsidies that year. In 2022, that spending increased to **$21.3 million**.[^ameren_swo_calc]

[^ameren_swo_calc]: Ameren SWO C3231 gross and CIAC. ICC Docket 23-0067, PIO Exhibit 12, [p. 13](https://www.documentcloud.org/documents/26188848-ameren-p2023-0067-pio-exhibit-12-96925852/#document/p13/a2674478) (2021–2022), [p. 14](https://www.documentcloud.org/documents/26188848-ameren-p2023-0067-pio-exhibit-12-96925852/#document/p14/a2674480) (2023–2024 forecast). Net figures: 2021 $21,486,023 − $1,355,566 = $20,130,457; 2022 $22,217,528 − $867,845 = $21,349,683. [@kilhoffer_Docket230067Ameren_2023]

This spending is particularly notable because unlike Nicor and Peoples Gas,
Ameren's customer base is shrinking. US Department of Transportation data shows
Ameren has lost approximately 1,000 to 2,000 service connections per year
consistently since 2017. The total number of connections has declined from
roughly 819,000 in 2017 to 809,000 in 2025.[^phmsa_ameren]

[^phmsa_ameren]: U.S. Department of Transportation, Pipeline and Hazardous Materials Safety Administration, Gas Distribution Annual Report (Form F7100.1-1), 2017–2025. [@phmsa_GasDistributionAnnual_2025]

In other words, Ameren's ratepayers are likely funding more than **$20 million**
annually of new infrastructure. That’s more spending for a pipeline network that 
serves fewer and fewer customers each year.

Ameren's annual LEA spending rose steadily from 2021 through 2024, even as its 
total service connections declined.(@fig-ameren-spending-trend)

:::{.column-page-inset-right}
{{< embed notebooks/analysis.qmd#fig-ameren-spending-trend >}}
:::

Ameren’s customer base has shrunk by roughly 10,000 connections since 2017, its 
but its LEA spending has continued to rise — meaning each remaining customer 
is effectively subsidizing a larger share of new connection costs.

### What does each Ameren connection cost?

Ameren reports an **average standard cost equivalent** per new connection.
These figures cover main extensions only. Because they do not include service
line extensions they may also underestimate the true cost of connecting a new
customer to Ameren's network.

Our best estimate is that Ameren's per connection cost rose from
**$5,539** in 2021 to **$7,331** in 2025 (for main extensions only -- service
lines were not separately reported).

## Why is LEA spending data so difficult to obtain?

No utility voluntarily published any of the data in this memo. Instead, we 
extracted it from discovery responses in ICC rate case filings.

Regrettably, even formal discovery procedures aren’t sufficient to give the 
public a complete picture of LEA spending. Utilities can object to requests for 
data on the grounds that isolating the information would require too much work. 
The legal term is “an objection on grounds of undue burden.” As a result, no 
utility reported a standardized LEA spending figure. Nicor provided the most 
detailed data, including per-connection costs, but no other utility replicated 
that level of disclosure. Peoples Gas does not track service line extension 
costs as a separate category, while Ameren has denied requests to report 
directly on LEA spending.

# Appendix

## Acknowledgments

The authors would like to thank:

- Bradley Cebulko, Current Energy Group

## Data and Methods

### Data collected from rate case filings

All source data used in this memo is available in consolidated form in our 
[public input data spreadsheet](https://docs.google.com/spreadsheets/d/1b6I4uyH8H2FiMoeM-trO7eAWk_X2AYhovzEbS9u1cfM/edit?gid=527161228#gid=527161228), 
which links each figure to its original ICC filing.

LEA spending figures in this memo come from rate cases for the three major 
Illinois gas utilities. No utility voluntarily publishes this data.

**Nicor Gas** ([ICC Rate Case 23-0066](https://www.documentcloud.org/documents/26188653-nicor-p2023-0066-pio-exhibit-12-30205654/)):

- *Main extension spending*: Reported as `extension_cost` by year
  (2018–2022 historical, 2023–2026 forecast), net of Contributions in Aid
  of Construction (CIAC) which Nicor reports separately as
  `customer_contribution`. Figures in this memo reflect the net amount
  (gross minus CIAC).
- *Service line spending*: Same structure as main extensions, with
  separate CIAC figures for service lines.
- *Per-connection main extension cost*: Reported directly by Nicor as
  `lea_cost_per_customer` (2019–2023). Nicor is the only utility in our
  sample that reports a standardized per-connection figure.
- *Service line per-project cost*: Not directly reported. We calculate it
  as total service line extension cost (net of CIAC) divided by the number
  of service line projects (`n_projects`). We interpret one project as
  one connection.
- *Per-foot main extension cost*: Reported directly by Nicor for
  high-pressure mains (2019–2023).
- *Extended forecast (2026–2029)*: ICC Docket 25-0055, PIO Cross Exhibit
  2.0 provides an additional source of forecast data covering 2026–2029.
  These figures are reported in aggregate only (no service/main
  breakdown), so they are shown as combined in @fig-nicor-lea-trend and
  in @tbl-data-availability.

**Peoples Gas** ([ICC Rate Case 23-0068](https://www.documentcloud.org/documents/26189740-peoplesgas-p2023-0068-pio-exhibit-12-64269107/); [ICC Rate Case 23-0069](https://www.documentcloud.org/documents/26189763-peoples-gas-p2023-0069-pio-cross-exhibit-100-95312469/)):

- *Main extension spending*: Reported as `lea_cost` by year (2018–2022
  historical, 2023–2024 forecast). Peoples' filing explicitly states main
  figures are "net of customer contributions."
- *Service line spending*: Reported as `lea_cost` by year with the same
  structure. Peoples does not separately disclose CIAC amounts for service
  lines. We therefore assume these figures are also net of customer
  contributions, consistent with the treatment of main extensions in the
  same filing.
- *Per-connection combined cost*: Reported as `cost_per_new_service`
  (2020–2022), covering single-family residential connections only.
  Multi-family connection costs were not disclosed.
- *Service line project counts*: Reported for 2018–2022. Peoples does not
  forecast service line project counts, describing this work as
  customer-driven.

**Ameren Illinois Gas** ([ICC Rate Case 23-0067](https://www.documentcloud.org/documents/26188848-ameren-p2023-0067-pio-exhibit-12-96925852/); [ICC Rate Case 25-0084](https://www.documentcloud.org/documents/26193410-ameren-p2025-0084-pio-exhibit-12-57399815/)):

- *LEA spending*: Ameren does not track LEA spending as a separate
  category. We use the distribution standing work order (SWO, internal
  code C3231) as the closest available proxy. Per Ameren's filing, C3231
  covers "gas mains, service lines, fittings and other material associated
  with new service requests." Figures are reported as
  `distribution_extensions_total` (gross) and
  `distribution_extensions_customer_contributions` (CIAC). This memo uses
  net figures. A separate transmission SWO (C3160) is excluded as it
  covers large-scale equipment rather than new-customer connections.
- *Per-connection cost*: Reported as the standard cost equivalent for 400
  feet of main extension (`max_lea_cost_per_connect`, 2021–2025). This
  covers main extensions only. Service line costs are not reported
  separately.

### Data availability

The data available for each utility varies substantially in both completeness 
and comparability. Nicor Gas provides the most detailed disclosures, including 
per-connection costs for main extensions and separate CIAC figures for both 
mains and service lines. Peoples Gas provides total spending figures but does 
not isolate service line CIAC amounts, and does not forecast service line 
project counts. Ameren Illinois does not track LEA spending as a distinct 
category at all. The SWO C3231 figures used in this memo are the closest 
available proxy.

```{r}
#| label: setup-gt-table
#| include: false

suppressPackageStartupMessages({
  library(gt)
  library(tidyverse)
})

load("report_vars.RData")

gt_table$`_data`[6, "Services Total"] <- "2023-2024"
gt_table <- gt_table |>
  tab_style(
    style = cell_fill(color = "#fff3cd"),
    locations = cells_body(columns = `Services Total`, rows = 6)
  )

gt_table <- gt_table |>
  tab_options(
    table.font.size = px(8),
    container.overflow.x = FALSE,
    container.overflow.y = FALSE
  )
```

```{r}
#| label: tbl-data-availability
#| tbl-cap: "Data availability by utility, metric, and period. Green indicates complete data; yellow indicates partial or caveated data; red indicates data not available or not provided. Peoples Gas forecast service line spending (2023–2024) is reported net of customer contributions in ICC Docket 23-0068, though the utility does not maintain forecasts for this category."
#| echo: false

gt_table
```

This inconsistency is itself a finding. Illinois ratepayers cannot make
informed decisions about gas system expansion when the three utilities
serving the state use three different accounting methods, and when one of
them does not separately track LEA spending at all.


### Estimated data

The following memo figures involve calculation or estimation beyond the
raw values in utility filings:

- **Nicor service line per-project cost ($3,716 in 2022)**. Calculated as
  net service line spending ($34,624,181) divided by the number of
  service line projects (9,318). Nicor does not directly report a
  cost-per-project figure for service lines.[^est_25]

  [^est_25]: See ICC Docket 23-0066, PIO Exhibit 12. [@northernillinoisgascompany_Docket230066Nicor_2023]

- **Nicor main extension spending, net of CIAC**. Calculated as
  `extension_cost` + `customer_contribution`, where CIAC is stored as a
  negative value. Gross and net figures differ by approximately 3–8
  percent in each year.[^est_26]

  [^est_26]: See ICC Docket 23-0066, PIO Exhibit 12. [@northernillinoisgascompany_Docket230066Nicor_2023]

- **Peoples Gas service line spending, CIAC treatment**. Peoples does not
  explicitly disclose whether its reported service line spending is gross
  or net of customer contributions. The same filing states that Peoples'
  main extension figures are net of customer contributions. We assume the
  same treatment applies to service lines. If service line figures are
  instead gross, the actual ratepayer burden would be approximately 3–8
  percent lower.[^est_27]

  [^est_27]: See ICC Docket 23-0068, PIO Exhibit 12. [@peoplesgas_Docket230068Peoples_2023]

- **Ameren LEA spending**. Estimated using the distribution SWO (C3231)
  net of CIAC. The SWO covers gas mains, service lines, fittings, and
  other materials associated with new service requests. This proxy may
  overstate LEA spending to the extent that C3231 includes costs beyond
  traditional line extensions. Conversely, it may understate LEA spending
  to the extent that relevant costs are charged to other work orders.
  Because Ameren does not track LEA spending as a separate category, no
  direct comparison is possible.[^est_28]

  [^est_28]: See ICC Docket 23-0067, PIO Exhibit 12. [@kilhoffer_Docket230067Ameren_2023]

- **Statewide 2022 total ($113.8 million)**. Calculated as the sum of
  each utility's 2022 net LEA spending: Nicor ($65.7M), Peoples ($26.7M), 
  Ameren ($21.3M). All three figures are net of customer contributions.[^est_29]

  [^est_29]: See ICC Dockets 23-0066, 23-0068, and 23-0067. [@northernillinoisgascompany_Docket230066Nicor_2023; @peoplesgas_Docket230068Peoples_2023; @kilhoffer_Docket230067Ameren_2023]

- **PHMSA-derived connection counts**. Statewide net growth (3,000–6,000
  new connections per year), gross new installations (27,500–51,000 per
  year), and per-utility net change figures (Ameren: -1,000 to -2,000 per
  year & Nicor: +5,000 to +7,000 per year) are calculated from U.S.
  Department of Transportation Pipeline and Hazardous Materials Safety
  Administration Gas Distribution Annual Reports (Form F7100.1-1),
  2017–2025. PHMSA data tracks all physical service line installations
  regardless of funding mechanism. It does not distinguish LEA-subsidized
  connections from customer-paid connections or from replacements of
  existing service lines. Gross installation counts from PHMSA are
  therefore not directly comparable to LEA-funded project counts from
  utility rate case filings.[^est_30]

  [^est_30]: U.S. Department of Transportation, Pipeline and Hazardous Materials Safety Administration, Gas Distribution Annual Report (Form F7100.1-1), 2017–2025. [@phmsa_GasDistributionAnnual_2025]

## References

::: {#refs}
:::
