---
title: ANALYSIS for ny_hp_rates

toc: true
reference-location: margin
fig-cap-location: bottom

appendix-style: default
citation-location: document
citation:
  container-title: Switchbox

format:
  html:
    page-layout: full
---


```{r}

library(tidyverse)
library(scales)

library(gt)
library(gtExtras)
library(jsonlite)
library(RSQLite)

source("/workspaces/reports/lib/ggplot/switchbox_theme.R")

```

# Introduction
## Context
- ConEd wants to charge a flat fee for HPs
-

## The rates
- first hot take: when the NYDPS published average bill increases, they should consider more than 600kWh/month
  - what's the average kWh/month for upgrade0? upgrade3? upgrade5?

### Where to find these rates?
- most are in ERP-2 [@coned_ConEdExhibitERP2_2025]
  - SC1 Rate i - Resi Basic -> ERP-2, Schedule 3, Table 1
  - SC1 Rate ii - Resi TOU -> ERP-2, Schedule 3, Table 2
  - SC1 Rate III - Resi TOU -> ERP-2, Schedule 3, Table 3
  - Rider Z - SC1 SPP -> ERP-2, Schedule 3, Table 48


### What about the kWh limits?
-

## Supply
- "If you buy electricity from Con Edison, the total supply charges on your bill are made up of Supply (described below), the Merchant Function Charge, the Clean Energy Standard Supply Surcharge, and Gross Receipts Taxes and other surcharges." [source](https://lite.coned.com/_external/cerates/supply_charges.asp)
- "The Supply Charge is the sum of the following charges:

## Data Sources

### Other relevant docs
- https://lite.coned.com/_external/cerates/elec.asp
- Supply rates
  - [@coned_ConEdSupplyRates_2025] gives the price that ConEd _pays_ for wholesale electricity


# Construct the rates
### TOU Rates (Hourly)
```{r}
# Create a data frame with all combinations of month and hour
month_hour_grid <- expand.grid(
  month = 1:12,
  hour = 0:23
) |>
  arrange(month, hour)

# Show dimensions
print(sprintf("Number of rows: %d", nrow(month_hour_grid)))
print(sprintf("Number of columns: %d", ncol(month_hour_grid)))

# Classify the months and hours into seasons and TOU periods
summer <- c(6,7,8,9)
winter <- c(1,2,3,4,5,10,11,12)
on_peak <- c(8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23)
off_peak <- c(0,1,2,3,4,5,6,7)

seasons_tou_grid <- month_hour_grid |>
  mutate(
    season = case_when(
      month %in% summer ~ "summer",
      month %in% winter ~ "winter"
    ),
    `tou_period` = case_when(
      hour %in% on_peak ~ "on-peak",
      hour %in% off_peak ~ "off-peak"
    )
  )

# Save the seasons and TOU grid to CSV
# Print current working directory
print(getwd())

write_csv(
  seasons_tou_grid,
  "/workspaces/reports/data/utility_rates/electric/ConEd/coned_seasons_tou_grid.csv"
)

```

### Supply Rates (Monthly)
```{r}
coned_supply_rates <- tribble(
  ~month, ~rate,
     1,    0.12,
     2,    0.12,
     3,    0.12,
     4,    0.12,
     5,    0.12,
     6,    0.12,
     7,    0.12,
     8,    0.12,
     9,    0.12,
    10,    0.12,
    11,    0.12,
    12,    0.12
)
```

```{r}
library(lubridate)

msc_supply_codes = c(
  "sc1_r1" = "001",
  "sc1_r3" = "001IPP",
  "sc1_r4" = "001VTO"
)

supply_rates = read_csv("/workspaces/reports/data/utility_rates/electric/ConEd/ConEd - LDS_APPND.csv", skip = 1) |>
  filter(TYPE == "1") |>
  rename(
    from_dt = `...2`,
    to_dt = `...3`,
    ALL_HOURS_H = `LT-ZONE H...5`
  ) |>
  mutate(
    from_dt = ymd(from_dt),
    to_dt = ymd(to_dt)
  )

View(supply_rates)

# Create scatter plot of supply rates over time
# ggplot(supply_rates, aes(x = from_dt, y = ALL_HOURS_H)) +
#   geom_point() +
#   labs(
#     title = "ConEd Supply Rates Over Time",
#     x = "Date",
#     y = "Rate ($/kWh)",
#   ) +
#   scale_y_continuous(labels = scales::dollar_format())
# # Display the plot
# print(last_plot())

# Create scatter plot of supply rates over time with all years shown
ggplot(supply_rates, aes(x = from_dt, y = ALL_HOURS_H)) +
  geom_point() +
  labs(
    title = "ConEd Supply Rates Over Time",
    x = "Date",
    y = "Rate ($/kWh)",
  ) +
  scale_x_date(
    date_breaks = "1 year",
    date_labels = "%Y"
  ) +
  scale_y_continuous(labels = scales::dollar_format())

# Create violin plot of supply rates by month
supply_rates |>
  mutate(month = month(from_dt)) |>
  ggplot(aes(x = factor(month), y = ALL_HOURS_H)) +
  geom_violin(fill = sb_colors[["sky"]], alpha = 0.7) +
  geom_boxplot(width = 0.1, fill = "white", alpha = 0.7) +
  labs(
    title = "Distribution of ConEd Supply Rates by Month",
    x = "Month",
    y = "Rate ($/kWh)"
  ) +
  scale_x_discrete(labels = month.abb) +
  scale_y_continuous(labels = scales::dollar_format())

supply_rates <- supply_rates |>
  filter(day(from_dt) == 1)

# Add year column and create multi-year plot
supply_rates |>
  mutate(year = year(from_dt)) |>
  ggplot(aes(x = month(from_dt), y = ALL_HOURS_H, color = factor(year), group = year)) +
  geom_line() +
  geom_point() +
  labs(
    title = "ConEd Supply Rates by Year",
    x = "Month",
    y = "Rate ($/kWh)",
    color = "Year"
  ) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_y_continuous(labels = scales::dollar_format()) +
  scale_color_gradient2(
    low = "grey",
    high = "red",
    mid = "pink",
    midpoint = mean(year(supply_rates$from_dt))
  )

```






### SC1, Rate I - Basic Residential
```{r}

```

### SC1, Rate III - Residential TOU
```{r}


# Apply rates to the seasons and TOU periods
sc1_r1_tou_energy <- tibble(
  winter_on_peak_rate = 0.19366,
  winter_off_peak_rate = 0.19366,
  summer_on_peak_rate = 0.22266,
  summer_off_peak_rate = 0.22266
)

sc1_r3_tou_energy <- tibble(
  winter_on_peak_rate = 0.20210,
  winter_off_peak_rate = 0.06170,
  summer_on_peak_rate = 0.32900,
  summer_off_peak_rate = 0.06170
)

# ----------------------------------------------------
# Assembling the rates...
sc1_r1_288 <- seasons_tou_grid |>
  mutate(
    rate = case_when(
      season == "winter" & tou_period == "on-peak" ~ sc1_r1_tou_energy$winter_on_peak_rate,
      season == "winter" & tou_period == "off-peak" ~ sc1_r1_tou_energy$winter_off_peak_rate,
      season == "summer" & tou_period == "on-peak" ~ sc1_r1_tou_energy$summer_on_peak_rate,
      season == "summer" & tou_period == "off-peak" ~ sc1_r1_tou_energy$summer_off_peak_rate
    )
  )

# Apply rates based on season and TOU period
sc1_r3_288 <- seasons_tou_grid |>
  mutate(
    rate = case_when(
      season == "winter" & tou_period == "on-peak" ~ sc1_r3_tou_energy$winter_on_peak_rate,
      season == "winter" & tou_period == "off-peak" ~ sc1_r3_tou_energy$winter_off_peak_rate,
      season == "summer" & tou_period == "on-peak" ~ sc1_r3_tou_energy$summer_on_peak_rate,
      season == "summer" & tou_period == "off-peak" ~ sc1_r3_tou_energy$summer_off_peak_rate
    )
  )

# Group by season and show summary statistics
sc1_r3_wide <- sc1_r3_288 |>
  group_by(hour, season) |>
  summarize(rate = first(rate), .groups = "drop") |>
  pivot_wider(
    names_from = season,
    values_from = rate
  )


print("Summary statistics by season:")
print(sc1_r3_wide)

# Create a line plot showing rates by hour for each season
ggplot(sc1_r3_wide, aes(x = hour)) +
  geom_line(aes(y = winter, color = "Winter"), linewidth = 1) +
  geom_line(aes(y = summer, color = "Summer"), linewidth = 1) +
  scale_color_manual(values = c(
    "Winter" = sb_colors[["sky"]],
    "Summer" = sb_colors[["carrot"]]
  )) +
  scale_x_continuous(breaks = seq(0, 23, 4)) +
  scale_y_continuous(
    labels = scales::dollar_format(accuracy = 0.001),
    limits = c(0, 0.35)
  ) +
  labs(
    title = "SC1 Rate III: Time-of-Use Rates by Season",
    x = "Hour of Day",
    y = "Rate ($/kWh)",
    color = "Season"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom"
  )

print(last_plot())

```






### SC1, Rate IV - Optional Demand Based
```{r}

sc1_r4 <- tibble(
  customer_charge = 34,
  winter_on_peak_rate = 24.88,
  winter_off_peak_rate = 8.87,
  summer_on_peak_rate = 32.34,
  summer_off_peak_rate = 8.87
)

sc1_r4_288 <- seasons_tou_grid |>
  mutate(
    rate = case_when(
      season == "winter" & tou_period == "on-peak" ~ sc1_r4$winter_on_peak_rate,
      season == "winter" & tou_period == "off-peak" ~ sc1_r4$winter_off_peak_rate,
      season == "summer" & tou_period == "on-peak" ~ sc1_r4$summer_on_peak_rate,
      season == "summer" & tou_period == "off-peak" ~ sc1_r4$summer_off_peak_rate
    )
  )

print(paste0("Winter On Peak/Off-Peak Ratio: ", sc1_r4$winter_on_peak_rate / sc1_r4$winter_off_peak_rate))
print(paste0("Summer On Peak/Off-Peak Ratio: ", sc1_r4$summer_on_peak_rate / sc1_r4$summer_off_peak_rate))

# Group by season and show summary statistics
sc1_r4_wide <- sc1_r4_288 |>
  group_by(hour, season) |>
  summarize(rate = first(rate), .groups = "drop") |>
  pivot_wider(
    names_from = season,
    values_from = rate
  )

print("Summary statistics by season:")
print(sc1_r4_wide)

# Create a line plot showing rates by hour for each season
ggplot(sc1_r4_wide, aes(x = hour)) +
  geom_line(aes(y = winter, color = "Winter"), linewidth = 1) +
  geom_line(aes(y = summer, color = "Summer"), linewidth = 1) +
  scale_color_manual(values = c(
    "Winter" = sb_colors[["sky"]],
    "Summer" = sb_colors[["carrot"]]
  )) +
  scale_x_continuous(breaks = seq(0, 23, 4)) +
  scale_y_continuous(
    labels = scales::dollar_format(accuracy = 0.01),
    limits = c(0, 40)
  ) +
  labs(
    title = "Time-of-Use Rates by Season",
    x = "Hour of Day",
    y = "Rate ($/kWh)",
    color = "Season"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom"
  )

print(last_plot())


```




# ------------------------------------------------------------------------------

# Load the hourly data
- load from the database
- get mean 24-hour load profile for each month

## Load the data

### Select N random buildings
```{r}

con <- dbConnect(duckdb::duckdb(), "/workspaces/reports/data/ResStock/2022_resstock_amy2018_release_1.1/rs_20250307.db")

get_new_sample = FALSE
N = 1000
if (get_new_sample) {

  # Get 20 random building IDs
  random_bldgs <- dbGetQuery(con, "
    SELECT DISTINCT bldg_id
    FROM NY_hourly
    WHERE upgrade = 4
    ORDER BY RANDOM()
    LIMIT 1000
  ")


  print(paste0("Randomly selected building IDs: ", paste(random_bldgs$bldg_id, collapse = ", ")))

  write_csv(random_bldgs, "/workspaces/reports/data/ResStock/2022_resstock_amy2018_release_1.1/random_bldgs.csv")
} else {
  random_bldgs <- read_csv("/workspaces/reports/data/ResStock/2022_resstock_amy2018_release_1.1/random_bldgs.csv")
  bldg_ids <- paste(random_bldgs$bldg_id, collapse = ", ")
}


# Get the hourly data for the N buildings
query_string <- paste0("
  SELECT
    bldg_id,
    month,
    hour,
    SUM(\"sum_out.electricity.total.energy_consumption\") as mh_energy_consumption
  FROM NY_hourly
  WHERE upgrade = 4
  AND bldg_id IN (", bldg_ids, ")
  GROUP BY bldg_id, month, hour
  ORDER BY bldg_id, month, hour
")

hourly_sample_bldgs <- dbGetQuery(con,query_string)

# Disconnect from database
dbDisconnect(con)

print(paste0("hourly_sample_bldgs has ", nrow(hourly_sample_bldgs), " rows"))
head(hourly_sample_bldgs, n=29)


# Get monthly totals for each of the N bldg_ids
monthly_sample_bldgs <- hourly_sample_bldgs |>
  group_by(bldg_id, month) |>
  summarize(
    m_energy_consumption = sum(mh_energy_consumption, na.rm = TRUE)
  )

print(paste0("monthly_sample_bldgs has ", nrow(monthly_sample_bldgs), " rows"))
head(monthly_sample_bldgs, n=29)
```


```{r}


# Calculate statistics by month and hour across all buildings
hourly_stats <- hourly_sample_bldgs |>
  group_by(month, hour) |>
  summarize(
    mean_consumption = mean(mh_energy_consumption, na.rm = TRUE),
    p90_consumption = quantile(mh_energy_consumption, 0.9, na.rm = TRUE),
    p10_consumption = quantile(mh_energy_consumption, 0.1, na.rm = TRUE)
  )

# Create faceted plot with one panel per month
ggplot(hourly_stats, aes(x = hour)) +
  geom_line(aes(y = mean_consumption, color = "Mean"), linewidth = 1) +
  geom_line(aes(y = p90_consumption, color = "90th Percentile"), linewidth = 0.8) +
  geom_line(aes(y = p10_consumption, color = "10th Percentile"), linewidth = 0.8) +
  facet_wrap(~month, nrow = 3, labeller = labeller(month = function(x) month.abb[as.numeric(x)])) +
  labs(
    x = "Hour of Day",
    y = "Energy Consumption (kWh)",
    title = "Hourly Energy Consumption Profiles by Month",
    color = "Statistic"
  ) +
  scale_x_continuous(breaks = seq(0, 23, 6)) +
  scale_color_manual(
    values = c("Mean" = "black",
              "90th Percentile" = "#E69F00",
              "10th Percentile" = "#56B4E9")
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "grey90"),
    strip.text = element_text(face = "bold")
  )

# Display the plot
print(last_plot())

# Save the plot
ggsave(
  "/workspaces/reports/src/ny_hp_rates/img/hourly_consumption.png",
  last_plot(),
  width = 12,
  height = 8,
  dpi = 300
)




```

### Peak Power
```{r}
# con <- dbConnect(duckdb::duckdb(), "/workspaces/reports/data/ResStock/2022_resstock_amy2018_release_1.1/rs_20250307.db")


# query_string <- paste0("
#   SELECT
#     bldg_id,
#     month,
#     MAX(\"sum_out.electricity.total.energy_consumption\") as m_peak_power
#   FROM NY_hourly
#   WHERE upgrade = 4
#   AND bldg_id IN (", bldg_ids, ")
#   GROUP BY bldg_id, month
#   ORDER BY bldg_id, month
# ")

# peak_power_sample_bldgs <- dbGetQuery(con,query_string)
# dbDisconnect(con)


# View(peak_power_sample_bldgs)
# # Disconnect from database

# # Create box plot of monthly peak power
# ggplot(peak_power_sample_bldgs, aes(x = factor(month), y = m_peak_power)) +
#   geom_boxplot() +
#   labs(
#     x = "Month",
#     y = "Monthly Peak Power (kW)",
#     title = "Distribution of Monthly Peak Power"
#   ) +
#   scale_x_discrete(labels = month.abb) +  # Use abbreviated month names
#   theme(
#     legend.position = "right",
#     panel.grid.minor = element_blank()
#   )

# # Display the plot
# print(last_plot())


```


### Average load profile across all 33676 buildings
```{r}

con <- dbConnect(duckdb::duckdb(), "/workspaces/reports/data/ResStock/2022_resstock_amy2018_release_1.1/rs_20250307.db")

# ------------------------------------------------------------------------------
#Get the average load profile across all 33676 buildings
# build the dbGetQuery string
query_string <- paste0("
  SELECT
    month,
    hour,
    AVG(\"sum_out.electricity.total.energy_consumption\") as mean,
    STDDEV(\"sum_out.electricity.total.energy_consumption\") as stdev
  FROM NY_hourly
  WHERE upgrade = 4
  AND bldg_id IN (", bldg_ids, ")
  GROUP BY month, hour
  ORDER BY month, hour
")

# Query to get mean and standard deviation by month and hour
hourly_stats <- dbGetQuery(con,query_string)

dbDisconnect(con)
```


# Combine load data with rates
## Plot the summer and winter rates separately
```{r}

summer <- c(6,7,8,9)
winter <- c(1,2,3,4,5,10,11,12)

# Create the plot
# Split data into winter and summer
winter_data <- hourly_stats |> filter(month %in% winter)
summer_data <- hourly_stats |> filter(month %in% summer)

# Create plots with shared axis limits
y_limits <- range(hourly_stats$mean)

winter_plot <- ggplot(winter_data, aes(x = hour, y = mean, color = factor(month), group = month)) +
  geom_line(linewidth = 1) +
  # Add rate line on secondary y-axis
  geom_line(data = sc1_r3_wide,
            aes(x = hour, y = winter * (max(winter_data$mean)/0.4), group = 1), # Scale the rate values
            color = "red", linewidth = 2.5,
            inherit.aes = FALSE) +
  labs(
    title = "Winter Months",
    x = "Hour of Day",
    y = "Average Electricity Consumption (kWh)",
    color = "Month"
  ) +
  # Add secondary y-axis for rates
  scale_y_continuous(
    sec.axis = sec_axis(~./(max(winter_data$mean)/0.4), # Transform back for display
                       name = "Rate ($/kWh)")
  ) +
  scale_x_continuous(breaks = seq(0, 23, 3)) +
  #scale_color_manual(values = rep(sb_colors['sky'], length(winter))) +
  coord_cartesian(ylim = c(0, 4)) +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank(),
    # Add axis title on right side
    axis.title.y.right = element_text(color = "red")
  )

summer_plot <- ggplot(summer_data, aes(x = hour, y = mean, color = factor(month), group = month)) +
  geom_line(linewidth = 1) +
  # Add rate line on secondary y-axis
  geom_line(data = sc1_r3_wide,
            aes(x = hour, y = summer * (max(summer_data$mean)/0.4), group = 1), # Scale the rate values
            color = "red", linewidth = 2.5,
            inherit.aes = FALSE) +
  labs(
    title = "Summer Months",
    x = "Hour of Day",
    y = "Average Electricity Consumption (kWh)",
    color = "Month"
  ) +
  # Add secondary y-axis for rates
  scale_y_continuous(
    sec.axis = sec_axis(~./(max(summer_data$mean)/0.4), # Transform back for display
                       name = "Rate ($/kWh)")
  ) +
  scale_x_continuous(breaks = seq(0, 23, 3)) +
  #scale_color_manual(values = rep(sb_colors['sky'], length(winter))) +
  coord_cartesian(ylim = c(0, 4)) +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank(),
    # Add axis title on right side
    axis.title.y.right = element_text(color = "red")
  )


# Arrange plots vertically with shared title
library(patchwork)
(winter_plot / summer_plot) +
  plot_annotation(
    title = "Average Daily Load Profile by Month",
    subtitle = paste0("Based on ", N, " buildings")
  )

```


## Apply TOU rates
### SC1, Rate I - Basic Residential
```{r}

sc1_r1_rate_info <- read_csv("/workspaces/reports/data/utility_rates/electric/ConEd/sc1_r1.csv")
supply_charge_columns = sc1_r1_rate_info |> filter(chargeClass == "supply" & !grepl("_tax", name)) |> pull(name)
delivery_charge_columns = sc1_r1_rate_info |> filter(chargeClass == "delivery" & !grepl("_tax", name)) |> pull(name)

# ------------------------------------------------------------------------------
# Part 1 - Charges that change hourly
# Join hourly consumption with rates and calculate cost
tou_charge_monthly <- hourly_sample_bldgs |>
  left_join(sc1_r1_288, by = c("month", "hour")) |>
  mutate(cost = mh_energy_consumption * rate) |>
  group_by(bldg_id, month) |>
  summarize(
    tou_delivery_charge = sum(cost, na.rm = TRUE)
  )
#View(tou_charge_monthly)



# ------------------------------------------------------------------------------
# Part 2 - Consumption-based charges
supply_monthly <- monthly_sample_bldgs |>
  left_join(coned_supply_rates, by = "month") |>
  mutate(
    variable_supply_charge = m_energy_consumption * rate
  )
#View(supply_monthly)


consumption_delivery_charge_names = sc1_r1_rate_info |> filter(chargeType == "consumption", chargeClass == "delivery", variability =="static") |> pull(name)
#View(consumption_delivery_charge_names)
consumption_delivery_monthly <- monthly_sample_bldgs |>
  left_join(sc1_r1_rate_info |>
    filter(name %in% consumption_delivery_charge_names) |>
    select(name, value) |>
    pivot_wider(names_from = name, values_from = value) |>
    slice(rep(1, 12)) |>
    mutate(month = 1:12),
    by = "month") |>
  mutate(across(all_of(consumption_delivery_charge_names),
         ~.x * m_energy_consumption))

#View(consumption_delivery_monthly)

# ------------------------------------------------------------------------------
# Part 3 - Static charges that are applied monthly_sample_bldgs
fixed_supply_charges <- sc1_r1_rate_info |>
  filter(chargeType == "fixed", chargeClass == "supply") |>
  select(name, value) |>
  pivot_wider(names_from = name, values_from = value) |>
  slice(rep(1, 12)) |>
  mutate(month = 1:12)
#View(fixed_supply_charges)

fixed_delivery_charges <- sc1_r1_rate_info |>
  filter(chargeType == "fixed", chargeClass == "delivery") |>
  select(name, value) |>
  pivot_wider(names_from = name, values_from = value) |>
  slice(rep(1, 12)) |>
  mutate(month = 1:12)
#View(fixed_delivery_charges)

# ------------------------------------------------------------------------------
# Part 4 - Combine all the charges

# SUPPLY
final_supply_charges <- c("variable_supply_charge", "supply_sales_tax", "supply_grt_and_other_tax_surcharges", "merchant_function_charge")
total_supply_charges <- supply_monthly |>
  left_join(fixed_supply_charges, by = "month")|>
  mutate(
    supply_sales_tax = (variable_supply_charge + merchant_function_charge + supply_grt_and_other_tax_surcharges) *
      (sc1_r1_rate_info |> filter(name == "supply_sales_tax") |> pull(value))
  ) |>
  select(bldg_id, month, all_of(final_supply_charges))
##View(total_supply_charges)

# DELIVERY
final_delivery_charges <- c("tou_delivery_charge", "basic_service_charge", "system_benefit_charge", "delivery_grt_and_other_tax_surcharges", "delivery_sales_tax")
total_delivery_charges <- tou_charge_monthly |>
  left_join(fixed_delivery_charges, by = "month") |>
  left_join(consumption_delivery_monthly, by = c("month", "bldg_id")) |>
  mutate(
    delivery_sales_tax = (basic_service_charge + tou_delivery_charge + system_benefit_charge + delivery_grt_and_other_tax_surcharges) *
      (sc1_r1_rate_info |> filter(name == "delivery_sales_tax") |> pull(value))
  ) |>
  select(bldg_id, month, all_of(final_delivery_charges))

#View(total_delivery_charges)

# COMBINED
all_billable_columns = c(final_supply_charges, final_delivery_charges)
sc1_r1_final_total_bill <- total_supply_charges |>
  left_join(total_delivery_charges, by = c("month", "bldg_id")) |>
  mutate(
      total_bill = rowSums(across(all_of(all_billable_columns)))
  )
#View(final_total_bill)

# Visualize monthly bill distribution
ggplot(sc1_r1_final_total_bill, aes(x = factor(month), y = total_bill)) +
  geom_boxplot() +
  labs(
    x = "Month",
    y = "Total Bill ($)",
    title = "Distribution of Total Bills by Month"
  ) +
  theme_minimal() +
  scale_x_discrete(labels = month.abb) # Use abbreviated month names

# Display the plot
print(last_plot())

```

### SC1, Rate III - Residential TOU
```{r}


sc1_r3_rate_info <- read_csv("/workspaces/reports/data/utility_rates/electric/ConEd/sc1_r3.csv")
supply_charge_columns = sc1_r3_rate_info |> filter(chargeClass == "supply" & !grepl("_tax", name)) |> pull(name)
delivery_charge_columns = sc1_r3_rate_info |> filter(chargeClass == "delivery" & !grepl("_tax", name)) |> pull(name)



# ------------------------------------------------------------------------------
# Part 1 - Charges that change hourly
# Join hourly consumption with rates and calculate cost
tou_charge_monthly <- hourly_sample_bldgs |>
  left_join(sc1_r3_288, by = c("month", "hour")) |>
  mutate(cost = mh_energy_consumption * rate) |>
  group_by(bldg_id, month) |>
  summarize(
    tou_delivery_charge = sum(cost, na.rm = TRUE)
  )
#View(tou_charge_monthly)



# ------------------------------------------------------------------------------
# Part 2 - Consumption-based charges
supply_monthly <- monthly_sample_bldgs |>
  left_join(coned_supply_rates, by = "month") |>
  mutate(
    variable_supply_charge = m_energy_consumption * rate
  )
#View(supply_monthly)


consumption_delivery_charge_names = sc1_r3_rate_info |> filter(chargeType == "consumption", chargeClass == "delivery", variability =="static") |> pull(name)
#View(consumption_delivery_charge_names)
consumption_delivery_monthly <- monthly_sample_bldgs |>
  left_join(sc1_r3_rate_info |>
    filter(name %in% consumption_delivery_charge_names) |>
    select(name, value) |>
    pivot_wider(names_from = name, values_from = value) |>
    slice(rep(1, 12)) |>
    mutate(month = 1:12),
    by = "month") |>
  mutate(across(all_of(consumption_delivery_charge_names),
         ~.x * m_energy_consumption))

#View(consumption_delivery_monthly)

# ------------------------------------------------------------------------------
# Part 3 - Static charges that are applied monthly_sample_bldgs
fixed_supply_charges <- sc1_r3_rate_info |>
  filter(chargeType == "fixed", chargeClass == "supply") |>
  select(name, value) |>
  pivot_wider(names_from = name, values_from = value) |>
  slice(rep(1, 12)) |>
  mutate(month = 1:12)
#View(fixed_supply_charges)

fixed_delivery_charges <- sc1_r3_rate_info |>
  filter(chargeType == "fixed", chargeClass == "delivery") |>
  select(name, value) |>
  pivot_wider(names_from = name, values_from = value) |>
  slice(rep(1, 12)) |>
  mutate(month = 1:12)
#View(fixed_delivery_charges)

# ------------------------------------------------------------------------------
# Part 4 - Combine all the charges

# SUPPLY
final_supply_charges <- c("variable_supply_charge", "supply_sales_tax", "supply_grt_and_other_tax_surcharges", "merchant_function_charge")
total_supply_charges <- supply_monthly |>
  left_join(fixed_supply_charges, by = "month")|>
  mutate(
    supply_sales_tax = (variable_supply_charge + merchant_function_charge + supply_grt_and_other_tax_surcharges) *
      (sc1_r3_rate_info |> filter(name == "supply_sales_tax") |> pull(value))
  ) |>
  select(bldg_id, month, all_of(final_supply_charges))
##View(total_supply_charges)

# DELIVERY
final_delivery_charges <- c("tou_delivery_charge", "basic_service_charge", "system_benefit_charge", "delivery_grt_and_other_tax_surcharges", "delivery_sales_tax")
total_delivery_charges <- tou_charge_monthly |>
  left_join(fixed_delivery_charges, by = "month") |>
  left_join(consumption_delivery_monthly, by = c("month", "bldg_id")) |>
  mutate(
    delivery_sales_tax = (basic_service_charge + tou_delivery_charge + system_benefit_charge + delivery_grt_and_other_tax_surcharges) *
      (sc1_r3_rate_info |> filter(name == "delivery_sales_tax") |> pull(value))
  ) |>
  select(bldg_id, month, all_of(final_delivery_charges))

#View(total_delivery_charges)

# COMBINED
all_billable_columns = c(final_supply_charges, final_delivery_charges)
sc1_r3_final_total_bill <- total_supply_charges |>
  left_join(total_delivery_charges, by = c("month", "bldg_id")) |>
  mutate(
      total_bill = rowSums(across(all_of(all_billable_columns)))
  )
#View(final_total_bill)

# Visualize monthly bill distribution
ggplot(sc1_r3_final_total_bill, aes(x = factor(month), y = total_bill)) +
  geom_boxplot() +
  labs(
    x = "Month",
    y = "Total Bill ($)",
    title = "Distribution of Total Bills by Month"
  ) +
  theme_minimal() +
  scale_x_discrete(labels = month.abb) # Use abbreviated month names

# Display the plot
print(last_plot())

```


### SC1, Rate IV - Residential TOU with Demand Charge
```{r}
sc1_r4_rate_info <- read_csv("/workspaces/reports/data/utility_rates/electric/ConEd/sc1_r4.csv")
supply_charge_columns = sc1_r4_rate_info |> filter(chargeClass == "supply" & !grepl("_tax", name)) |> pull(name)
delivery_charge_columns = sc1_r4_rate_info |> filter(chargeClass == "delivery" & !grepl("_tax", name)) |> pull(name)






# ------------------------------------------------------------------------------
# Part 1 - Charges that change hourly
# Join hourly consumption with rates and calculate cost
demand_charge_monthly <- hourly_sample_bldgs |>
  left_join(sc1_r4_288, by = c("month", "hour")) |>
  mutate(demand_charge = mh_energy_consumption * rate)

View(demand_charge_monthly)

# Group by month and find maximum on-peak and off-peak values
monthly_peaks <- demand_charge_monthly |>
  group_by(bldg_id, month, tou_period) |>
  summarize(
    max_demand_charge = max(demand_charge, na.rm = TRUE)
  )

# View the first few rows
head(monthly_peaks)




```


# ------------------------------------------------------------------------------
# Part 5 - Analysis

## Histograms of annual bills
```{r}

sc1_r1_annual_bills <- sc1_r1_final_total_bill |>
  group_by(bldg_id) |>
  summarize(
    annual_bill = sum(total_bill)
  )

sc1_r3_annual_bills <- sc1_r3_final_total_bill |>
  group_by(bldg_id) |>
  summarize(
    annual_bill = sum(total_bill)
  )

# Combine the two datasets and add a rate identifier
combined_bills <- bind_rows(
  sc1_r3_annual_bills |> mutate(rate = "sc1_r3"),
  sc1_r1_annual_bills |> mutate(rate = "sc1_r1")
)

# Calculate and print median annual bills
sc1_r1_median <- median(sc1_r1_annual_bills$annual_bill)
sc1_r3_median <- median(sc1_r3_annual_bills$annual_bill)

print(paste0("SC1 Rate I median annual bill: $", round(sc1_r1_median)))
print(paste0("SC1 Rate III median annual bill: $", round(sc1_r3_median)))


# Create histograms for each rate
bin_width = 250
ggplot(combined_bills) +

  # SC1, Rate I
  geom_histogram(data = subset(combined_bills, rate == "sc1_r1"),
                aes(x = annual_bill, fill = "SC1, Rate I"),
                binwidth = bin_width, alpha = 0.7) +

  # Add median line for SC1, Rate I
  geom_vline(data = subset(combined_bills, rate == "sc1_r1"),
             aes(xintercept = median(annual_bill)),
             linetype = "dashed", color = sb_colors["sky"], linewidth = 1) +

  # Add text annotation for SC1, Rate I median
  # annotate("text",
  #          x = median(subset(combined_bills, rate == "sc1_r1")$annual_bill),
  #          y = 40,
  #          label = paste0("$", round(median(subset(combined_bills, rate == "sc1_r1")$annual_bill))),
  #          color = sb_colors["sky"],
  #          hjust = -0.2) +

  # SC1, Rate III
  geom_histogram(data = subset(combined_bills, rate == "sc1_r3"),
                aes(x = annual_bill, fill = "SC1, Rate III"),
                binwidth = bin_width, alpha = 0.7) +

  geom_vline(data = subset(combined_bills, rate == "sc1_r3"),
             aes(xintercept = median(annual_bill)),
             linetype = "dashed", color = sb_colors["carrot"], linewidth = 1) +

  facet_wrap(~rate, ncol = 1, scales = "free_y",
             labeller = labeller(rate = c(sc1_r1 = "SC1, Rate I",
                                        sc1_r3 = "SC1, Rate III"))) +
  labs(
    x = "Annual Bill ($)",
    y = "Count",
    title = "Distribution of Annual Bills by Rate",
    fill = "Rate Type"
  ) +
  scale_fill_manual(values = setNames(
      c(sb_colors["sky"], sb_colors["carrot"]),
      c("SC1, Rate I", "SC1, Rate III")
  )) +
  scale_x_continuous(limits = c(0, 10000)) +
  theme(legend.position = "none")

# Display the plot
print(last_plot())

# Save the plot
ggsave(
  "/workspaces/reports/src/ny_hp_rates/img/bill_hists.png",
  last_plot(),
  width = 4,
  height = 4,
  dpi = 300
)

```


```{r}

# Calculate annual bill for each building and rate
annual_by_bldg <- combined_bills |>
  group_by(bldg_id, rate) |>
  summarize(annual_bill = sum(annual_bill)) |>
  ungroup() |>
  pivot_wider(
    names_from = rate,
    values_from = annual_bill
  ) |>
  # Calculate savings (positive means sc1_r3 is cheaper)
  mutate(
    annual_savings = sc1_r1 - sc1_r3,
    pct_savings = (annual_savings / sc1_r1) * 100
  )

# Print summary statistics
cat("Summary of annual savings with SC1 Rate III vs Rate I:\n")
cat("Median absolute savings: $", round(median(annual_by_bldg$annual_savings)), "\n")
cat("Median percent savings: ", round(median(annual_by_bldg$pct_savings), 1), "%\n")

# Create histogram of savings
ggplot(annual_by_bldg, aes(x = pct_savings)) +
  geom_histogram(fill = sb_colors["pistachio"], binwidth = 0.5, alpha = 0.7) +
  geom_vline(aes(xintercept = median(annual_savings)),
             linetype = "dashed", color = "black", linewidth = 1) +
  annotate(
    "text",
    x = median(annual_by_bldg$annual_savings)+140,
    y = max(ggplot_build(last_plot())$data[[1]]$count),
    label = paste0("Median savings: $", round(median(annual_by_bldg$annual_savings)), " (", round(median(annual_by_bldg$pct_savings), 1), "%)"),
    vjust = 0.5
  ) +

  labs(
    x = "Annual Savings (%)",
    y = "Count",
    title = "Distribution of Annual Savings",
    subtitle = "SC1 Rate III vs Rate I (positive means Rate III is cheaper)"
  ) +
  scale_x_continuous(limits = c(NA, 20), labels = function(x) paste0(x, "%")) +
  theme(panel.grid.minor = element_blank())

# Save the plot
ggsave(
  "/workspaces/reports/src/ny_hp_rates/img/savings_hist.png",
  last_plot(),
  width = 6,
  height = 4,
  dpi = 300
)





```


## Box-and-whisker plot of the bills by rate
```{r}
# First combine the two datasets and add a rate identifier
combined_bills <- bind_rows(
  sc1_r3_final_total_bill |> select(bldg_id, month, total_bill) |> mutate(rate = "sc1_r3"),
  sc1_r1_final_total_bill |> select(bldg_id, month, total_bill) |> mutate(rate = "sc1_r1")
)
View(combined_bills)

# Create the plot with side-by-side boxplots
ggplot(combined_bills, aes(x = factor(month), y = total_bill, fill = rate)) +
  geom_boxplot(position = position_dodge(width = 0.8), outlier.size = 0.5) +
  labs(
    x = "Month",
    y = "Total Bill ($)",
    title = "Distribution of Total Bills by Month and Rate",
    subtitle = paste0("Random sample of ", N, " buildings"),
    fill = "Rate Type"
  ) +
  scale_x_discrete(labels = month.abb) +  # Use abbreviated month names
  scale_fill_manual(values = setNames(
      c(sb_colors["sky"], sb_colors["carrot"]),
      c("sc1_r1", "sc1_r3")  # Use your color palette
  )) +
  scale_y_continuous(limits = c(0, 2000)) +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank()
  )

# Display the plot
print(last_plot())

# Save the plot
ggsave(
  "/workspaces/reports/src/ny_hp_rates/img/bill_boxplot.png",
  last_plot(),
  width = 8,
  height = 4,
  dpi = 300
)

```
