---
title: analysis - ConEd HP Rates

toc: true
reference-location: margin
fig-cap-location: bottom

appendix-style: default
citation-location: document
citation:
  container-title: Switchbox

format:
  html:
    page-layout: full
---

```{r}
library(lubridate)
library(patchwork)
library(tidyverse)

# Switchbox Libraries
source("lib/ggplot/switchbox_theme.R")
source("lib/rates_analysis/heat_pump_rate_funcs.R")
```


# Basic Steps
1. Use `ny_delivery_tariffs.qmd` to clean the delivery tariffs and save them to the database
2. Use `ny_supply_rates.qmd` to clean the supply rates and save them to the database
3. Select a supply rate type ('dynamic' or 'all_hours') and a year (eg 2022)
    3.1 - use the get_month_hour_supply() function to get the month-hour supply rates
    3.2 - add that table to the database


4. Back in this notebook (`analysis.qmd`), we do that rate calculations
    - most of that should happen in the database, as much as possible

5.


# Delivery Tariffs
```{r}
path_rs_db <- "data/ResStock/2022_resstock_amy2018_release_1.1/rs_ny_coned_hp_rates.db"
con <- DBI::dbConnect(duckdb::duckdb(), path_rs_db)
delivery_tariff_month_hour_energy <- DBI::dbReadTable(con, "delivery_tariff_month_hour_energy")
DBI::dbDisconnect(con)

```


# Supply Rates
```{r}
source("lib/rates_analysis/heat_pump_rate_funcs.R")
# path_to_supply_rates <- "/workspaces/reports/data/utility_rates/electric/ConEd/coned_supply_rates.csv"
# supply_rates <- read_csv(path_to_supply_rates)

path_rs_db <- "data/ResStock/2022_resstock_amy2018_release_1.1/rs_ny_coned_hp_rates.db"
con <- DBI::dbConnect(duckdb::duckdb(), path_rs_db)
supply_rates <- DBI::dbReadTable(con, "supply_rates")
DBI::dbDisconnect(con)

# Monthly Supply Rates are used parts of the bill calculation
month_supply_2014 <- get_month_supply(supply_rates, target_year = 2014, target_name = "all_hours_h")
month_supply_2018 <- get_month_supply(supply_rates, target_year = 2018, target_name = "all_hours_h")

# # Month-Hour Supply Rates required for TOU tariffs
month_hour_supply_h_2014 <- get_month_hour_supply(supply_rates, start_year = 2014, end_year = 2014, dynamic_or_all_hours = "dynamic", all_hours_name = "all_hours_h", on_peak_name = "on_peak_h", off_peak_name = "off_peak_h")
month_hour_supply_h_2018 <- get_month_hour_supply(supply_rates, start_year = 2018, end_year = 2018, dynamic_or_all_hours = "dynamic", all_hours_name = "all_hours_h", on_peak_name = "on_peak_h", off_peak_name  = "off_peak_h")

```

## Supply Plots
```{r}
source("lib/rates_analysis/heat_pump_rate_funcs.R")

# plot each year as a separate line against 12 months
spaghetti_plot <- plot_supply_rates(supply_rates = supply_rates,
                  y = "all_hours_h",
                  start_year = 2014,
                  end_year = 2024,
                  highlight_years = c("2020", "2024"))
print(spaghetti_plot)

# Ribbon plot - shows the peak/off-peak spread and all-hours rate
ribbon_plot <- plot_supply_rates_ribbon(supply_rates = supply_rates,
                        first_year = 2014,
                        last_year = 2024,
                        ribbon_top = "on_peak_h",
                        ribbon_bottom = "off_peak_h",
                        middle = "all_hours_h")
print(ribbon_plot)
```


# Compute bills

```{r}
print("calculating monthly_load_r3_2014")
monthly_load_r3_2014 <- calculate_monthly_load_with_tariffs(path_rs_db,
                                                      month_hour_r3,
                                                      month_hour_supply_h_2014,
                                                      "r3")


```
