---
title: rates_prototype

toc: true
reference-location: margin
fig-cap-location: bottom

appendix-style: default
citation-location: document
citation:
  container-title: Switchbox

format:
  html:
    page-layout: full
---


```{r}
source("lib/ggplot/switchbox_theme.R")
library(lubridate)
library(patchwork)
library(tidyverse)

```

# Rates Prototype
### Delivery Tariff Functions
```{r}
assign_tariffs_to_month_hour_grid <- function(tariffs) {
  # Build the month-hour grid
  month_hour_grid <- expand.grid(
    month = 1:12,
    hour = 0:23
  ) |>
  arrange(month, hour)

  # Process each tariff and create a lookup table
  lookup_table <- tariffs |>
    # Must operate rowwise()
    rowwise() |>
    # create lists for months and hours
    mutate(
      months = list(if (start_month <= end_month) {
        start_month:end_month
      } else {
        c(start_month:12, 1:end_month)
      }),
      hours = list(if (start_hour <= end_hour) {
        start_hour:end_hour
      } else {
        c(start_hour:23, 0:end_hour)
      })
    ) |>
    # Expand the months and hours
    unnest(months) |>
    unnest(hours) |>
    # Keep relevant columns including tariff_name
    select(month = months, hour = hours, tariff_name, value)

  # First pivot to wide format by tariff_name
  month_hour_tariff_wide <- month_hour_grid |>
    left_join(lookup_table, by = c("month", "hour")) |>
    pivot_wider(
      names_from = tariff_name,
      values_from = value
    )

  # Then pivot back to long format
  month_hour_tariff_long <- month_hour_tariff_wide |>
    pivot_longer(
      cols = c(r3, r4),  # specify the rate columns
      names_to = "tariff_name",
      values_to = "rate"
    )

  return(month_hour_tariff_long)
}


#
#
assign_tariffs_to_month_grid <- function(tariffs) {
  # Create a lookup table by expanding the months for each tariff
  lookup_table <- tariffs |>
    # Must operate rowwise()
    rowwise() |>

    # Create a list of months covered by this tariff
    mutate(
      months = list(if (start_month <= end_month) {
        start_month:end_month
      } else {
        c(start_month:12, 1:end_month)
      })
    ) |>

    # Expand the months
    unnest(months) |>

    # Select only the columns we need for the final output
    select(
      utility, version, customer_class, tariff_name,
      domain, class, type, name, value,
      month = months
    )

  # Check for duplicate month combinations
  duplicate_check <- lookup_table |>
    group_by(utility, version, customer_class, tariff_name, type, month) |>
    summarize(
      count = n(),
      values = paste(unique(value), collapse = ", "),
      .groups = "drop"
    ) |>
    filter(count > 1)

  if (nrow(duplicate_check) > 0) {
    warning("Found duplicate values for some month combinations:")
    print(duplicate_check)

    # Use the first value for each month combination
    lookup_table <- lookup_table |>
      group_by(utility, version, customer_class, tariff_name, type, month) |>
      slice(1) |>
      ungroup()
  }

  return(lookup_table)
}
```


### Supply Rate Functions
```{r}
# assign_supply_rates_to_month_hour_grid <- function(supply_rates) {
#   # Build the month-hour grid
#   month_hour_grid <- expand.grid(
#     month = 1:12,
#     hour = 0:23
#   ) |>
#   arrange(month, hour)

#   # Process each tariff and create a lookup table
#   lookup_table <- supply_rates |>
#     # Must operate rowwise()
#     rowwise() |>
#     # create lists for months and hours
#     mutate(
#       months = list(if (start_month <= end_month) {
#         start_month:end_month
#       } else {
#         c(start_month:12, 1:end_month)
#       }),
#       hours = list(if (start_hour <= end_hour) {
#         start_hour:end_hour
#       } else {
#         c(start_hour:23, 0:end_hour)
#       })
#     ) |>
#     # Expand the months and hours
#     unnest(months) |>
#     unnest(hours) |>
#     # Keep relevant columns including tariff_name
#     select(month = months, hour = hours, tariff_name, value)

#   # First pivot to wide format by tariff_name
#   month_hour_tariff_wide <- month_hour_grid |>
#     left_join(lookup_table, by = c("month", "hour")) |>
#     pivot_wider(
#       names_from = tariff_name,
#       values_from = value
#     )

#   # Then pivot back to long format
#   month_hour_tariff_long <- month_hour_tariff_wide |>
#     pivot_longer(
#       cols = c(r3, r4),  # specify the rate columns
#       names_to = "tariff_name",
#       values_to = "rate"
#     )

#   return(month_hour_tariff_long)
# }


#
#
# assign_tariffs_to_month_grid <- function(tariffs) {
#   # Create a lookup table by expanding the months for each tariff
#   lookup_table <- tariffs |>
#     # Must operate rowwise()
#     rowwise() |>

#     # Create a list of months covered by this tariff
#     mutate(
#       months = list(if (start_month <= end_month) {
#         start_month:end_month
#       } else {
#         c(start_month:12, 1:end_month)
#       })
#     ) |>

#     # Expand the months
#     unnest(months) |>

#     # Select only the columns we need for the final output
#     select(
#       utility, version, customer_class, tariff_name,
#       domain, class, type, name, value,
#       month = months
#     )

#   # Check for duplicate month combinations
#   duplicate_check <- lookup_table |>
#     group_by(utility, version, customer_class, tariff_name, type, month) |>
#     summarize(
#       count = n(),
#       values = paste(unique(value), collapse = ", "),
#       .groups = "drop"
#     ) |>
#     filter(count > 1)

#   if (nrow(duplicate_check) > 0) {
#     warning("Found duplicate values for some month combinations:")
#     print(duplicate_check)

#     # Use the first value for each month combination
#     lookup_table <- lookup_table |>
#       group_by(utility, version, customer_class, tariff_name, type, month) |>
#       slice(1) |>
#       ungroup()
#   }

#   return(lookup_table)
# }


```


# Gather Data: Loads, Supply Rates, Tariffs
```{r}
# coned building ids
source("/workspaces/reports/lib/utility_mapping.R")

coned_bldg_ids <- get_bldg_by_utility(state_code = "NY", utility_electric = "coned", utility_gas = NULL) |>
  pull(bldg_id)

glimpse(coned_bldg_ids)

print(paste("coned_bldg_ids Count:", length(coned_bldg_ids)))

# Define path to the ResStock database
path_rs_db <-"/workspaces/reports/data/ResStock/2022_resstock_amy2018_release_1.1/rs_20250307.db"

```

## Loads
- We don't actually use the loads loaded here in any calculations, they're just for visualization
```{r}
con <- DBI::dbConnect(duckdb::duckdb(), path_rs_db)

# Get the hourly data for the N buildings
query <- paste0("
  SELECT
    bldg_id,
    month,
    hour,
    SUM(\"sum_out.electricity.total.energy_consumption\") as mh_energy_consumption
  FROM NY_hourly
  WHERE upgrade = 4 AND month in (1,8)
  AND bldg_id IN (", paste(coned_bldg_ids, collapse = ", "), ")
  GROUP BY bldg_id, month, hour
  ORDER BY bldg_id, month, hour
")

hourly_loads <- DBI::dbGetQuery(con, query)

# Disconnect from database
DBI::dbDisconnect(con)

print(paste0("hourly_loads has ", nrow(hourly_loads), " rows"))
head(hourly_loads, n=29)

```


## ConEd Supply Rates
### Data Loading and Cleaning
```{r}
# supply_rates_monthly_coned_2024 <- tibble(
#   year = rep(2024, 12),
#   month = 1:12,
#   utility = rep("coned", 12),
#   supply_rate = rep(0.12, 12)
# )

# supply_rates_raw = read_csv("/workspaces/reports/data/utility_rates/electric/ConEd/ConEd - LDS_APPND.csv", skip = 1) |>
#   janitor::clean_names() |>
#   filter(type == "1") |>
#   rename(
#     from_dt = x2,
#     to_dt = x3,
#     trip = x4,
#     all_hours_h = `lt_zone_h_5`,
#     on_peak_h = `lt_zone_h_7`,
#     off_peak_h = `lt_zone_h_9`,
#     all_hours_i = `lt_zone_i_11`,
#     on_peak_i = `lt_zone_i_13`,
#     off_peak_i = `lt_zone_i_15`,
#     all_hours_j = `lt_zone_j_17`,
#     on_peak_j = `lt_zone_j_19`,
#     off_peak_j = `lt_zone_j_21`
#   ) |>
#   mutate(
#     from_dt = ymd(from_dt),
#     to_dt = ymd(to_dt)
#   )|>
#   mutate(
#     start_year = year(from_dt),
#     start_month = month(from_dt),
#     start_day = day(from_dt)
#   ) |>
#   mutate(
#     utility = "coned"
#   )

# # ------------------------------------------------------------
# # Mean by month
# supply_rates <- supply_rates_raw |>
#   select(
#     all_hours_h,
#     on_peak_h,
#     off_peak_h,
#     all_hours_i,
#     on_peak_i,
#     off_peak_i,
#     all_hours_j,
#     on_peak_j,
#     off_peak_j,
#     start_month,
#     start_year
#   ) |>
#   group_by(start_month, start_year) |>
#   summarize(
#     all_hours_h = mean(all_hours_h),
#     on_peak_h = mean(on_peak_h),
#     off_peak_h = mean(off_peak_h),
#     all_hours_i = mean(all_hours_i),
#     on_peak_i = mean(on_peak_i),
#     off_peak_i = mean(off_peak_i),
#     all_hours_j = mean(all_hours_j),
#     on_peak_j = mean(on_peak_j),
#     off_peak_j = mean(off_peak_j),
#     .groups = "drop"
#   ) |>
#   arrange(start_year, start_month)

# glimpse(supply_rates)


# # ------------------------------------------------------------
# # get these into the month_hour_grid format
# # that means "long", with columns:
# # - month
# # - hour
# # - rate
# # - tariff_name

# get_month_hour_supply <- function(supply_rates, year) {
#   # Extract supply rates for given year
#   supply_rates_year <- supply_rates |>
#     filter(start_year == as.numeric(year)) |>
#     select(
#       month = start_month,
#       hour = 0,
#       all_hours = all_hours_h,
#       on_peak = on_peak_h,
#       off_peak = off_peak_h
#     )

#   # Define peak/off-peak hours
#   on_peak_hours <- c(8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23)
#   off_peak_hours <- c(0,1,2,3,4,5,6,7)

#   # Create month-hour grid
#   month_hour_grid <- expand.grid(
#     month = 1:12,
#     hour = 0:23
#   ) |>
#   arrange(month, hour)

#   # Join and assign rates based on peak/off-peak hours
#   month_hour_supply <- month_hour_grid |>
#     left_join(supply_rates_year, by = "month") |>
#     mutate(
#       supply_rate = case_when(
#         hour %in% on_peak_hours ~ on_peak,
#         hour %in% off_peak_hours ~ off_peak,
#         TRUE ~ all_hours # Fallback to all-hours rate
#       )
#     ) |>
#     select(month, hour, supply_rate)

#   return(month_hour_supply)
# }


# # Simpler monthly supply rates
# get_month_supply <- function(supply_rates, year) {
#   supply_rates_year <- supply_rates |>
#     filter(start_year == as.numeric(year)) |>
#     mutate(
#       month = start_month,
#       supply_rate = all_hours_h
#     ) |>
#     select(month, supply_rate)
#   return(supply_rates_year)
# }

# # Get simple monthly supply rates
# month_supply_2014 <- get_month_supply(supply_rates, 2014)
# month_supply_2018 <- get_month_supply(supply_rates, 2018)


# # Month-Hour Supply Rates required for TOU tariffs
# month_hour_supply_h_2014 <- get_month_hour_supply(supply_rates, 2014)
# month_hour_supply_h_2018 <- get_month_hour_supply(supply_rates, 2018)




```

### 12-Month supply rates plots
```{r}
# ------------------------------------------------------------
#
# Highlight Plot
# Define years to highlight
highlight_years <- c("2021","2022", "2023", "2024", "2025")

# Create line plot
ggplot(supply_rates, aes(x = start_month, y = all_hours_h, group = start_year)) +

  # Add grey lines for years not in highlight years
  geom_line(data = filter(supply_rates, !as.character(start_year) %in% highlight_years),
            color = "grey80") +
  geom_point(data = filter(supply_rates, !as.character(start_year) %in% highlight_years),
             color = "grey80", size = 0.1) +

  # Add colored lines for highlight years
  geom_line(data = filter(supply_rates, as.character(start_year) %in% highlight_years),
            aes(color = as.factor(start_year))) +
  geom_point(data = filter(supply_rates, as.character(start_year) %in% highlight_years),
             aes(color = as.factor(start_year)), size = 0.25) +

  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_y_continuous(labels = scales::label_dollar()) +
  scale_color_viridis_d() +
  labs(
    title = "ConEd Supply Rates by Year",
    x = "",
    y = "Wholesale Supply Cost ($/kWh)",
    color = "Year"
  ) +
  theme(legend.position = "right")

# ggsave(
#   "src/ny_hp_rates/img/coned_supply_rates.png",
#   width = 5,
#   height = 2.5,
#   dpi = 300
# )


# ------------------------------------------------------------
# All 3 Zones Plot
#
# Create base plot function to avoid repetition
create_rate_plot <- function(data, y_var, title) {
  ggplot(data, aes(x = start_month, y = {{y_var}}, group = start_year, color = as.factor(start_year))) +
    geom_line() +
    geom_point(size = 0.25) +
    geom_text(
      data = filter(data, start_month == 1),
      aes(label = start_year),
      hjust = 1.2,
      size = 3
    ) +
    scale_x_continuous(breaks = 1:12, labels = month.abb) +
    scale_y_continuous(limits = c(0, 0.25), labels = scales::label_dollar()) +
    scale_color_viridis_d() +
    labs(
      title = title,
      x = "",
      y = "Wholesale Supply Cost ($/kWh)",
      color = "Year"
    ) +
    theme(legend.position = "none")
}

# Create the three plots
p1 <- create_rate_plot(supply_rates, all_hours_h, "ConEd Supply Rates - All Hours, Zone H")
p2 <- create_rate_plot(supply_rates, all_hours_i, "ConEd Supply Rates - All Hours, Zone I")
p3 <- create_rate_plot(supply_rates, all_hours_j, "ConEd Supply Rates - All Hours, Zone J")

# Combine plots using patchwork
p1 / p2 / p3

```

### On-Peak and Off-Peak supply rates plots
- These plots show the on-peak: off-peak spread for each year, as well as the all-hours rate
- This is useful for "just picking" two years for comparison
- We can see that the on-peak: off-peak spread has been increasing over time
  - 2014: January was really expensive
  - 2017: Mild costs overall, and typical summer spike
```{r}

# Filter for a recent year to show the peak/off-peak spread
# Create a color scale using viridis for years 2011-2024
year_colors <- viridis::viridis(14, begin = 0, end = 1, direction = -1) |>
  setNames(2011:2024)


first_year <- 2012
last_year <- 2024

peak_data <- supply_rates |>
  filter(start_year >= first_year & start_year <= last_year)

# Create plot showing peak/off-peak spread with all-hours line
# Create datetime sequence for x-axis
peak_data <- peak_data |>
  mutate(date = make_date(start_year, start_month, 1))

ggplot(peak_data, aes(x = date, group = start_year)) +
  # Add ribbon between on and off peak
  geom_ribbon(aes(ymin = off_peak_h, ymax = on_peak_h, fill = as.factor(start_year)),
              alpha = 0.3) +
  # Add all hours line
  geom_line(aes(y = all_hours_h, color = as.factor(start_year)),
            linewidth = 1) +
  # Formatting
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = scales::label_dollar()) +
  scale_fill_manual(values = year_colors) +
  scale_color_manual(values = year_colors) +
  labs(
    title = paste("ConEd Zone H Supply Rates:", first_year, "-", last_year),
    subtitle = "Showing peak/off-peak spread and all-hours rate",
    x = "",
    y = "Supply Rate ($/kWh)",
    fill = "Year",
    color = "Year"
  ) +
  theme(legend.position = "None")
```

## Delivery Tariffs
### ConEd Delivery Tariffs
```{r}
library(googlesheets4)

url_sheet <- "https://docs.google.com/spreadsheets/d/1HZOZfyuCLaAWvTndVjkhhuZe6Blcp1tcajYwcHBRN3M/edit?gid=1947524045#gid=1947524045"
googlesheets4::gs4_deauth()
tariffs_raw <- read_sheet(url_sheet, sheet = "SC1_ERB2") |>
  mutate(tariff_id = paste0(utility, "_", version)) |>
  select(-source)

# Combine some categories for convenience
tariffs_coned_delivery <- tariffs_raw |>
  # Group by tariff_name, type, and all other columns except name and value
  group_by(
    utility, version, customer_class, tariff_name, domain,
    class, type, variability, season,
    start_month, end_month, period, start_hour, end_hour, tariff_id
  ) %>%
  # Summarize to combine the values and use type as the new name
  summarize(
    name = first(type),  # Use the type as the new name
    value = sum(value),
    .groups = "drop"
  ) |>
  # Apply the name mapping
  mutate(
    name = case_when(
      name == "customer_charge" ~ "Customer Charge",
      name == "merchant_function_rate" ~ "Merchant Function Charge",
      name == "delivery_rate" ~ "Energy Delivery Rate",
      name == "adjustment_rate" ~ "Adjustment Charge",
      name == "program_rate" ~ "Program Charge",
      name == "sales_tax_rate" ~ "Sales Tax Rate",
      TRUE ~ name  # Keep original name if no mapping exists
    )
  )


```

### Assign Tariffs to Month-Hour Grid and Month Grid
```{r}
# Get hourly tariffs
hourly_tariffs_coned_delivery <- tariffs_coned_delivery |> filter(variability == "hourly")
glimpse(hourly_tariffs_coned_delivery)

month_hour_tariff_energy <- assign_tariffs_to_month_hour_grid(hourly_tariffs_coned_delivery)

month_hour_r3 <- month_hour_tariff_energy |> filter(tariff_name == "r3")

month_hour_r4 <- month_hour_tariff_energy |> filter(tariff_name == "r4")

# Expand monthly and yearly tariffs to month grid
monthly_yearly_tariffs <- tariffs_coned_delivery |> filter(variability %in% c("monthly", "yearly"))
month_tariff <- assign_tariffs_to_month_grid(monthly_yearly_tariffs)
head(month_tariff, n=30)

```


### Apply rates INSIDE the db
```{r}

calculate_monthly_load_with_tariffs <- function(path_rs_db,
                                                month_hour_tariff,
                                                month_hour_supply,
                                                tariff_name) {
  # Connect to database
  con <- DBI::dbConnect(duckdb::duckdb(), path_rs_db)

  # Add month_hour_tariff to the db as temporary table
  DBI::dbWriteTable(con, "month_hour_tariff", month_hour_tariff, overwrite = TRUE)
  DBI::dbWriteTable(con, "month_hour_supply", month_hour_supply, overwrite = TRUE)
  # Query using the existing bldg_ids table
  query <- "
SELECT
    nyh.bldg_id,
    nyh.upgrade,
    nyh.month,
    SUM(nyh.\"sum_out.electricity.total.energy_consumption\") AS monthly_elec_consumption,
    SUM(nyh.\"sum_out.electricity.total.energy_consumption\" * mht.rate) AS monthly_tou_energy_charge,
    MAX(nyh.\"sum_out.electricity.total.energy_consumption\" * mht.rate) AS monthly_tou_power_charge,
    SUM(nyh.\"sum_out.electricity.total.energy_consumption\" * mhs.supply_rate) AS monthly_tou_supply_charge

FROM
    NY_hourly AS nyh
JOIN
    bldg_ids AS bid ON nyh.bldg_id = bid.bldg_id
LEFT JOIN
    month_hour_tariff AS mht
ON
    nyh.month = mht.month AND
    nyh.hour = mht.hour
LEFT JOIN
    month_hour_supply AS mhs
ON
    nyh.month = mhs.month AND
    nyh.hour = mhs.hour
GROUP BY
    nyh.bldg_id,
    nyh.upgrade,
    nyh.month
ORDER BY
    nyh.bldg_id,
    nyh.month"

  monthly_load_with_tariffs <- DBI::dbGetQuery(con, query)

  # Close connection
  DBI::dbDisconnect(con)

  monthly_load_with_tariffs <- monthly_load_with_tariffs |>
    mutate(
      utility = "coned",
      tariff_name = tariff_name
    )

  return(monthly_load_with_tariffs)
}

# First, add the building IDs to the database
con <- DBI::dbConnect(duckdb::duckdb(), path_rs_db)
bldg_ids_df <- data.frame(bldg_id = coned_bldg_ids)
DBI::dbWriteTable(con, "bldg_ids", bldg_ids_df, overwrite = TRUE)
DBI::dbDisconnect(con)


# -----------------------------------------
# R3
# R3, 2014
print("calculating monthly_load_r3_2014")
monthly_load_r3_2014 <- calculate_monthly_load_with_tariffs(path_rs_db,
                                                      month_hour_r3,
                                                      month_hour_supply_h_2014,
                                                      "r3")
monthly_load_r3_2014 <- monthly_load_r3_2014 |>
  select(-monthly_tou_power_charge)


# R3, 2018
print("calculating monthly_load_r3_2018")
monthly_load_r3_2018 <- calculate_monthly_load_with_tariffs(path_rs_db,
                                                      month_hour_r3,
                                                      month_hour_supply_h_2018,
                                                      "r3")
monthly_load_r3_2018 <- monthly_load_r3_2018 |>
  select(-monthly_tou_power_charge)

# -----------------------------------------
# R4
# R4, 2018
print("calculating monthly_load_r4_2018")
monthly_load_r4_2018 <- calculate_monthly_load_with_tariffs(path_rs_db,
                                                      month_hour_r4,
                                                      month_hour_supply_h_2018,
                                                      "r4")
monthly_load_r4_2018 <- monthly_load_r4_2018 |>
  select(-monthly_tou_energy_charge)


# R4, 2014
print("calculating monthly_load_r4_2014")
monthly_load_r4_2014 <- calculate_monthly_load_with_tariffs(path_rs_db,
                                                      month_hour_r4,
                                                      month_hour_supply_h_2014,
                                                      "r4")
monthly_load_r4_2014 <- monthly_load_r4_2014 |>
  select(-monthly_tou_energy_charge)


# -----------------------------------------
# R1
print("calculating monthly_load_r1_2014")
monthly_load_r1_2014 <- monthly_load_r3_2014 |>
  left_join(
    month_supply_2014,
    by = c("month" = "month")
  ) |>
  mutate(
    tariff_name = "r1"
  ) |>
  select(bldg_id, upgrade, month, monthly_elec_consumption, utility, supply_rate, tariff_name)


print("calculating monthly_load_r1_2018")
monthly_load_r1_2018 <- monthly_load_r3_2018 |>
  left_join(
    month_supply_2018,
    by = c("month" = "month")
  ) |>
  mutate(
    tariff_name = "r1"
  ) |>
  select(bldg_id, upgrade, month, monthly_elec_consumption, utility, supply_rate, tariff_name)
```

# Calculate Bills
## Monthly Bills

### New Version of Calculate Monthly Bills
```{r}
# First reshape the tariffs data to wide format

tariffs_wide <- month_tariff |>
  pivot_wider(
    id_cols = c(month, utility, version, tariff_name),
    names_from = type,
    values_from = value
  )


calculate_monthly_bills <- function(monthly_load, tariffs_wide, tariff_name) {
  tariffs_wide <- tariffs_wide |>
   filter(tariff_name == tariff_name)


  monthly_bills <- monthly_load |>
    left_join(
      tariffs_wide,
      by = c("month" = "month", "utility" = "utility", "tariff_name" = "tariff_name")
    )

  if (tariff_name == "r1") {
    monthly_bills <- monthly_bills |>
      mutate(
        delivery_charge = monthly_elec_consumption * delivery_rate,
        supply_charge = monthly_elec_consumption * supply_rate,
        merchant_function_charge = monthly_elec_consumption * merchant_function_rate,
        adjustment_charge = monthly_elec_consumption * adjustment_rate,

        total_pretax_bill = delivery_charge + customer_charge + merchant_function_charge + adjustment_charge + supply_charge,

        sales_tax_charge = total_pretax_bill * sales_tax_rate,

        total_bill = total_pretax_bill + sales_tax_charge
      ) |>
      arrange(bldg_id, month, upgrade)
  }

  if (tariff_name == "r3") {
    monthly_bills <- monthly_bills |>
      mutate(
        delivery_charge = monthly_tou_energy_charge,
        supply_charge = monthly_tou_supply_charge,
        merchant_function_charge = monthly_elec_consumption * merchant_function_rate,
        adjustment_charge = monthly_elec_consumption * adjustment_rate,

        total_pretax_bill = delivery_charge + customer_charge + merchant_function_charge + adjustment_charge + supply_charge,

        sales_tax_charge = total_pretax_bill * sales_tax_rate,

        total_bill = total_pretax_bill + sales_tax_charge
      ) |>
      arrange(bldg_id, month, upgrade)
  }

  if (tariff_name == "r4") {
    monthly_bills <- monthly_bills |>
      mutate(
        delivery_charge = monthly_tou_power_charge,
        supply_charge = monthly_tou_supply_charge,
        merchant_function_charge = monthly_elec_consumption * merchant_function_rate,

        # **********
        # NOTE: r4 does not have a adjustment charge  !!!
        adjustment_charge = monthly_elec_consumption * adjustment_rate,
        # **********

        total_pretax_bill = delivery_charge + customer_charge + merchant_function_charge + supply_charge,

        sales_tax_charge = total_pretax_bill * sales_tax_rate,

        total_bill = total_pretax_bill + sales_tax_charge
      ) |>
      arrange(bldg_id, month, upgrade)
  }

  return(monthly_bills)
}

print("calculating monthly_bills_r1_2014")
monthly_bills_r1_2014 <- calculate_monthly_bills(monthly_load_r1_2014, tariffs_wide, "r1")

print("calculating monthly_bills_r3_2014")
monthly_bills_r3_2014 <- calculate_monthly_bills(monthly_load_r3_2014, tariffs_wide, "r3")

print("calculating monthly_bills_r4_2014")
monthly_bills_r4_2014 <- calculate_monthly_bills(monthly_load_r4_2014, tariffs_wide, "r4")


print("calculating monthly_bills_r1_2018")
monthly_bills_r1_2018 <- calculate_monthly_bills(monthly_load_r1_2018, tariffs_wide, "r1")

print("calculating monthly_bills_r3_2018")
monthly_bills_r3_2018 <- calculate_monthly_bills(monthly_load_r3_2018, tariffs_wide, "r3")

print("calculating monthly_bills_r4_2018")
monthly_bills_r4_2018 <- calculate_monthly_bills(monthly_load_r4_2018, tariffs_wide, "r4")

```


## Annual Bills
```{r}

calculate_annual_bills_from_monthly <- function(monthly_bills) {
  monthly_bills |>
    group_by(bldg_id, upgrade, version, tariff_name) |>
    summarize(
      customer_charge = sum(customer_charge),
      delivery_charge = sum(delivery_charge),
      supply_charge = sum(supply_charge),
      merchant_function_charge = sum(merchant_function_charge),
      adjustment_charge = sum(adjustment_charge),
      sales_tax_charge = sum(sales_tax_charge),
      total_bill = sum(total_bill)
    ) |>
    pivot_longer(
      cols = c(customer_charge, delivery_charge, supply_charge, merchant_function_charge, adjustment_charge, sales_tax_charge, total_bill),
      names_to = "charge_type",
      values_to = "amount"
    ) |>
    collect()
}


#annual_bills_r1_2014 <- calculate_annual_bills_from_monthly(monthly_bills_r1_2014)

annual_bills <- bind_rows(
  # 2014 data
  bind_rows(
    calculate_annual_bills_from_monthly(monthly_bills_r1_2014),
    calculate_annual_bills_from_monthly(monthly_bills_r3_2014),
    calculate_annual_bills_from_monthly(monthly_bills_r4_2014)
  ) |>
  mutate(year = 2014),

  # 2018 data
  bind_rows(
    calculate_annual_bills_from_monthly(monthly_bills_r1_2018),
    calculate_annual_bills_from_monthly(monthly_bills_r3_2018),
    calculate_annual_bills_from_monthly(monthly_bills_r4_2018)
  ) |>
  mutate(year = 2018)
)


```


```{r}
# Calculate medians for each rate
median_r1_2014 <- median(annual_bills$amount[annual_bills$charge_type == "total_bill" & annual_bills$tariff_name == "r1" & annual_bills$year == 2014])
median_r3_2014 <- median(annual_bills$amount[annual_bills$charge_type == "total_bill" & annual_bills$tariff_name == "r3" & annual_bills$year == 2014])
median_r4_2014 <- median(annual_bills$amount[annual_bills$charge_type == "total_bill" & annual_bills$tariff_name == "r4" & annual_bills$year == 2014])

median_r1_2018 <- median(annual_bills$amount[annual_bills$charge_type == "total_bill" & annual_bills$tariff_name == "r1" & annual_bills$year == 2018])
median_r3_2018 <- median(annual_bills$amount[annual_bills$charge_type == "total_bill" & annual_bills$tariff_name == "r3" & annual_bills$year == 2018])
median_r4_2018 <- median(annual_bills$amount[annual_bills$charge_type == "total_bill" & annual_bills$tariff_name == "r4" & annual_bills$year == 2018])

# Print medians and ratios
cat("\n2014 Median annual bills:\n")
cat(sprintf("R1: %s\n", scales::dollar(median_r1_2014)))
cat(sprintf("R3: %s\n", scales::dollar(median_r3_2014)))
cat(sprintf("R4: %s\n", scales::dollar(median_r4_2014)))
cat("\nRatios compared to R1:\n")
cat(sprintf("R3/R1: %.2f\n", median_r3_2014/median_r1_2014))
cat(sprintf("R4/R1: %.2f\n", median_r4_2014/median_r1_2014))

# Print medians and ratios
cat("\n2018 Median annual bills:\n")
cat(sprintf("R1: %s\n", scales::dollar(median_r1_2018)))
cat(sprintf("R3: %s\n", scales::dollar(median_r3_2018)))
cat(sprintf("R4: %s\n", scales::dollar(median_r4_2018)))
cat("\nRatios compared to R1:\n")
cat(sprintf("R3/R1: %.2f\n", median_r3_2018/median_r1_2018))
cat(sprintf("R4/R1: %.2f\n", median_r4_2018/median_r1_2018))

```


# Plots
## Single Year, 3 Rates
```{r}
library(patchwork)  # Add this at the top with other library imports

# Tune some plot parameters
min_bill <- 0
max_bill <- 7500

bin_width <- 100
y_max <- 1500
y_text_pos <- 0.88 * y_max

year <- 2014

if (year == 2014) {
  median_r1 <- median_r1_2014
  median_r3 <- median_r3_2014
  median_r4 <- median_r4_2014
} else {
  median_r1 <- median_r1_2018
  median_r3 <- median_r3_2018
  median_r4 <- median_r4_2018
}



p1 <- annual_bills |>
  filter(tariff_name == "r4" & year == year & charge_type == "total_bill") |>
  ggplot(aes(x = amount)) +
  geom_histogram(fill = sb_colors["midnight"], alpha = 0.8, binwidth = bin_width) +
  geom_vline(aes(xintercept = median_r4), color = sb_colors["pistachio"], linewidth = 1) +
  annotate("text",
          x = 1.005 * median_r4,
          hjust = 0.0,
          y = y_text_pos,
          label = paste0(scales::dollar(median_r4), " (", round((1 - median_r4/median_r1) * 100, 1), "% savings)"),
          vjust = -0.5) +
  labs(title = "SC1_R4 - TOU Demand-Based", x = NULL, y = "Homes Count") +
  scale_x_continuous(
    labels = scales::dollar_format(),
    limits = c(min_bill, max_bill)
  ) +
  scale_y_continuous(limits = c(0, y_max)) +
  theme(plot.title = element_text(hjust = 0.5))

p2 <- annual_bills |>
  filter(tariff_name == "r3" & year == year & charge_type == "total_bill") |>
  ggplot(aes(x = amount)) +
  geom_histogram(fill = sb_colors["carrot"], alpha = 0.8, binwidth = bin_width) +
  geom_vline(aes(xintercept = median_r3), color = sb_colors["pistachio"], linewidth = 1) +
  annotate("text",
          x = 1.005 * median_r3,
          hjust = 0.0,
          y = y_text_pos,
          label = paste0(scales::dollar(median_r3), " (", round((1 - median_r3/median_r1) * 100, 1), "% savings)"),
          vjust = -0.5) +
  labs(title = "SC1_R3 - TOU Volumetric", x = NULL, y = "Homes Count") +
  scale_x_continuous(
    labels = scales::dollar_format(),
    limits = c(min_bill, max_bill)
  ) +
  scale_y_continuous(limits = c(0, y_max)) +
  theme(plot.title = element_text(hjust = 0.5))

p3 <- annual_bills |>
  filter(tariff_name == "r1" & year == year & charge_type == "total_bill") |>
  ggplot(aes(x = amount)) +
  geom_histogram(fill = sb_colors["sky"], alpha = 0.8, binwidth = bin_width) +
  geom_vline(aes(xintercept = median_r1), color = sb_colors["pistachio"], linewidth = 1) +
  annotate("text",
          x = 1.005 * median_r1,
          hjust = 0.0,
          y = y_text_pos,
          label = paste0(scales::dollar(median_r1), " median annual bill"),
          vjust = -0.5) +
  labs(title = "SC1_R1 - Flat Volumetric", x = "Annual Bill ($)", y = "Homes Count") +
  scale_x_continuous(
    labels = scales::dollar_format(),
    limits = c(min_bill, max_bill)
  ) +
  scale_y_continuous(limits = c(0, y_max)) +
  theme(plot.title = element_text(hjust = 0.5))

# Combine plots vertically using patchwork
p1 / p2 / p3 +
  plot_layout(heights = c(1, 1, 1)) +
  plot_annotation(
    title = "Annual Bills",
    subtitle = paste("Supply Rates:", year),
    theme = theme(plot.margin = margin(10, 10, 10, 10))
  )

ggsave(
  "src/ny_hp_rates/img/annual_bills_hist.png",
  width = 3,
  height = 4,
  dpi = 300
)

```

## Comparison of 2 Years, 3 Rates
```{r}
library(patchwork)  # Add this at the top with other library imports

# Tune some plot parameters
min_bill <- 0
max_bill <- 7500

bin_width <- 100
y_max <- 1000
y_text_pos <- 0.88 * y_max

year <- 2018

if (year == 2014) {
  median_r1 <- median_r1_2014
  median_r3 <- median_r3_2014
  median_r4 <- median_r4_2014
} else {
  median_r1 <- median_r1_2018
  median_r3 <- median_r3_2018
  median_r4 <- median_r4_2018
}



p1 <- annual_bills |>
  filter(tariff_name == "r4" & year %in% c(2014, 2018) & charge_type == "total_bill") |>
  ggplot(aes(x = amount, fill = as.factor(year))) +
  geom_histogram(position = "identity", alpha = 0.6, binwidth = bin_width) +
  geom_vline(aes(xintercept = median_r4_2014), color = year_colors["2014"], linewidth = 1) +
  geom_vline(aes(xintercept = median_r4_2018), color = year_colors["2018"], linewidth = 1) +
  annotate("text",
          x = 1.005 * median_r4_2014,
          y = y_text_pos,
          hjust = 0,
          label = paste0("2014: ", scales::dollar(median_r4_2014)),
          color = year_colors["2014"]) +
  annotate("text",
          x = 0.99 * median_r4_2018,
          y = y_text_pos * 1.05,
          hjust = 1,
          label = paste0("2018: ", scales::dollar(median_r4_2018)),
          color = year_colors["2018"]) +
  labs(title = "SC1_R4 - TOU Demand-Based", x = NULL, y = "Homes Count") +
  scale_x_continuous(
    labels = scales::dollar_format(),
    limits = c(min_bill, max_bill)
  ) +
  scale_y_continuous(limits = c(0, y_max)) +
  scale_fill_manual(values = year_colors) +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "none")

p2 <- annual_bills |>
  filter(tariff_name == "r3" & year %in% c(2014, 2018) & charge_type == "total_bill") |>
  ggplot(aes(x = amount, fill = as.factor(year))) +
  geom_histogram(position = "identity", alpha = 0.6, binwidth = bin_width) +
  geom_vline(aes(xintercept = median_r3_2014), color = year_colors["2014"], linewidth = 1) +
  geom_vline(aes(xintercept = median_r3_2018), color = year_colors["2018"], linewidth = 1) +
  annotate("text",
          x = 1.005 * median_r3_2014,
          y = y_text_pos,
          hjust = 0,
          label = paste0("2014: ", scales::dollar(median_r3_2014)),
          color = year_colors["2014"]) +
  annotate("text",
          x = 0.99 * median_r3_2018,
          y = y_text_pos * 1.05,
          hjust = 1,
          label = paste0("2018: ", scales::dollar(median_r3_2018)),
          color = year_colors["2018"]) +
  labs(title = "SC1_R3 - TOU Volumetric", x = NULL, y = "Homes Count") +
  scale_x_continuous(
    labels = scales::dollar_format(),
    limits = c(min_bill, max_bill)
  ) +
  scale_y_continuous(limits = c(0, y_max)) +
  scale_fill_manual(values = year_colors) +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "none")

p3 <- annual_bills |>
  filter(tariff_name == "r1" & year %in% c(2014, 2018) & charge_type == "total_bill") |>
  ggplot(aes(x = amount, fill = as.factor(year))) +
  geom_histogram(position = "identity", alpha = 0.6, binwidth = bin_width) +
  geom_vline(aes(xintercept = median_r1_2014), color = year_colors["2014"], linewidth = 1) +
  geom_vline(aes(xintercept = median_r1_2018), color = year_colors["2018"], linewidth = 1) +
  annotate("text",
          x = 1.005 * median_r1_2014,
          y = y_text_pos,
          hjust = 0,
          label = paste0("2014: ", scales::dollar(median_r1_2014)),
          color = year_colors["2014"]) +
  annotate("text",
          x = 0.99 * median_r1_2018,
          y = y_text_pos * 1.05,
          hjust = 1,
          label = paste0("2018: ", scales::dollar(median_r1_2018)),
          color = year_colors["2018"]) +
  labs(title = "SC1_R1 - Flat Volumetric", x = "Annual Bill ($)", y = "Homes Count") +
  scale_x_continuous(
    labels = scales::dollar_format(),
    limits = c(min_bill, max_bill)
  ) +
  scale_y_continuous(limits = c(0, y_max)) +
  scale_fill_manual(values = year_colors) +
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "none")

# Combine plots vertically using patchwork
p1 / p2 / p3 +
  plot_layout(heights = c(1, 1, 1)) +
  plot_annotation(
    title = "Annual Bills",
    subtitle = "Supply Rates: 2014 vs 2018",
    theme = theme(plot.margin = margin(10, 10, 10, 10))
  )

ggsave(
  "src/ny_hp_rates/img/annual_bills_hist_comparison.png",
  width = 3,
  height = 4,
  dpi = 300
)

```


## Bill Stacks

```{r}

# Find bldg_id with median total bill for 2018
median_bldg_2018 <- annual_bills |>
  filter(charge_type == "total_bill", tariff_name == "r1", year == 2018) |>
  ungroup() |>  # Add this to remove any grouping
  arrange(amount) |>
  slice(floor(n()/2)) |>
  pull(bldg_id)

print(paste0("Building ID with median 2018 total bill: ", median_bldg_2018))


# Get unique components in order they appear
charge_types <- rev(unique(annual_bills$charge_type))
print(charge_types)
components <- c("sales_tax_charge", "adjustment_charge", "delivery_charge", "merchant_function_charge", "supply_charge", "customer_charge")


median_bill_2018 <- annual_bills |>
  filter(
    bldg_id == median_bldg_2018,
    charge_type != "total_bill",  # This will give us just the component charges
    year == 2018
  ) |>
  # First make charge_type a factor with our desired order
  mutate(
    charge_type = factor(charge_type, levels = components)
  ) |>
  # Then arrange by this factor to ensure the correct stacking order
  arrange(charge_type)


extra_colors <- c(
  "Deep Ocean" = "#0A4B68",
  "Azure" = "#2A6D8A",
  "Cerulean" = "#4996B2",
  "Grey" = "#808080"
)

bill_colors <- c(
  "delivery_charge" = sb_colors[["carrot"]],
  "supply_charge" = extra_colors[["Cerulean"]],
  "merchant_function_charge" = extra_colors[["Azure"]],
  "adjustment_charge" = extra_colors[["Deep Ocean"]],
  "sales_tax_charge" = sb_colors[["pistachio"]],
  "customer_charge" = extra_colors[["Grey"]]
)

legend_aliases <- c(
  "delivery_charge" = "Delivery Charge",
  "supply_charge" = "Supply Charge",
  "merchant_function_charge" = "Merchant Function Charge",
  "adjustment_charge" = "Adjustment Charge",
  "sales_tax_charge" = "Sales Tax Charge",
  "customer_charge" = "Customer Charge"
)


# First, ensure charge_type is a factor with the correct order
median_bill_2018 <- median_bill_2018 %>%
  mutate(charge_type = factor(charge_type, levels = components))


# Plot parameters
y_max <- 2500
plot_components <- rev(components)
y_annotation_offset <- 50

supply_period <- "2018"


for (i in 0:length(plot_components)) {
  # Take only first i components
  subset_bills <- median_bill_2018 %>%
    filter(charge_type %in% plot_components[0:i])

  # Calculate subtotals for each rate up to current component
  r1_component <- subset_bills$amount[subset_bills$tariff_name == "r1"]
  r1_subtotal <- sum(subset_bills$amount[subset_bills$tariff_name == "r1"])

  r3_component <- subset_bills$amount[subset_bills$tariff_name == "r3"]
  r3_subtotal <- sum(subset_bills$amount[subset_bills$tariff_name == "r3"])

  r4_component <- subset_bills$amount[subset_bills$tariff_name == "r4"]
  r4_subtotal <- sum(subset_bills$amount[subset_bills$tariff_name == "r4"])

  # Create a dummy dataset with all components to set consistent legend space
  dummy_data <- median_bill_2018 %>%
    complete(tariff_name, charge_type) %>%
    replace_na(list(amount = 0))

  ggplot() +
    # Plot the dummy data with zero alpha to reserve legend space
    geom_bar(data = dummy_data,
             aes(x = factor(tariff_name, levels = c("r1", "r3", "r4")),
                 y = amount, fill = charge_type),
             stat = "identity", position = "stack", width = 0.6,
             alpha = 0) +
    # Plot the actual subset data
    geom_bar(data = subset_bills,
             aes(x = factor(tariff_name, levels = c("r1", "r3", "r4")),
                 y = amount, fill = charge_type),
             stat = "identity", position = "stack", width = 0.6) +
    scale_fill_manual(
      values = bill_colors,
      breaks = components,  # This ensures legend order matches stacking order
      labels = legend_aliases[components]
    ) +
    scale_x_discrete(labels = c("r1" = "R1 - Flat Volumetric", "r3" = "R3 - TOU Volumetric", "r4" = "R4 - TOU Demand-Based")) +
    scale_y_continuous(limits = c(0, y_max), labels = scales::dollar_format()) +
    labs(
      title = "Annual Bill Comparison: r1 vs r3 vs r4",
      subtitle = paste("Supply Period:", supply_period),
      x = "",
      y = "Median Annual Bill ($)",
      fill = "Bill Component"
    ) +
    theme(
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      legend.position = "right",
      legend.box = "vertical",
      legend.background = element_rect(fill = "grey90")
    ) +
    # Add subtotal annotations for current components
    annotate(
      "text",
      x = "r1",
      y = (r1_subtotal + y_annotation_offset),
      label = paste0("$", format(round(r1_subtotal, 0), big.mark=",")),
    ) +
    annotate(
      "text",
      x = "r3",
      y = (r3_subtotal + y_annotation_offset),
      label = paste0("$", format(round(r3_subtotal, 0), big.mark=",")),
    ) +
    annotate(
      "text",
      x = "r4",
      y = (r4_subtotal + y_annotation_offset),
      label = paste0("$", format(round(r4_subtotal, 0), big.mark=",")),
    )

  ggsave(
    paste0("src/ny_hp_rates/img/bills_annual_stacked_", i, ".png"),
    width = 4,
    height = 3,
    dpi = 300
  )
}

```


## Plot of hourly loads

```{r}
# Calculate statistics for January
january_stats <- hourly_loads |>
  filter(month == 1) |>
  group_by(hour) |>
  summarize(
    mean_load = mean(mh_energy_consumption),
    p05_load = quantile(mh_energy_consumption, 0.05),
    p25_load = quantile(mh_energy_consumption, 0.25),
    p75_load = quantile(mh_energy_consumption, 0.75),
    p95_load = quantile(mh_energy_consumption, 0.95)
  )

january_rates <- month_hour_tariff_energy |>
  filter(month == 1, tariff_name %in% c("r3"))

# Calculate statistics for August
august_stats <- hourly_loads |>
  filter(month == 8) |>
  group_by(hour) |>
  summarize(
    mean_load = mean(mh_energy_consumption),
    p05_load = quantile(mh_energy_consumption, 0.05),
    p25_load = quantile(mh_energy_consumption, 0.25),
    p75_load = quantile(mh_energy_consumption, 0.75),
    p95_load = quantile(mh_energy_consumption, 0.95)
  )

august_rates <- month_hour_tariff_energy |>
  filter(month == 8, tariff_name %in% c("r3"))

# -----------------------------------------
# Plotting

# Create load profile plot
create_load_profile_plot <- function(stats, rates, month_name, y2_max = 0.5, show_rates = TRUE) {
  # Find the maximum value for the primary y-axis
  y_max <- max(stats$p95_load)

  # Create the transformation coefficient
  # This is the key fix: we need to create a transformation that maps from the primary y-axis to the secondary y-axis
  coef <- y_max / y2_max

  p <- ggplot(stats, aes(x = hour)) +
    # Load profile elements
    geom_ribbon(aes(ymin = p05_load, ymax = p95_load),
                fill = "grey80", alpha = 0.5) +
    geom_ribbon(aes(ymin = p25_load, ymax = p75_load),
                fill = "grey60", alpha = 0.5) +
    geom_line(aes(y = mean_load),
              color = "darkblue", linewidth = 1)

  if (show_rates) {
    # The key fix: we need to scale the rate values to match the primary y-axis
    rates_scaled <- rates
    rates_scaled$rate_scaled <- rates$rate * coef

    p <- p +
      # Plot rates with scaled values
      geom_line(data = rates_scaled,
                aes(y = rate_scaled, color = tariff_name),
                linewidth = 1, linetype = "dashed") +
      scale_y_continuous(
        labels = scales::label_number(suffix = " kWh"),
        # Use sec.axis with transformation
        sec.axis = sec_axis(
          ~./coef,  # Transform back to original rate values
          name = "Rate ($/kWh)",
          labels = scales::label_dollar()
        ),
        # Set the limits for the primary y-axis
        limits = c(0, y_max)
      ) +
      scale_color_manual(
        values = c("r3" = "red", "r4" = "orange"),
        labels = c("r3" = "R3 Rate", "r4" = "R4 Rate"),
        name = "Tariff"
      ) +
      theme(
        axis.title.y.right = element_text(color = "red"),
        axis.text.y.right = element_text(color = "red"),
        legend.position = "bottom"
      )
  }

  p + labs(
    title = paste(month_name, "Load Profile", if(show_rates) "and TOU Rates" else ""),
    x = "Hour of Day",
    y = "Energy Consumption"
  ) +
  theme(
    panel.grid.minor = element_blank()
  )
}

# Usage:
p1_with_rates <- create_load_profile_plot(january_stats, january_rates, "January", y2_max = 0.4, show_rates = TRUE)
p2_with_rates <- create_load_profile_plot(august_stats, august_rates, "August", y2_max = 0.4, show_rates = TRUE)

p1_without_rates <- create_load_profile_plot(january_stats, january_rates, "January", y2_max = 0.4, show_rates = FALSE)
p2_without_rates <- create_load_profile_plot(august_stats, august_rates, "August", y2_max = 0.4, show_rates = FALSE)

# Combine plots
p1_with_rates + p2_with_rates +
  plot_annotation(
    title = "Hourly Load Profiles: Winter vs Summer",
    subtitle = "Showing mean (line) and percentile ranges (25-75% darker, 5-95% lighter)"
  )

p1_without_rates + p2_without_rates +
  plot_annotation(
    title = "Hourly Load Profiles: Winter vs Summer",
    subtitle = "Showing mean (line) and percentile ranges (25-75% darker, 5-95% lighter)"
  )
```
