---
title: Who saves when switching to heat pumps, and why?

toc: true
reference-location: margin
fig-cap-location: bottom

appendix-style: default
citation-location: document
citation:
  container-title: Switchbox

format:
  html:
    page-layout: full
---

# ------------------------------------------------------------------------------
# Setup and Data Loading
# ------------------------------------------------------------------------------
## Setup
```{r}
#| label: setup

# =============================================================================
# PATH DEFINITIONS
# =============================================================================
# Project root
repo_root <- "/workspaces/reports2"

# Base directory paths
path_to_lib <- file.path(repo_root, "lib")
path_to_data <- file.path(repo_root, "data")
path_to_report <- file.path(repo_root, "reports", "hp_who_saves")
path_to_cache <- file.path(path_to_report, "cache")

# Library paths
path_switchbox_theme <- file.path(path_to_lib, "ggplot", "switchbox_theme.R")
path_heat_pump_funcs <- file.path(path_to_lib, "rates_analysis", "heat_pump_rate_funcs.R")
path_heat_pump_plots <- file.path(path_to_lib, "rates_analysis", "heat_pump_rate_plots.R")
path_create_housing_units <- file.path(path_to_lib, "rates_analysis", "create_sb_housing_units.R")

# Data paths
path_monthly_data <- file.path(path_to_data, "resstock", "2024_release2_tmy3", "load_curve_monthly")
path_metadata_dir <- file.path(path_to_data, "resstock", "2024_release2_tmy3", "metadata")
path_fuel_oil_supply_rates <- file.path(path_to_data, "eia", "heating_oil", "ri_eia_heating_oil_prices_monthly.parquet")
path_propane_supply_rates <- file.path(path_to_data, "eia", "propane", "ri_eia_propane_prices_monthly.parquet")

# Cache paths
path_monthly_consumption <- file.path(path_to_cache, "monthly_consumption.RData")
path_calculated_bills_electric <- file.path(path_to_cache, "calculated_bills_electric.RData")
path_calculated_bills_fuel_oil <- file.path(path_to_cache, "calculated_bills_fuel_oil.RData")
path_calculated_bills_propane <- file.path(path_to_cache, "calculated_bills_propane.RData")
path_calculated_bills_gas <- file.path(path_to_cache, "calculated_bills_gas.RData")
path_report_variables <- file.path(path_to_cache, "report_variables.RData")

# =============================================================================
# LIBRARY LOADING
# =============================================================================
library(googlesheets4)
library(arrow)
library(dplyr)
library(ggplot2)
library(GGally)
library(lubridate)
library(patchwork)
library(ggrepel)

# Switchbox Libraries
source(path_switchbox_theme)
source(path_heat_pump_funcs)
source(path_heat_pump_plots)
source(path_create_housing_units)
```

## Parameters
```{r}
#| label: parameters
supply_year = 2024
round_percent = 1
round_dollars <- 0

# Define which monthly data and which upgrades to use
# path_monthly_data <- file.path(project_root, "data", "resstock", "2024_release2_tmy3", "load_curve_monthly_10")
use_these_upgrades <- c("00", "01", "02", "03")

use_these_states <- c("RI")

# Define seasons
winter_months <- c(1,2,3,10,11,12)
summer_months <- c(4,5,6,7,8,9)

hp_option <- "hp_high"

if (hp_option == "hp_low") {
  upgrade_option <- 1
} else if (hp_option == "hp_high") {
  upgrade_option <- 2
} else if (hp_option == "hp_best") {
  upgrade_option <- 3
}

# Tariffs and other sheets from GDrive
# Load Delivery Rates and associated LMI Thresholds
url_delivery_rates <- "https://docs.google.com/spreadsheets/d/17mTVgMnZaROOimWJBaVp-M-p5P96bK9g_iyjkKTJ2Mk/edit?gid=1792930160#gid=1792930160"

# SMI thresholds
url_smi_thresholds <- "https://docs.google.com/spreadsheets/d/1bVbr8jDF8YgT6-yiEEpg-RFvMdV7veoapFzhZYVcaWg/edit?gid=0#gid=0"

load_tariffs_from_cache <- TRUE
load_thresholds_from_cache <- TRUE
load_monthly_consumption_from_cache <- TRUE

```

## Load ri_hp_rates report variables
```{r}
#| label: load ri_hp_rates report variables from cache

# if this isn't found...
# cd ../ri_hp_rates
# just all
path_ri_hp_rates_results_cache <- file.path(repo_root, "reports", "ri_hp_rates", "cache", "report_variables.RData")
print(paste0("Loading ri_hp_rates report variables from: ", path_ri_hp_rates_results_cache))
load(file = path_ri_hp_rates_results_cache)

```


## Color Definitions
```{r}
#| label: color definitions
scenario_colors = c(
  "baseline_rates_current_hvac" = "grey50",

  "baseline" = "#023047",
  "baseline_rates_low_hp" = "#023047",
  "baseline_rates_mid_hp" = "#023047",
  "baseline_rates_high_hp" = "#023047",

  "hpr_approved" = "#FC9706",
  "approved_low_hp" = "#FC9706",
  "approved_mid_hp" = "#FC9706",
  "approved_high_hp" = "#FC9706",

  "hpr_doer" = "#68BED8",
  "doer_low_hp" = "#68BED8",
  "doer_mid_hp" = "#68BED8",
  "doer_high_hp" = "#68BED8"
)

heating_type_colors = c(
  "Natural Gas" = "#8C510A",        # brown
  "Fuel Oil" = "#D8B365",           # light brown
  "Propane" = "#F6E8C3",            # very light brown
  "Electric Resistance" = "#5AB4AC",         # BrBG palette
  "Heat Pump" = "#80CDC1",          # BrBG palette
  "Other/None" = "grey50"          # BrBG palette
)

cooling_type_colors = c(
  "Central AC" = "#8C510A",        # brown
  "Room AC" = "#D8B365",           # light brown
  "Heat Pump" = "#80CDC1",            # very light brown
  "Other Cooling" = "#5AB4AC",         # BrBG palette
  "None" = "grey50"        # BrBG palette
)

building_type_colors = c(
  "Single-Family" = "#234267",    # darkest dusty blue
  "2-4 Units" = "#6383A9",        # medium dusty blue
  "5+ Units" = "#B2C7DF",         # lightest dusty blue
  "Other" = "#80CDC1"             # BrBG palette for other
)

occupants_group_colors = c(
  "Vacant" = "#424242",        # dark grey
  "Single" = "#D1BFE6",        # light dusty purple
  "Couple" = "#A285C8",        # medium dusty purple
  "3-4 Occupants" = "#7C579A", # darker dusty purple
  "5+ Occupants" = "#4C266A"   # darkest dusty purple
)

smi_tier_colors = c(
  "No Income" = "#424242",             # dark grey
  "Low Income" = "#B4C6A6",            # light dusty green
  "Moderate Income" = "#7D9D81",       # medium dusty green
  "Not LMI" = "#475D4B"            # dark dusty green
)

smi_tier_label_colors = c(
  "No Income" = "#424242",             # dark grey
  "Low Income (<60% SMI)" = "#B4C6A6",            # light dusty green
  "Moderate Income (60–100% SMI)" = "#7D9D81",       # medium dusty green
  "Not LMI (>100% SMI)" = "#475D4B"            # dark dusty green
)

dollar_tier_colors = c(
  "No Income" = "#424242",             # dark grey
  "Low Income" = "#B4C6A6",            # light dusty green
  "Moderate Income" = "#7D9D81",       # medium dusty green
  "Not LMI" = "#475D4B"            # dark dusty green
)

bill_component_colors = c(
  "annual_gas_bill" = "grey50", # grey
  "annual_fuel_oil_bill" = "grey50", # grey
  "supply_charge" = "#A0AF12",
  "delivery_charge" = "#023047",
  "customer_charge" = "#68BED8",
  "pre_hp_delivery_total" = "#FC9706"
)

savings_colors = c(
  "dark_red" = "#db8b87", # dark red
  "light_red" = "#eaada9", # light red
  "light_green" = "#8ebd85", # light green
  "dark_green" = "#5e8a5e" # dark green
)

energy_burden_colors = c(
  "low_energy_burden" = "#5e8a5e", # dark green
  "moderate_energy_burden" = "#8ebd85", # light green
  "high_energy_burden" = "#eaada9", # light red
  "very_high_energy_burden" = "#db8b87" # dark red
)

```

## def load_or_download function
```{r}
#| label: load delivery tariffs and supply rates

# Delivery Tariffs and LMI Discount Thresholds
load_or_download <- function(table_name, url, path_to_cache, load_from_cache=FALSE) {
  # Construct file path for this table
  file_path <- paste0(path_to_cache, "/", table_name, ".RData")

  # Try to load from cache first
  if (load_from_cache && file.exists(file_path)) {
    print(paste0("Attempting to load from cache: ", file_path))
    load(file_path, envir = .GlobalEnv)
    assign(table_name, data, envir = .GlobalEnv)
    print(paste0("    Loaded cached ", table_name))

  } else {
    # Download if not found
    googlesheets4::gs4_deauth()
    print(paste0("    Downloading ", table_name))
    # Read the sheet and throw out columns-for-humans
    data <- read_sheet(url, sheet = table_name) |>
      select(-source, -notes)

    # if billing_unit is anything other than "dollars", "dollars_per_kwh", or "percentage", we convert it to "dollars_per_kwh"
    if (str_detect(table_name, "tariffs")) {
      print("    ...converting billing_unit from dollar_per_therm to dollars_per_kwh")
      data <- data |>
        mutate(
          value = case_when(
            billing_unit == "dollars_per_therm" ~ value / 29.3,
            TRUE ~ value
          ),
          billing_unit = case_when(
            billing_unit == "dollars_per_therm" ~ "dollars_per_kwh",
            TRUE ~ billing_unit
          )
        )
    }

    # tariffs are specified by season, so we need to expand them out to the monthly grid
    if (str_detect(table_name, "tariffs")) {
      print("    ...expanding tariffs out to the monthly grid")
      # The tariffs are specified by season, so we need to expand them out to the monthly grid
      data <- data |>
      # Create a mapping of months to each season
      cross_join(
        tibble(
          month = 1:12,
          season_mapping = case_when(
            month %in% winter_months ~ "winter",
            month %in% summer_months ~ "summer"
          )
        )
      ) |>
      # Keep only the rows where the season matches the month's season
      filter(season == season_mapping) |>
      # Convert month to int32
      mutate(month = as.integer(month)) |>
      # Remove the helper columns
      select(-season, -season_mapping) |>
      # Arrange for better organization
      arrange(class, type, tariff_name, month)
    }


    # if table_name is gas_delivery_tariffs, we need combine several rows to get an all-in delivery rate
    if (str_detect(table_name, "gas_delivery_tariffs")) {
      print("    ...combining several rows to get an all-in delivery rate")


      # let's add a "bill_component" column to the data based on the class and type columns
      # fixed = everything where class == "fixed"
      # delivery = everything where class == "volumetric", expect if type == "supply"
      # tax = everything where type == "tax"
      data <- data |>
        mutate(bill_component = case_when(
          class == "fixed" ~ "customer_charge",
          class == "volumetric" & type != "supply" ~ "delivery_rate",
          class == "volumetric" & type == "supply" ~ "supply_rate",
          class == "tax" ~ "sales_tax_rate",
          TRUE ~ "other"
        ))

      # now let's group by the bill_component column
      data <- data |>
        group_by(bill_component, tariff_name, gas_utility, heat_non_heat, version, month) |>
        summarise(
          value = sum(value, na.rm = TRUE),
          type = first(bill_component),
          # Keep other columns from the first row of each group
          class = first(class),
          customer_class = first(customer_class),
          tariff_name = first(tariff_name),
          billing_unit = first(billing_unit),
          lmi = first(lmi),
          effective_date = first(effective_date),
          .groups = "drop"
        ) |>
        select(-bill_component)
    }


    # supply rates are specified by year
    if (str_detect(table_name, "supply")) {
      print("    ...converting supply rates to int32")
      data <- data |>
        mutate(
          month = as.integer(month),
          year = as.integer(year)
        )
    }


    # Save to cache
    save(data, file = file_path)
    assign(table_name, data, envir = .GlobalEnv)
    print(paste0("    Downloaded and cached ", table_name))
  }
}
```

## Housing Units
### Load and Transform Housing Units
```{r}
#| label: load housing units
use_these_in_columns <- c(
  "bldg_id",
  "upgrade",
  "in.sqft",
  "in.representative_income",
  "in.income",
  "in.city",
  "in.geometry_building_type_acs",
  "in.heating_fuel",
  "in.hvac_cooling_type",
  "in.hvac_heating_efficiency",
  "in.hvac_heating_type_and_fuel",
  "in.occupants",
  "in.state",
  "in.vintage",
  "in.weather_file_city",
  "in.vacancy_status"
)

use_these_out_columns <- c(
  "out.electricity.total.energy_consumption.kwh",
  "out.fuel_oil.total.energy_consumption.kwh",
  "out.natural_gas.total.energy_consumption.kwh",
  "out.propane.total.energy_consumption.kwh"
)

# Create pattern to match only the desired upgrade files
upgrade_patterns <- paste0("upgrade", use_these_upgrades, "_metadata_and_annual_results\\.parquet$")
upgrade_pattern <- paste(upgrade_patterns, collapse = "|")

# Get only the upgrade parquet files that match our criteria
parquet_files <- list.files(
  path_metadata_dir,
  pattern = upgrade_pattern,
  full.names = TRUE
)

# Create a lazy dataset from selected parquet files
housing_units <- open_dataset(parquet_files) |>
  select(all_of(use_these_in_columns), all_of(use_these_out_columns)) |>
  filter(`in.state` %in% use_these_states) |>
  collect()

#-----------------------------------------------------------------
# Simple stuff: trim to a max of 4000 sqft
#-----------------------------------------------------------------
housing_units <- housing_units |>
  filter(in.sqft <= 5000)

#-----------------------------------------------------------------
# Preferred labels for heating type, cooling type, building type, and occupants group
#-----------------------------------------------------------------
# Add baseline heating type and set as factor
housing_units <- get_baseline_heating_type(housing_units) |> right_join(housing_units, by = c("bldg_id"))
baseline_heating_type_levels <- c("Natural Gas", "Fuel Oil", "Propane", "Electric Resistance", "Heat Pump", "Other/None")
housing_units$baseline_heating_type <- forcats::fct(housing_units$baseline_heating_type, levels = baseline_heating_type_levels)

# Add baseline heating efficiency and set as factor
housing_units <- get_baseline_heating_efficiency(housing_units) |> right_join(housing_units, by = c("bldg_id"))
baseline_heating_efficiency_levels <- c("ASHP, SEER 10, 6.2 HSPF", "ASHP, SEER 13, 7.7 HSPF", "ASHP, SEER 15, 8.5 HSPF", "Electric Baseboard, 100% Efficiency", "Electric Boiler, 100% AFUE", "Electric Furnace, 100% AFUE", "Electric Wall Furnace, 100% AFUE", "Fuel Boiler, 76% AFUE", "Fuel Boiler, 80% AFUE", "Fuel Boiler, 90% AFUE", "Fuel Furnace, 60% AFUE", "Fuel Furnace, 76% AFUE", "Fuel Furnace, 80% AFUE", "Fuel Furnace, 92.5% AFUE", "Fuel Wall/Floor Furnace, 60% AFUE", "Fuel Wall/Floor Furnace, 68% AFUE", "MSHP, SEER 14.5, 8.2 HSPF", "MSHP, SEER 29.3, 14 HSPF", "None", "Shared Heating", "Other")
housing_units$baseline_heating_efficiency <- forcats::fct(housing_units$baseline_heating_efficiency, levels = baseline_heating_efficiency_levels)

# Add baseline cooling type and set as factor
housing_units <- get_baseline_cooling_type(housing_units) |> right_join(housing_units, by = c("bldg_id"))
baseline_cooling_type_levels <- c("Room AC", "Central AC",  "Heat Pump", "Other Cooling", "None")
housing_units$baseline_cooling_type <- forcats::fct(housing_units$baseline_cooling_type, levels = baseline_cooling_type_levels)

# Add building type group
housing_units <- add_building_type_group(housing_units) |> right_join(housing_units, by = c("bldg_id"))
building_type_group_levels <- c("Single-Family", "2-4 Units", "5+ Units", "Other")
housing_units$building_type_group <- forcats::fct(housing_units$building_type_group, levels = building_type_group_levels)


# Add occupants group
housing_units <- change_occupants_to_number(housing_units) # changes occupants to number
housing_units <- add_occupants_group(housing_units) |> right_join(housing_units, by = c("bldg_id"))
occupants_group_levels <- c("Vacant", "Single", "Couple", "3-4 Occupants", "5+ Occupants")
housing_units$occupants_group <- forcats::fct(housing_units$occupants_group, levels = occupants_group_levels)

housing_units <- add_hvac_primary(housing_units) |> right_join(housing_units, by = c("bldg_id", "upgrade"))
hvac_primary_levels <- c(baseline_heating_type_levels, "current", "hp_low", "hp_high", "hp_best", "hp_geo")
housing_units$hvac_primary <- forcats::fct(housing_units$hvac_primary, levels = hvac_primary_levels)

housing_units <- add_hvac_backup(housing_units) |> right_join(housing_units, by = c("bldg_id", "upgrade"))
hvac_backup_levels <- c(baseline_heating_type_levels, "electric_resistance", "none")
housing_units$hvac_backup <- forcats::fct(housing_units$hvac_backup, levels = hvac_backup_levels)

housing_units <- add_appliances(housing_units) |> right_join(housing_units, by = c("bldg_id", "upgrade"))
appliances_levels <- c("Baseline", "All Electric")
housing_units$appliances <- forcats::fct(housing_units$appliances, levels = appliances_levels)

housing_units <- add_shell(housing_units) |> right_join(housing_units, by = c("bldg_id", "upgrade"))
shell_levels <- c("Baseline", "All Electric")
housing_units$shell <- forcats::fct(housing_units$shell, levels = shell_levels)

# Add heat_non_heat flag
housing_units <- add_heat_non_heat_flag(housing_units) |> right_join(housing_units, by = c("bldg_id", "upgrade"))
heat_non_heat_levels <- c("heat", "non_heat")
housing_units$heat_non_heat <- forcats::fct(housing_units$heat_non_heat, levels = heat_non_heat_levels)

#-----------------------------------------------------------------
# Income
#-----------------------------------------------------------------

# Inflate income to 2024
housing_units <- inflate_income_to_2024(housing_units, from_year = 2019)

# 2024 Rhode Island state median income is $92,290 (https://fred.stlouisfed.org/series/MEHOINUSRIA646N)
# 60% -> $55,374
# 80% -> $73,832
# 100% -> $92,290
housing_units <- group_income_by_dollars(housing_units, dollar_tiers = c(1000, 55374, 92290))
dollar_tier_levels <- c("No Income", "Low Income", "Moderate Income", "Not LMI")
housing_units$dollar_tier <- forcats::fct(housing_units$dollar_tier, levels = dollar_tier_levels)

# Add SMI tiers
housing_units <- group_income_by_smi(housing_units, table_name = "smi_thresholds_2024", url_smi_thresholds = url_smi_thresholds, smi_tiers = c(0.6, 1.0))
smi_tier_levels <- c("No Income", "Low Income", "Moderate Income", "Not LMI")
housing_units$smi_tier <- forcats::fct(housing_units$smi_tier, levels = smi_tier_levels)

#-----------------------------------------------------------------
# Extensions: add elec and gas utility assignments, then LMI discounts
#-----------------------------------------------------------------
# Assign electricity utility
housing_units <- assign_utilities(housing_units)
# Add LMI discounts
load_or_download("electric_lmi_thresholds", url_delivery_rates, path_to_cache, load_from_cache = load_thresholds_from_cache)
load_or_download("gas_lmi_thresholds", url_delivery_rates, path_to_cache, load_from_cache = load_thresholds_from_cache)
housing_units <- add_lmi_discount(housing_units, electric_lmi_thresholds, gas_lmi_thresholds)


# Add fuel oil utility
housing_units <- housing_units |>
  mutate(
    fuel_oil_utility = "generic_retail",
    discount_rate_fuel_oil = 0
  )

# Add propane utility
housing_units <- housing_units |>
  mutate(
    propane_utility = "generic_retail",
    discount_rate_propane = 0
  )


# Print column names of housing_units
print("Column names in housing_units:")
print(colnames(housing_units))

n_total_homes <- housing_units |> filter(upgrade == 0) |> nrow()

housing_units <- housing_units |>
  filter(occupants_group != "Vacant")

n_non_vacant_homes <- housing_units |> filter(upgrade == 0) |> filter(occupants_group != "Vacant") |> nrow()
```

# ------------------------------------------------------------------------------
# Load Supply Rates and Delivery Tariffs
# ------------------------------------------------------------------------------
```{r}
#| label: supply rates and delivery tariffs
# Electric
url_electric_supply_rates <- "https://docs.google.com/spreadsheets/d/1oq0WxARD6cY6gPE4s2TH59O4pHY0sreZNKpHFayn36g/edit?gid=0#gid=0"
# TODO return the data, do not save as a side effect.
load_or_download("electric_supply_rates", url_electric_supply_rates, path_to_cache, load_from_cache = load_tariffs_from_cache)
load_or_download("electric_delivery_tariffs", url_delivery_rates, path_to_cache, load_from_cache = load_tariffs_from_cache)

# Fuel Oil Supply Rates
fuel_oil_supply_rates <- open_dataset(path_fuel_oil_supply_rates) |>
  select(year, month, supply_rate, fuel_oil_utility) |>
  filter(year == supply_year) |>
  collect()

# Delivery Tariffs
load_or_download("fuel_oil_delivery_tariffs", url_delivery_rates, path_to_cache, load_from_cache = load_tariffs_from_cache)
load_or_download("fuel_oil_lmi_thresholds", url_delivery_rates, path_to_cache, load_from_cache = load_tariffs_from_cache)

# Propane Supply Rates
propane_supply_rates <- open_dataset(path_propane_supply_rates) |>
  select(year, month, supply_rate, propane_utility) |>
  filter(year == supply_year) |>
  collect()

# Delivery Tariffs
load_or_download("propane_delivery_tariffs", url_delivery_rates, path_to_cache, load_from_cache = load_tariffs_from_cache)
load_or_download("propane_lmi_thresholds", url_delivery_rates, path_to_cache, load_from_cache = load_tariffs_from_cache)


# Gas Supply Rates
url_gas_supply_rates <- "https://docs.google.com/spreadsheets/d/1_2u9S3r2JWKREhD_pnWXfyQKqjc4IsImbbTdHHN7xqM/edit?gid=0#gid=0"
load_or_download("gas_supply_rates", url_gas_supply_rates, path_to_cache, load_from_cache = load_tariffs_from_cache)
load_or_download("gas_delivery_tariffs", url_delivery_rates, path_to_cache, load_from_cache = load_tariffs_from_cache)
load_or_download("gas_lmi_thresholds", url_delivery_rates, path_to_cache, load_from_cache = load_tariffs_from_cache)

```

# ------------------------------------------------------------------------------
# Heating Bills
# ------------------------------------------------------------------------------
## parameters
```{r}
cols_from_housing_units <- c("bldg_id", "upgrade", "hvac_primary", "baseline_heating_type", "baseline_heating_efficiency")
```

## Consumption by End Use Group
```{r}
#| label: consumption by end use group
if (file.exists(path_monthly_consumption)) {
  load(file = path_monthly_consumption)
} else {

  # Get the end use groups
  path_end_use_groups_feather <- "/workspaces/reports2/lib/resstock/2024/end_use_groups.feather"
  end_use_groups <- arrow::read_feather(path_end_use_groups_feather)

  # Monthly Load Data
  monthly_consumption <- get_monthly_consumption_many(
    path_monthly_data = path_monthly_data,
    fuels = c("electricity", "natural_gas", "fuel_oil", "propane"),
    functional_groups = c("appliance", "cooking", "cooling", "heating", "hot_water", "total"),
    use_these_states = "RI",
    use_these_upgrades = use_these_upgrades
  )

  # Filter out any vacant homes
  monthly_consumption <- monthly_consumption |>
    left_join(housing_units |> filter(upgrade == 0)|> select(bldg_id, occupants_group, baseline_heating_type), by = c("bldg_id")) |>
    filter(occupants_group != "Vacant") |>
    select(-occupants_group)

  # Save to cache
  save(monthly_consumption, file = path_monthly_consumption)
}


#View(monthly_consumption)


```


# Heating Bill Calculations
```{r}
#| label: heating bills

# Monthly Bills
# Electricity
monthly_bills_heating_elec <- calc_monthly_bills(
  monthly_consumption = monthly_consumption,
  consumption_column = "electricity_heating_consumption_kwh",
  fuel_type = "electricity",
  delivery_tariffs = electric_delivery_tariffs,
  supply_rates = electric_supply_rates,
  housing_units = housing_units,
  supply_year = supply_year,
  state = "RI",
  enable_lmi_discount = TRUE,
  use_these_upgrades = use_these_upgrades
) |>
mutate(
  volumetric_rate = delivery_rate + supply_rate,
  volumetric_bill = !!sym("electricity_heating_consumption_kwh") * volumetric_rate
)

monthly_changes_heating_elec <- calc_monthly_changes(monthly_bills_heating_elec, "electricity")

annual_changes_heating_elec <- calc_annual_change_from_monthly(monthly_changes_heating_elec, "electricity")

# Gas
monthly_bills_heating_gas <- calc_monthly_bills(
  monthly_consumption = monthly_consumption,
  consumption_column = "natural_gas_heating_consumption_kwh",
  fuel_type = "natural_gas",
  delivery_tariffs = gas_delivery_tariffs,
  supply_rates = gas_supply_rates,
  housing_units = housing_units,
  supply_year = supply_year,
  state = "RI",
  enable_lmi_discount = TRUE,
  use_these_upgrades = use_these_upgrades
) |>
mutate(
  volumetric_rate = delivery_rate + supply_rate,
  volumetric_bill = !!sym("natural_gas_heating_consumption_kwh") * volumetric_rate
)

monthly_changes_heating_gas <- calc_monthly_changes(monthly_bills_heating_gas, "natural_gas")

annual_changes_heating_gas <- calc_annual_change_from_monthly(monthly_changes_heating_gas, "natural_gas")

# Fuel Oil
monthly_bills_heating_fuel_oil <- calc_monthly_bills(
  monthly_consumption = monthly_consumption,
  consumption_column = "fuel_oil_heating_consumption_kwh",
  fuel_type = "fuel_oil",
  delivery_tariffs = fuel_oil_delivery_tariffs,
  supply_rates = fuel_oil_supply_rates,
  housing_units = housing_units,
  supply_year = supply_year,
  state = "RI",
  enable_lmi_discount = TRUE,
  use_these_upgrades = use_these_upgrades
) |>
mutate(
  volumetric_rate = delivery_rate + supply_rate,
  volumetric_bill = !!sym("fuel_oil_heating_consumption_kwh") * volumetric_rate
)

monthly_changes_heating_fuel_oil <- calc_monthly_changes(monthly_bills_heating_fuel_oil, "fuel_oil")

annual_changes_heating_fuel_oil <- calc_annual_change_from_monthly(monthly_changes_heating_fuel_oil, "fuel_oil")

# Propane
monthly_bills_heating_propane <- calc_monthly_bills(
  monthly_consumption = monthly_consumption,
  consumption_column = "propane_heating_consumption_kwh",
  fuel_type = "propane",
  delivery_tariffs = propane_delivery_tariffs,
  supply_rates = propane_supply_rates,
  housing_units = housing_units,
  supply_year = supply_year,
  state = "RI",
  enable_lmi_discount = TRUE,
  use_these_upgrades = use_these_upgrades
) |>
mutate(
  volumetric_rate = delivery_rate + supply_rate,
  volumetric_bill = !!sym("propane_heating_consumption_kwh") * volumetric_rate
)

monthly_changes_heating_propane <- calc_monthly_changes(monthly_bills_heating_propane, "propane")

annual_changes_heating_propane <- calc_annual_change_from_monthly(monthly_changes_heating_propane, "propane")

annual_changes_heating <- calc_annual_changes_total(
  annual_changes_heating_elec,
   annual_changes_heating_gas,
  annual_changes_heating_fuel_oil,
  annual_changes_heating_propane
)|>
left_join(housing_units |> select(bldg_id, upgrade, afue_value, hvac_primary, baseline_heating_type), by = c("bldg_id", "upgrade")) |>
filter(afue_value != 'NA')

# used summarize to print the counts and percentages of the unique values of afue_value
annual_changes_heating |>
summarize(
  counts = n(),
  percentages = counts / n(),
  .by = afue_value
) |>
arrange(desc(afue_value)) |>
print()
```

# Heating bill changes
```{r}
#| label: fig-hist_1_annual_change_heating_baseline_hp_high
#| fig-cap: "Distribution of annual _heating_ bill changes for homes switching to heat pumps from natural gas, all under baseline rates."

result <- hist_for_single_rate_version(
  annual_change_table = annual_changes_heating,
  bill_type = 'annual_change_total',
  baseline_heating_type_option = 'Natural Gas',
  version_elec = "baseline",
  hvac_option = "hp_high",
  supply_year = supply_year,
  title = "Change in Total Annual HEATING Bills for Homes Switching to Heat Pumps from Natural Gas",
  second_subtitle = "Rhode Island",
  x_limits = c(-3000, 3000),
  y_limits = c(0, 200),
  binwidth = 100
)

annual_change_stats_heating_baseline_hp_high <- result[[2]]
print(result[[1]])

```

# Heating bill changes by AFUE value
```{r}
#| label: fig-hist_1_annual_change_heating_baseline_hp_high_by_afue
#| fig-cap: "Distribution of annual _heating_ bill changes for homes switching to heat pumps from natural gas, all under baseline rates, by AFUE value."

x_limits <- c(-2500, 2500)
y_limits <- c(0, 15000) / 242.13
binwidth <- 100

# Plot 1: AFUE = 92.5
# -----------------------------------------------
result_1 <- hist_for_single_rate_version(
  annual_change_table = annual_changes_heating |> filter(afue_value == 92.5),
  bill_type = 'annual_change_total',
  version_elec = "baseline",
  baseline_heating_type_option = 'Natural Gas',
  hvac_option = "hp_high",
  supply_year = supply_year,
  title = 'auto',
  second_subtitle = "Rhode Island",
  x_limits = x_limits,
  y_limits = y_limits,
  binwidth = 100
)

p1 <- result_1[[1]] +
  labs(title = NULL) + # Remove title
  ylab(NULL) + # Remove y-axis label
  theme(
    axis.title.x = element_blank(), # Remove x axis title from top two plots
    axis.text.x = element_blank(), # Remove x tick labels from top two plots
    axis.ticks.x = element_blank(), # Remove x ticks from top two plots
    plot.margin = margin(5, 5, 0, 5), # Consistent margins (top, right, bottom, left)
    axis.title.y = element_blank(), # Remove y title from top plot
    aspect.ratio = 0.18
  ) +
  # Add elegant version label on right side
  annotate(
    "text",
    x = x_limits[2] * 0.95,
    y = y_limits[2] * 0.7,
    label = "AFUE = 92.5",
    hjust = 1,
    fontface = "bold",
    size = 4,
    color = "#023047"
  )

# Plot 2: AFUE = 92.5
# -----------------------------------------------
result_2 <- hist_for_single_rate_version(
  annual_change_table = annual_changes_heating |> filter(afue_value == 80),
  bill_type = 'annual_change_total',
  version_elec = "baseline",
  baseline_heating_type_option = 'Natural Gas',
  hvac_option = "hp_high",
  supply_year = supply_year,
  title = 'auto',
  second_subtitle = "Rhode Island",
  x_limits = x_limits,
  y_limits = y_limits,
  binwidth = 100
)

p2 <- result_2[[1]] +
  labs(title = NULL) + # Remove title
  ylab(NULL) + # Remove y-axis label
  theme(
    axis.title.x = element_blank(), # Remove x axis title from top two plots
    axis.text.x = element_blank(), # Remove x tick labels from top two plots
    axis.ticks.x = element_blank(), # Remove x ticks from top two plots
    plot.margin = margin(5, 5, 0, 5), # Consistent margins (top, right, bottom, left)
    axis.title.y = element_blank(), # Remove y title from top plot
    aspect.ratio = 0.18
  ) +
  # Add elegant version label on right side
  annotate(
    "text",
    x = x_limits[2] * 0.95,
    y = y_limits[2] * 0.7,
    label = "AFUE = 80",
    hjust = 1,
    fontface = "bold",
    size = 4,
    color = "#023047"
  )

# Plot 3: AFUE = 92.5
# -----------------------------------------------
result_3 <- hist_for_single_rate_version(
  annual_change_table = annual_changes_heating |> filter(afue_value == 68),
  bill_type = 'annual_change_total',
  version_elec = "baseline",
  baseline_heating_type_option = 'Natural Gas',
  hvac_option = "hp_high",
  supply_year = supply_year,
  title = 'auto',
  second_subtitle = "Rhode Island",
  x_limits = x_limits,
  y_limits = y_limits,
  binwidth = 100
)

p3 <- result_3[[1]] +
  labs(title = NULL) + # Remove title
  ylab(NULL) + # Remove y-axis label
  theme(
    axis.title.x = element_blank(), # Remove x axis title from top two plots
    axis.text.x = element_blank(), # Remove x tick labels from top two plots
    axis.ticks.x = element_blank(), # Remove x ticks from top two plots
    plot.margin = margin(5, 5, 0, 5), # Consistent margins (top, right, bottom, left)
    axis.title.y = element_blank(), # Remove y title from top plot
    aspect.ratio = 0.18
  ) +
  # Add elegant version label on right side
  annotate(
    "text",
    x = x_limits[2] * 0.95,
    y = y_limits[2] * 0.7,
    label = "AFUE = 60",
    hjust = 1,
    fontface = "bold",
    size = 4,
    color = "#023047"
  )

# Combine plots with shared x and y axes
combined_plot <- p1 / p2 / p3 +
plot_layout(heights = c(1, 1, 1)) + plot_annotation(title = "Change in Total Annual HEATING Bills for Homes Switching to Heat Pumps from Natural Gas", subtitle = "Rhode Island") +
theme(
  plot.title = element_text(hjust = 0.5, size = 10, face = "bold"),
  plot.subtitle = element_text(hjust = 0.5, size = 8)
)

print(combined_plot)
# save the plot to cache as an image
ggsave(file.path(path_to_cache, "heating_bill_changes_by_afue_plot.png"), combined_plot, width = 10, height = 8)

```

## Annual changes by AFUE value - Line plot
```{r}
#| label: annual changes by afue value

annual_changes_y  <- annual_changes_total |>
  filter(upgrade == 2) |>
  filter(version_elec == "baseline") |>
  right_join(housing_units |> filter(upgrade==0)|> filter(baseline_heating_type == "Natural Gas")|> select(bldg_id, afue_value), by = c("bldg_id"))

#View(annual_changes_y)

annual_changes_x <- annual_changes_y |>
  summarize(
    counts = n(),
    median_annual_change = median(annual_change_total),
    .by = c(afue_value)
  )

#View(annual_changes_x)

bill_change_vs_afue_plot <- ggplot(annual_changes_x, aes(x = afue_value/100, y = median_annual_change)) +
  #shade light green above y=0
  geom_rect(aes(xmin = -Inf, xmax = Inf, ymin = 0, ymax = Inf), fill = savings_colors["light_red"], alpha = 0.5) +
  #shade light red below y=0
  geom_rect(aes(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = 0), fill = savings_colors["light_green"], alpha = 0.5) +
  labs(x = "Baseline Gas Heating Efficiency (AFUE Value)", y = "Median Annual Total Bill Change") +
  theme_minimal()+
  scale_x_continuous(breaks = seq(0, 1, by = 0.1), labels = scales::percent_format(accuracy = 1)) +
  scale_y_continuous(limits = c(-20, 500), breaks = seq(-1000, 1000, by = 100), labels = scales::dollar_format()) +
  geom_point(size=4, color = sb_colors["midnight"]) +
  geom_line(size=3, color = sb_colors["midnight"]) +
  # use ggrepel to write the counts on the points
  # add a small offset to the counts
  ggrepel::geom_text_repel(aes(label = counts), size = 8, color = sb_colors["sky"], nudge_y = 15) +

  # add a text that says "Count of Bldg IDs" in sky color
  annotate("text", x = 0.62, y = 50, label = "Count of Bldg IDs", color = sb_colors["sky"], size = 8)

print(bill_change_vs_afue_plot)

# save the plot to cache as an image
ggsave(file.path(path_to_cache, "bill_change_vs_afue_plot.png"), bill_change_vs_afue_plot, width = 10, height = 8)
```

## Natural Gas consumption by functional group and month
```{r}
#| label: natural gas consumption by functional group and month

baseline_heating_option <- "Natural Gas"

heating_type_to_fuel_lookup <- tibble(
  heating_type = c("Electric Resistance", "Natural Gas"),
  fuel = c("electricity", "natural_gas")
)

fuel <- heating_type_to_fuel_lookup |> filter(heating_type == baseline_heating_option) |> pull(fuel)


# get the 10th, 50th, and 90th percentiles of consumption by month
monthly_consumption_percentiles <- monthly_consumption |>
  filter(upgrade == 0) |>
  filter(baseline_heating_type == baseline_heating_option) |>
  select(-baseline_heating_type) |>
  # get the quantiles for each column other than bldg_id, upgrade, and month
  reframe(
    across(everything(), ~ quantile(.x, c(0.1, 0.5, 0.9))),
    .by = c(month, upgrade)
  ) |>
  mutate(
    percentile = rep(c("p10", "p50", "p90"),
                   times = n() / 3)
  )|>
  pivot_longer(cols = -c(month, upgrade, percentile), names_to = "functional_group", values_to = "consumption_kwh") |>
  filter(str_detect(functional_group, fuel)) |>


  mutate(
    functional_group = case_when(
      str_detect(functional_group, "appliance") ~ "Appliance",
      str_detect(functional_group, "cooling") ~ "Cooling",
      str_detect(functional_group, "heating") ~ "Heating",
      str_detect(functional_group, "hot_water") ~ "Hot Water",
      str_detect(functional_group, "total") ~ "Total",
      TRUE ~ "Other"
    )
  )
#View(monthly_consumption_percentiles)

# plot the average consumption by month
colors = c(
  "Appliance" = "grey50",
  "Cooling" = "blue",
  "Heating" = "red",
  "Hot Water" = "green",
  "Total" = "black"
)

stack_order <- c("Heating", "Appliance", "Cooling", "Hot Water")  # exclude "Total" since we're filtering that group

plot_p50 <- monthly_consumption_percentiles |>
  filter(percentile == "p50") |>
  filter(functional_group != "Total") |>
  mutate(functional_group = factor(functional_group, levels = stack_order)) |>
  arrange(month, functional_group) |>
  ggplot(aes(x = month, y = consumption_kwh, fill = functional_group)) +
  geom_area(position = "stack", alpha = 0.8) +
  labs(title = paste("Average", baseline_heating_option, "Consumption by Functional Group and Month"),
       x = "Month",
       y = "Consumption (kWh)") +
  scale_x_continuous(breaks = 1:12, labels = month.abb, limits = c(0, 0.05)) +
  scale_y_continuous(labels = scales::comma_format()) +
  scale_fill_manual(values = colors, breaks = stack_order)

print(plot_p50)
# save the plot to cache as an image
ggsave(file.path(path_to_cache, "natural_gas_consumption_by_functional_group_and_month_plot.png"), plot_p50, width = 10, height = 8)

```

```{r}
#path_metadata_upgrade2 = "data/ResStock/2024_release2_tmy3/metadata/upgrade02_metadata_and_annual_results.parquet"
path_metadata_upgrade2 <- file.path(path_to_data, "resstock", "2024_release2_tmy3", "metadata", "upgrade02_metadata_and_annual_results.parquet")
metadata_upgrade2 <- read_parquet(path_metadata_upgrade2) |> filter(in.state == "RI")

# for each row (bldg_id), get the ratio of out.electricity.heating_hp_bkup.energy_consumption to out.electricity.heating.energy_consumption
heating_change_vs_bkup <- annual_changes_heating |>
left_join(metadata_upgrade2 |> select(bldg_id, out.electricity.heating_hp_bkup.energy_consumption.kwh, out.electricity.heating.energy_consumption.kwh), by = c("bldg_id")) |>
mutate(
  bkup_ratio = !!sym("out.electricity.heating_hp_bkup.energy_consumption.kwh") / !!sym("out.electricity.heating.energy_consumption.kwh")
)

# plot a scatter diagram: x = bkup_ratio, y = annual_change_total
p <- ggplot(heating_change_vs_bkup, aes(x = bkup_ratio, y = annual_change_total)) +
  geom_point(alpha = 0.2) +
  labs(x = "Backup Ratio", y = "Annual Change Total")
print(p)

```


```{r}


```
