---
title: "BRM Sandbox"

toc: true
reference-location: margin
fig-cap-location: bottom

format:
  html:
    page-layout: full
---

Import libraries.

```{r}
# Clear environment first
rm(list = ls())

library(tidyverse)
library(lubridate)
library(scales)
library(gt)
library(gtExtras)
library(jsonlite)  # Make sure to load this explicitly

# Check that it's alive
current_time <- Sys.time()
print(current_time)
```

1.  Open the JSON and quick inspection

```{r}

library(tidyverse)
library(jsonlite)

# Read the JSON
unitil_ma_rd1 <- fromJSON("data/arcadia/tariffs/unitil_ma_rd1_20250108.json")
#unitil_ma_rd1 <- fromJSON("data/arcadia/tariffs/ng_ny_scv1_new.json")


# Convert the rates array directly to a dataframe and select columns
# Extract the data with rateAmount from the first rate band
unitil_ma_rd1_table <- tibble(
  rateGroupName = unitil_ma_rd1$rates$rateGroupName,
  rateName = unitil_ma_rd1$rates$rateName,
  chargeType = unitil_ma_rd1$rates$chargeType,
  chargeClass = unitil_ma_rd1$rates$chargeClass,
  rateAmount = sapply(unitil_ma_rd1$rates$rateBands, function(x) x$rateAmount[1])
)

# Save to CSV
write_csv(unitil_ma_rd1_table, "data/ma_hp_rates/electric/unitil_ma_rd1_20250108.csv")


# Print to console
print(n=75, unitil_ma_rd1_table)

```

2.  Group the rates by type

```{r}
# Get sums of rateAmount for each unique combination
unitil_ma_rd1_grouped <- unitil_ma_rd1_table %>%
  group_by(chargeType, chargeClass) %>%
  summarise(
    total_rate = sum(rateAmount),
    .groups = 'drop'
  ) %>%
  arrange(chargeType, chargeClass)

# Print to console
print(unitil_ma_rd1_grouped)

# 1. Get the fully-loaded per-kWh rate
# Get the sum of consumption-based rates from Unitil table
consumption_sum <- unitil_ma_rd1_grouped %>%
  filter(chargeType == "CONSUMPTION_BASED") %>%
  summarise(total_consumption_rate = sum(total_rate, na.rm = TRUE)) %>%
  pull(total_consumption_rate)

print(paste("Total consumption-based rate:", consumption_sum))
```

To calculate an electric bill we need to....

1.  Get the fully-burdened per-kWh portion (`CONSUMPTION_BASED`)
    -   Typically this will be two parts
        -   `DISTRIBUTION`
        -   `SUPPLY, CONTRACTED`
2.  Get any demand-based charges
    -   This will be based on the max load in a given timeframe?
3.  Get the fixed portion of the bill (`FIXED_PRICE`)

```{r}
# source("src/rates_api/rates_modules/00_outline.R")

# #distribution_8760 = transform_consumption_rates_to_8760(unitil_ma_rd1_grouped, "CONSUMPTION_BASED", "DISTRIBUTION")

# rate_summary_table <- unitil_ma_rd1_grouped

# rate_value <- rate_summary_table %>%
#   filter(chargeClass == "DISTRIBUTION", chargeType == "CONSUMPTION_BASED") %>%
#   pull(total_rate)

# print(rate_value)
# # Create 8760-length vector with the rate value repeated
# hourly_rates <- rep(rate_value, 8760)

# print(hourly_rates)

# tibble(
#   timestamp = 1:8760,
#   rate = hourly_rates
# )


```

# 2 - Apply Consumption rates to ResStock homes

1.  Load a table of ResStock electric load profiles (8760 rows x N homes)
2.  `mutate` to apply the all-in consumption rate
3.  Get a 12xN tibble for the consumption component of the bill

```{r}
source("src/rates_api/rates_modules/00_outline.R")
consumption_distribution_8760 = transform_flat_consumption_rates_to_8760(unitil_ma_rd1_grouped, "CONSUMPTION_BASED", "DISTRIBUTION")

glimpse(consumption_distribution_8760)

consumption_supply_8760 = transform_flat_consumption_rates_to_8760(unitil_ma_rd1_grouped, "CONSUMPTION_BASED", "SUPPLY,CONTRACTED")
```

```{r}
electricity_consumption <- as_tibble(read_csv("data/ResStock/ma_hp/combined_electricity_consumption.csv"))

# # Show tibble structure
# glimpse(electricity_consumption)

# # Get dimensions of the tibble
# dim(electricity_consumption)

# Apply the
hourly_cost <- electricity_consumption %>%
  mutate(across(everything(), ~.x * consumption_distribution_8760$rate))



hourly_cost <- hourly_cost %>%
  mutate(datetime = as_datetime("2018-01-01") + hours(row_number() - 1))

glimpse(hourly_cost)

monthly_consumption_costs <- hourly_cost %>%
  group_by(month = floor_date(datetime, "month")) %>%
  summarize(
    bldg_10_total = sum(bldg_10),
    bldg_1000_total = sum(bldg_1000)
  )

monthly_consumption_costs

```

# 3 - Get the Fixed Price portion

```{r}
rate_value <- unitil_ma_rd1_grouped %>%
  filter(chargeType == "FIXED_PRICE") %>%
  pull(total_rate)

fixed_12 = transform_fixed_price_to_12(unitil_ma_rd1_grouped, "FIXED_PRICE")


```

# 4 - Get the minimum

```{r}
rate_value <- unitil_ma_rd1_grouped %>%
  filter(chargeType == "MINIMUM_CHARGE") %>%
  pull(total_rate)

minimum_12 = transform_fixed_price_to_12(unitil_ma_rd1_grouped, "MINIMUM_CHARGE")
```

# 5 - Add them together to get a bill

```{r}

total_bill_by_month <- monthly_consumption_costs %>%
  mutate(across(everything(), ~.x + fixed_12$rate))

total_bill_by_month


```
