---
title: ANALYSIS - ma_hp_rates

toc: true
reference-location: margin
fig-cap-location: bottom

appendix-style: default
citation-location: document
citation:
  container-title: Switchbox

format:
  html:
    page-layout: full
---

# Introduction
How does Unitil's heat pump friendly rate change the bill for heat-pump-having homeowners?

## Steps

### 1.  Get the monthly total energy from all RS 2022.1 homes in

-   specifically `out.electricity.total.energy_consumption`
-   this is accomplished by downloading and repacking (see `data/ResStock/Justfile`)

### 2.  Get the rates

-   Unitil's rates are provided in ...
-   Supply rates are saved at `/data/ma_hp_rates/rates/supply_250227.csv`

### 3. Multiply the monthly energy usage totals by the rates


### R Boilerplate
```{r}
library(tidyverse)
library(scales)

library(gt)
library(gtExtras)
library(jsonlite)
library(RSQLite)

source("/workspaces/reports/lib/ggplot/switchbox_theme.R")

source("/workspaces/reports/lib/rates_api/arcadia_utils.R")
```

```{r}
library(systemfonts)
library(showtext)
font_add(
  family = "IBM-Plex-Sans",
  regular = "/workspaces/reports/docs/fonts/ibm_plex_sans/IBMPlexSans-Regular.otf",
  bold = "/workspaces/reports/docs/fonts/ibm_plex_sans/IBMPlexSans-Bold.otf"
)
showtext_auto()

theme_set(theme_minimal())
theme_update(
  panel.background = element_rect(fill = "white", color = "white"),
  legend.title = element_text(hjust = 0.5), # Centers the legend title
  axis.line = element_line(linewidth = 0.5),
  axis.ticks = element_line(color = "black"),
  # panel.grid.minor.x = element_blank(),
  text = element_text(family = "IBM-Plex-Sans", size = 12),
  axis.text = element_text(
    family = "IBM-Plex-Sans",
    size = 12
  ),
  axis.title = element_text(
    family = "IBM-Plex-Sans",
    size = 12
  ),
  strip.text = element_text(
    size = 12, # Font size
    family = "IBM-Plex-Sans",
  ),
  axis.title.x = element_text(margin = margin(t = 15)),
  axis.title.y = element_text(margin = margin(r = 15))
)


sb_colors <- c(
  "sky" = "#68BED8", # primary
  "midnight" = "#023047", # primary
  "carrot" = "#FC9706", # primary
  "saffron" = "#FFC729", # secondary
  "pistachio" = "#A0AF12", # secondary
  "black" = "#000000", # utilitarian
  "white" = "#FFFFFF", # utilitarian
  "midnight_text" = "#0B6082", # lighter, text only
  "pistachio_text" = "#546800" # darker, text only
)


ma_hp_rates_colors <- c(
  "Deep Ocean" = "#0A4B68",
  "Azure" = "#2A6D8A",
  "Cerulean" = "#4996B2"
)

```


# Step 1. Get the monthly total energy from all RS 2022.1 homes
```{r}
con <- dbConnect(duckdb::duckdb(), "/workspaces/reports/data/ResStock/2022_resstock_amy2018_release_1.1/rs_20250227.db")

# Read only the electricity consumption column from monthly_2022 table
monthly_2022 <- dbGetQuery(con, "SELECT bldg_id, month, upgrade, \"sum_out.electricity.total.energy_consumption\", \"sum_out.electricity.heating.energy_consumption\" FROM monthly_2022 WHERE upgrade IN (3,4)")

# Disconnect from database
dbDisconnect(con)

# Show first few rows and structure
#glimpse(monthly_2022)
#head(monthly_2022, n=100)

# Median TOTAL monthly electricity consumption by upgrade
total_monthly_2022 <- monthly_2022 |>
  group_by(month, upgrade) |>
  summarize(
    median_electricity = median(`sum_out.electricity.total.energy_consumption`),
    .groups = 'drop'
  ) |>
  pivot_wider(
    names_from = upgrade,
    values_from = median_electricity,
    names_prefix = "upgrade_"
  )

total_monthly_2022 <- total_monthly_2022 |>
  mutate(
    relative_diff = (`upgrade_4` - `upgrade_3`) / `upgrade_3` * 100
  )

print("Median TOTAL monthly electricity consumption by upgrade:")
print(total_monthly_2022, n=24)

# Median HEATING monthly electricity consumption by upgrade
heating_monthly_2022 <- monthly_2022 |>
  select(bldg_id, month, upgrade, `sum_out.electricity.heating.energy_consumption`) |>
  group_by(month, upgrade) |>
  summarize(
    median_heating = median(`sum_out.electricity.heating.energy_consumption`),
    .groups = 'drop'
  ) |>
  pivot_wider(
    names_from = upgrade,
    values_from = median_heating,
    names_prefix = "upgrade_"
  )

heating_monthly_2022 <- heating_monthly_2022 |>
  mutate(
    relative_diff = (`upgrade_4` - `upgrade_3`) / `upgrade_3` * 100
  )

print("Median HEATING monthly electricity consumption by upgrade:")
print(heating_monthly_2022, n=24)
```


```{r}

total_u4_2022 <- monthly_2022 |>
  filter(upgrade == 4) |>
  select(bldg_id, month, `sum_out.electricity.total.energy_consumption`) |>
  group_by(month)

glimpse(total_u4_2022)

# Get 20 random unique building IDs
bldgs_u4_2022 <- total_u4_2022 |>
  ungroup() |>
  distinct(bldg_id)

# Join the sampled buildings back to the total data
subset_u4_2022 <- total_u4_2022 |>
  inner_join(bldgs_u4_2022, by = "bldg_id") |>
  pivot_wider(
    names_from = bldg_id,
    values_from = `sum_out.electricity.total.energy_consumption`
  ) |>
  select(month, everything())
View(subset_u4_2022)

# Get a single building
bldg40261_u4_2022 <- total_u4_2022 |>
  filter(bldg_id == 40261)
```

# Step 2. Get the rates

## Baseline Rate
Rates are just a bunch of sub-charges that basically fall into 2 classes:
- Fixed charges
- Consumption-based charges

We can see all the rate components for Unitil's baseline "RD-1" rate alongside the Heat Pump Rate in the table below.

```{r}

# Read in the rates CSV
unitil_rates_2024 <- read_csv("data/ma_hp_rates/electric/unitil_rates_2024.csv") |>
  as_tibble()

print(unitil_rates_2024, n = nrow(unitil_rates_2024))

```

## Heat Pump Rate
- The heat pump rate is nothing more than a 64% discount on the per-kWh Distribution Charge, as described on (this Unitil webpage)[https://unitil.com/ma-heat-pump-rate/en].
  - See page 400 MADPU ruling on Unitil Rate case from July 2024 [@madpu_MAUnitilRate_2024]
  - $0.0958/kWh -> $0.0345/kWh
  - This discount only applies in the months of November through April (Nov 1 - Apr 30)
- 64% sounds great, but the the distribution charge is only part of the all-in per-kWh delivery rate, not to mention the per-kWh supply rate.
- It's really only a discount on the all-in per-kWh _delivery_ rate (see below)

```{r}

rate_components <- tibble(
  component = c("Fixed Charge", "Transmission", "Distribution", "Adjustment", "Programs", "Supply"),
  `RD-1` = c(
    sum(ifelse(unitil_rates_2024$chargeType == "FIXED_PRICE" &
               unitil_rates_2024$chargeClass == "FIXED_CHARGE",
               unitil_rates_2024$`RD-1`, 0)),
    sum(ifelse(unitil_rates_2024$chargeType == "CONSUMPTION_BASED" &
               unitil_rates_2024$chargeClass == "TRANSMISSION",
               unitil_rates_2024$`RD-1`, 0)),
    sum(ifelse(unitil_rates_2024$chargeType == "CONSUMPTION_BASED" &
               unitil_rates_2024$chargeClass == "DISTRIBUTION",
               unitil_rates_2024$`RD-1`, 0)),
    sum(ifelse(unitil_rates_2024$chargeType == "CONSUMPTION_BASED" &
               unitil_rates_2024$chargeClass == "ADJUSTMENT",
               unitil_rates_2024$`RD-1`, 0)),
    sum(ifelse(unitil_rates_2024$chargeType == "CONSUMPTION_BASED" &
               unitil_rates_2024$chargeClass == "PROGRAMS",
               unitil_rates_2024$`RD-1`, 0)),
    sum(ifelse(unitil_rates_2024$chargeType == "CONSUMPTION_BASED" &
               unitil_rates_2024$chargeClass == "SUPPLY",
               unitil_rates_2024$`RD-1`, 0))
  )
)

print("--------------------------------")
print("Rate components:")
print(rate_components)

# Create consumption_components by filtering out "Fixed Distribution"
consumption_components <- rate_components |>
  filter(component != "Fixed Charge")

# Display results
# Calculate sum of delivery components (excluding supply)
delivery_sum <- sum(consumption_components$`RD-1`[consumption_components$component  %in% c("Transmission", "Distribution", "Adjustment", "Programs")])
print(paste("Sum of delivery components:", delivery_sum))

delivery_pct <- consumption_components$`RD-1`[consumption_components$component == "Distribution"] / delivery_sum
print(paste("RD-1 delivery pct:", delivery_pct))

```


```{r}
# Transform consumption_components into a 12x5 tibble with monthly rows
# 12 rows
# 5 columns (Transmission, Distribution, Adjustment, Programs, Supply)
consumption_RD1_12rows <- consumption_components |>
  pivot_wider(
    names_from = component,
    values_from = `RD-1`
  ) |>
  slice(rep(1, 12)) |>
  mutate(month = 1:12) |>
  select(month, everything())

# HPR is the same as RD1 except for the Distribution charge, which is discounted by 64% in Nov-Apr
consumption_HPR_12rows <- consumption_RD1_12rows |>
  mutate(Distribution = ifelse(month %in% c(11,12,1,2,3,4), Distribution * 0.36, Distribution))


print(consumption_RD1_12rows)
print(consumption_HPR_12rows)

```


### Savings Shortcut
- Note that we have a savings shortcut here...it's just the per-kWh difference for the Distribution Charge, $0.0958 - $0.0345 = $0.0613
- The MADPU likes to use 600kWh a month as a benchmark for consumption. 600kWh x $0.0613/kWh = $36.78/month
- But homes that adopt a heat pump will use more than 600kWh a month. For these homes, both the total bill and the savings will be higher.
- With ResStock we can do even better by multiplying the charge difference by the heating signal
  - `out.heating.total.energy_consumption` * $0.0613/kWh
  - This is the most direct way to estimates "how much cheaper is it to operate a heat pump when adoption this new rate?" because it removes the rest of the home's energy use from the equation.

```{r}
# Step 1. Create a monthly all-in per-kWh rate
hpr_savings <- consumption_HPR_12rows$Distribution - consumption_RD1_12rows$Distribution

# Step 2. Multiply the all-in per-kWh rate by the heating signal
monthly_savings <- tibble(
  month = 1:12,
  savings = hpr_savings * heating_monthly_2022$upgrade_4
)

print(paste("Total savings: $", round(sum(monthly_savings$savings), 2)))
```

### Supply Rates
- (Unitil historical rates)[https://unitil.com/suppliers/energy-supplier-resources#historical]
  - That page provides an annoying Excel file, which we transform into a tidy CSV format
  - tranformation: `ma_hp_rates/notebooks/03_ma_supply_rates.qmd`
  - tidy csv: `data/ma_hp_rates/electric/unitil_supply_rates.csv`

- The discount is eroded when we consider the supply rate.
- For simplicity we'll consider the customer option of buying energy directly from Unitil. Those are published....
-

```{r}
supply_rates <- read_csv("data/ma_hp_rates/electric/unitil_supply_rates.csv")

supply_period = "flat15"

if (supply_period == "2024/2025") {
  # We'll consider this past year, so April 2024 - March 2025
  # Get 12 months of supply rates starting from Nov 2024
  # supply_rates_12mo <- supply_rates |>
  #   filter(
  #     (Year == 2024 & Month >= 8) |
  #     (Year == 2025 & Month < 8)
  #   ) |>
  #   select(Year, Month, Monthly) |>
  #   arrange(Month)

} else if (supply_period == "2022") {
  supply_rates_12mo <- supply_rates |>
    filter(Year == 2022)|>
    select(Year, Month, Monthly) |>
  arrange(Month)

} else if (supply_period == "flat15") {
  supply_rates_12mo <- tibble(
    Year = 2022,
    Month = 1:12,
    Monthly = 0.1500
  )
}

print(supply_rates_12mo)

```

- Now replace the Supply column with the actually monthly-varying supply rate
```{r}
consumption_RD1_12 <- consumption_RD1_12rows |>
  mutate(
    Supply = supply_rates_12mo$Monthly,
    total = rowSums(across(c(Supply, Transmission, Adjustment, Programs, Distribution)))
  )

options(digits = 3)
print(consumption_RD1_12, n=24)

consumption_HPR_12 <- consumption_HPR_12rows |>
  mutate(
    Supply = supply_rates_12mo$Monthly,
    total = rowSums(across(c(Supply, Transmission, Adjustment, Programs, Distribution)))
  )

print(consumption_HPR_12, n=24)


```

```{r}
all_in_consumption_RD1 <- consumption_RD1_12 |>
  mutate(
    all_in = rowSums(across(c(Supply, Transmission, Adjustment, Programs, Distribution)))
  ) |>
  select(month, all_in)

print("All-in consumption under RD-1 rate:")
print(all_in_consumption_RD1, n=12)


all_in_consumption_HPR <- consumption_HPR_12 |>
  mutate(
    all_in = rowSums(across(c(Supply, Transmission, Adjustment, Programs, Distribution)))
  ) |>
  select(month, all_in)

print("All-in consumption under HPR rate:")
print(all_in_consumption_HPR, n=12)

```



# 3. Multiply the monthly energy usage totals by the rates to get the monthly bill
- This gives a sample of monthly bills, including the distribution of who pays more and who pays less
- We can even review that set for things like income distribution, etc.

```{r}
# Create a new dataframe with the same structure as consumption_RD1_12
# but multiply each rate component by the monthly consumption
monthly_bills_RD1 <- consumption_RD1_12 |>
  mutate(
    Fixed = rate_components$`RD-1`[rate_components$component == "Fixed Charge"],
    Supply = Supply * total_monthly_2022$upgrade_4,
    Transmission = Transmission * total_monthly_2022$upgrade_4,
    Adjustment = Adjustment * total_monthly_2022$upgrade_4,
    Programs = Programs * total_monthly_2022$upgrade_4,
    Distribution = Distribution * total_monthly_2022$upgrade_4,
  ) |>
  select(Fixed, Supply, Transmission, Adjustment, Programs, Distribution)


print("Monthly bills under RD-1 rate:")
print(monthly_bills_RD1, n=24)


# Heat pump Rate
monthly_bills_HPR <- consumption_HPR_12 |>
  mutate(
    Fixed = rate_components$`RD-1`[rate_components$component == "Fixed Charge"],
    Supply = Supply * total_monthly_2022$upgrade_4,
    Transmission = Transmission * total_monthly_2022$upgrade_4,
    Adjustment = Adjustment * total_monthly_2022$upgrade_4,
    Programs = Programs * total_monthly_2022$upgrade_4,
    Distribution = Distribution * total_monthly_2022$upgrade_4,
  ) |>
  select(Fixed, Supply, Transmission, Adjustment, Programs, Distribution)

print("Monthly bills under Heat Pump Rate:")
print(monthly_bills_HPR, n=24)


monthly_bills_RD1_long <- monthly_bills_RD1 |>
  mutate(Month = 1:12) |>
  pivot_longer(
    cols = -Month,
    names_to = "Component",
    values_to = "Amount"
  ) |>
  mutate(Rate = "RD-1")

monthly_bills_HPR_long <- monthly_bills_HPR |>
  mutate(Month = 1:12) |>
  pivot_longer(
    cols = -Month,
    names_to = "Component",
    values_to = "Amount"
  ) |>
  mutate(Rate = "HPR")

# Combine the data
monthly_bills_combined <- bind_rows(monthly_bills_RD1_long, monthly_bills_HPR_long)

# Set the order of components
monthly_bills_combined$Component <- factor(
  monthly_bills_combined$Component,
  levels = c("Distribution", "Programs", "Adjustment", "Transmission", "Supply", "Fixed")
)

```

```{r}
# Calculate annual totals by component and rate
annual_bills <- monthly_bills_combined |>
  group_by(Rate, Component) |>
  summarize(
    Amount = sum(Amount),
    .groups = 'drop'
  )

print("Annual Bills:")
print(annual_bills)


# Print total annual bills by rate
annual_totals <- annual_bills |>
  group_by(Rate) |>
  summarize(
    Total = sum(Amount),
    .groups = 'drop'
  )

print("Annual totals by rate:")
print(annual_totals)

# Calculate and print the difference
rd1_total <- annual_totals |> filter(Rate == "RD-1") |> pull(Total)
hpr_total <- annual_totals |> filter(Rate == "HPR") |> pull(Total)
savings <- rd1_total - hpr_total

print(paste("\nAnnual savings with Heat Pump Rate: $", round(savings, 2)))

# Print full component breakdown
print("\nFull component breakdown:")
print(annual_bills, n=24)


```


### Annual bill for every bldg_id
```{r}

glimpse(subset_u4_2022)

# Get building IDs (column names except 'month')
rs_upgrade <- "4, High Efficiency"

if (rs_upgrade == "4, High Efficiency") {
  bldg_ids <- names(subset_u4_2022)[names(subset_u4_2022) != "month"]
} else if (rs_upgrade == "3, Minimum Efficiency") {
  bldg_ids <- names(subset_u3_2022)[names(subset_u3_2022) != "month"]
}
# Create empty list to store annual bills
annual_bills_RD1 <- list()
annual_bills_HPR <- list()

# Calculate annual bill for each building
for (bldg_id in bldg_ids) {
  # Get monthly consumption for this building
  monthly_consumption <- subset_u4_2022[[bldg_id]]

  # Multiply by monthly rates and sum
  annual_bill <- sum(monthly_consumption * all_in_consumption_RD1$all_in)

  # Store in list
  annual_bills_RD1[[bldg_id]] <- annual_bill
}

for (bldg_id in bldg_ids) {
  # Get monthly consumption for this building
  monthly_consumption <- subset_u4_2022[[bldg_id]]

  # Multiply by monthly rates and sum
  annual_bill <- sum(monthly_consumption * all_in_consumption_HPR$all_in)

  # Store in list
  annual_bills_HPR[[bldg_id]] <- annual_bill
}


# Convert to tibble for easier viewing
annual_bills_RD1_df <- tibble(
  bldg_id = names(annual_bills_RD1),
  annual_bill = unlist(annual_bills_RD1)
)

annual_bills_HPR_df <- tibble(
  bldg_id = names(annual_bills_HPR),
  annual_bill = unlist(annual_bills_HPR)
)

print("Annual bills under RD-1 rate for sample buildings:")
print(annual_bills_RD1_df)

print("Annual bills under HPR rate for sample buildings:")
print(annual_bills_HPR_df)

```

```{r}

# Calculate percent savings
savings_df <- annual_bills_RD1_df |>
  left_join(annual_bills_HPR_df, by = "bldg_id") |>
  mutate(
    savings = annual_bill.y - annual_bill.x,
    percent_savings = (annual_bill.y - annual_bill.x) / annual_bill.x * 100
  ) |>
  select(bldg_id, savings, percent_savings)

print("Savings under HPR vs RD-1:")
print(savings_df)

# Calculate summary statistics
savings_summary <- savings_df |>
  summarise(
    mean_savings = mean(percent_savings),
    median_savings = median(percent_savings),
    min_savings = min(percent_savings),
    max_savings = max(percent_savings)
  )

print("\nSavings Summary Statistics:")
print(savings_summary)





```




# 4. Bar Chart

## Monthly Bar Chart
```{r}
source("/workspaces/reports/lib/ggplot/switchbox_theme.R")


# Create a position variable for side-by-side bars
monthly_bills_combined <- monthly_bills_combined |>
  mutate(pos = as.numeric(factor(Month)) + ifelse(Rate == "RD-1", -0.2, 0.2))

# Create the plot with stacked bars side-by-side
ggplot(monthly_bills_combined, aes(x = pos, y = Amount, fill = Component)) +
  geom_bar(stat = "identity", position = "stack", width = 0.4) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_fill_brewer(palette = "Set3") +
  labs(
    title = "Monthly Bills Comparison: RD-1 vs HPR",
    subtitle = paste("Supply Period:", supply_period),
    x = "",
    y = "Amount ($)",
    fill = "Bill Component"
  ) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = c(.5, .7),
    legend.box = "vertical",
    legend.background = element_rect(fill = "grey90")
  )+

  # Add text annotations for rate types
  annotate(
    "text",
    x = 1.2, # Position for RD-1 label (middle of first half)
    y = monthly_bills_combined |> filter(Month == 1, Rate == "RD-1") |> pull(Amount) |> sum() * 1.03,
    label = "RD-1 (Baseline)",
    family = "IBM-Plex-Sans",
    size = 4
  ) +
  annotate(
    "text",
    x = 1.7, # Position for HPR label (middle of second half)
    y = monthly_bills_combined |> filter(Month == 1, Rate == "HPR") |> pull(Amount) |> sum() * 1.03,
    label = "Heat Pump Rate",
    family = "IBM-Plex-Sans",
    size = 4
  )


```

## Annual Bar Chart

```{r}


# Create the annual comparison plot
ggplot(annual_bills, aes(x = factor(Rate, levels = c("RD-1", "HPR")), y = Amount, fill = Component)) +
  geom_bar(stat = "identity", position = "stack", width = 0.6) +
  scale_fill_brewer(palette = "Set3") +
  scale_x_discrete(labels = c("RD-1" = "RD-1 (Default)", "HPR" = "Heat Pump Rate")) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Annual Bill Comparison: RD-1 vs HPR",
    subtitle = paste("Supply Period:", supply_period),
    x = "",
    y = "Median Annual Bill ($)",
    fill = "Bill Component"
  ) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "right",
    legend.box = "vertical",
    legend.background = element_rect(fill = "grey90")
  ) +
  # Add total amount annotations
  annotate(
    "text",
    x = "RD-1",
    y = rd1_total * 1.02,
    label = paste0("$", format(round(rd1_total, 0), big.mark=",")),
  ) +
  annotate(
    "text",
    x = "HPR",
    y = hpr_total * 1.02,
    label = paste0("$", format(round(hpr_total, 0), big.mark=",")),
  )
  # Add percent difference annotation
  # annotate(
  #   "text",
  #   x = 1.5,
  #   y = max(rd1_total, hpr_total) * 0.97,
  #   label = paste0(round((hpr_total - rd1_total) / rd1_total * 100, 1), "% difference ↓"),
  # )

  # ggsave(
  #   "src/ma_hp_rates/img/bills_annual.png",
  #   width = 10,
  #   height = 8,
  #   dpi = 300
  # )

```

## Annual Stacked Bar, Looped
```{r}

bill_colors <- c(
  "Distribution" = sb_colors[["carrot"]],
  "Programs" = ma_hp_rates_colors[["Cerulean"]],
  "Adjustment" = ma_hp_rates_colors[["Azure"]],
  "Transmission" = ma_hp_rates_colors[["Deep Ocean"]],
  "Supply" = sb_colors[["pistachio"]],
  "Fixed" = sb_colors[["midnight"]]
)

y_annotation_offset <- 50

# Get unique components in order they appear
components <- rev(unique(annual_bills$Component))
print(components)

# Loop through number of components
for (i in 0:length(components)) {
  # Take only first i components
  subset_bills <- annual_bills |>
    filter(Component %in% components[0:i])

  # Calculate subtotals for each rate up to current component
  rd1_component <- subset_bills$Amount[subset_bills$Rate == "RD-1"]
  rd1_subtotal <- sum(subset_bills$Amount[subset_bills$Rate == "RD-1"])
  hpr_component <- subset_bills$Amount[subset_bills$Rate == "HPR"]
  hpr_subtotal <- sum(subset_bills$Amount[subset_bills$Rate == "HPR"])

  # Create a dummy dataset with all components to set consistent legend space
  dummy_data <- annual_bills |>
    complete(Rate, Component) |>
    replace_na(list(Amount = 0))

  ggplot() +
    # Plot the dummy data with zero alpha to reserve legend space
    geom_bar(data = dummy_data,
             aes(x = factor(Rate, levels = c("RD-1", "HPR")),
                 y = Amount, fill = Component),
             stat = "identity", position = "stack", width = 0.6,
             alpha = 0) +
    # Plot the actual subset data
    geom_bar(data = subset_bills,
             aes(x = factor(Rate, levels = c("RD-1", "HPR")),
                 y = Amount, fill = Component),
             stat = "identity", position = "stack", width = 0.6) +
    scale_fill_manual(values = bill_colors) +
    scale_x_discrete(labels = c("RD-1" = "RD-1 (Default)", "HPR" = "Heat Pump Rate")) +
    scale_y_continuous(limits = c(0, 4500), labels = scales::dollar_format()) +
    labs(
      title = "Annual Bill Comparison: RD-1 vs HPR",
      subtitle = paste("Supply Period:", supply_period),
      x = "",
      y = "Median Annual Bill ($)",
      fill = "Bill Component"
    ) +
    theme(
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      legend.position = "right",
      legend.box = "vertical",
      legend.background = element_rect(fill = "grey90")
    ) +
    # Add subtotal annotations for current components
    annotate(
      "text",
      x = "RD-1",
      y = (rd1_subtotal + y_annotation_offset),
      label = paste0("$", format(round(rd1_subtotal, 0), big.mark=",")),
    ) +
    annotate(
      "text",
      x = "HPR",
      y = (hpr_subtotal + y_annotation_offset),
      label = paste0("$", format(round(hpr_subtotal, 0), big.mark=",")),
    )

  ggsave(
    paste0("src/ma_hp_rates/img/bills_annual_stacked_", i, ".png"),
    width = 4,
    height = 3,
    dpi = 300
  )
}

```




# 5. Histograms

```{r}
# Combine the data and create density plots
annual_bills_combined <- bind_rows(
  mutate(annual_bills_HPR_df, rate = "HPR"),
   mutate(annual_bills_RD1_df, rate = "RD-1")
)


print(mean(annual_bills_combined$annual_bill))
print(sd(annual_bills_combined$annual_bill))

# Calculate mean and standard deviation
mean_bill <- mean(annual_bills_combined$annual_bill)
sd_bill <- sd(annual_bills_combined$annual_bill)

# Filter to within 3 standard deviations
annual_bills_combined <- annual_bills_combined |>
  filter(annual_bill >= mean_bill - 3*sd_bill,
         annual_bill <= mean_bill + 3*sd_bill)


print(annual_bills_combined)

# Create density plot
ggplot(annual_bills_combined, aes(x = annual_bill, color = rate)) +
  geom_density(size = 1) +
  # geom_vline(data = annual_bills_combined |> group_by(rate) |>
  #            summarize(mean_bill = mean(annual_bill)),
  #            aes(xintercept = mean_bill, color = rate),
  #            linetype = "dashed") +
  annotate(
    "text",
    x = 1000,
    y = 0.00015,
    label = "HPR",
    size = 18/.pt,
    color = sb_colors[["carrot"]],
    fontface = "bold"
  ) +
  annotate(
    "text",
    x = 2600,
    y = 0.00013,
    label = "RD-1",
    size = 18/.pt,
    color = sb_colors[["midnight"]],
    fontface = "bold"
  ) +
  geom_vline(xintercept = 6000, color = "red", linetype = "solid") +
  annotate(
    "text",
    x = 6200,
    y = 0.00012,
    label = "Median Energy Burden Threshold in MA",
    size = 14/.pt,
    color = "red",
    fontface = "bold",
    hjust = -0.03
  ) +
  scale_color_manual(values = c(
    "RD-1" = sb_colors[["midnight"]],
    "HPR" = sb_colors[["carrot"]]
  )) +
  xlim(0, 20000) +
  labs(
    title = "Distribution of Annual Bills",
    subtitle = sprintf(
      "Supply Period: %s | Upgrade: %s",
      as.character(supply_period),
      as.character(rs_upgrade)
    ),
    x = "Annual Bill Amount ($)",
    y = "Density",
    color = "Rate Type"
  ) +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5)
  )

ggsave(
  "src/ma_hp_rates/img/bills_density.png",
  width = 5,
  height = 3,
  dpi = 300
)

```


# 6. Box-and-Whisker Plot
```{r}

# Calculate quartiles for each rate type
quartiles <- annual_bills_combined |>
  group_by(rate) |>
  summarize(
    q1 = quantile(annual_bill, 0.25),
    median = median(annual_bill),
    q3 = quantile(annual_bill, 0.75)
  )

print("Quartiles by rate type:")
print(quartiles)

# Create box plot
ggplot(annual_bills_combined, aes(x = factor(rate, levels = c("RD-1", "HPR")), y = annual_bill, fill = rate)) +
  geom_boxplot(outlier.size = 0.5) +
  scale_y_continuous(limits = c(0, 20000), labels = scales::dollar_format()) +
  scale_x_discrete(labels = c("RD-1" = "RD-1 (Default)", "HPR" = "Heat Pump Rate")) +
  scale_fill_manual(values = c(
    "RD-1" = sb_colors[["midnight"]],
    "HPR" = sb_colors[["carrot"]]
  )) +
  geom_text(data = quartiles,
            aes(x = rate, y = q1, label = sprintf("Q1: $%.0f", q1)),
            vjust = 1.7, hjust = 0, position = position_nudge(x = 0.05), color = "black", size = 12/.pt) +
  geom_text(data = quartiles,
            aes(x = rate, y = median, label = sprintf("Median: $%.0f", median)),
            vjust = -.8, hjust = 0, position = position_nudge(x = 0.05), color = c("black", sb_colors[["sky"]]), size = 12/.pt) +
  geom_text(data = quartiles,
            aes(x = rate, y = q3, label = sprintf("Q3: $%.0f", q3)),
            vjust = -0.7, hjust = 0, position = position_nudge(x = 0.05), color = "black", size = 12/.pt) +
  labs(
    title = "Distribution of Annual Bills",
    subtitle = sprintf(
      "Supply Period: %s | Upgrade: %s",
      as.character(supply_period),
      as.character(rs_upgrade)
    ),
    x = "Rate Type",
    y = "Annual Bill Amount ($)"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "none"
  )
ggsave(
  "src/ma_hp_rates/img/bills_boxplot.png",
  width = 6,
  height = 3,
  dpi = 300
)

```


```{r}
library(patchwork)
print(annual_bills_combined)

title_x = 0
title_y = 90
title_hjust = 0.1

# Create stacked histogram
# Create separate plots for each rate type
rd1_plot <- ggplot(subset(annual_bills_combined, rate == "RD-1"),
                   aes(x = annual_bill)) +
  geom_histogram(fill = sb_colors[["midnight"]], binwidth = 500) +
  scale_x_continuous(labels = scales::dollar_format()) +
  labs(y = "Count") +
  annotate(
    "text",
    x = title_x,
    y = title_y,
    label = "RD-1 (Default)",
    size = 18/.pt,
    color = sb_colors[["midnight"]],
    fontface = "bold",
    hjust = title_hjust
  ) +
  geom_vline(xintercept = median(subset(annual_bills_combined, rate == "RD-1")$annual_bill), color = sb_colors[["midnight"]], linetype = "solid") +
  annotate(
    "text",
    x = median(subset(annual_bills_combined, rate == "RD-1")$annual_bill),
    y = 80,
    label = "Median RD-1 Bill: $4,269",
    size = 14/.pt,
    color = sb_colors[["midnight"]],
    fontface = "bold",
    hjust = -.05
  )
  # geom_vline(xintercept = 6000, color = "red", linetype = "solid") +
  # annotate(
  #   "text",
  #   x = 6100,
  #   y = 60,
  #   label = "Median Energy Burden Threshold in MA",
  #   size = 14/.pt,
  #   color = "red",
  #   fontface = "bold",
  #   hjust = 0
  # )


hpr_plot <- ggplot(subset(annual_bills_combined, rate == "HPR"),
                   aes(x = annual_bill)) +
  geom_histogram(fill = sb_colors[["carrot"]], binwidth = 500) +
  scale_x_continuous(labels = scales::dollar_format()) +
  labs(x = "Annual Bill Amount ($)",
       y = "Count") +
  annotate(
    "text",
    x = title_x,
    y = title_y,
    label = "Heat Pump Rate",
    size = 18/.pt,
    color = sb_colors[["carrot"]],
    fontface = "bold",
    hjust = title_hjust
  ) +
  geom_vline(xintercept = median(subset(annual_bills_combined, rate == "HPR")$annual_bill), color = sb_colors[["carrot"]], linetype = "solid") +
  annotate(
    "text",
    x = median(subset(annual_bills_combined, rate == "HPR")$annual_bill),
    y = 80,
    label = "Median HPR Bill: $3,838",
    size = 14/.pt,
    color = sb_colors[["carrot"]],
    fontface = "bold",
    hjust = -0.05
  )
  # geom_vline(xintercept = 6000, color = "red", linetype = "solid") +
  # annotate(
  #   "text",
  #   x = 6100,
  #   y = 60,
  #   label = "Median Energy Burden Threshold in MA",
  #   size = 14/.pt,
  #   color = "red",
  #   fontface = "bold",
  #   hjust = 0
  # )

# Combine plots using patchwork
combined_plot <- rd1_plot / hpr_plot +
  plot_annotation(
    title = "Distribution of Annual Bills",
    subtitle = sprintf(
      "Supply Period: %s | Upgrade: %s",
      as.character(supply_period),
      as.character(rs_upgrade)
    )
  )

combined_plot


# Save the combined plot
ggsave(
  "src/ma_hp_rates/img/bills_stacked_histograms.png",
  plot = combined_plot,
  width = 5,
  height = 3,
  dpi = 300
)




```


```{r}

# Create histogram of percent savings
ggplot(savings_df, aes(x = abs(percent_savings))) +
  geom_histogram(
    fill = sb_colors[["pistachio"]],
    color = sb_colors[["black"]],
    binwidth = 0.5
  ) +
  labs(
    title = "Distribution of Absolute Percent Savings",
    subtitle = sprintf(
      "Supply Period: %s | Upgrade: %s",
      as.character(supply_period),
      as.character(rs_upgrade)
    ),
    x = "Percent Savings (%)",
    y = "Count"
  ) +
  scale_x_continuous(labels = function(x) paste0(x, "%")) +
  theme(
    plot.title = element_text(
      family = "IBM-Plex-Sans",
      size = 16,
      face = "bold"
    ),
    plot.subtitle = element_text(
      family = "IBM-Plex-Sans",
      size = 12
    )
  )

# Save the plot
ggsave(
  "src/ma_hp_rates/img/savings_histogram.png",
  width = 5,
  height = 3,
  dpi = 300
)


```


# Next round

- Histograms
  3 layers
    Default
    HPR
    DOER

- Savings
  - default vs HPR
- default vs doer

- bill breakdown
  - 3 categories
    - fixed
    - delviery
    - supply

- (maybe)

- ----
- comparing utiliies
  - hist values 3x (by utility)




Sprint 2: static switching analyis
  - get gas bills
  - also geographic parsin

Sprint 3: Supply variance
  - just use the same 3 rates, but different years' supply rates


-
