---
title: ANALYSIS - ma_hp_rates

toc: true
reference-location: margin
fig-cap-location: bottom

appendix-style: default
citation-location: document
citation:
  container-title: Switchbox

format:
  html:
    page-layout: full
---

# Introduction
How does Unitil's heat pump friendly rate change the bill for heat-pump-having homeowners?

## Steps

### 1.  Get the monthly total energy from all RS 2022.1 homes in

-   specifically `out.electricity.total.energy_consumption`
-   this is accomplished by downloading and repacking (see `data/ResStock/Justfile`)

### 2.  Get the rates

-   Unitil's rates are provided in ...
-   Supply rates are saved at `/data/ma_hp_rates/rates/supply_250227.csv`

### 3. Multiply the monthly energy usage totals by the rates


### R Boilerplate
```{r}
library(tidyverse)
library(scales)

library(gt)
library(gtExtras)
library(jsonlite)
library(RSQLite)

library(patchwork)

source("/workspaces/reports/lib/ggplot/switchbox_theme.R")

source("/workspaces/reports/lib/rates_api/arcadia_utils.R")
```

### SB Theme
```{r}
library(systemfonts)
library(showtext)
font_add(
  family = "IBM-Plex-Sans",
  regular = "/workspaces/reports/docs/fonts/ibm_plex_sans/IBMPlexSans-Regular.otf",
  bold = "/workspaces/reports/docs/fonts/ibm_plex_sans/IBMPlexSans-Bold.otf"
)
showtext_auto()

theme_set(theme_minimal())
theme_update(
  panel.background = element_rect(fill = "white", color = "white"),
  legend.title = element_text(hjust = 0.5), # Centers the legend title
  axis.line = element_line(linewidth = 0.5),
  axis.ticks = element_line(color = "black"),
  # panel.grid.minor.x = element_blank(),
  text = element_text(family = "IBM-Plex-Sans", size = 12),
  axis.text = element_text(
    family = "IBM-Plex-Sans",
    size = 12
  ),
  axis.title = element_text(
    family = "IBM-Plex-Sans",
    size = 12
  ),
  strip.text = element_text(
    size = 12, # Font size
    family = "IBM-Plex-Sans",
  ),
  axis.title.x = element_text(margin = margin(t = 15)),
  axis.title.y = element_text(margin = margin(r = 15))
)


sb_colors <- c(
  "sky" = "#68BED8", # primary
  "midnight" = "#023047", # primary
  "carrot" = "#FC9706", # primary
  "saffron" = "#FFC729", # secondary
  "pistachio" = "#A0AF12", # secondary
  "black" = "#000000", # utilitarian
  "white" = "#FFFFFF", # utilitarian
  "midnight_text" = "#0B6082", # lighter, text only
  "pistachio_text" = "#546800" # darker, text only
)


ma_hp_rates_colors <- c(
  "Deep Ocean" = "#0A4B68",
  "Azure" = "#2A6D8A",
  "Cerulean" = "#4996B2"
)

```


# Step 1. Get the monthly total energy from all RS 2022.1 homes
```{r}

# Read from the database
con <- dbConnect(duckdb::duckdb(), "/workspaces/reports/data/ResStock/2022_resstock_amy2018_release_1.1/rs_20250227.db")
monthly_2022 <- dbGetQuery(con, "SELECT bldg_id, month, upgrade, \"sum_out.electricity.total.energy_consumption\" FROM monthly_2022 WHERE upgrade IN (4)")

monthly_gas_2022 <- dbGetQuery(con, "SELECT bldg_id, month, upgrade, \"sum_out.natural_gas.total.energy_consumption\" FROM monthly_2022 WHERE upgrade IN (0)")



dbDisconnect(con)

View(monthly_2022)
View(monthly_gas_2022)

# Median TOTAL monthly electricity consumption by upgrade
total_monthly_2022 <- monthly_2022 |>
  group_by(month, upgrade) |>
  summarize(
    median_electricity = median(`sum_out.electricity.total.energy_consumption`),
    .groups = 'drop'
  ) |>
  pivot_wider(
    names_from = upgrade,
    values_from = median_electricity,
    names_prefix = "upgrade_"
  )
View(total_monthly_2022)

print("Median TOTAL monthly electricity consumption by upgrade:")
print(total_monthly_2022, n=24)
```

```{r}

total_u4_2022 <- monthly_2022 |>
  filter(upgrade == 4) |>
  select(bldg_id, month, `sum_out.electricity.total.energy_consumption`) |>
  group_by(month)

glimpse(total_u4_2022)

# Get 20 random unique building IDs
bldgs_u4_2022 <- total_u4_2022 |>
  ungroup() |>
  distinct(bldg_id)

# Join the sampled buildings back to the total data
subset_u4_2022 <- total_u4_2022 |>
  inner_join(bldgs_u4_2022, by = "bldg_id") |>
  pivot_wider(
    names_from = bldg_id,
    values_from = `sum_out.electricity.total.energy_consumption`
  ) |>
  select(month, everything())
View(subset_u4_2022)

# Get a single building
bldg40261_u4_2022 <- total_u4_2022 |>
  filter(bldg_id == 40261)
```

# Step 2. Get the rates

## Electric Rates
### Baseline Rate
Rates are just a bunch of sub-charges that basically fall into 2 classes:
- Fixed charges
- Consumption-based charges

We can see all the rate components for Unitil's baseline "RD-1" rate alongside the Heat Pump Rate in the table below.

```{r}

# Expand exhibit_c_rates to have 12 months of data
exhibit_c_rates <- exhibit_c_rates |>
  # First add month=1 to all rows
  mutate(month = 1) |>
  # Then repeat each row 12 times and assign months 1-12
  slice(rep(1:n(), times = 12)) |>
  # Update month numbers for the repeated rows
  group_by(class, type, season, utility, customer_class, name, version, value) |>
  mutate(month = row_number()) |>
  ungroup()

# summer <- (6,7,8,9)
# winter <- (1,2,3,4,5,10,11,12)

# Filter out mismatched season/month combinations
exhibit_c_rates <- exhibit_c_rates |>
  filter(
    !(season == "summer" & !(month %in% c(5,6,7,8,9,10))) &
    !(season == "winter" & !(month %in% c(1,2,3,4,11,12)))
  )
View(exhibit_c_rates)
```

- Plot the 3 tariff options, by month, for a single utility
```{r}
# Create three versions of the plot with progressively more rate versions
for (i in 1:3) {
  # Create filtered dataset based on iteration
  plot_data <- exhibit_c_rates |>
    filter(utility == "unitil", type == "delivery_charge")

  if (i == 1) {
    # First iteration: Only baseline, with dummy data for other versions
    plot_data <- plot_data |>
      filter(version == "baseline") |>
      bind_rows(
        plot_data |>
        filter(version == "baseline") |>
        mutate(version = "hpr_approved", value = 0),
        plot_data |>
        filter(version == "baseline") |>
        mutate(version = "hpr_doer", value = 0)
      )
  } else if (i == 2) {
    # Second iteration: Baseline and HPR Approved
    plot_data <- plot_data |>
      filter(version %in% c("baseline", "hpr_approved")) |>
      bind_rows(
        plot_data |>
        filter(version == "baseline") |>
        mutate(version = "hpr_doer", value = 0)
      )
  } else {
    # Third iteration: All versions
    plot_data <- plot_data
  }

  # Create plot
  ggplot(plot_data, aes(x = month, y = value, fill = version)) +
    geom_bar(stat = "identity", position = "dodge") +
    scale_x_continuous(breaks = 1:12, labels = month.abb) +
    scale_y_continuous(labels = scales::dollar_format()) +
    scale_fill_manual(values = c(
      "baseline" = sb_colors[["midnight"]],
      "hpr_approved" = sb_colors[["carrot"]],
      "hpr_doer" = sb_colors[["sky"]]
    )) +
    labs(
      title = "Unitil Delivery Charges by Month",
      y = "Delivery Charge ($/kWh)",
      fill = "Rate Version"
    ) +
    theme(
      legend.position = c(0.85, 1.25),
      legend.direction = "horizontal"
    ) +
    theme(axis.title.x = element_blank())

  # Save plot
  ggsave(
    sprintf("src/ma_hp_rates/img/unitil_delivery_monthly_%d.png", i),
    width = 5,
    height = 2,
    dpi = 300
  )
}

```

### Heat Pump Rate
- The heat pump rate is nothing more than a 64% discount on the per-kWh Distribution Charge, as described on (this Unitil webpage)[https://unitil.com/ma-heat-pump-rate/en].
  - We have 3 rates for each of the 3 utilities
- | Utility | Default | HPR | DOER |
  - | --- | --- | --- | --- |
  - | Unitil | $0.0958 | $0.0345 | $0.0345 |
  - | Eversource | $0.0958 | $0.0345 | $0.0345 |
  - | National Grid | $0.0958 | $0.0345 | $0.0345 |

### Supply Rates
- (Unitil historical rates)[https://unitil.com/suppliers/energy-supplier-resources#historical]
  - That page provides an annoying Excel file, which we transform into a tidy CSV format
  - tranformation: `ma_hp_rates/notebooks/03_ma_supply_rates.qmd`
  - tidy csv: `data/ma_hp_rates/electric/unitil_supply_rates.csv`

- The discount is eroded when we consider the supply rate.
- For simplicity we'll consider the customer option of buying energy directly from Unitil. Those are published....
-
```{r}
supply_rates <- read_csv("data/ma_hp_rates/electric/unitil_supply_rates.csv")

supply_period = "flat15"

if (supply_period == "2024/2025") {
  # We'll consider this past year, so April 2024 - March 2025
  # Get 12 months of supply rates starting from Nov 2024
  # supply_rates_12mo <- supply_rates |>
  #   filter(
  #     (Year == 2024 & Month >= 8) |
  #     (Year == 2025 & Month < 8)
  #   ) |>
  #   select(Year, Month, Monthly) |>
  #   arrange(Month)

} else if (supply_period == "2022") {
  supply_rates_12mo <- supply_rates |>
    filter(Year == 2022)|>
    select(Year, Month, Monthly) |>
  arrange(Month)

} else if (supply_period == "flat15") {
  supply_rates_12mo <- tibble(
    year = 2022,
    month = 1:12,
    monthly = 0.1500
  )
}

View(supply_rates_12mo)

```

- Now replace the Supply column with the actually monthly-varying supply rate
```{r}
# Join supply_rates_12mo with exhibit_c_rates to update supply charges
exhibit_c_rates <- exhibit_c_rates |>
  left_join(supply_rates_12mo, by = "month") |>
  mutate(
    value = ifelse(type == "supply_charge", monthly, value)
  ) |>
  select(-monthly, -year) # Remove the joined columns we no longer need
View(exhibit_c_rates)
```

# Gas Rates
## Delivery
```{r}
# Read in the rates CSV
ma_gas_rates <- read_csv("/workspaces/reports/data/utility_rates/gas/ma/madpu/ma_gas_rates.csv") |>
  as_tibble() |>
  select(-source)

ma_gas_rates <- ma_gas_rates |>
  # Convert volumetric gas rates from $/therm to $/ccf
  mutate(value = ifelse(class == "volumetric", value / 29.307, value)) |>

  mutate(month = 1) |>
  # Then repeat each row 12 times and assign months 1-12
  slice(rep(1:n(), times = 12)) |>
  # Update month numbers for the repeated rows
  group_by(class, type, season, utility, customer_class, name, version, value) |>
  mutate(month = row_number()) |>
  ungroup()
View(ma_gas_rates)
```


## Gas Supply
- (Unitil historical rates)[https://unitil.com/suppliers/energy-supplier-resources#historical]
  - Peak period (or winter) rate: November 1 through April 30.
  - Off-peak period (or summer) rate: May 1 through October 31.

- [Rates here](https://www.mass.gov/info-details/information-on-gas-supply-and-delivery-charges#gas-supply-rate-information-(gaf)-)



```{r}
library(tidyverse)
resi_codes <- c("R-1", "Residential", "All Residential")

# Read in the gas supply rates CSV
ma_gas_supply_rates <- read_csv("/workspaces/reports/data/utility_rates/gas/ma/madpu/gas_supply_rates_tidy.csv") |>
  as_tibble() |>
  filter(Rate Class %in% resi_codes)

View(ma_gas_supply_rates)


```


# 3. Multiply the monthly energy usage totals by the rates to get the monthly bill
- This gives a sample of monthly bills, including the distribution of who pays more and who pays less
- We can even review that set for things like income distribution, etc.

## Electricity Bill Impacts Calculation
- this is the core calculation for the report
```{r}
# Create a new dataframe with the same structure as consumption_RD1_12
# but multiply each rate component by the monthly consumption
# exhibit_c_unitil <- exhibit_c_rates |>
#   filter(utility == "unitil")
# View(exhibit_c_unitil)


monthly_bill_by_component <- monthly_2022 |>
  left_join(exhibit_c_rates, by = "month") |>
  mutate(
    bill_component = case_when(
      class == "volumetric" ~ value * `sum_out.electricity.total.energy_consumption`,
      TRUE ~ value
    )
  )
View(monthly_bill_by_component)

monthly_bills <- monthly_2022 |>
  left_join(exhibit_c_rates, by = "month") |>
  mutate(
    bill_component = case_when(
      class == "volumetric" ~ value * `sum_out.electricity.total.energy_consumption`,
      TRUE ~ value
    )
  ) |>
  group_by(bldg_id, month, utility, version) |>
  summarize(
    total_bill = sum(bill_component)
  )
View(monthly_bills)


annual_bills <- monthly_bills |>
  group_by(bldg_id, utility, version) |>
  summarize(
    annual_bill = sum(total_bill)
  )
View(annual_bills)

# print median bills by utility and version
annual_bills |>
  group_by(utility, version) |>
  summarize(
    median_bill = median(annual_bill),
    .groups = 'drop'
  ) |>
  mutate(
    median_bill = round(median_bill, 2)
  ) |>
  pivot_wider(
    names_from = version,
    values_from = median_bill
  ) |>
  mutate(
    approved_savings = baseline - hpr_approved,
    doer_savings = baseline - hpr_doer,
    approved_savings_pct = round(approved_savings / baseline * 100, 1),
    doer_savings_pct = round(doer_savings / baseline * 100, 1)
  ) |>
  print()


# Calculate percentage of bills above $6000 for each version
annual_bills |>
  filter(utility == "unitil") |>
  group_by(version) |>
  summarize(
    pct_above_6000 = mean(annual_bill > 6000) * 100,
    .groups = 'drop'
  ) |>
  mutate(
    pct_above_6000 = round(pct_above_6000, 1)
  ) |>
  print()


```

## Gas Bill Impacts Calculation
```{r}

View(monthly_gas_2022)


# Filter out homes that don't use any natural gas
ng_threshold <- 100
monthly_gas_2022 <- monthly_gas_2022 |>
  group_by(bldg_id) |>
  filter(sum(`sum_out.natural_gas.total.energy_consumption`) >= ng_threshold) |>
  ungroup()
View(monthly_gas_2022)

monthly_gas_by_component <- monthly_gas_2022 |>
  left_join(ma_gas_rates, by = "month") |>
  mutate(
    bill_component = case_when(
      class == "volumetric" ~ value * `sum_out.natural_gas.total.energy_consumption`,
      TRUE ~ value
    )
  ) |>
  filter(`sum_out.natural_gas.total.energy_consumption` >= 1e-5)

View(monthly_gas_by_component)

monthly_gas_bills <- monthly_gas_2022 |>
  left_join(ma_gas_rates, by = "month") |>
  mutate(
    bill_component = case_when(
      class == "volumetric" ~ value * `sum_out.natural_gas.total.energy_consumption`,
      TRUE ~ value
    )
  ) |>
  group_by(bldg_id, month, utility, version) |>
  summarize(
    total_bill = sum(bill_component)
  )
View(monthly_gas_bills)

annual_gas_bills <- monthly_gas_bills |>
  group_by(bldg_id, utility, version) |>
  summarize(
    annual_bill = sum(total_bill)
  )
View(annual_gas_bills)

```

# 4. Electric Bill Analysis (Plots)
## Bar Plot:Average monthly bills by utility, version and month
```{r}
monthly_bills_avg <- monthly_bills |>
  filter(utility == "unitil") |>
  group_by(month, version) |>
  summarize(
    avg_bill = mean(total_bill),
    .groups = 'drop'
  )

# Create the bar chart
ggplot(monthly_bills_avg, aes(x = factor(month), y = avg_bill, fill = version)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_x_discrete(labels = month.abb) +
  scale_y_continuous(labels = scales::dollar_format()) +
  scale_fill_manual(values = c(
    "baseline" = sb_colors[["midnight"]],
    "hpr_approved" = sb_colors[["carrot"]],
    "hpr_doer" = sb_colors[["sky"]]
  )) +
  labs(
    title = "Average Monthly Bills by Rate Type",
    x = "Month",
    y = "Average Bill Amount ($)",
    fill = "Rate Type"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )

ggsave(
  "src/ma_hp_rates/img/unitil_delivery_monthly.png",
  width = 5,
  height = 2,
  dpi = 300
)

```

## Histogram of Unitil bills
```{r}
# Create separate plots for each version
bin_width <- 500
x_max <- 15000
y_max <- 125
ma_median_energy_burden <- 6000
median_x_bump <- 3200
median_y <- 100

margin_theme <- theme(
  plot.margin = margin(0, 0, 0, 0, "cm"),
  axis.title.x = element_text(margin = margin(t = 1, b = 0)),
  axis.title.y = element_text(margin = margin(r = 1, l = 0))
)

p1 <- annual_bills |>
  filter(utility == "unitil", version == "hpr_doer") |>
  ggplot(aes(x = annual_bill)) +
  geom_histogram(binwidth = bin_width, fill = sb_colors[["sky"]]) +
  scale_y_continuous(limits = c(0, y_max)) +
  scale_x_continuous(labels = scales::dollar_format(), limits = c(0, x_max)) +
  geom_vline(xintercept = ma_median_energy_burden, color = "red", linetype = "dashed", linewidth = 0.5) +
  annotate(
    "text",
    x = ma_median_energy_burden + median_x_bump,
    y = median_y,
    label = "Median Energy Burden Threshold in MA",
    size = 14/.pt,
    color = "red",
    fontface = "bold"
  ) +
  labs(
    title = "Baseline Rate (RD-1)",
    x = "",
    y = "Density"
  ) +
  margin_theme

p2 <- annual_bills |>
  filter(utility == "unitil", version == "hpr_approved") |>
  ggplot(aes(x = annual_bill)) +
  geom_histogram(binwidth = bin_width, fill = sb_colors[["carrot"]]) +
  scale_y_continuous(limits = c(0, y_max)) +
  scale_x_continuous(labels = scales::dollar_format(), limits = c(0, x_max)) +
  geom_vline(xintercept = ma_median_energy_burden, color = "red", linetype = "dashed", linewidth = 0.5) +
  annotate(
    "text",
    x = ma_median_energy_burden + median_x_bump,
    y = median_y,
    label = "Median Energy Burden Threshold in MA",
    size = 14/.pt,
    color = "red",
    fontface = "bold"
  ) +
  labs(
    title = "Heat Pump Rate (Approved)",
    y = "Density",
    x = ""
  ) +
  margin_theme

p3 <- annual_bills |>
  filter(utility == "unitil", version == "baseline") |>
  ggplot(aes(x = annual_bill)) +
  geom_histogram(binwidth = bin_width, fill = sb_colors[["midnight"]]) +
  scale_y_continuous(limits = c(0, y_max)) +
  scale_x_continuous(labels = scales::dollar_format(), limits = c(0, x_max)) +
  geom_vline(xintercept = ma_median_energy_burden, color = "red", linetype = "dashed", linewidth = 0.5) +
  annotate(
    "text",
    x = ma_median_energy_burden + median_x_bump,
    y = median_y,
    label = "Median Energy Burden Threshold in MA",
    size = 14/.pt,
    color = "red",
    fontface = "bold"
  ) +
  labs(
    title = "Heat Pump Rate (DOER)",
    x = "Annual Bill Amount ($)",
    y = "Density"
  ) +
  margin_theme

# Combine plots vertically
combined_plot <- (p1 / p2 / p3)
print(combined_plot)

scale <- 1
ggsave(
  "src/ma_hp_rates/img/unitil_bills_histogram.png",
  combined_plot,
  width = 4,  # Reduced width to prevent text scaling issues
  height = 4, # Increased height to maintain proportions
  dpi = 300
)
```

## Savings Histograms
```{r}
View(annual_bills)

margin_theme <- theme(
  plot.margin = margin(0, 0, 0, 0, "cm"),
  axis.title.x = element_text(margin = margin(t = 1, b = 0)),
  axis.title.y = element_text(margin = margin(r = 1, l = 0))
)

# Calculate savings (differences) from baseline for both HPR versions
savings_comparison <- annual_bills |>
  filter(utility == "unitil") |>
  select(bldg_id, version, annual_bill) |>
  pivot_wider(
    names_from = version,
    values_from = annual_bill
  ) |>
  mutate(
    approved_savings = (baseline - hpr_approved)/baseline,
    doer_savings = (baseline - hpr_doer)/baseline
  ) |>
  select(bldg_id, approved_savings, doer_savings)

# View summary statistics
View(savings_comparison)

print(median(savings_comparison$doer_savings))


savings_bin_width <- .005
savings_x_max <- 0.4
# Create histograms for approved and doer savings
# Create the two histograms
p1 <- ggplot(savings_comparison, aes(x = doer_savings)) +
  geom_histogram(binwidth = savings_bin_width, fill = sb_colors[["sky"]]) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, savings_x_max)) +
  geom_vline(xintercept = median(savings_comparison$doer_savings), color = "black", linetype = "dashed", linewidth = 0.6) +
  # annotate(
  #   "text",
  #   x = median(savings_comparison$doer_savings)+0.05,
  #   y = 50,
  #   label = "Median Savings",
  #   size = 14/.pt,
  #   color = sb_colors[["sky"]],
  #   fontface = "bold"
  # ) +
  labs(
    title = "Savings from DOER HPR",
    x = "Savings (%)",
    y = "Count"
  ) +
  margin_theme

p2 <- ggplot(savings_comparison, aes(x = approved_savings)) +
  geom_histogram(binwidth = savings_bin_width, fill = sb_colors[["carrot"]]) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1), limits = c(0, savings_x_max)) +
  geom_vline(xintercept = median(savings_comparison$approved_savings), color = "black", linetype = "dashed", linewidth = 0.6) +
  # annotate(
  #   "text",
  #   x = median(savings_comparison$approved_savings) + 0.05,
  #   y = 108,
  #   label = "Median Savings",
  #   size = 14/.pt,
  #   color = sb_colors[["carrot"]],
  #   fontface = "bold"
  # ) +
  labs(
    title = "Savings from Approved HPR",
    x = "Savings (%)",
    y = "Count"
  ) +
  margin_theme

# Combine plots vertically
combined_plot <- (p1 / p2)
print(combined_plot)

scale <- 1
ggsave(
  "src/ma_hp_rates/img/unitil_savings_histogram.png",
  combined_plot,
  width = 4,
  height = 3,
  dpi = 300
)
```

## Bill Breakdown (Stacked Bar) - Single Utility
```{r}
#View(monthly_bill_by_component)
y_annotation_offset <- 50
x_label_offset <- 3.5

component_order <- c(
  "delivery_charge",   # Will be on top
  "supply_charge",     # Will be in middle
  "customer_charge"    # Will be on bottom
)
components <- factor(component_order, levels = (component_order))

# Create the color mapping using the same order
bill_component_colors <- setNames(
  c(sb_colors[["carrot"]],      # delivery (top)
    sb_colors[["pistachio"]],   # supply (middle)
    sb_colors[["midnight"]]),   # customer (bottom)
  component_order
)

# Use the order in your data processing
unitil_bill_by_component <- monthly_bill_by_component |>
  filter(utility == "eversource") |>
  group_by(bldg_id, version, type) |>
  summarize(
    annual_bill = sum(bill_component),
    .groups = 'drop'
  ) |>
  group_by(version, type) |>
  summarize(
    median_annual_bill = median(annual_bill),
    .groups = 'drop'
  ) |>
  arrange(match(type, component_order))

View(unitil_bill_by_component)




for (i in 0:length(components)) {
  # Take only first i components
  # Take only first i components
  subset_bills <- unitil_bill_by_component |>
    mutate(type = factor(type, levels =  component_order)) |>
    filter(type %in% tail(components, i))

  # Calculate subtotals for each rate up to current component
  baseline_sum <- sum(subset_bills$median_annual_bill[subset_bills$version == "baseline"])
  hpr_approved_sum <- sum(subset_bills$median_annual_bill[subset_bills$version == "hpr_approved"])
  hpr_doer_sum <- sum(subset_bills$median_annual_bill[subset_bills$version == "hpr_doer"])

  # Create dummy data with same factor levels, ensuring correct stacking order
  dummy_data <- unitil_bill_by_component |>
    complete(version, type) |>
    mutate(type = factor(type, levels =  component_order)) |>
    replace_na(list(median_annual_bill = 0))

  ggplot() +
    # Plot the dummy data with zero alpha to reserve legend space
    geom_bar(data = dummy_data,
             aes(x = factor(version), y = median_annual_bill, fill = type),
             stat = "identity",
             alpha = 0) +
    # Plot the actual subset data
    geom_bar(data = subset_bills,
             aes(x = factor(version), y = median_annual_bill, fill = type),
             position = "stack",
             stat = "identity") +
    scale_fill_manual(
      values = bill_component_colors,
      guide = "none"
    ) +
    scale_x_discrete(labels = c("baseline" = "RD-1", "hpr_approved" = "Approved", "hpr_doer" = "DOER")) +
    scale_y_continuous(labels = scales::dollar_format(), limits = c(0, 4700), breaks = seq(0, 5000, by = 500)) +
    labs(
      title = paste("Bill Components by Rate Version"),
      x = "",
      y = "Median Annual Bill ($)"
    ) +
    coord_cartesian(clip = "off") +  # Add this line before theme
    theme(
      text = element_text(family = "IBM-Plex-Sans"),
      axis.title.x = element_text(margin = margin(t = 3)),
      axis.title.y = element_text(margin = margin(r = 3)),
      legend.position = "right",
      plot.margin = margin(t = 3, r = 50, b = 0, l = 10, unit = "pt")
    ) +

    # Add labels of the stack types
    {
      if (i >= 1) annotate(
        "text",
        x = x_label_offset,
        y = 150,
        label = "Customer Charge",
        hjust = 0,  # 0 means left-justified, text will start at x=3.1
        family = "IBM-Plex-Sans",
        color = bill_component_colors["customer_charge"],
        size = 6,
        fontface = "bold"
      )
    } +
    {
      if (i >= 2) annotate(
        "text",
        x = x_label_offset,
        y = 800,
        label = "Supply Charge",
        hjust = 0,  # 0 means left-justified
        family = "IBM-Plex-Sans",
        color = bill_component_colors["supply_charge"],
        size = 6,
        fontface = "bold"
      )
    } +
    {
      if (i >= 3) annotate(
        "text",
        x = x_label_offset,
        y = 2500,
        label = "Delivery Charge",
        hjust = 0,  # 0 means left-justified
        family = "IBM-Plex-Sans",
        color = bill_component_colors["delivery_charge"],
        size = 6,
        fontface = "bold"
      )
    } +

    # Add subtotal annotations for current components
    annotate(
      "text",
      x = "baseline",
      y = (baseline_sum + y_annotation_offset),
      label = paste0("$", format(round(baseline_sum, 0), big.mark=",")),
    ) +
    annotate(
      "text",
      x = "hpr_approved",
      y = (hpr_approved_sum + y_annotation_offset),
      label = paste0("$", format(round(hpr_approved_sum, 0), big.mark=",")),
    ) +
    annotate(
      "text",
      x = "hpr_doer",
      y = (hpr_doer_sum + y_annotation_offset),
      label = paste0("$", format(round(hpr_doer_sum, 0), big.mark=",")),
    )

  ggsave(
    paste0("src/ma_hp_rates/img/bills_stacked_", i, ".png"),
    width = 3.5,
    height = 3,
    dpi = 300
  )
}
```

## Bill Breakdown (Stacked Bar) - All Utilities
```{r}
component_order <- c(
  "delivery_charge",   # Will be on top
  "supply_charge",     # Will be in middle
  "customer_charge"    # Will be on bottom
)
components <- factor(component_order, levels = (component_order))

# Calculate annual bills for each building, utility, and version
annual_bills_by_component <- monthly_bill_by_component |>
  group_by(bldg_id, version, utility, type) |>
  summarize(
    annual_bill = sum(bill_component),
    .groups = 'drop'
  ) |>
  group_by(version, utility, type) |>
  summarize(
    median_annual_bill = median(annual_bill),
    .groups = 'drop'
  ) |>
  arrange(match(type, component_order))

# First, set the correct order for utilities and versions in totals
totals <- annual_bills_by_component %>%
  group_by(utility, version) |>
  summarize(total = round(sum(median_annual_bill), 0)) |>
  ungroup() |>
  # Make sure utility and version are factors with the correct order
  mutate(
    utility = factor(utility, levels = c("unitil", "eversource", "national_grid")),
    version = factor(version, levels = c("baseline", "hpr_approved", "hpr_doer"))
  ) |>
  # Sort by these factors
  arrange(utility, version)

savings <- totals |>
  filter(version %in% c("baseline", "hpr_approved")) |>
  summarize(
    approved_savings = sum(total[version == "baseline"]) - sum(total[version == "hpr_approved"]),
    doer_savings = sum(total[version == "baseline"]) - sum(total[version == "hpr_doer"])
  )
View(savings)

savings_pct <- savings |>
  mutate(
    approved_savings_pct = approved_savings / sum(savings$approved_savings),
    doer_savings_pct = doer_savings / sum(savings$doer_savings)
  )
View(savings_pct)


# Pivot totals wider and calculate savings
totals_wide <- totals |>
  pivot_wider(
    names_from = utility,
    values_from = total
  ) |>
  # Add savings rows
  bind_rows(
    summarize(
      filter(totals, version %in% c("baseline", "hpr_approved")) |>
        pivot_wider(names_from = utility, values_from = total),
      version = "approved_savings",
      across(-version, ~ first(.x) - last(.x))
    ),
    summarize(
      filter(totals, version %in% c("baseline", "hpr_doer")) |>
        pivot_wider(names_from = utility, values_from = total),
      version = "doer_savings",
      across(-version, ~ first(.x) - last(.x))
    )
  ) |>
  # savings as a percentage
  bind_rows(
    summarize(
      filter(totals, version %in% c("baseline", "hpr_approved")) |>
        pivot_wider(names_from = utility, values_from = total),
      version = "approved_savings_pct",
      across(-version, ~ (first(.x) - last(.x)) / first(.x))
    ),
    summarize(
      filter(totals, version %in% c("baseline", "hpr_doer")) |>
        pivot_wider(names_from = utility, values_from = total),
      version = "doer_savings_pct",
      across(-version, ~ (first(.x) - last(.x)) / first(.x))
    )
  )
View(totals_wide)

# Convert totals_wide to long format for plotting
totals_long <- totals_wide |>
  pivot_longer(
    cols = c(unitil, eversource, national_grid),
    names_to = "utility",
    values_to = "total"
  ) |>
  # Keep only the rows we want to plot
  filter(version %in% c("baseline", "hpr_approved", "hpr_doer", "approved_savings", "doer_savings", "approved_savings_pct", "doer_savings_pct")) |>
  # Make sure utility and version are factors with the correct order
  mutate(
    utility = factor(utility, levels = c("unitil", "eversource", "national_grid")),
    version = factor(version, levels = c("baseline", "hpr_approved", "hpr_doer", "approved_savings", "doer_savings", "approved_savings_pct", "doer_savings_pct"))
  )
View(totals_long)



# Set the order of utilities and versions
annual_bills_by_component$utility <- factor(
  annual_bills_by_component$utility,
  levels = c("unitil", "eversource", "national_grid")
)

annual_bills_by_component$version <- factor(
  annual_bills_by_component$version,
  levels = c("baseline", "hpr_approved", "hpr_doer")
)

annual_bills_by_component$type <- factor(
  annual_bills_by_component$type,
  levels = c("delivery_charge", "supply_charge", "customer_charge")
)

# Create the color mapping using the same order
bill_component_colors <- setNames(
  c(sb_colors[["carrot"]],      # delivery (top)
    sb_colors[["pistachio"]],   # supply (middle)
    sb_colors[["midnight"]]),   # customer (bottom)
  component_order
)

# Create the plot
ggplot(annual_bills_by_component,
       aes(x = version, y = median_annual_bill, fill = type)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_grid(~utility, scales = "free_x", space = "free",
            labeller = as_labeller(c(
              "unitil" = "Unitil",
              "eversource" = "Eversource",
              "national_grid" = "National Grid"
            ))) +
  scale_fill_manual(
    values = bill_component_colors,
    labels = c(
      "customer_charge" = "Customer Charge",
      "supply_charge" = "Supply Charge",
      "delivery_charge" = "Delivery Charge"
    )
  ) +
  scale_x_discrete(labels = c(
    "baseline" = "Default",
    "hpr_approved" = "Approved",
    "hpr_doer" = "DOER"
  )) +
  scale_y_continuous(
    labels = scales::dollar_format(),
    breaks = seq(0, 5000, by = 1000)
  ) +
  labs(
    title = "",
    y = "Annual Bill ($)",
    x = "",
    fill = "Bill Component"
  ) +
  theme(
    panel.spacing = unit(1, "lines"),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    plot.margin = margin(t = 3, r = 10, b = 0, l = 50, unit = "pt"),
    legend.position = c(-0.15,0.2),
    legend.box = "vertical",
    legend.margin = margin(r = 20)  # Add some right margin to the legend
  ) +
  geom_text(
    data = totals,
    mapping = aes(x = version, y = total, label = scales::dollar(total)),
    inherit.aes = FALSE,  # Add this line to prevent inheriting the fill aesthetic
    vjust = -0.5,
    hjust = 0.5,
    family = "IBM-Plex-Sans",
    size = 5,
    fontface = "bold"
  )

ggsave(
  "src/ma_hp_rates/img/utility_comparison_stacked.png",
  width = 5,
  height = 3,
  dpi = 300
)

```

### Version with savings arrows
```{r}
# Create the plot
ggplot(annual_bills_by_component,
       aes(x = version, y = median_annual_bill, fill = type)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_grid(~utility, scales = "free_x", space = "free",
            labeller = as_labeller(c(
              "unitil" = "Unitil",
              "eversource" = "Eversource",
              "national_grid" = "National Grid"
            ))) +
  scale_fill_manual(
    values = bill_component_colors,
    labels = c(
      "customer_charge" = "Customer Charge",
      "supply_charge" = "Supply Charge",
      "delivery_charge" = "Delivery Charge"
    )
  ) +
  scale_x_discrete(labels = c(
    "baseline" = "Default",
    "hpr_approved" = "Approved",
    "hpr_doer" = "DOER"
  )) +
  scale_y_continuous(
    labels = scales::dollar_format(),
    breaks = seq(0, 5000, by = 1000)
  ) +
  labs(
    title = "",
    y = "Annual Bill ($)",
    x = "",
    fill = "Bill Component"
  ) +
  theme(
    panel.spacing = unit(1, "lines"),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    plot.margin = margin(t = 3, r = 10, b = 0, l = 50, unit = "pt"),
    legend.position = c(-0.15,0.2),
    legend.box = "vertical",
    legend.margin = margin(r = 20)  # Add some right margin to the legend
  ) +
  geom_text(
    data = totals,
    mapping = aes(x = version, y = total, label = scales::dollar(total)),
    inherit.aes = FALSE,  # Add this line to prevent inheriting the fill aesthetic
    vjust = -0.5,
    hjust = 0.5,
    family = "IBM-Plex-Sans",
    size = 5,
    fontface = "bold"
  ) +
  geom_segment(
    data = totals %>% filter(utility == "unitil"),
    aes(
      x = "hpr_approved",
      xend = "hpr_approved",
      y = total[version == "baseline"],
      yend = total[version == "hpr_approved"]
    ),
    arrow = arrow(length = unit(0.3, "cm")),
    color = sb_colors[["sky"]],
    linewidth = 0.8,
    inherit.aes = FALSE
  ) +
  geom_segment(
    data = totals %>% filter(utility == "eversource"),
    aes(
      x = "hpr_approved",
      xend = "hpr_approved",
      y = total[version == "baseline"],
      yend = total[version == "hpr_approved"]
    ),
    arrow = arrow(length = unit(0.3, "cm")),
    color = sb_colors[["sky"]],
    linewidth = 0.8,
    inherit.aes = FALSE
  ) +
  geom_segment(
    data = totals %>% filter(utility == "national_grid"),
    aes(
      x = "hpr_approved",
      xend = "hpr_approved",
      y = total[version == "baseline"],
      yend = total[version == "hpr_approved"]
    ),
    arrow = arrow(length = unit(0.3, "cm")),
    color = sb_colors[["sky"]],
    linewidth = 0.8,
    inherit.aes = FALSE
  )

ggsave(
  "src/ma_hp_rates/img/utility_comparison_stacked.png",
  width = 5,
  height = 3,
  dpi = 300
)

```



# 5. Gas Bill Analysis
### Monthly Box plot of Gas Bills
```{r}
ggplot(monthly_gas_bills, aes(x = factor(month), y = total_bill)) +
  geom_boxplot() +
  scale_x_discrete(
    labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
               "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
  ) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Monthly Gas Bills Distribution",
    x = "Month",
    y = "Total Bill ($)"
  ) +
  theme_minimal() +
  theme(
    panel.spacing = unit(1, "lines"),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 0, hjust = 0.5)
  )
```

### Histogram of gas bills
```{r}
binwidth <- 500
xmax <- 10000
ggplot(annual_gas_bills, aes(x = annual_bill)) +
  geom_histogram(binwidth = binwidth, fill = sb_colors[["carrot"]]) +
  scale_x_continuous(labels = scales::dollar_format(), limits = c(0, xmax)) +
  labs(
    title = "Histogram of Gas Bills",
    x = "Total Bill ($)",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    panel.spacing = unit(1, "lines"),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 0, hjust = 0.5)
  )
```


# Switching Analysis
### Unitil Switching

```{r}
View(annual_bills)
View(annual_gas_bills)

# Check if there are matching building IDs
print("\nNumber of unique building IDs in annual_bills:")
length(unique(annual_bills$bldg_id))
print("\nNumber of unique building IDs in annual_gas_bills:")
length(unique(annual_gas_bills$bldg_id))

# Check for any overlapping building IDs
print("\nNumber of overlapping building IDs:")
length(intersect(annual_bills$bldg_id, annual_gas_bills$bldg_id))

annual_bills_both_fuels <- annual_gas_bills |>
  left_join(annual_bills, by = c("bldg_id", "utility"), suffix = c("_gas", "_electric")) |>
  select(-version_gas) |>
  rename(version = version_electric) |>
  mutate(
    difference = annual_bill_electric - annual_bill_gas
  )
View(annual_bills_both_fuels)

median_bill_differences <- annual_bills_both_fuels |>
  group_by(version) |>
  summarize(median_difference = round(median(difference))) |>
  arrange(desc(median_difference))

median_bill_differences <- annual_bills_both_fuels |>
  group_by(version) |>
  summarize(
    median_difference = round(median(difference)),
    pct_savings = round(mean(difference < 0) * 100, 1)
  ) |>
  arrange(desc(median_difference))

print("\nPercentage of customers with bill savings by version:")
median_bill_differences |>
  select(version, pct_savings) |>
  mutate(
    summary = paste0(version, ": ", pct_savings, "% of customers save money")
  ) |>
  pull(summary) |>
  cat(sep = "\n")


```

### Histgram of switching impact
```{r}
bin_width <- 500
x_min <- -2000
x_max <- 5000
y_max <- 125

x_bump <- 50

View(annual_bills_both_fuels)

baseline_median_difference <- median_bill_differences |> filter(version == "baseline") |> pull(median_difference)
hpr_approved_median_difference <- median_bill_differences |> filter(version == "hpr_approved") |> pull(median_difference)
hpr_doer_median_difference <- median_bill_differences |> filter(version == "hpr_doer") |> pull(median_difference)




p1 <- annual_bills_both_fuels |>
  filter(version == "hpr_doer") |>
  ggplot(aes(x = difference)) +
  geom_histogram(binwidth = bin_width, fill = sb_colors[["sky"]]) +
  scale_y_continuous(limits = c(0, y_max)) +
  scale_x_continuous(labels = scales::dollar_format(), limits = c(x_min, x_max)) +
  geom_vline(xintercept = hpr_doer_median_difference, color = sb_colors[["pistachio"]], linetype = "dashed", linewidth = 0.5) +
  annotate(
    "text",
    x = hpr_doer_median_difference + x_bump,
    y = 120,
    label = paste("Median Bill Difference: ", scales::dollar(hpr_doer_median_difference)),
    size = 14/.pt,
    color = sb_colors[["sky"]],
    fontface = "bold",
    hjust = 0
  ) +
  labs(
    title = "Heat Pump Rate (DOER)",
    x = "",
    y = "Density"
  ) +
  margin_theme

p2 <- annual_bills_both_fuels |>
  filter(version == "hpr_approved") |>
  ggplot(aes(x = difference)) +
  geom_histogram(binwidth = bin_width, fill = sb_colors[["carrot"]]) +
  scale_y_continuous(limits = c(0, y_max)) +
  scale_x_continuous(labels = scales::dollar_format(), limits = c(x_min, x_max)) +
  geom_vline(xintercept = hpr_approved_median_difference, color = sb_colors[["pistachio"]], linetype = "dashed", linewidth = 0.5) +
  annotate(
    "text",
    x = hpr_approved_median_difference + x_bump,
    y = 120,
    label = paste("Median Bill Difference: ", scales::dollar(hpr_approved_median_difference)),
    size = 14/.pt,
    color = sb_colors[["carrot"]],
    fontface = "bold",
    hjust = 0
  ) +
  labs(
    title = "Heat Pump Rate (Approved)",
    y = "Density",
    x = ""
  ) +
  margin_theme

p3 <- annual_bills_both_fuels |>
  filter(version == "baseline") |>
  ggplot(aes(x = difference)) +
  geom_histogram(binwidth = bin_width, fill = sb_colors[["midnight"]]) +
  scale_y_continuous(limits = c(0, y_max)) +
  scale_x_continuous(labels = scales::dollar_format(), limits = c(x_min, x_max)) +
  geom_vline(xintercept = baseline_median_difference, color = sb_colors[["pistachio"]], linetype = "dashed", linewidth = 0.5) +
  annotate(
    "text",
    x = baseline_median_difference + x_bump,
    y = 120,
    label = paste("Median Bill Difference: ", scales::dollar(baseline_median_difference)),
    size = 14/.pt,
    color = sb_colors[["midnight"]],
    fontface = "bold",
    hjust = 0
  ) +
  labs(
    title = "Baseline Rate (RD-1)",
    x = "Difference Between Electric and Gas Bills ($)",
    y = "Density"
  ) +
  margin_theme

# Combine plots vertically
combined_plot <- (p1 / p2 / p3)
print(combined_plot)

ggsave(
  "src/ma_hp_rates/img/bill_difference_histogram.png",
  combined_plot,
  width = 4,
  height = 4,
  dpi = 300
)
```
