---
title: rates_prototype

toc: true
reference-location: margin
fig-cap-location: bottom

appendix-style: default
citation-location: document
citation:
  container-title: Switchbox

format:
  html:
    page-layout: full
---
# R Boilerploate
```{r}
library(tidyverse)
library(scales)

library(gt)
library(gtExtras)
library(jsonlite)
library(RSQLite)

library(patchwork)

source("/workspaces/reports/lib/ggplot/switchbox_theme.R")
source("/workspaces/reports/lib/utility_mapping.R")
source("/workspaces/reports/lib/rates_analysis/heat_pump_rate_funcs.R")
```


# Gather Delivery Tariffs and Supply Rates
## Supply Rates
### MA Electric Supply Rates
```{r}
supply_rates_elec_ma_monthly <- read_csv("/workspaces/reports/data/utility_rates/electric/madpu/ma_supply_rates_monthly_20250329.csv")

con <- DBI::dbConnect(duckdb::duckdb(), "/workspaces/reports/data/ResStock/2022_resstock_amy2018_release_1.1/rs_20250227.db")
# Write the supply rates to the database
DBI::dbWriteTable(
  con,
  "supply_rates_elec_ma_monthly",
  supply_rates_elec_ma_monthly,
  overwrite = TRUE
)
DBI::dbDisconnect(con)
```

### MA Gas Supply Rates
```{r}
resi_codes <- c("R-1", "R-1 ", "Residential", "All Residential")

# Read in the gas supply rates CSV
supply_rates_gas_ma_monthly <- read_csv("/workspaces/reports/data/utility_rates/gas/ma/madpu/gas_supply_rates_tidy_filled.csv") |>
  as_tibble() |>
  #filter(utility %in% c("liberty_blackstone", "liberty_north_attleboro")) |>
  filter(rate_class %in% resi_codes)

save_path <- "/workspaces/reports/src/ma_hp_rates/img/gas_supply_rates.png"
plot_gas_supply_rates(supply_rates_gas_ma_monthly, save_path)
```

```{r}
# plot Eversource's gas and electric supply rates
library(lubridate)
first_year <- 2017
# First create the date column for both datasets
supply_rates_elec <- supply_rates_elec_ma_monthly |>
  filter(electric_utility %in% c("eversource_nstar", "eversource_wmeco")) |>
  filter(year >= first_year) |>
  mutate(
    date = ymd(paste(year, month, "01", sep = "-")),
    type = "electricity"
  ) |>
  arrange(electric_utility, date)  # This will put nstar before wmeco alphabetically

supply_rates_gas <- supply_rates_gas_ma_monthly |>
  filter(gas_utility %in% c("eversource_nstar","eversource_bay_state")) |>
  filter(year >= first_year) |>
  mutate(
    date = ymd(paste(year, month, "01", sep = "-")),
    type = "gas"
  ) |>
  arrange(gas_utility, date)  # This will put nstar before bay_state alphabetically


show_elec = FALSE
show_gas = TRUE
# Create the plot
ggplot() +
  # Electric supply rates
  if (show_elec) {
    geom_line(data = supply_rates_elec,
              aes(x = date, y = supply_rate,
                  color = paste(electric_utility, "Electric")),
              linewidth = 0.5) +
    geom_point(data = supply_rates_elec,
              aes(x = date, y = supply_rate,
                  color = paste(electric_utility, "Electric")),
              size = 0.2)
  } +

  # Gas supply rates
  geom_line(data = supply_rates_gas,
            aes(x = date, y = gaf,
                color = paste(gas_utility, "Gas")),
            linewidth = 0.5) +
  geom_point(data = supply_rates_gas,
             aes(x = date, y = gaf,
                 color = paste(gas_utility, "Gas")),
             size = 0.2) +

  # Single color scale for all utilities
  scale_color_manual(
    values = c(
      "eversource_nstar Electric" = "#023047",    # midnight
      "eversource_wmeco Electric" = "#FC9706",    # carrot
      "eversource_nstar Gas" = "#023047",         # midnight
      "eversource_bay_state Gas" = "#68BED8"      # sky
    ),
    labels = c(
      "eversource_nstar Electric" = "Eversource NStar Electric",
      "eversource_wmeco Electric" = "Eversource WMeCo Electric",
      "eversource_nstar Gas" = "Eversource NStar Gas",
      "eversource_bay_state Gas" = "Eversource Bay State Gas"
    ),
    breaks = c(
      "eversource_nstar Electric",
      "eversource_wmeco Electric",
      "eversource_nstar Gas",
      "eversource_bay_state Gas"
    ),
    name = "Utility"
  ) +

  # Axes and labels
  scale_y_continuous(
    name = "Supply Rate ($/unit)",
    labels = scales::dollar_format()
  ) +
  scale_x_date(
    date_breaks = "1 year",
    date_labels = "%Y"
  ) +

  # Title and theme
  labs(
    title = "Eversource Energy Supply Rates",
    x = "Date"
  ) +
  guides(
    color = guide_legend(nrow = 2, ncol = 2)
  ) +
  theme(
    legend.position = c(0.02, 0.98),  # Position in the upper left
    legend.justification = c(0, 1),    # Anchor point at top left
    legend.box = "vertical",
    legend.background = element_rect(fill = "grey90", color = NA),
    legend.spacing.y = unit(0.02, "cm"),  # Reduce vertical spacing between legend items
    legend.key.height = unit(0.3, "lines"),  # Reduce height of legend keys
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "white", color = "white"),
    axis.line = element_line(linewidth = 0.5),
    axis.ticks = element_line(color = "black")
  )

save_path <- "/workspaces/reports/src/ma_hp_rates/img/eversource_supply_rates.png"
ggsave(save_path,
      width = 5,
      height = 4,
      dpi = 300,
      bg = "white")

```


```{r}
# for eversource_nstar, plot the "spark gap" ... = [p_elec/p_gas]

supply_rates_eversource_nstar <- supply_rates_gas |>
  filter(gas_utility == "eversource_nstar") |>
  select(date, gaf) |>
  inner_join(
    supply_rates_elec |>
      filter(electric_utility == "eversource_nstar"),
      select(date, supply_rate),
    by = "date"
  ) |>
  mutate(spark_gap = supply_rate / gaf) |>
  pivot_longer(
    cols = c(supply_rate, gaf, spark_gap),
    names_to = "rate_type",
    values_to = "rate"
  )

ggplot(
  supply_rates_eversource_nstar |> filter(rate_type == "spark_gap"),
  aes(x = date, y = rate)
) +
  geom_line(color = "green") +
  geom_point(color = "green") +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(
    title = "Eversource NStar Spark Gap",
    x = "Date",
    y = "Spark Gap [Electric kWh / Gas kWh]"
  ) +
  theme(
    legend.position = "none"
  )


ggplot(
  supply_rates_eversource_nstar |> filter(rate_type == "spark_gap"),
  aes(x = date, y = rate)
) +
  geom_line(color = "#023047", linewidth = 0.5) +
  geom_point(color = "#023047", size = 0.3) +
  #geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed") +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(limits = c(0, 30)) +
  labs(
    title = "Eversource NStar Spark Gap",
    x = "Date",
    y = "Spark Gap [Electric kWh / Gas kWh]"
  ) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
  )

save_path <- "/workspaces/reports/src/ma_hp_rates/img/eversource_nstar_spark_gap.png"
ggsave(save_path,
      width = 5,
      height = 4,
      dpi = 300,
      bg = "white")

```



## MA Electric Delivery Tariffs
```{r}
library(googlesheets4)

url_sheet_delivery_tariffs_elec_ma <- "https://docs.google.com/spreadsheets/d/1efE96zgziHTaLX-lPV1vRKVZze9gcKuD-nJyB-QFOQY/edit?gid=658606865#gid=658606865"
googlesheets4::gs4_deauth()
tariffs_elec_raw <- read_sheet(url_sheet_delivery_tariffs_elec_ma, sheet = "tariffs") |>
  mutate(tariff_id = paste0(electric_utility, "_", version)) |>
  select(-source)

summer_months <- c(5,6,7,8,9,10)
winter_months <- c(11,12,1,2,3,4)

# The tariffs are specified by season, so we need to expand them out to the monthly grid
delivery_tariffs_elec <- tariffs_elec_raw |>
  mutate(
    month = if_else(season == "summer",
                    list(summer_months),
                    list(winter_months))
  ) |>
  select(-season) |>
  unnest(month) |>
  arrange(electric_utility, month)

con <- DBI::dbConnect(duckdb::duckdb(), "/workspaces/reports/data/ResStock/2022_resstock_amy2018_release_1.1/rs_20250227.db")
DBI::dbWriteTable(
  con,
  "ma_elec_delivery_tariffs",
  delivery_tariffs_elec,
  overwrite = TRUE
)
DBI::dbDisconnect(con)
```


## MA Gas Delivery Tariffs
- MADPU has a page for each utility (link)[https://www.mass.gov/info-details/natural-gas-tariffs-and-delivery-rates#current-natural-gas-tariffs-and-delivery-rates-]
-

```{r}
library(googlesheets4)

url_sheet_delivery_tariffs_gas_ma <- "https://docs.google.com/spreadsheets/d/1ikBwsX3UF4RheRrEoQzAHFqwXbI07Z-iXEiIEYraFNI/edit?gid=0#gid=0"
googlesheets4::gs4_deauth()
tariffs_gas_raw <- read_sheet(url_sheet_delivery_tariffs_gas_ma, sheet = "delivery_tariffs_gas_ma") |>
  mutate(tariff_id = paste0(gas_utility, "_", tariff_name)) |>

  # 1 therm = 29.3 kWh...ResStock data is in kWh
  mutate(
    value = if_else(billing_unit == "therm", value / 29.3, value)
  ) |>
  select(-source)

summer_months <- c(5,6,7,8,9,10)
winter_months <- c(11,12,1,2,3,4)

# The tariffs are specified by season, so we need to expand them out to the monthly grid
delivery_tariffs_gas <- tariffs_gas_raw |>
  mutate(
    month = if_else(season == "summer",
                    list(summer_months),
                    list(winter_months))
  ) |>
  select(-season) |>
  unnest(month) |>
  arrange(gas_utility, month)

```


# Calculate Bills

## Loads
```{r}

# -------------
electric_utility = 'eversource'
gas_utility = 'eversource'
gas_supply_year <- 2024
# -------------


path_to_database <- "/workspaces/reports/data/ResStock/2022_resstock_amy2018_release_1.1/rs_20250329.db"
con <- DBI::dbConnect(duckdb::duckdb(), path_to_database)

# Mapping bldg_ids to utilities
ma_hh_utilities <- read_csv("/workspaces/reports/data/ResStock/utility_lookups/MA_hh_utilities.csv")
#get a count of bldg_ids by electric_utility
ma_hh_utilities |>
  group_by(electric_utility) |>
  summarize(n_bldgs = n_distinct(bldg_id))

# get a count of bldg_ids by gas_utility
ma_hh_utilities |>
  group_by(gas_utility) |>
  summarize(n_bldgs = n_distinct(bldg_id))

# get a count of bldg_ids that have both {electric_utility} electric and {gas_utility} gas
ma_hh_utilities |>
  filter(electric_utility == !!electric_utility & gas_utility == !!gas_utility) |>
  summarize(n_bldgs = n_distinct(bldg_id)) |>
  pull(n_bldgs)



all_bldg_ids <- ma_hh_utilities |>
    filter(electric_utility == !!electric_utility & gas_utility == !!gas_utility) |>
    pull(bldg_id)

static_sample_bldg_ids <- c(349309, 353457, 374636, 168141, 210903, 93188, 39967, 47874, 515247, 16373)


use_subset <- FALSE
if (use_subset) {
  if (new_sample) {
    sample_bldg_ids <- all_bldg_ids |>
      sample(10)
  } else {
    sample_bldg_ids <- static_sample_bldg_ids
  }
} else {
  sample_bldg_ids <- all_bldg_ids
}


# ---------
#### Electricity
# to-do: use dyplyer!!!!!
# make it a function
# probably all these can be into one giant long table


monthly_load_elec_baseline <- DBI::dbGetQuery(
  con,
  sprintf(
    "SELECT bldg_id, month, upgrade, \"sum_out.electricity.total.energy_consumption\" as elec_consumption
     FROM ma_monthly
     WHERE upgrade IN (0) AND bldg_id IN ('%s')",
    paste(sample_bldg_ids, collapse = "','")
  )
)
monthly_load_elec_baseline <- monthly_load_elec_baseline |>
  left_join(ma_hh_utilities, by = "bldg_id")
#View(monthly_load_elec_baseline)

monthly_load_elec_upgrade4 <- DBI::dbGetQuery(
  con,
  sprintf(
    "SELECT bldg_id, month, upgrade, \"sum_out.electricity.total.energy_consumption\" as elec_consumption
     FROM ma_monthly
     WHERE upgrade IN (4) AND bldg_id IN ('%s')",
    paste(sample_bldg_ids, collapse = "','")
  )
)
monthly_load_elec_upgrade4 <- monthly_load_elec_upgrade4 |>
  left_join(ma_hh_utilities, by = "bldg_id")
#View(monthly_load_elec_upgrade4)

# ---------
#### Gas
monthly_load_gas_baseline <- DBI::dbGetQuery(
  con,
  sprintf(
    "SELECT bldg_id, month, upgrade, \"sum_out.natural_gas.total.energy_consumption\" as gas_consumption
     FROM ma_monthly
     WHERE upgrade IN (0) AND bldg_id IN ('%s')",
    paste(sample_bldg_ids, collapse = "','")
  )
)
monthly_load_gas_baseline <- monthly_load_gas_baseline |>
  left_join(ma_hh_utilities, by = "bldg_id")
#View(monthly_load_gas_baseline)

monthly_load_gas_upgrade4 <- DBI::dbGetQuery(
  con,
  sprintf(
    "SELECT bldg_id, month, upgrade, \"sum_out.natural_gas.total.energy_consumption\" as gas_consumption
     FROM ma_monthly
     WHERE upgrade IN (4) AND bldg_id IN ('%s')",
    paste(sample_bldg_ids, collapse = "','")
  )
)
monthly_load_gas_upgrade4 <- monthly_load_gas_upgrade4 |>
  left_join(ma_hh_utilities, by = "bldg_id")
#View(monthly_load_gas_upgrade4)

# ---------
# Disconnect from the database
DBI::dbDisconnect(con)

```


## Monthly Electricity Bills
```{r}

calculate_elec_bills_monthly <- function(monthly_elec_load, delivery_tariffs, supply_rates) {
  monthly_elec_load |>
    select(bldg_id, month, elec_consumption, electric_utility) |>

    # Part 1: Delivery Tariffs -> 3 joins for customer_charge, delivery_rate, sales_tax
    left_join(
      delivery_tariffs |>
        filter(type == "customer_charge") |>
        select(month, electric_utility, value, version, tariff_id) |>
        rename(customer_charge = value),
        by = c("month" = "month", "electric_utility" = "electric_utility")  # Explicit join
    ) |>
    left_join(
      delivery_tariffs |>
        filter(type == "delivery_rate") |>
        select(month, electric_utility, value, version, tariff_id) |>
        rename(delivery_rate = value),
        by = c("month" = "month", "electric_utility" = "electric_utility", "version" = "version", "tariff_id" = "tariff_id")  # Explicit join
    ) |>
    left_join(
      delivery_tariffs |>
        filter(type == "sales_tax_rate") |>
        select(month, electric_utility, value, version, tariff_id) |>
        rename(sales_tax_rate = value),
        by = c("month" = "month", "electric_utility" = "electric_utility", "version" = "version", "tariff_id" = "tariff_id")  # Explicit join
    ) |>

    # Part 2: Supply rates -> join on utility and month
    left_join(
      supply_rates |>
        select(-year),
      by = c("month" = "month", "electric_utility" = "electric_utility")  # Explicit join
    ) |>

    # Part 3: Calculate monthly bill
    mutate(
      delivery_charge = elec_consumption * delivery_rate,
      supply_charge = elec_consumption * supply_rate,
      total_pretax_bill = supply_charge + delivery_charge + customer_charge,
      sales_tax_charge = total_pretax_bill * sales_tax_rate,
      total_bill = total_pretax_bill + sales_tax_charge
    ) |>
    arrange(bldg_id, month, electric_utility, version, tariff_id)
}


# Explicity choose the supply year
monthly_elec_bills_baseline_2018 <- calculate_elec_bills_monthly(
                                      monthly_load_elec_baseline |> filter(electric_utility == !!electric_utility),
                                      delivery_tariffs_elec,
                                      supply_rates_elec_ma_monthly |> filter(year == 2018)
                                      )
monthly_elec_bills_baseline_2024 <- calculate_elec_bills_monthly(
                                      monthly_load_elec_baseline |> filter(electric_utility == !!electric_utility),
                                      delivery_tariffs_elec,
                                      supply_rates_elec_ma_monthly |> filter(year == 2024)
                                      )
#View(monthly_elec_bills_baseline_2024)


monthly_elec_bills_upgrade4_2018 <- calculate_elec_bills_monthly(
                                      monthly_load_elec_upgrade4 |> filter(electric_utility == !!electric_utility),
                                      delivery_tariffs_elec,
                                      supply_rates_elec_ma_monthly |> filter(year == 2018)
                                      )
monthly_elec_bills_upgrade4_2024 <- calculate_elec_bills_monthly(
                                      monthly_load_elec_upgrade4 |> filter(electric_utility == !!electric_utility),
                                      delivery_tariffs_elec,
                                      supply_rates_elec_ma_monthly |> filter(year == 2024)
                                      )
#View(monthly_elec_bills_upgrade4_2024)
```


## Monthly Gas Bills
```{r}

calculate_gas_bills_monthly <- function(monthly_gas_load, delivery_tariffs, supply_rates) {
  monthly_gas_load |>
    select(bldg_id, month, gas_consumption, gas_utility) |>

    # Part 1: Delivery Tariffs -> 3 joins for customer_charge, delivery_rate, sales_tax
    left_join(
      delivery_tariffs |>
        filter(type == "customer_charge") |>
        select(month, gas_utility, value, version, tariff_id) |>
        rename(customer_charge = value),
        by = c("month" = "month", "gas_utility" = "gas_utility")  # Explicit join
    ) |>
    left_join(
      delivery_tariffs |>
        filter(type == "distribution_rate") |>
        select(month, gas_utility, value, version, tariff_id) |>
        rename(distribution_rate = value),
        by = c("month" = "month", "gas_utility" = "gas_utility", "version" = "version", "tariff_id" = "tariff_id")  # Explicit join
    ) |>
    left_join(
      delivery_tariffs |>
        filter(type == "sales_tax_rate") |>
        select(month, gas_utility, value, version, tariff_id) |>
        rename(sales_tax_rate = value),
        by = c("month" = "month", "gas_utility" = "gas_utility", "version" = "version", "tariff_id" = "tariff_id")  # Explicit join
    ) |>

    # Part 2: Supply rates -> join on utility and month
    left_join(
      supply_rates |>
        select(-year),
      by = c("month" = "month", "gas_utility" = "gas_utility")  # Explicit join
    ) |>

    # Part 3: Calculate monthly bill
    mutate(
      distribution_charge = gas_consumption * distribution_rate,
      supply_charge = gas_consumption * gaf,
      total_pretax_bill = supply_charge + distribution_charge + customer_charge,
      sales_tax_charge = total_pretax_bill * sales_tax_rate,
      total_bill = total_pretax_bill + sales_tax_charge
    ) |>
    arrange(bldg_id, month, gas_utility, tariff_id)
}

monthly_gas_bills_baseline_2018 <- calculate_gas_bills_monthly(monthly_load_gas_baseline |> filter(gas_utility == !!gas_utility),
                                                              delivery_tariffs_gas |> filter(tariff_name == "r3_resi_heat"),
                                                              supply_rates_gas_ma_monthly |> filter(year == 2018))

monthly_gas_bills_upgrade4_2018 <- calculate_gas_bills_monthly(monthly_load_gas_upgrade4 |> filter(gas_utility == !!gas_utility),
                                                              delivery_tariffs_gas |> filter(tariff_name == "r1_resi_nonheat"),
                                                              supply_rates_gas_ma_monthly |> filter(year == 2018))

monthly_gas_bills_baseline_2024 <- calculate_gas_bills_monthly(monthly_load_gas_baseline |> filter(gas_utility == !!gas_utility),
                                                              delivery_tariffs_gas |> filter(tariff_name == "r3_resi_heat"),
                                                              supply_rates_gas_ma_monthly |> filter(year == 2024))

monthly_gas_bills_upgrade4_2024 <- calculate_gas_bills_monthly(monthly_load_gas_upgrade4 |> filter(gas_utility == !!gas_utility),
                                                              delivery_tariffs_gas |> filter(tariff_name == "r1_resi_nonheat"),
                                                              supply_rates_gas_ma_monthly |> filter(year == 2024))


#View(monthly_gas_bills_baseline_2018)
#View(monthly_gas_bills_upgrade4_2018)
#View(monthly_gas_bills_baseline_2024)
#View(monthly_gas_bills_upgrade4_2024)
```



## Annual Bills
```{r}

calculate_annual_elec_bills_from_monthly <- function(monthly_bills) {
  monthly_bills   |>
    group_by(bldg_id, electric_utility, version, tariff_id) |>
    summarize(
      annual_bill = sum(total_bill)
    ) |>
    ungroup()
}

annual_elec_bills_baseline_2018 <- calculate_annual_elec_bills_from_monthly(monthly_elec_bills_baseline_2018)
annual_elec_bills_upgrade4_2018 <- calculate_annual_elec_bills_from_monthly(monthly_elec_bills_upgrade4_2018)

annual_elec_bills_baseline_2024 <- calculate_annual_elec_bills_from_monthly(monthly_elec_bills_baseline_2024)
annual_elec_bills_upgrade4_2024 <- calculate_annual_elec_bills_from_monthly(monthly_elec_bills_upgrade4_2024)


calculate_annual_gas_bills_from_monthly <- function(monthly_bills) {
  monthly_bills   |>
    group_by(bldg_id, gas_utility, version, tariff_id) |>
    summarize(
      annual_bill = sum(total_bill)
    ) |>
    ungroup()
}

annual_gas_bills_baseline_2018 <- calculate_annual_gas_bills_from_monthly(monthly_gas_bills_baseline_2018)
annual_gas_bills_upgrade4_2018 <- calculate_annual_gas_bills_from_monthly(monthly_gas_bills_upgrade4_2018)

annual_gas_bills_baseline_2024 <- calculate_annual_gas_bills_from_monthly(monthly_gas_bills_baseline_2024)
annual_gas_bills_upgrade4_2024 <- calculate_annual_gas_bills_from_monthly(monthly_gas_bills_upgrade4_2024)

```


# Analysis
## Annual Bills
```{r}

#pre-electrification
annual_bills_baseline_2018_pre <- annual_gas_bills_baseline_2018 |>
  rename(annual_gas_bill = annual_bill, tariff_id_gas = tariff_id, version_gas = version) |>
  left_join(
    annual_elec_bills_baseline_2018 |>
      rename(annual_elec_bill = annual_bill, tariff_id_elec = tariff_id, version_elec = version),
    by = c("bldg_id")
  ) |>
  mutate(
    annual_total_bill = annual_elec_bill + annual_gas_bill
  ) |>
  filter(version_elec == "baseline")

# Post-electrification (upgrade 4)
annual_bills_upgrade4_2018_post <- annual_gas_bills_upgrade4_2018 |>
  rename(annual_gas_bill = annual_bill, tariff_id_gas = tariff_id, version_gas = version) |>
  left_join(
    annual_elec_bills_upgrade4_2018 |>
      rename(annual_elec_bill = annual_bill, tariff_id_elec = tariff_id, version_elec = version),
    by = c("bldg_id")
  ) |>
  mutate(
    annual_total_bill = annual_elec_bill + annual_gas_bill
  ) |>
  filter(version_elec == "baseline")


annual_bills_upgrade4_2018_hpr_approved <- annual_gas_bills_upgrade4_2018 |>
  rename(annual_gas_bill = annual_bill, tariff_id_gas = tariff_id, version_gas = version) |>
  left_join(
    annual_elec_bills_upgrade4_2018 |>
      rename(annual_elec_bill = annual_bill, tariff_id_elec = tariff_id, version_elec = version),
    by = c("bldg_id")
  ) |>
  mutate(
    annual_total_bill = annual_elec_bill + annual_gas_bill
  ) |>
  filter(version_elec == "hpr_approved")


annual_bills_upgrade4_2018_hpr_doer <- annual_gas_bills_upgrade4_2018 |>
  rename(annual_gas_bill = annual_bill, tariff_id_gas = tariff_id, version_gas = version) |>
  left_join(
    annual_elec_bills_upgrade4_2018 |>
      rename(annual_elec_bill = annual_bill, tariff_id_elec = tariff_id, version_elec = version),
    by = c("bldg_id")
  ) |>
  mutate(
    annual_total_bill = annual_elec_bill + annual_gas_bill
  ) |>
  filter(version_elec == "hpr_doer")



# Check input table sizes
cat("Baseline pre rows:", nrow(annual_bills_baseline_2018_pre), "\n")
cat("Baseline post rows:", nrow(annual_bills_upgrade4_2018_post), "\n")
cat("HPR approved rows:", nrow(annual_bills_upgrade4_2018_hpr_approved), "\n")
  cat("HPR DOER rows:", nrow(annual_bills_upgrade4_2018_hpr_doer), "\n")

annual_bills_2022_all_scenarios <- bind_rows(
  annual_bills_baseline_2018_pre |> mutate(scenario = "baseline_pre"),
  annual_bills_upgrade4_2018_post |> mutate(scenario = "baseline_post"),
  annual_bills_upgrade4_2018_hpr_approved |> mutate(scenario = "hpr_approved"),
  annual_bills_upgrade4_2018_hpr_doer |> mutate(scenario = "hpr_doer")
)

cat("\nCombined table rows:", nrow(annual_bills_2022_all_scenarios), "\n")
View(annual_bills_2022_all_scenarios)


# -----------------------
# Print medians of annual total bills for each scenario
cat("Pre-Electrification median annual bill 2018: $",
    format(round(median(annual_bills_baseline_2018_pre$annual_total_bill), 2),
           big.mark=","), "\n")

cat("Post-Electrification median annual bill 2018: $",
    format(round(median(annual_bills_upgrade4_2018_post$annual_total_bill), 2),
           big.mark=","), "\n")

cat("HPR Approved median annual bill 2018: $",
    format(round(median(annual_bills_upgrade4_2018_hpr_approved$annual_total_bill), 2),
           big.mark=","), "\n")

cat("HPR DOER median annual bill 2018: $",
    format(round(median(annual_bills_upgrade4_2018_hpr_doer$annual_total_bill), 2),
           big.mark=","), "\n")

# -----------------------
# Savings

annual_bills_2018 <- annual_bills_baseline_2018_pre |>
  select(bldg_id, annual_total_bill) |>
  rename(baseline_pre_bill = annual_total_bill) |>
  left_join(
    annual_bills_upgrade4_2018_post |>
      select(bldg_id, annual_total_bill) |>
      rename(baseline_post_bill = annual_total_bill),
    by = "bldg_id"
  ) |>
  left_join(
    annual_bills_upgrade4_2018_hpr_approved |>
      select(bldg_id, annual_total_bill) |>
      rename(hpr_approved_bill = annual_total_bill),
    by = "bldg_id"
  ) |>
  left_join(
    annual_bills_upgrade4_2018_hpr_doer |>
      select(bldg_id, annual_total_bill) |>
      rename(hpr_doer_bill = annual_total_bill),
    by = "bldg_id"
  ) |>
  mutate(
    post_savings = baseline_post_bill - baseline_pre_bill,
    post_savings_pct = post_savings / baseline_pre_bill,
    approved_savings = hpr_approved_bill - baseline_pre_bill,
    approved_savings_pct = approved_savings / baseline_pre_bill,
    doer_savings = hpr_doer_bill - baseline_pre_bill,
    doer_savings_pct = doer_savings / baseline_pre_bill
  ) |>
  mutate(
    supply_year = 2018
  )

View(annual_bills_2018)
```

```{r}

#pre-electrification
annual_bills_baseline_2024_pre <- annual_gas_bills_baseline_2024 |>
  rename(annual_gas_bill = annual_bill, tariff_id_gas = tariff_id, version_gas = version) |>
  left_join(
    annual_elec_bills_baseline_2024 |>
      rename(annual_elec_bill = annual_bill, tariff_id_elec = tariff_id, version_elec = version),
    by = c("bldg_id")
  ) |>
  mutate(
    annual_total_bill = annual_elec_bill + annual_gas_bill
  ) |>
  filter(version_elec == "baseline")
# Post-electrification (upgrade 4)
annual_bills_upgrade4_2024_post <- annual_gas_bills_upgrade4_2024 |>
  rename(annual_gas_bill = annual_bill, tariff_id_gas = tariff_id, version_gas = version) |>
  left_join(
    annual_elec_bills_upgrade4_2024 |>
      rename(annual_elec_bill = annual_bill, tariff_id_elec = tariff_id, version_elec = version),
    by = c("bldg_id")
  ) |>
  mutate(
    annual_total_bill = annual_elec_bill + annual_gas_bill
  ) |>
  filter(version_elec == "baseline")

annual_bills_upgrade4_2024_hpr_approved <- annual_gas_bills_upgrade4_2024 |>
  rename(annual_gas_bill = annual_bill, tariff_id_gas = tariff_id, version_gas = version) |>
  left_join(
    annual_elec_bills_upgrade4_2024 |>
      rename(annual_elec_bill = annual_bill, tariff_id_elec = tariff_id, version_elec = version),
    by = c("bldg_id")
  ) |>
  mutate(
    annual_total_bill = annual_elec_bill + annual_gas_bill
  ) |>
  filter(version_elec == "hpr_approved")

annual_bills_upgrade4_2024_hpr_doer <- annual_gas_bills_upgrade4_2024 |>
  rename(annual_gas_bill = annual_bill, tariff_id_gas = tariff_id, version_gas = version) |>
  left_join(
    annual_elec_bills_upgrade4_2024 |>
      rename(annual_elec_bill = annual_bill, tariff_id_elec = tariff_id, version_elec = version),
    by = c("bldg_id")
  ) |>
  mutate(
    annual_total_bill = annual_elec_bill + annual_gas_bill
  ) |>
  filter(version_elec == "hpr_doer")

# Check input table sizes
cat("Baseline pre rows:", nrow(annual_bills_baseline_2024_pre), "\n")
cat("Baseline post rows:", nrow(annual_bills_upgrade4_2024_post), "\n")
cat("HPR approved rows:", nrow(annual_bills_upgrade4_2024_hpr_approved), "\n")
  cat("HPR DOER rows:", nrow(annual_bills_upgrade4_2024_hpr_doer), "\n")

annual_bills_2022_all_scenarios <- bind_rows(
  annual_bills_baseline_2024_pre |> mutate(scenario = "baseline_pre"),
  annual_bills_upgrade4_2024_post |> mutate(scenario = "baseline_post"),
  annual_bills_upgrade4_2024_hpr_approved |> mutate(scenario = "hpr_approved"),
  annual_bills_upgrade4_2024_hpr_doer |> mutate(scenario = "hpr_doer")
)

cat("\nCombined table rows:", nrow(annual_bills_2022_all_scenarios), "\n")


# -----------------------
# Print medians of annual total bills for each scenario
cat("Pre-Electrification median annual bill 2024: $",
    format(round(median(annual_bills_baseline_2024_pre$annual_total_bill), 2),
           big.mark=","), "\n")

cat("Post-Electrification median annual bill 2024: $",
    format(round(median(annual_bills_upgrade4_2024_post$annual_total_bill), 2),
           big.mark=","), "\n")

cat("HPR Approved median annual bill 2024: $",
    format(round(median(annual_bills_upgrade4_2024_hpr_approved$annual_total_bill), 2),
           big.mark=","), "\n")

cat("HPR DOER median annual bill 2024: $",
    format(round(median(annual_bills_upgrade4_2024_hpr_doer$annual_total_bill), 2),
           big.mark=","), "\n")

# -----------------------
# Savings
median_baseline_pre_2024 <- median(annual_bills_baseline_2024_pre$annual_total_bill)
print(median_baseline_pre_2024)

annual_bills_2024 <- annual_bills_baseline_2024_pre |>
  select(bldg_id, annual_total_bill) |>
  rename(baseline_pre_bill = annual_total_bill) |>
  left_join(
    annual_bills_upgrade4_2024_post |>
      select(bldg_id, annual_total_bill) |>
      rename(baseline_post_bill = annual_total_bill),
    by = "bldg_id"
  ) |>
  left_join(
    annual_bills_upgrade4_2024_hpr_approved |>
      select(bldg_id, annual_total_bill) |>
      rename(hpr_approved_bill = annual_total_bill),
    by = "bldg_id"
  ) |>
  left_join(
    annual_bills_upgrade4_2024_hpr_doer |>
      select(bldg_id, annual_total_bill) |>
      rename(hpr_doer_bill = annual_total_bill),
    by = "bldg_id"
  ) |>
  mutate(
    post_savings = baseline_post_bill - baseline_pre_bill,
    post_savings_pct = post_savings / baseline_pre_bill,
    approved_savings = hpr_approved_bill - baseline_pre_bill,
    approved_savings_pct = approved_savings / baseline_pre_bill,
    doer_savings = hpr_doer_bill - baseline_pre_bill,
    doer_savings_pct = doer_savings / baseline_pre_bill
  ) |>
  mutate(
    supply_year = 2024
  ) |>
  filter(post_savings >= 0) |>
  filter(approved_savings >= 0) |>
  filter(doer_savings >= 0)


cat("Median annual bill impact (post-electrification): $",
    format(round(median(annual_bills_2024$post_savings), 2),
           big.mark=","), "\n")

cat("Median annual bill impact (HPR Approved): $",
    format(round(median(annual_bills_2024$approved_savings), 2),
           big.mark=","), "\n")

cat("Median annual bill impact (HPR DOER): $",
    format(round(median(annual_bills_2024$doer_savings), 2),
           big.mark=","), "\n")

```


# Plots (for sanity checking)
## 4 bars x 2 years
```{r}
annual_bills_all <- bind_rows(
  annual_bills_2018,
  annual_bills_2024
) |>
  mutate(
    supply_year = as.factor(supply_year)
  )

annual_bills_all |>
  pivot_longer(
    cols = c(baseline_pre_bill, baseline_post_bill, hpr_approved_bill, hpr_doer_bill),
    names_to = "rate_type",
    values_to = "annual_bill"
  ) |>
  mutate(
    rate_type = factor(rate_type,
                      levels = c("baseline_pre_bill", "baseline_post_bill",
                               "hpr_approved_bill", "hpr_doer_bill"),
                      labels = c("Pre HP Install\nBaseline Rates",
                               "Post HP Install\nBaseline Rates",
                               "Post HP Install\nApproved HP Rates",
                               "Post HP Install\nDOER HP Rates"))
  ) |>
  group_by(rate_type, supply_year) |>
  summarize(median_bill = median(annual_bill)) |>
  ggplot(aes(x = rate_type, y = median_bill, fill = supply_year)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    title = "Median Annual Bills by Rate Type and Supply Year",
    x = NULL,
    y = "Median Annual Bill",
    fill = "Supply Year"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

ggsave(
  "src/ma_hp_rates/img/annual_bills_4bars.png",
  width = 5,
  height = 4,
  dpi = 300,
  bg = "white"
)

```






## 4 Histograms: baseline, hpr_approved, hpr_doer
```{r}
library(patchwork)  # Add this at the top with other library imports

median_hpr_doer_2018 <- median(annual_bills_2018_hpr_doer$annual_total_bill)
median_hpr_approved_2018 <- median(annual_bills_2018_hpr_approved$annual_total_bill)
median_baseline_pre_2018 <- median(annual_bills_2018_baseline_pre$annual_total_bill)
median_baseline_post_2018 <- median(annual_bills_2018_baseline_post$annual_total_bill)


# Get the overall x-axis limits from all data
min_bill <- 0
max_bill <- 7500
median_bump <- 200
y_max = 125
y_limits <- c(0, y_max)

y_median <- y_max*0.90
binwidth <- 250

p1 <- annual_bills_2018_hpr_doer |>
  ggplot(aes(x = annual_total_bill)) +
  geom_histogram(fill = "#68BED8", alpha = 0.8, binwidth = binwidth) +
  geom_vline(aes(xintercept = median(annual_total_bill)), color = "black", linetype = "solid") +
  annotate("text",
           x = median(annual_bills_2018_hpr_doer$annual_total_bill) + median_bump, # offset text to the right
           y = y_median,
           label = paste0("Median Annual Bill: $", round(median(annual_bills_2018_hpr_doer$annual_total_bill), 0)),
           hjust = 0) +
  labs(title = paste0("Post HP Install, DOER HP Rates"), x = NULL, y = "Homes Count") +
  scale_y_continuous(labels = scales::comma, limits = y_limits) +
  scale_x_continuous(
    labels = scales::dollar_format(),
    limits = c(min_bill, max_bill)
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid.major = element_line(linewidth = 0.2),
    panel.grid.minor = element_blank()
  )

p2 <- annual_bills_2018_hpr_approved |>
  ggplot(aes(x = annual_total_bill)) +
  geom_histogram(fill = "#FC9706", alpha = 0.8, binwidth = binwidth) +
  geom_vline(aes(xintercept = median(annual_total_bill)), color = "black", linetype = "solid") +
  annotate("text",
           x = median(annual_bills_2018_hpr_approved$annual_total_bill) + median_bump,
           y = y_median,
           label = paste0("Median Annual Bill: $", round(median(annual_bills_2018_hpr_approved$annual_total_bill), 0)),
           hjust = 0) +
  labs(title = paste0("Post HP Install, Approved HP Rates"), x = NULL, y = "Homes Count") +
  scale_y_continuous(labels = scales::comma, limits = y_limits) +
  scale_x_continuous(
    labels = scales::dollar_format(),
    limits = c(min_bill, max_bill)
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid.major = element_line(linewidth = 0.2),
    panel.grid.minor = element_blank()
  )

p3 <- annual_bills_2018_baseline_post |>
  ggplot(aes(x = annual_total_bill)) +
  geom_histogram(fill = "#023047", alpha = 0.8, binwidth = binwidth) +
  geom_vline(aes(xintercept = median(annual_total_bill)), color = "black", linetype = "solid") +
  annotate("text",
           x = median(annual_bills_2018_baseline_post$annual_total_bill) + median_bump,
           y = y_median,
           label = paste0("Median Annual Bill: $", round(median(annual_bills_2018_baseline_post$annual_total_bill), 0)),
           hjust = 0) +
  labs(title = paste0("Post HP Install, Default Rates"), x = NULL, y = "Homes Count" ) +
  scale_y_continuous(labels = scales::comma, limits = y_limits) +
  scale_x_continuous(
    labels = scales::dollar_format(),
    limits = c(min_bill, max_bill)
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid.major = element_line(linewidth = 0.2),
    panel.grid.minor = element_blank()
  )

p4 <- annual_bills_2018_baseline_pre |>
  ggplot(aes(x = annual_total_bill)) +
  geom_histogram(fill = "grey80", alpha = 0.8, binwidth = binwidth) +
  geom_vline(aes(xintercept = median(annual_total_bill)), color = "black", linetype = "solid") +
  annotate("text",
           x = median(annual_bills_2018_baseline_pre$annual_total_bill) + median_bump,
           y = y_median,
           label = paste0("Median Annual Bill: $", round(median(annual_bills_2018_baseline_pre$annual_total_bill), 0)),
           hjust = 0) +
  labs(title = paste0("Pre HP Install, Default Rates"), x = "Annual Bill ($)", y = "Homes Count" ) +
  scale_y_continuous(labels = scales::comma, limits = y_limits) +
  scale_x_continuous(
    labels = scales::dollar_format(),
    limits = c(min_bill, max_bill)
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid.major = element_line(linewidth = 0.2),
    panel.grid.minor = element_blank()
  )

# Combine plots vertically using patchwork
p1 / p2 / p3 / p4 +
  plot_layout(heights = c(1, 1, 1, 1)) +
  plot_annotation(
    title = "Heat Pump (HP) Rate Impacts on Total Household Bills",
    subtitle = glue::glue("Electric Utility: {electric_utility} | Gas Utility: {gas_utility} | 2024 Gas Supply Rates"),
    theme = theme(
      plot.margin = margin(3, 3, 3, 3),
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 12)
    )
  )

# Display the plot
# print(p1 / p2 / p3 +
#   plot_layout(heights = c(1, 1, 1)) +
#   plot_annotation(
#     theme = theme(plot.margin = margin(10, 10, 10, 10))
#   ))


ggsave(
  "src/ma_hp_rates/img/annual_bills_4hists.png",
  width = 5,
  height = 4,
  dpi = 300,
  bg = "white"
)


```


## Plot: Baseline_post vs Baseline_pre, HPR Approved vs Baseline_pre, HPR DOER vs Baseline_pre
```{r}
pct_below_zero <- function(x) {
  mean(x < 0) * 100
}
data.frame(
  rate_type = c("DOER", "Approved", "Post"),
  pct_negative = c(
    pct_below_zero(annual_bills_2018$doer_savings),
    pct_below_zero(annual_bills_2018$approved_savings),
    pct_below_zero(annual_bills_2018$post_savings)
  )
) |>
  print()




x_axis_limits <- c(-3000, 3000)
y_max = 175
y_axis_limits <- c(0, y_max)
median_bump <- 200

# Create the two histograms
p1 <- ggplot(annual_bills_2024, aes(x = doer_savings)) +
  geom_histogram(fill = "#68BED8", alpha = 0.8, binwidth = 100) +
  #geom_vline(aes(xintercept = median(doer_savings)), color = "black", linetype = "solid") +
  # annotate("text",
  #          x = median(annual_bills_2018$doer_savings) + median_bump,
  #          y = Inf,
  #          label = paste0("Median Annual Bill Impact: $", round(median(annual_bills_2018$doer_savings), 0)),
  #          hjust = 0,
  #          vjust = 2) +
  labs(title = paste0("Post HP Install, DOER Rates"), x = NULL, y= "Homes Count") +
  scale_x_continuous(labels = scales::dollar_format(), limits = x_axis_limits) +
  scale_y_continuous(labels = scales::comma, limits = y_axis_limits) +
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid.major = element_line(linewidth = 0.2),
    panel.grid.minor = element_blank()
  )

p2 <- ggplot(annual_bills_2024, aes(x = approved_savings)) +
  geom_histogram(fill = "#FC9706", alpha = 0.8, binwidth = 100) +
  #geom_vline(aes(xintercept = median(approved_savings)), color = "black", linetype = "solid") +
  # annotate("text",
  #          x = median(annual_bills_2018$approved_savings) + median_bump,
  #          y = Inf,
  #          label = paste0("Median Annual Bill Impact: $", round(median(annual_bills_2018$approved_savings), 0)),
  #          hjust = 0,
  #          vjust = 2) +
  labs(title = paste0("Post HP Install, Approved Rates"), x = NULL, y= "Homes Count") +
  scale_x_continuous(labels = scales::dollar_format(), limits = x_axis_limits) +
  scale_y_continuous(labels = scales::comma, limits = y_axis_limits) +
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid.major = element_line(linewidth = 0.2),
    panel.grid.minor = element_blank()
  )

p3 <- ggplot(annual_bills_2024, aes(x = post_savings)) +
  geom_histogram(fill = "#023047", alpha = 0.8, binwidth = 100) +
  #geom_vline(aes(xintercept = median(post_savings)), color = "black", linetype = "solid") +
  # annotate("text",
  #          x = median(annual_bills_2018$post_savings) + median_bump,
  #          y = Inf,
  #          label = paste0("Median Annual Bill Impact: $", round(median(annual_bills_2018$post_savings), 0)),
  #          hjust = 0,
  #          vjust = 2) +
  labs(title = paste0("Post HP Install, Default Rates"), x = "Impact on Annual Bill ($)", y= "Homes Count") +
  scale_x_continuous(labels = scales::dollar_format(), limits = x_axis_limits) +
  scale_y_continuous(labels = scales::comma, limits = y_axis_limits) +
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid.major = element_line(linewidth = 0.2),
    panel.grid.minor = element_blank()
  )

# Combine plots vertically using patchwork
p1 / p2 / p3 +
  plot_layout(heights = c(1, 1, 1)) +
  plot_annotation(
    title = "Heat Pump Rate Impacts on Total Household Bills",
    subtitle = glue::glue("Electric Utility: {electric_utility} | Gas Utility: {gas_utility} | 2024 Supply Rates"),
    theme = theme(
      plot.margin = margin(3, 3, 3, 3),
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 12)
    )
  )

# Save the plot
ggsave(
  "src/ma_hp_rates/img/annual_savings_hists.png",
  width = 4.5,
  height = 4,
  dpi = 300,
  bg = "white"
)
```
