---
title: analysis - MA HP Rates

toc: true
reference-location: margin
fig-cap-location: bottom

appendix-style: default
citation-location: document
citation:
  container-title: Switchbox

format:
  html:
    page-layout: full
---
# Setup
```{r}
#| label: setup
library(lubridate)
library(patchwork)
library(tidyverse)
library(ggrepel)

# Switchbox Libraries
source("/workspaces/reports/lib/ggplot/switchbox_theme.R")
source("/workspaces/reports/lib/rates_analysis/heat_pump_rate_funcs.R")

# Global parameters
exclude_municipal <- TRUE
supply_year = 2024
path_rs_db <- "/workspaces/reports/data/ResStock/2022_resstock_amy2018_release_1.1/rs_ma_hp_rates_2.db"

hp_eff_for_report <- "mid_hp"
hvac_for_utility_bills <- "mid_hp"
round_dollars <- 0
round_percent <- 1

bldg_id_for_debugging = 2913

# Run calcs.qmd to generate the data for this notebook
# This will be stored in /cache, but removed
load("/workspaces/reports/src/ma_hp_rates/cache/output_for_analysis.RData")





```

## Color Definitions
```{r}
scenario_colors = c(
  "baseline_rates_current_hvac" = "grey50",

  "baseline" = "#023047",
  "baseline_rates_low_hp" = "#023047",
  "baseline_rates_mid_hp" = "#023047",
  "baseline_rates_high_hp" = "#023047",

  "hpr_approved" = "#FC9706",
  "approved_low_hp" = "#FC9706",
  "approved_mid_hp" = "#FC9706",
  "approved_high_hp" = "#FC9706",

  "hpr_doer" = "#68BED8",
  "doer_low_hp" = "#68BED8",
  "doer_mid_hp" = "#68BED8",
  "doer_high_hp" = "#68BED8"
)

library(RColorBrewer)


heating_type_colors = c(
  "Natural Gas" = "#8C510A",        # brown
  "Fuel Oil" = "#D8B365",           # light brown
  "Propane" = "#F6E8C3",            # very light brown
  "Resistance" = "#5AB4AC",         # BrBG palette
  "Heat Pump" = "#80CDC1",          # BrBG palette
  "Other/None" = "#C7EAE5"          # BrBG palette
)

housing_type_colors = c(
  "Single-Family" = "#1B9E77",     # dark2[1] - teal
  "2-4 Units" = "#D95F02",         # dark2[2] - orange
  "5+ Units" = "#7570B3"           # dark2[3] - purple
)

```


## plot function prototypes
### hist_for_single_rate_version()
```{r}
#| label: hist_for_single_rate_version() prototype
hist_for_single_rate_version <- function(
  annual_change_table,
  version_elec,
  x_limits = c(-3000, 3000),
  y_limits = c(0, 1000),
  binwidth = 100,
  show_category_labels = TRUE
) {

  # --------------------------------------
  # Plot parameters
  x_min <- x_limits[1]
  x_max <- x_limits[2]

  # Define the breaks for binning
  breaks <- seq(floor(x_min/binwidth)*binwidth, ceiling(x_max/binwidth)*binwidth, by = binwidth)

  # More balanced color palette with less contrast
  dark_red <- "#db8b87"        # dark red
  light_red <- "#eaada9"       # light red
  light_green <- "#8ebd85"     # light green
  dark_green <- "#5e8a5e"      # dark green
  # --------------------------------------

  # Filter data for mid_hp
  plot_data <- annual_change_table |>
    filter(hvac == "mid_hp", version_elec == !!version_elec) |>
    # Add category column before binning with new cutoffs
    mutate(bill_category = case_when(
      annual_bill_change <= -1000 ~ "large_savings",
      annual_bill_change < 0 ~ "small_savings",
      annual_bill_change < 1000 ~ "small_increase",
      TRUE ~ "large_increase"
    ))

  # Calculate percentages for each category
  category_percentages <- plot_data |>
    group_by(bill_category) |>
    summarise(count = n(), .groups = "drop") |>
    mutate(percentage = count / sum(count) * 100)

  # Get percentages for annotation
  pct_large_savings <- category_percentages |>
    filter(bill_category == "large_savings") |>
    pull(percentage) |>
    round(1)
  if(length(pct_large_savings) == 0) pct_large_savings <- 0

  pct_small_savings <- category_percentages |>
    filter(bill_category == "small_savings") |>
    pull(percentage) |>
    round(1)
  if(length(pct_small_savings) == 0) pct_small_savings <- 0

  pct_small_increase <- category_percentages |>
    filter(bill_category == "small_increase") |>
    pull(percentage) |>
    round(1)
  if(length(pct_small_increase) == 0) pct_small_increase <- 0

  pct_large_increase <- category_percentages |>
    filter(bill_category == "large_increase") |>
    pull(percentage) |>
    round(1)
  if(length(pct_large_increase) == 0) pct_large_increase <- 0

  # Manually bin the data and calculate counts - this preserves the bill_category
  binned_data <- plot_data |>
    mutate(bin = cut(annual_bill_change,
                     breaks = breaks,
                     include.lowest = TRUE,
                     right = FALSE)) |>
    group_by(bin, bill_category) |>
    summarise(count = n(), .groups = "drop") |>
    # Add bin midpoint for plotting
    mutate(bin_mid = as.numeric(as.character(
      sapply(bin, function(b) {
        # Extract bin boundaries and calculate midpoint
        vals <- as.numeric(gsub("\\(|\\]|\\[|\\)", "", strsplit(as.character(b), ",")[[1]]))
        return(mean(vals))
      })
    )))

  # Position the bars at the very top and text just below
  bar_y_pos <- y_limits[2]  # Exactly at the top
  text_y_pos <- y_limits[2] * 0.95  # Just below the colored line

  # Calculate x positions for percentage labels (center of each category)
  x_large_savings <- mean(c(x_min, -1000))
  x_small_savings <- mean(c(-1000, 0))
  x_small_increase <- mean(c(0, 1000))
  x_large_increase <- mean(c(1000, x_max))

  # Create the plot using geom_col() with binned data
  px_hist_by_rate_version <- ggplot(binned_data, aes(x = bin_mid, y = count, fill = bill_category)) +
    # Add vertical dotted lines at the category boundaries
    geom_vline(xintercept = c(-1000, 0, 1000), linetype = "dotted", color = "gray50", size = 0.5) +

    # Add the bars
    geom_col(position = "stack", width = binwidth * 0.9) +

    # Set the fill colors manually with expanded categories
    scale_fill_manual(
      values = c(
        "large_savings" = "#5e8a5e",      # dark green
        "small_savings" = "#8ebd85",      # light green
        "small_increase" = "#eaada9",     # light red
        "large_increase" = "#db8b87"      # dark red
      ),
      guide = "none"
    ) +

    # Add horizontal lines at the very top of the plot
    # With colors matching their respective categories
    annotate("segment",
             x = x_min, xend = -1000,
             y = bar_y_pos, yend = bar_y_pos,
             color = dark_green, size = 1) +

    annotate("segment",
             x = -1000, xend = 0,
             y = bar_y_pos, yend = bar_y_pos,
             color = light_green, size = 1) +

    annotate("segment",
             x = 0, xend = 1000,
             y = bar_y_pos, yend = bar_y_pos,
             color = light_red, size = 1) +

    annotate("segment",
             x = 1000, xend = x_max,
             y = bar_y_pos, yend = bar_y_pos,
             color = dark_red, size = 1) +

    # Add percentage labels for each category directly below the lines
    annotate("text",
             x = x_large_savings,
             y = text_y_pos * 0.95,
             label = paste0(pct_large_savings, "%"),
             size = 3,
             fontface = "bold") +

    annotate("text",
             x = x_small_savings,
             y = text_y_pos * 0.95,
             label = paste0(pct_small_savings, "%"),
             size = 3,
             fontface = "bold") +

    annotate("text",
             x = x_small_increase,
             y = text_y_pos * 0.95,
             label = paste0(pct_small_increase, "%"),
             size = 3,
             fontface = "bold") +

    annotate("text",
             x = x_large_increase,
             y = text_y_pos * 0.95,
             label = paste0(pct_large_increase, "%"),
             size = 3,
             fontface = "bold") +

    # Rest of the plot styling
    labs(
      title = glue::glue("Change in Total Winter Energy Bills after Switching to Heat Pumps"),
      x = "Annual Bill Change ($)",
      y = "# of Homes"
    ) +
    scale_y_continuous(
      labels = function(x) paste0(round(x * 242.13 * 0.001), "k"),
      name = "# of Homes",
      limits = y_limits,
      breaks = seq(0, y_limits[2]*242.13, by=50000)/242.13
    ) +
    scale_x_continuous(
      labels = scales::dollar_format(),
      limits = x_limits,
      breaks = seq(x_min, x_max, by = 500)
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 10, face = "bold"),
      panel.grid.major = element_line(linewidth = 0.2),
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "none",
      aspect.ratio = 0.6
    ) +
    {
      if (show_category_labels) {
        list(
          annotate("text", x = x_large_savings,   y = y_limits[2]*0.97,
                   label = "savings > $1k",  size = 3, fontface = "bold"),
          annotate("text", x = x_small_savings,   y = y_limits[2]*0.97,
                   label = "savings $0–1k", size = 3, fontface = "bold"),  # make bold
          annotate("text", x = x_small_increase,  y = y_limits[2]*0.97,
                   label = "losses  $0–1k", size = 3, fontface = "bold"),  # make bold
          annotate("text", x = x_large_increase,  y = y_limits[2]*0.97,
                   label = "losses  > $1k", size = 3, fontface = "bold")
        )
      }
    } +
    coord_cartesian(clip = "off")

  return(list(px_hist_by_rate_version, category_percentages))
}

```

### plot_bill_change_histograms()
```{r}
#| label: plot_bill_change_histograms() prototype

plot_bill_change_histograms <- function(
  annual_change_table,
  bill_type,
  baseline_heating_type_option,
  electric_utility_option,
  supply_year,
  second_subtitle = NULL,
  x_limits = c(-2500, 2500),
  y_limits = c(0, 100000)/242.13,
  binwidth = 100
) {


  # Create nicer version labels
  version_labels <- c(
    "baseline" = "Default Rates",
    "hpr_approved" = "1.0 Rates",
    "hpr_doer" = "2.0 Rates"
  )

  # Create three histograms without individual titles
  result_baseline <- hist_for_single_rate_version(
    annual_change_table = annual_change_table,
    version_elec = "baseline",
    x_limits = x_limits,
    y_limits = y_limits,
    binwidth = binwidth,
    show_category_labels = FALSE
  )

  p1_baseline <- result_baseline[[1]] +
    labs(title = NULL) +  # Remove title
    ylab(NULL) +                 # Remove y-axis label
    theme(
      axis.title.x = element_blank(),  # Remove x axis title from top two plots
      axis.text.x = element_blank(),   # Remove x tick labels from top two plots
      axis.ticks.x = element_blank(),  # Remove x ticks from top two plots
      plot.margin = margin(5, 5, 0, 5),  # Consistent margins (top, right, bottom, left)
      axis.title.y = element_blank()     # Remove y title from top plot
    ) +
    # Add elegant version label on right side
    annotate("text",
             x = x_limits[2] * 0.95,
             y = y_limits[2] * 0.7,
             label = version_labels["baseline"],
             hjust = 1,
             fontface = "bold",
             size = 4,
             color = "#023047")

  result_approved <- hist_for_single_rate_version(
    annual_change_table = annual_change_table,
    version_elec = "hpr_approved",
    x_limits = x_limits,
    y_limits = y_limits,
    binwidth = binwidth,
    show_category_labels = FALSE
  )

  p2_approved <- result_approved[[1]] +
    labs(title = NULL) +  # Remove title
    ylab(NULL) +
    theme(
      axis.title.x = element_blank(),  # Remove x axis title from top two plots
      axis.text.x = element_blank(),   # Remove x tick labels from top two plots
      axis.ticks.x = element_blank(),  # Remove x ticks from top two plots
      axis.title.y = element_blank(),   # Remove y title from middle plot
      plot.margin = margin(0, 5, 0, 5)  # Consistent margins (no top/bottom margin)
    ) +
    # Add elegant version label on right side
    annotate("text",
             x = x_limits[2] * 0.95,
             y = y_limits[2] * 0.7,
             label = version_labels["hpr_approved"],
             hjust = 1,
             fontface = "bold",
             size = 4,
             color = "#FC9706")

  result_doer <- hist_for_single_rate_version(
    annual_change_table = annual_change_table,
    version_elec = "hpr_doer",
    x_limits = x_limits,
    y_limits = y_limits,
    binwidth = binwidth,
    show_category_labels = FALSE
  )

  p3_doer <- result_doer[[1]] +
    labs(title = NULL) +  # Remove title
    ylab(NULL) +
    theme(
      axis.title.y = element_blank(),  # Remove y title from bottom plot
      plot.margin = margin(0, 5, 5, 5),  # Consistent margins
      axis.title.x = element_text(size = 9),
    ) +
    # Add elegant version label on right side
    annotate("text",
             x = x_limits[2] * 0.95,
             y = y_limits[2] * 0.7,
             label = version_labels["hpr_doer"],
             hjust = 1,
             fontface = "bold",
             size = 4,
             color = "#68BED8")

  # print(result_baseline[[1]])
  # print(result_approved[[1]])
  # print(result_doer[[1]])

  # Combine plots with shared x and y axes
  combined_plot <- p1_baseline / p2_approved / p3_doer +
    plot_layout(heights = c(1, 1, 1)) +  # Equal heights for all plots
    plot_annotation(
      title = case_when(
        bill_type == "annual_bill_change_elec" ~ glue::glue("Switching from {baseline_heating_type_option} to Heat Pump: Change in Annual Electric Bill"),
        bill_type == "annual_bill_change_energy" ~ glue::glue("Change in Total Winter Energy Bills after Switching to Heat Pump from {baseline_heating_type_option}"),
        bill_type == "annual_bill_change_gas" ~ glue::glue("Switching from {baseline_heating_type_option} to Heat Pump: Change in Annual Gas Bill"),
        TRUE ~ "Heat Pump Rate Bill Changes" # Default case
      ),
      subtitle = second_subtitle,
      theme = theme(
        plot.margin = margin(10, 25, 10, 10),  # Add more right margin for labels
        plot.title = element_text(hjust = 0.5, size = 10, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 8),
      )
    ) &
    # Apply consistent theming to all plots
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_line(linewidth = 0.2, color = "gray90"),
      panel.grid.major.y = element_line(linewidth = 0.2, color = "gray90"),
      aspect.ratio = 0.18,
    )

  # Wrap the combined plot to add a shared y-axis label
  combined_plot <- patchwork::wrap_elements(combined_plot) +
    labs(tag = "# of Homes") +
    theme(
      plot.tag = element_text(angle = 90, vjust = 0.5, hjust = 0.5, face = "plain", size = 9),
      plot.tag.position = "left"
    )

  return(list(combined_plot, result_baseline[[2]], result_approved[[2]], result_doer[[2]]))
}


```

## table-delivery-tariffs-elec
```{r}
#| label: tbl-delivery-tariffs-elec
round_tariffs <- 4

# Eversource
delivery_rate_eversource_summary_baseline <- delivery_tariffs_elec |>
  filter(electric_utility == "eversource", version == "baseline", type == "delivery_rate") |>
  pull(value) |>
  unique() |>
  round(round_tariffs)

delivery_rate_eversource_summary_approved <- delivery_tariffs_elec |>
  filter(electric_utility == "eversource", version == "hpr_approved", type == "delivery_rate") |>
  pull(value) |>
  min() |>
  round(round_tariffs)

delivery_rate_eversource_summary_doer <- delivery_tariffs_elec |>
  filter(electric_utility == "eversource", version == "hpr_doer", type == "delivery_rate") |>
  pull(value) |>
  min() |>
  round(round_tariffs)

# National Grid
delivery_rate_nationalgrid_summary_baseline <- delivery_tariffs_elec |>
  filter(electric_utility == "nationalgrid", version == "baseline", type == "delivery_rate") |>
  pull(value) |>
  unique() |>
  round(round_tariffs)

delivery_rate_nationalgrid_summary_approved <- delivery_tariffs_elec |>
  filter(electric_utility == "nationalgrid", version == "hpr_approved", type == "delivery_rate") |>
  pull(value) |>
  min() |>
  round(round_tariffs)

delivery_rate_nationalgrid_summary_doer <- delivery_tariffs_elec |>
  filter(electric_utility == "nationalgrid", version == "hpr_doer", type == "delivery_rate") |>
  pull(value) |>
  min() |>
  round(round_tariffs)

# Eversource
delivery_rate_unitil_summary_baseline <- delivery_tariffs_elec |>
  filter(electric_utility == "unitil", version == "baseline", type == "delivery_rate") |>
  pull(value) |>
  unique() |>
  round(round_tariffs)

delivery_rate_unitil_summary_approved <- delivery_tariffs_elec |>
  filter(electric_utility == "unitil", version == "hpr_approved", type == "delivery_rate") |>
  pull(value) |>
  min() |>
  round(round_tariffs)

delivery_rate_unitil_summary_doer <- delivery_tariffs_elec |>
  filter(electric_utility == "unitil", version == "hpr_doer", type == "delivery_rate") |>
  pull(value) |>
  min() |>
  round(round_tariffs)


# Municipal
delivery_rate_municipal_summary_baseline <- delivery_tariffs_elec |>
  filter(electric_utility == "municipal", version == "baseline", type == "delivery_rate") |>
  pull(value) |>
  unique() |>
  round(round_tariffs)

delivery_rate_municipal_summary_approved <- delivery_tariffs_elec |>
  filter(electric_utility == "municipal", version == "hpr_approved", type == "delivery_rate") |>
  pull(value) |>
  min() |>
  round(round_tariffs)

delivery_rate_municipal_summary_doer <- delivery_tariffs_elec |>
  filter(electric_utility == "municipal", version == "hpr_doer", type == "delivery_rate") |>
  pull(value) |>
  min() |>
  round(round_tariffs)


# Create summary table of delivery rates
delivery_rates_summary <- tibble(
  Utility = c(
    "Eversource", "Eversource", "Eversource",
    "National Grid", "National Grid", "National Grid",
    "Unitil", "Unitil", "Unitil",
    "Municipal", "Municipal", "Municipal"
  ),
  Version = rep(c("Baseline", "HPR Approved", "HPR DOER"), 4),
  `Delivery Rate` = c(
    delivery_rate_eversource_summary_baseline,
    delivery_rate_eversource_summary_approved,
    delivery_rate_eversource_summary_doer,
    delivery_rate_nationalgrid_summary_baseline,
    delivery_rate_nationalgrid_summary_approved,
    delivery_rate_nationalgrid_summary_doer,
    delivery_rate_unitil_summary_baseline,
    delivery_rate_unitil_summary_approved,
    delivery_rate_unitil_summary_doer,
    delivery_rate_municipal_summary_baseline,
    delivery_rate_municipal_summary_approved,
    delivery_rate_municipal_summary_doer
  )
)

# Display the table
knitr::kable(delivery_rates_summary,
             caption = "Summary of Delivery Rates by Utility and Version",
             format = "pipe",
             digits = 4)
```


# ------------------------------------------------------------------------------
# Background
# ------------------------------------------------------------------------------

## fig-stacked-bar-components-delivery-on-top
```{r}
#| label: fig-stacked-bar-components-delivery-on-top
#| fig-cap: "Median electric bill for Massachusetts homes that install air-source heat pumps, under default, 1.0 and 2.0 heat pump rates."

bill_component_colors = c(
  "annual_gas_bill" = "grey50", # grey
  "annual_elec_supply_charge" = "#A0AF12",
  "annual_elec_delivery_charge" = "#023047",
  "annual_elec_customer_charge" = "#68BED8",
  "pre_hp_delivery_total" = "#FC9706"
)


stack_order_delivery_on_top = c(
  "annual_gas_bill",
  "annual_elec_delivery_charge", # top
  "annual_elec_customer_charge", # middle
  "annual_elec_supply_charge" # bottom
)

components <- factor(stack_order_delivery_on_top, levels = stack_order_delivery_on_top)

annual_bills_by_component_plot <- annual_bills_by_component |>
  arrange(match(charge_type, components)) |>
  filter(hvac == "mid_hp")


# Ugly, but here we get the balues from the stacked bar chart
y_baseline_customer_charge <- annual_bills_by_component_plot |> filter(version == "baseline", hvac == "mid_hp", charge_type == "annual_elec_customer_charge") |> pull(amount)
y_baseline_delivery_charge <- annual_bills_by_component_plot |> filter(version == "baseline", hvac == "mid_hp", charge_type == "annual_elec_delivery_charge") |> pull(amount)
y_baseline_supply_charge <- annual_bills_by_component_plot |> filter(version == "baseline", hvac == "mid_hp", charge_type == "annual_elec_supply_charge") |> pull(amount)

y_approved_customer_charge <- annual_bills_by_component_plot |> filter(version == "hpr_approved", hvac == "mid_hp", charge_type == "annual_elec_customer_charge") |> pull(amount)
y_approved_delivery_charge <- annual_bills_by_component_plot |> filter(version == "hpr_approved", hvac == "mid_hp", charge_type == "annual_elec_delivery_charge") |> pull(amount)
y_approved_supply_charge <- annual_bills_by_component_plot |> filter(version == "hpr_approved", hvac == "mid_hp", charge_type == "annual_elec_supply_charge") |> pull(amount)

y_doer_customer_charge <- annual_bills_by_component_plot |> filter(version == "hpr_doer", hvac == "mid_hp", charge_type == "annual_elec_customer_charge") |> pull(amount)
y_doer_delivery_charge <- annual_bills_by_component_plot |> filter(version == "hpr_doer", hvac == "mid_hp", charge_type == "annual_elec_delivery_charge") |> pull(amount)
y_doer_supply_charge <- annual_bills_by_component_plot |> filter(version == "hpr_doer", hvac == "mid_hp", charge_type == "annual_elec_supply_charge") |> pull(amount)
y_doer_natural_gas <- annual_bills_by_component_plot |> filter(version == "hpr_doer", hvac == "mid_hp", charge_type == "annual_gas_bill") |> pull(amount)


# Savings and percentage savings
fig_stacked_bar_components_delivery_on_top_savings_approved <- y_approved_delivery_charge - y_baseline_delivery_charge
fig_stacked_bar_components_delivery_on_top_savings_approved_pct <- fig_stacked_bar_components_delivery_on_top_savings_approved / y_baseline_delivery_charge

fig_stacked_bar_components_delivery_on_top_savings_doer <- y_doer_delivery_charge - y_baseline_delivery_charge
fig_stacked_bar_components_delivery_on_top_savings_doer_pct <- fig_stacked_bar_components_delivery_on_top_savings_doer / y_baseline_delivery_charge

y_baseline_bill <- y_baseline_customer_charge + y_baseline_delivery_charge + y_baseline_supply_charge

fig_stacked_bar_components_delivery_on_top_savings_approved_pct_of_total_bill <- (fig_stacked_bar_components_delivery_on_top_savings_approved / y_baseline_bill * -100) |> round(0)

fig_stacked_bar_components_delivery_on_top_savings_doer_pct_of_total_bill <- (fig_stacked_bar_components_delivery_on_top_savings_doer / y_baseline_bill * -100) |> round(0)


y_annotation_offset <- 0
x_label_offset <- 3.5


fig_bill_stacked_bar_delivery_on_top <- ggplot(annual_bills_by_component_plot |>
  mutate(charge_type = factor(charge_type, levels = components)),
       aes(x = interaction(hvac, version), y = amount, fill = charge_type)) +
  geom_bar(stat = "identity") +
  labs(title = "Components of Annual Electricity Bill",
       x = "",
       y = "Annual Cost ($)",
       fill = "Charge Type") +
  scale_fill_manual(values = bill_component_colors) +
  scale_y_continuous(labels = scales::dollar_format(), limits = c(0, 4500), breaks = seq(0, 5000, by = 1000), expand = c(0,0)) +
  scale_x_discrete(labels = c("Current HVAC\nCurrent Rates",
                           "Mid-Efficiency HP\nCurrent Rates",
                           "Mid-Efficiency HP\n1.0 Rates",
                           "Mid-Efficiency HP\n2.0 Rates")) +
  coord_cartesian(clip = "off") +
  theme(
    text = element_text(family = "IBM-Plex-Sans"),
    axis.title.x = element_text(margin = margin(t = 3)),
    axis.title.y = element_text(margin = margin(r = 3)),
    legend.position = "none",
    plot.margin = margin(t = 3, r = 200, b = 0, l = 10, unit = "pt")
  ) +

  # instead of a legend, write annotation for the 3 categories to the right of the right-most bar
  annotate(
    "text",
    x = x_label_offset,
    y = (y_doer_supply_charge/2) + y_annotation_offset,
    label = "Supply Charge",
    hjust = 0,
    family = "IBM-Plex-Sans",
    color = bill_component_colors["annual_elec_supply_charge"],
    size = 6,
    fontface = "bold"
  ) +
  annotate(
    "text",
    x = x_label_offset,
    y = (y_doer_customer_charge/1.6) + y_doer_supply_charge + y_annotation_offset,
    label = "Customer Charge",
    hjust = 0,
    family = "IBM-Plex-Sans",
    color = bill_component_colors["annual_elec_customer_charge"],
    size = 6,
    fontface = "bold"
  ) +
  annotate(
    "text",
    x = x_label_offset,
    y = (y_doer_delivery_charge/2) + y_doer_customer_charge + y_doer_supply_charge + y_annotation_offset,
    label = "Delivery Charge",
    hjust = 0,
    family = "IBM-Plex-Sans",
    color = bill_component_colors["annual_elec_delivery_charge"],
    size = 6,
    fontface = "bold"
  ) +

  # No legend
  theme(legend.position = "none")

fig_bill_stacked_bar_delivery_on_top



```


```{r}
#| label: fig-stacked-bar-components-delivery-on-top-pre
#| fig-cap: "Median electric bill for Massachusetts homes, with current HVAC vs. mid-efficiency heat pump, under default delvery rates and 2024 supply rates."

fig_bill_stacked_bar_delivery_on_top

```

# ------------------------------------------------------------------------------
# Findings
# ------------------------------------------------------------------------------

## data-all-fuel-types
```{r}
#| label: data-all-fuel-types
########################################################
# Data
########################################################

# Parameters
bill_type <- 'annual_bill_change_energy'
electric_utility_option <- 'statewide'
baseline_heating_type_option <- 'All Fuels'

annual_change_all_fuel_types <- annual_change |>
  filter(electric_utility != "municipal") |>
  filter(!is.na(annual_bill_change) & is.finite(annual_bill_change),
        bill_type == !!bill_type,
        if (electric_utility_option != "statewide") electric_utility == !!electric_utility_option else TRUE,
        if (baseline_heating_type_option != "All Fuels") baseline_heating_type == !!baseline_heating_type_option else TRUE,
        year == !!supply_year
        )

# Print these to console
scenario_stats_all_fuel_types_low_hp <- calc_stats_by_rate_version(annual_change_all_fuel_types |> filter(electric_utility != "municipal"), hp_eff_for_stats = "low_hp", print_table = FALSE)
scenario_stats_all_fuel_types_mid_hp <- calc_stats_by_rate_version(annual_change_all_fuel_types |> filter(electric_utility != "municipal"), hp_eff_for_stats = "mid_hp", print_table = FALSE)
scenario_stats_all_fuel_types_high_hp <- calc_stats_by_rate_version(annual_change_all_fuel_types |> filter(electric_utility != "municipal"), hp_eff_for_stats = "high_hp", print_table = FALSE)
#
# These are the values used in the report!!!
# Calculate statistics for each version and set to the values used in the report
hp_eff_for_report <- "mid_hp"
scenario_stats_all_fuel_types <- calc_stats_by_rate_version(annual_change_all_fuel_types |> filter(electric_utility != "municipal"),
                                                          hp_eff_for_stats = hp_eff_for_report,
                                                          print_table = FALSE)

# Save as variables to be exported then used in the report
versions <- list(
  list(version = "baseline", prefix = "default"),
  list(version = "hpr_approved", prefix = "approved"),
  list(version = "hpr_doer", prefix = "doer")
)

# Loop through versions to calculate statistics
for (v in versions) {
  # Get the stats for this version
  stats <- scenario_stats_all_fuel_types |>
    filter(version_elec == v$version)

  # Calculate and assign all statistics for this version
  # eg "baseline_median_bill_change", "approved_pct_that_save"...
  assign(paste0("median_bill_change_", v$prefix, "_all_fuels"),
         round(stats |> pull(median_change), round_dollars))
  assign(paste0("pct_that_save_", v$prefix, "_all_fuels"),
         round(stats |> pull(pct_that_save), round_percent))
  assign(paste0("pct_that_save_big_", v$prefix, "_all_fuels"),
         round(stats |> pull(pct_that_save_big), round_percent))
  assign(paste0("pct_that_lose_", v$prefix, "_all_fuels"),
         round(stats |> pull(pct_that_lose), round_percent))
  assign(paste0("pct_that_lose_big_", v$prefix, "_all_fuels"),
         round(stats |> pull(pct_that_lose_big), round_percent))
  assign(paste0("median_savings_", v$prefix, "_all_fuels"),
         round(stats |> pull(median_savings), round_dollars))
  assign(paste0("median_loss_", v$prefix, "_all_fuels"),
         round(stats |> pull(median_loss), round_dollars))
}
```

## ---------- How many homes would lower their bills by switching to heat pumps today? ----------
### fig-baseline-hist-all-fuel-types
```{r}
#| label: fig-baseline-hist-all-fuel-types
result_baseline_all_fuels <- hist_for_single_rate_version(
  annual_change_table = annual_change_all_fuel_types,
  version_elec = "baseline"
)

plot_all_fuels <- result_baseline_all_fuels[[1]]


fig_baseline_hist_all_fuel_types_pct_large_savings <- result_baseline_all_fuels[[2]] |>
  filter(bill_category == "large_savings") |>
  pull(percentage)

fig_baseline_hist_all_fuel_types_pct_small_savings <- result_baseline_all_fuels[[2]] |>
  filter(bill_category == "small_savings") |>
  pull(percentage)

fig_baseline_hist_all_fuel_types_pct_large_increase <- result_baseline_all_fuels[[2]] |>
  filter(bill_category == "large_increase") |>
  pull(percentage)

fig_baseline_hist_all_fuel_types_pct_that_save <- result_baseline_all_fuels[[2]] |>
  filter(bill_category == "small_savings" | bill_category == "large_savings") |>
  pull(percentage) |>
  sum()

print(plot_all_fuels)
```

### data-hists-all-fuel-types-default-and-1.0
```{r}
#| label: fig-hists-all-fuel-types-default-and-1.0
#| fig-cap: "Change in total winter energy bills after switching to heat pumps, for every home in MA served by an investor-owned electric utility, under default vs. 1.0 heat pump electric delivery rates. Assumes default gas delivery rates, 2024 supply rates for gas and electricity, and 100% participation in low-income discount programs by eligible households."


# Parameters that match the 3-panel call above
x_limits  <- c(-2500, 2500)
y_limits  <- c(0, 200000)/242.13
binwidth  <- 100

# Default-rates histogram (mid-efficiency HPs)
p_default <- hist_for_single_rate_version(
  annual_change_table = annual_change_all_fuel_types,
  version_elec        = "baseline",
  x_limits            = x_limits,
  y_limits            = y_limits,
  binwidth            = binwidth,
  show_category_labels = FALSE   # keep consistent with 3-panel
)[[1]]

# remove x-axis from top panel
p_default <- p_default +
  labs(title = NULL, y = NULL) +
  theme(axis.title.x = element_blank(),
        # axis.text.x  = element_blank(),
        # axis.ticks.x = element_blank(),
        axis.title.y = element_blank()) +
  annotate("text", x = x_limits[2]*0.95, y = y_limits[2]*0.7,
           label = "Default Rates", hjust = 1, fontface = "bold", size = 4, color = "#023047")

# 1.0-rates histogram
p_1rates <- hist_for_single_rate_version(
  annual_change_table = annual_change_all_fuel_types,
  version_elec        = "hpr_approved",
  x_limits            = x_limits,
  y_limits            = y_limits,
  binwidth            = binwidth,
  show_category_labels = FALSE
)[[1]]

# remove subplot title and y-label from bottom histogram
p_1rates <- p_1rates + labs(title = NULL, y = NULL) +
  theme(axis.title.y = element_blank()) +
  annotate("text", x = x_limits[2]*0.95, y = y_limits[2]*0.7,
           label = "1.0 Rates", hjust = 1, fontface = "bold", size = 4, color = "#FC9706")

# ------------------------------------------------------------------------------
# Stack and combiner

plot_all_fuels_default_1 <- p_default / p_1rates +
  plot_layout(heights = c(1, 1)) +  # Equal heights for all plots
  plot_annotation(
    title = case_when(
      bill_type == "annual_bill_change_elec" ~ glue::glue("Switching from {baseline_heating_type_option} to Heat Pump: Change in Annual Electric Bill"),
      bill_type == "annual_bill_change_energy" ~ glue::glue("Change in Total Winter Energy Bills after Switching to Heat Pump from {baseline_heating_type_option}"),
      bill_type == "annual_bill_change_gas" ~ glue::glue("Switching from {baseline_heating_type_option} to Heat Pump: Change in Annual Gas Bill"),
      TRUE ~ "Heat Pump Rate Bill Changes" # Default case
    ),
    theme = theme(
      plot.margin = margin(10, 25, 10, 10),  # Add more right margin for labels
      plot.title = element_text(hjust = 0.5, size = 10, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 8),
    )
  ) &
  # Apply consistent theming to all plots
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(linewidth = 0.2, color = "gray90"),
    panel.grid.major.y = element_line(linewidth = 0.2, color = "gray90"),
    aspect.ratio = 0.18,
  )

# Wrap the combined plot to add a shared y-axis label
plot_all_fuels_default_1 <- patchwork::wrap_elements(plot_all_fuels_default_1) +
  labs(tag = "# of Homes") +
  theme(
    plot.tag = element_text(angle = 90, vjust = 0.5, hjust = 0.5, face = "plain", size = 9),
    plot.tag.position = "left"
  )

print(plot_all_fuels_default_1)
```

###  Bill impact of 1.0 seasonal heat pump rates (fig-hists-all-fuel-types)
```{r}
#| label: fig-hists-all-fuel-types
#| fig-cap: "Change in total winter energy bills after switching to heat pumps, for every home in MA served by an investor-owned electric utility, under default vs. 1.0 vs. 2.0 heat pump electric delivery rates. Assumes default gas delivery rates, 2024 supply rates for gas and electricity, and 100% participation in low-income discount programs by eligible households."

########################################################
# Plots
########################################################
result_all_fuels <- plot_bill_change_histograms(
  annual_change_table = annual_change_all_fuel_types,
  bill_type = bill_type,
  baseline_heating_type_option = baseline_heating_type_option,
  electric_utility_option = electric_utility_option,
  supply_year = supply_year,
  second_subtitle = NULL,
  x_limits = c(-2500, 2500),
  y_limits = c(0, 200000)/242.13,
  binwidth = 100
)

# First return item is the plot
plot_all_fuels <- result_all_fuels[[1]]


# results[2-4] are the category percentages for baseline, approved, and doer, respectively.
# Here we assigned them to workspace variables for easy access in the report.
rate_versions <- list(
  list(name = "baseline", result_index = 2, prefix = "baseline"),
  list(name = "approved", result_index = 3, prefix = "approved"),
  list(name = "doer", result_index = 4, prefix = "doer")
)

# Loop through each rate version
for (version in rate_versions) {
  # Get the result for this version
  result <- result_all_fuels[[version$result_index]]

  # Extract large savings percentage
  assign(paste0("fig_hists_all_fuel_type_", version$prefix, "_pct_large_savings"),
         round(result |> filter(bill_category == "large_savings") |> pull(percentage), round_percent))

  # Extract small savings percentage
  assign(paste0("fig_hists_all_fuel_type_", version$prefix, "_pct_small_savings"),
         round(result |> filter(bill_category == "small_savings") |> pull(percentage), round_percent))

  # Calculate total savings percentage
  assign(paste0("fig_hists_all_fuel_type_", version$prefix, "_pct_that_save"),
         result |>
           filter(bill_category == "small_savings" | bill_category == "large_savings") |>
           pull(percentage) |>
           sum() |>
           round(round_percent))
}

print(plot_all_fuels)
```

## ---------- How do savings vary by fuel? ----------
### fig-bar-fuel-types
```{r}
#| label: fig-bar-fuel-types
#| fig-cap: "Proportion of MA homes by primary heating fuel."
#| fig-width: 8
#| fig-height: 2

baseline_heating_type_counts_rs_raw <- get_housing_units_column_counts("electric_utility", "statewide")

# ------------------------------------------------------------------------------
# housing_units (ResStock)
baseline_heating_type_counts_rs_raw <- get_housing_units_column_counts("baseline_heating_type", "statewide")


baseline_heating_type_counts_rs <- baseline_heating_type_counts_rs_raw |>
  mutate(rs_heating_type_group = case_when(
    baseline_heating_type %in% c("Natural Gas") ~ "Natural Gas",
    baseline_heating_type %in% c("Fuel Oil") ~ "Fuel Oil",
    baseline_heating_type %in% c("Propane") ~ "Propane",
    baseline_heating_type %in% c("Resistance") ~ "Resistance",
    baseline_heating_type %in% c("Heat Pump") ~ "Heat Pump",
    baseline_heating_type %in% c("Other Fuel", "None") ~ "Other/None",
    TRUE ~ "Other"
  )) |>
  mutate(rs_heating_type_group = factor(rs_heating_type_group,
                                   levels = c("Natural Gas", "Propane", "Fuel Oil",  "Heat Pump", "Resistance", "Other/None"))) |>
  group_by(rs_heating_type_group, .drop = FALSE) |>
  summarise(count = sum(count), .groups = "drop") |>
  mutate(pct = count / sum(count),
         cum_pct = cumsum(pct),
         pct_label = paste0(rs_heating_type_group, ": ", scales::percent(pct, accuracy=1))
         )

pct_homes_natural_gas <- baseline_heating_type_counts_rs |> filter(rs_heating_type_group == "Natural Gas") |> pull(pct)
pct_homes_fuel_oil <- baseline_heating_type_counts_rs |> filter(rs_heating_type_group == "Fuel Oil") |> pull(pct)
pct_homes_propane <- baseline_heating_type_counts_rs |> filter(rs_heating_type_group == "Propane") |> pull(pct)
pct_homes_electric_resistance <- baseline_heating_type_counts_rs |> filter(rs_heating_type_group == "Resistance") |> pull(pct)

baseline_heating_type_counts_for_plot <- baseline_heating_type_counts_rs |>
  arrange((cum_pct)) |>
  mutate(
    # Start position is the cumulative percentage of previous segments
    start_pos = lag(cum_pct, default = 0),
    # End position is the cumulative percentage including current segment
    end_pos = cum_pct,
    # Center position for text labels
    label_pos = start_pos + pct/2
  )

ggplot() +
  geom_rect(
    data = baseline_heating_type_counts_for_plot,
    aes(xmin = start_pos, xmax = end_pos, ymin = 0.85, ymax = 1.15, fill = rs_heating_type_group)
  ) +
  ggrepel::geom_text_repel(
    data = baseline_heating_type_counts_for_plot,
    aes(x = label_pos, y = 1, label = pct_label, color = rs_heating_type_group),
    direction = "y",
    nudge_y = 0.15,
    nudge_x = 0.005,
    size = 3,
    segment.size = 0.0,
    hjust = 0,
    vjust = 0,
    angle = 30
  ) +
  xlim(0, 1.2) +
  ylim(0.8, 1.4) +
  scale_fill_manual(values = heating_type_colors) +
  scale_color_manual(values = heating_type_colors) +
  theme_void() +
  theme(legend.position = "none")
```



### CRUFT: Compare to EIA
```{r}
# # group to match EIA
# heating_type_counts_rs <- baseline_heating_type_counts_rs_raw |>
#   mutate(heating_type = case_when(
#     baseline_heating_type %in% c("Natural Gas") ~ "Natural Gas",
#     baseline_heating_type %in% c("Fuel Oil") ~ "Fuel Oil",
#     baseline_heating_type %in% c("Propane") ~ "Propane",
#     baseline_heating_type %in% c("Resistance", "Heat Pump") ~ "Electricity",
#     baseline_heating_type %in% c("Other Fuel", "None") ~ "Other/None",
#     TRUE ~ "Other"
#   )) |>
#   group_by(heating_type) |>
#   summarise(count = sum(count), .groups = "drop") |>
#   mutate(pct = count / sum(count)) |>
#   mutate(source = "rs")



# # ------------------------------------------------------------------------------
# # EIA
# # Load EIA state profile data
# heating_type_counts_eia <- read.csv("/workspaces/reports/data/eia/state_energy_profiles/ma_heating_sources_20250517.csv")
# heating_type_counts_eia <- heating_type_counts_eia |>
#   rename(heating_type = Energy.Source,
#          pct = MA,
#          pct_usa = U.S..Average) |>
#   mutate(source = "eia")

# # ------------------------------------------------------------------------------
# # Bind
# heating_type_counts <- bind_rows(
#   heating_type_counts_eia |> select(heating_type, pct, source),
#   heating_type_counts_rs |> select(heating_type, pct, source)
# )
# View(heating_type_counts)

# # heating_type_counts <- heating_type_counts |>
# #   mutate(source = factor(source, levels = c("eia", "rs")))


# plot_order = c("Natural Gas", "Fuel Oil",  "Electricity", "Propane", "Other/None")
# heating_type_counts_plot <- ggplot(
#   heating_type_counts |>
#     filter(source == "eia") |>
#     mutate(heating_type = factor(heating_type, levels = plot_order)),
#   aes(x = heating_type, y = pct)) +
#   geom_bar(stat = "identity", fill = sb_colors["midnight"]) +
#   geom_text(aes(label = sprintf("%.1f%%", pct*100),
#                 y = pct + 0.02),
#             size = 3) +
#   labs(
#     title = "Distribution of Heating Types in MA",
#     x = NULL,
#     y = "Percent of Buildings in MA"
#   ) +
#   scale_y_continuous(labels = scales::percent) +
#   theme(
#     axis.text.x = element_text(angle = 0, hjust = 0.5),
#     panel.grid.major.x = element_blank()
#   )

# print(heating_type_counts_plot)

```

### fig-hists-natural-gas
```{r}
#| label: fig-hists-natural-gas
#| fig-cap: "Change in total winter energy bills after switching to heat pumps, for every natural gas-heated home in MA served by an investor-owned electric utility, under default vs. 1.0 vs. 2.0 heat pump electric delivery rates. Assumes default gas delivery rates, 2024 supply rates for gas and electricity, and 100% participation in low-income programs by eligible households."

########################################################
# Data
########################################################
# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Parameters
bill_type <- 'annual_bill_change_energy'
electric_utility_option <- 'statewide'
baseline_heating_type_option <- 'Natural Gas'
# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
annual_change_natural_gas <- annual_change |>
  filter(electric_utility != "municipal") |>
  filter(!is.na(annual_bill_change) & is.finite(annual_bill_change),
        bill_type == !!bill_type,
        if (electric_utility_option != "statewide") electric_utility == !!electric_utility_option else TRUE,
        if (baseline_heating_type_option != "All Fuels") baseline_heating_type == !!baseline_heating_type_option else TRUE,
        year == !!supply_year
  )

# Print these to console
scenario_stats_natural_gas_low_hp <- calc_stats_by_rate_version(annual_change_natural_gas, "low_hp", print_table = FALSE)
scenario_stats_natural_gas_mid_hp <- calc_stats_by_rate_version(annual_change_natural_gas, "mid_hp", print_table = FALSE)
scenario_stats_natural_gas_high_hp <- calc_stats_by_rate_version(annual_change_natural_gas, "high_hp", print_table = FALSE)
#
#
#
# These are the values used in the report!!!
# Calculate statistics for each version and set to the values used in the report
hp_eff_for_report <- "mid_hp"
scenario_stats_natural_gas <- calc_stats_by_rate_version(annual_change_natural_gas, hp_eff_for_report, print_table = FALSE)

versions <- list(
  list(version = "baseline", prefix = "default"),
  list(version = "hpr_approved", prefix = "approved"),
  list(version = "hpr_doer", prefix = "doer")
)

# Loop through versions to calculate statistics
for (v in versions) {
  # Get the stats for this version
  stats <- scenario_stats_natural_gas |>
    filter(version_elec == v$version)

  # Calculate and assign all statistics for this version
  assign(paste0("median_bill_change_", v$prefix, "_natural_gas"),
         round(stats |> pull(median_change), round_dollars))
  assign(paste0("pct_that_save_", v$prefix, "_natural_gas"),
         round(stats |> pull(pct_that_save), round_percent))
  assign(paste0("median_savings_", v$prefix, "_natural_gas"),
         round(stats |> pull(median_savings), round_dollars))
  assign(paste0("median_loss_", v$prefix, "_natural_gas"),
         round(stats |> pull(median_loss), round_dollars))
}

result_natural_gas <- plot_bill_change_histograms(
  annual_change_table = annual_change_natural_gas,
  bill_type = bill_type,
  baseline_heating_type_option = baseline_heating_type_option,
  electric_utility_option = electric_utility_option,
  supply_year = supply_year,
  #second_subtitle = "Natural Gas Heating",
  x_limits = c(-2500, 2500),
  y_limits = c(0, 200000)/242.13,
  binwidth = 100
)

# First return item is the plot
plot_natural_gas <- result_natural_gas[[1]]


# results[2-4] are the category percentages for baseline, approved, and doer, respectively.
# Here we assigned them to workspace variables for easy access in the report.
rate_versions <- list(
  list(name = "baseline", result_index = 2, prefix = "baseline"),
  list(name = "approved", result_index = 3, prefix = "approved"),
  list(name = "doer", result_index = 4, prefix = "doer")
)

# Loop through each rate version
for (version in rate_versions) {
  # Get the result for this version
  result <- result_natural_gas[[version$result_index]]

  # Extract large savings percentage
  assign(paste0("fig_hists_natural_gas_", version$prefix, "_pct_large_savings"),
         round(result |> filter(bill_category == "large_savings") |> pull(percentage), round_percent))

  # Extract small savings percentage
  assign(paste0("fig_hists_natural_gas_", version$prefix, "_pct_small_savings"),
         round(result |> filter(bill_category == "small_savings") |> pull(percentage), round_percent))

  # Calculate total savings percentage
  assign(paste0("fig_hists_natural_gas_", version$prefix, "_pct_that_save"),
         result |>
           filter(bill_category == "small_savings" | bill_category == "large_savings") |>
           pull(percentage) |>
           sum() |>
           round(round_percent))
}

print(plot_natural_gas)
```

### fig-hists-heating-oil
```{r}
#| label: fig-hists-heating-oil
#| fig-cap: "Change in total winter energy bills after switching to heat pumps, for every heating oil-heated home in MA served by an investor-owned electric utility, under default vs. 1.0 vs. 2.0 heat pump electric delivery rates. Assumes 2024 supply rates for electricity, 2024 heating oil prices, and 100% participation in electric low-income programs by eligible households"

########################################################
# Data
########################################################
# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Parameters
bill_type <- 'annual_bill_change_energy'
electric_utility_option <- 'statewide'
baseline_heating_type_option <- 'Fuel Oil'
# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
annual_change_heating_oil <- annual_change |>
  filter(electric_utility != "municipal") |>
  filter(!is.na(annual_bill_change) & is.finite(annual_bill_change),
        bill_type == !!bill_type,
        if (electric_utility_option != "statewide") electric_utility == !!electric_utility_option else TRUE,
        if (baseline_heating_type_option != "all") baseline_heating_type == !!baseline_heating_type_option else TRUE,
        year == supply_year)

# Print these to console
scenario_stats_natural_gas_low_hp <- calc_stats_by_rate_version(annual_change_heating_oil, "low_hp", print_table = FALSE)
scenario_stats_natural_gas_mid_hp <- calc_stats_by_rate_version(annual_change_heating_oil, "mid_hp", print_table = FALSE)
scenario_stats_natural_gas_high_hp <- calc_stats_by_rate_version(annual_change_heating_oil, "high_hp", print_table = FALSE)
#
#
#
# These are the values used in the report!!!
# Calculate statistics for each version and set to the values used in the report
hp_eff_for_report <- "mid_hp"
scenario_stats_heating_oil <- calc_stats_by_rate_version(annual_change_heating_oil, hp_eff_for_report, print_table = FALSE)

versions <- list(
  list(version = "baseline", prefix = "default"),
  list(version = "hpr_approved", prefix = "approved"),
  list(version = "hpr_doer", prefix = "doer")
)

# Loop through versions to calculate statistics
for (v in versions) {
  # Get the stats for this version
  stats <- scenario_stats_heating_oil |>
    filter(version_elec == v$version)

  # Calculate and assign all statistics for this version
  assign(paste0("median_bill_change_", v$prefix, "_heating_oil"),
         round(stats |> pull(median_change), round_dollars))
  assign(paste0("pct_that_save_", v$prefix, "_heating_oil"),
         round(stats |> pull(pct_that_save), round_percent))
  assign(paste0("median_savings_", v$prefix, "_heating_oil"),
         round(stats |> pull(median_savings), round_dollars))
  assign(paste0("median_loss_", v$prefix, "_heating_oil"),
         round(stats |> pull(median_loss), round_dollars))
}


########################################################
# Plots
########################################################
result_heating_oil <- plot_bill_change_histograms(
  annual_change_table = annual_change_heating_oil,
  bill_type = bill_type,
  baseline_heating_type_option = baseline_heating_type_option,
  electric_utility_option = electric_utility_option,
  supply_year = supply_year,
  #second_subtitle = "Fuel Oil Heating",
  x_limits = c(-2500, 2500),
  y_limits = c(0, 50000)/242.13,
  binwidth = 100
)

# First return item is the plot
plot_heating_oil <- result_heating_oil[[1]]


# results[2-4] are the category percentages for baseline, approved, and doer, respectively.
# Here we assigned them to workspace variables for easy access in the report.
rate_versions <- list(
  list(name = "baseline", result_index = 2, prefix = "baseline"),
  list(name = "approved", result_index = 3, prefix = "approved"),
  list(name = "doer", result_index = 4, prefix = "doer")
)

# Loop through each rate version
for (version in rate_versions) {
  # Get the result for this version
  result <- result_heating_oil[[version$result_index]]

  # Extract large savings percentage
  assign(paste0("fig_hists_heating_oil_", version$prefix, "_pct_large_savings"),
         round(result |> filter(bill_category == "large_savings") |> pull(percentage), round_percent))

  # Extract small savings percentage
  assign(paste0("fig_hists_heating_oil_", version$prefix, "_pct_small_savings"),
         round(result |> filter(bill_category == "small_savings") |> pull(percentage), round_percent))

  # Calculate total savings percentage
  assign(paste0("fig_hists_heating_oil_", version$prefix, "_pct_that_save"),
         result |>
           filter(bill_category == "small_savings" | bill_category == "large_savings") |>
           pull(percentage) |>
           sum() |>
           round(round_percent))
}

print(plot_heating_oil)
```

### fig-hists-propane
```{r}
#| label: fig-hists-propane
#| fig-cap: "Change in total winter energy bills after switching to heat pumps, for every propane-heated home in MA served by an investor-owned electric utility, under default vs. 1.0 vs. 2.0 heat pump electric delivery rates. Assumes 2024 supply rates for electricity, 2024 propane prices, and 100% participation in electric low-income programs by eligible households"

########################################################
# Data
########################################################
# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Parameters
bill_type <- 'annual_bill_change_energy'
electric_utility_option <- 'statewide'
baseline_heating_type_option <- 'Propane'
# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
annual_change_propane <- annual_change |>
  filter(electric_utility != "municipal") |>
  filter(!is.na(annual_bill_change) & is.finite(annual_bill_change),
        bill_type == !!bill_type,
        if (electric_utility_option != "statewide") electric_utility == !!electric_utility_option else TRUE,
        if (baseline_heating_type_option != "all") baseline_heating_type == !!baseline_heating_type_option else TRUE,
        year == supply_year)

# Print these to console
scenario_stats_propane_low_hp <- calc_stats_by_rate_version(annual_change_propane, "low_hp", print_table = FALSE)
scenario_stats_propane_mid_hp <- calc_stats_by_rate_version(annual_change_propane, "mid_hp", print_table = FALSE)
scenario_stats_propane_high_hp <- calc_stats_by_rate_version(annual_change_propane, "high_hp", print_table = FALSE)
#
#
#
# These are the values used in the report!!!
# Calculate statistics for each version and set to the values used in the report
hp_eff_for_report <- "mid_hp"
scenario_stats_propane <- calc_stats_by_rate_version(annual_change_propane, hp_eff_for_report, print_table = FALSE)

versions <- list(
  list(version = "baseline", prefix = "default"),
  list(version = "hpr_approved", prefix = "approved"),
  list(version = "hpr_doer", prefix = "doer")
)

# Loop through versions to calculate statistics
for (v in versions) {
  # Get the stats for this version
  stats <- scenario_stats_propane |>
    filter(version_elec == v$version)

  # Calculate and assign all statistics for this version
  assign(paste0("median_bill_change_", v$prefix, "_propane"),
         round(stats |> pull(median_change), round_dollars))
  assign(paste0("pct_that_save_", v$prefix, "_propane"),
         round(stats |> pull(pct_that_save), round_percent))
  assign(paste0("median_savings_", v$prefix, "_propane"),
         round(stats |> pull(median_savings), round_dollars))
  assign(paste0("median_loss_", v$prefix, "_propane"),
         round(stats |> pull(median_loss), round_dollars))
}

result_propane <- plot_bill_change_histograms(
  annual_change_table = annual_change_propane,
  bill_type = bill_type,
  baseline_heating_type_option = baseline_heating_type_option,
  electric_utility_option = electric_utility_option,
  supply_year = supply_year,
  #second_subtitle = "Natural Gas Heating",
  x_limits = c(-2500, 2500),
  y_limits = c(0, 10000)/242.13,
  binwidth = 100
)

# First return item is the plot
plot_propane <- result_propane[[1]]


# results[2-4] are the category percentages for baseline, approved, and doer, respectively.
# Here we assigned them to workspace variables for easy access in the report.
rate_versions <- list(
  list(name = "baseline", result_index = 2, prefix = "baseline"),
  list(name = "approved", result_index = 3, prefix = "approved"),
  list(name = "doer", result_index = 4, prefix = "doer")
)

# Loop through each rate version
for (version in rate_versions) {
  # Get the result for this version
  result <- result_propane[[version$result_index]]

  # Extract large savings percentage
  assign(paste0("fig_hists_propane_", version$prefix, "_pct_large_savings"),
         round(result |> filter(bill_category == "large_savings") |> pull(percentage), round_percent))

  # Extract small savings percentage
  assign(paste0("fig_hists_propane_", version$prefix, "_pct_small_savings"),
         round(result |> filter(bill_category == "small_savings") |> pull(percentage), round_percent))

  # Calculate total savings percentage
  assign(paste0("fig_hists_propane_", version$prefix, "_pct_that_save"),
         result |>
           filter(bill_category == "small_savings" | bill_category == "large_savings") |>
           pull(percentage) |>
           sum() |>
           round(round_percent))
}

print(plot_propane)
```

### fig-hists-resistance
```{r}
#| label: fig-hists-resistance
#| fig-cap: "Change in total winter energy bills after switching to heat pumps, for every electric resistance-heated home in MA served by an investor-owned electric utility, under default vs. 1.0 vs. 2.0 heat pump electric delivery rates. Assumes default gas delivery rates, 2024 supply rates for gas and electricity, and 100% participation in low-income programs by eligible households"

########################################################
# Data
########################################################
# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Parameters
bill_type <- 'annual_bill_change_energy'
electric_utility_option <- 'statewide'
baseline_heating_type_option <- 'Resistance'
# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
annual_change_resistance <- annual_change |>
  filter(electric_utility != "municipal") |>
  # NOTE!!!! Deliberately filtering out positive bill changes
  filter(!is.na(annual_bill_change) & is.finite(annual_bill_change) & annual_bill_change < 0,
        bill_type == !!bill_type,
        if (electric_utility_option != "statewide") electric_utility == !!electric_utility_option else TRUE,
        if (baseline_heating_type_option != "all") baseline_heating_type == !!baseline_heating_type_option else TRUE,
        year == supply_year)

# Print these to console
scenario_stats_resistance_low_hp <- calc_stats_by_rate_version(annual_change_resistance, "low_hp", print_table = FALSE)
scenario_stats_resistance_mid_hp <- calc_stats_by_rate_version(annual_change_resistance, "mid_hp", print_table = FALSE)
scenario_stats_resistance_high_hp <- calc_stats_by_rate_version(annual_change_resistance, "high_hp", print_table = FALSE)
#
#
#
# These are the values used in the report!!!
# Calculate statistics for each version and set to the values used in the report
hp_eff_for_report <- "mid_hp"
scenario_stats_resistance <- calc_stats_by_rate_version(annual_change_resistance, hp_eff_for_report, print_table = FALSE)

versions <- list(
  list(version = "baseline", prefix = "default"),
  list(version = "hpr_approved", prefix = "approved"),
  list(version = "hpr_doer", prefix = "doer")
)

# Loop through versions to calculate statistics
for (v in versions) {
  # Get the stats for this version
  stats <- scenario_stats_resistance |>
    filter(version_elec == v$version)

  # Calculate and assign all statistics for this version
  assign(paste0("median_bill_change_", v$prefix, "_resistance"),
         round(stats |> pull(median_change), round_dollars))
  assign(paste0("pct_that_save_", v$prefix, "_resistance"),
         round(stats |> pull(pct_that_save), round_percent))
  assign(paste0("median_savings_", v$prefix, "_resistance"),
         round(stats |> pull(median_savings), round_dollars))
  assign(paste0("median_loss_", v$prefix, "_resistance"),
         round(stats |> pull(median_loss), round_dollars))
}

result_resistance <- plot_bill_change_histograms(
  annual_change_table = annual_change_resistance,
  bill_type = bill_type,
  baseline_heating_type_option = baseline_heating_type_option,
  electric_utility_option = electric_utility_option,
  supply_year = supply_year,
  #second_subtitle = "Natural Gas Heating",
  x_limits = c(-2500, 2500),
  y_limits = c(0, 50000)/242.13,
  binwidth = 100
)

# First return item is the plot
plot_resistance <- result_resistance[[1]]


# results[2-4] are the category percentages for baseline, approved, and doer, respectively.
# Here we assigned them to workspace variables for easy access in the report.
rate_versions <- list(
  list(name = "baseline", result_index = 2, prefix = "baseline"),
  list(name = "approved", result_index = 3, prefix = "approved"),
  list(name = "doer", result_index = 4, prefix = "doer")
)

# Loop through each rate version
for (version in rate_versions) {
  # Get the result for this version
  result <- result_resistance[[version$result_index]]

  # Extract large savings percentage
  assign(paste0("fig_hists_resistance_", version$prefix, "_pct_large_savings"),
         round(result |> filter(bill_category == "large_savings") |> pull(percentage), round_percent))

  # Extract small savings percentage
  assign(paste0("fig_hists_resistance_", version$prefix, "_pct_small_savings"),
         round(result |> filter(bill_category == "small_savings") |> pull(percentage), round_percent))

  # Calculate total savings percentage
  assign(paste0("fig_hists_resistance_", version$prefix, "_pct_that_save"),
         result |>
           filter(bill_category == "small_savings" | bill_category == "large_savings") |>
           pull(percentage) |>
           sum() |>
           round(round_percent))
}

print(plot_resistance)
```

## ---------- How do savings vary by building type? ----------
### data-building-type
```{r}
#| label: data-building-type
annual_change_bldg_type <- annual_change |>
  mutate(
    building_type_group = case_when(
      building_type %in% c("Single-Family Detached", "Mobile Home", "Single-Family Attached") ~ "Single-Family",
      building_type %in% c("2 Unit", "3 or 4 Unit") ~ "2-4 Units",
      building_type %in% c("5 to 9 Unit", "10 to 19 Unit", "20 to 49 Unit", "50 or more Unit") ~ "5+ Units",
      TRUE ~ "Other"
    ),
  )

# Create box plots by building type and version_elec
building_type_group_order = c("Single-Family", "2-4 Units", "5+ Units")

building_type_group_names = c("Single-Family", "2-4 Units", "5+ Units")

data_box_building_type <- annual_change_bldg_type |>
  filter(!is.na(annual_bill_change) & is.finite(annual_bill_change),
        bill_type == "annual_bill_change_energy",
        raw_or_discounted == "discounted",
        building_type_group %in% building_type_group_order)

# Convert building_type to factor with specified order
data_box_building_type <- data_box_building_type |>
  mutate(building_type_group = factor(building_type_group, levels = building_type_group_order))

# Calculate custom boxplot stats for each group
box_building_type_stats <- data_box_building_type |>
  group_by(building_type_group, version_elec) |>
  summarise(
    data_min   = min(annual_bill_change, na.rm = TRUE),                # lower whisker
    lower  = quantile(annual_bill_change, 0.1, na.rm = TRUE),     # lower hinge (Q1)
    middle = median(annual_bill_change, na.rm = TRUE),             # median
    upper  = quantile(annual_bill_change, 0.9, na.rm = TRUE),     # upper hinge (Q3)
    data_max   = max(annual_bill_change, na.rm = TRUE),                # upper whisker
    pct_that_save = mean(annual_bill_change < 0, na.rm = TRUE),
    median_savings = median(annual_bill_change[annual_bill_change < 0], na.rm = TRUE),  # median of negative values (savings)
    .groups = "drop",
  )

```

### fig-stacked-bar-building-type
```{r}
#| label: fig-stacked-bar-building-type
#| fig-cap: "Proportion of MA homes by building type."
#| fig-width: 8
#| fig-height: 1

# Get counts of buildings by type
building_counts <- data_box_building_type |>
  group_by(building_type_group) |>
  summarise(n = n_distinct(bldg_id)) |>
  mutate(pct = n / sum(n) * 100) |>
  arrange(match(building_type_group, building_type_group_order))

# Calculate placement for labels
building_counts <- building_counts |>
  mutate(
    pct = n / sum(n),
    pct_label = paste0(building_type_group, ": ", scales::percent(pct, accuracy=1)),
    cum_pct = cumsum(pct),
    placement = cum_pct
  )

building_counts_for_plot <- building_counts |>
  mutate(
    # Start position is the cumulative percentage of previous segments
    start_pos = lag(cum_pct, default = 0),
    # End position is the cumulative percentage including current segment
    end_pos = cum_pct,
    # Center position for text labels
    center_pos = (start_pos + end_pos) / 2
  )

ggplot() +
  geom_rect(
    data = building_counts_for_plot,
    aes(xmin = start_pos, xmax = end_pos, ymin = 0.85, ymax = 1.15, fill = building_type_group)
  ) +
  ggrepel::geom_text_repel(
    data = building_counts_for_plot,
    aes(x = end_pos-0.005, y = 1, label = pct_label, color = building_type_group),
    direction = "y",
    nudge_y = 0.2,
    nudge_x = -0.05,
    size = 3,
    segment.size = 0.0,
    hjust = .8,
    vjust = 0.5
  ) +
  xlim(0, 1.2) +
  ylim(0.8, 1.2) +
  scale_fill_manual(values = housing_type_colors) +
  scale_color_manual(values = housing_type_colors) +
  theme_void() +
  theme(legend.position = "none")

```


### fig-hists-single-family-home
```{r}
#| label: fig-hists-single-family-home
#| fig-cap: "Change in total winter energy bills after switching to heat pumps, for every single-family home in MA served by an investor-owned electric utility, under default vs. 1.0 vs. 2.0 heat pump electric delivery rates. Assumes default gas delivery rates, 2024 supply rates for gas and electricity, and 100% participation in low-income programs by eligible households"

bill_type <- 'annual_bill_change_energy'
electric_utility_option <- 'Statewide EDC'
baseline_heating_type_option <- 'All Fuels'

building_type_group <- "Single-Family"

annual_change_single_family_home <- annual_change_bldg_type |>
  filter(electric_utility != "municipal") |>
  filter(!is.na(annual_bill_change) & is.finite(annual_bill_change),
        bill_type == !!bill_type,
        if (electric_utility_option != "Statewide EDC") electric_utility == !!electric_utility_option else TRUE,
        if (baseline_heating_type_option != "All Fuels") baseline_heating_type == !!baseline_heating_type_option else TRUE,
        year == supply_year,
        building_type_group == !!building_type_group
        )



result_single_family_home <- plot_bill_change_histograms(
  annual_change_table = annual_change_single_family_home,
  bill_type = bill_type,
  baseline_heating_type_option = baseline_heating_type_option,
  electric_utility_option = electric_utility_option,
  supply_year = supply_year,
  second_subtitle = "Single-Family Homes",
  x_limits = c(-2500, 2500),
  y_limits = c(0, 100000)/242.13,
  binwidth = 100
)

# First return item is the plot
plot_single_family_home <- result_single_family_home[[1]]


# results[2-4] are the category percentages for baseline, approved, and doer, respectively.
# Here we assigned them to workspace variables for easy access in the report.
rate_versions <- list(
  list(name = "baseline", result_index = 2, prefix = "baseline"),
  list(name = "approved", result_index = 3, prefix = "approved"),
  list(name = "doer", result_index = 4, prefix = "doer")
)

# Loop through each rate version
for (version in rate_versions) {
  # Get the result for this version
  result <- result_single_family_home[[version$result_index]]

  # Extract large savings percentage
  assign(paste0("fig_hists_single_family_home_", version$prefix, "_pct_large_savings"),
         round(result |> filter(bill_category == "large_savings") |> pull(percentage), round_percent))

  # Extract small savings percentage
  assign(paste0("fig_hists_single_family_home_", version$prefix, "_pct_small_savings"),
         round(result |> filter(bill_category == "small_savings") |> pull(percentage), round_percent))

  # Calculate total savings percentage
  assign(paste0("fig_hists_single_family_home_", version$prefix, "_pct_that_save"),
         result |>
           filter(bill_category == "small_savings" | bill_category == "large_savings") |>
           pull(percentage) |>
           sum() |>
           round(round_percent))
}

print(plot_single_family_home)
```

### fig-hists-2-4-units
```{r}
#| label: fig-hists-2-4-units
#| fig-cap: "Change in total winter energy bills after switching to heat pumps, for every MA home in a 2-4 unit building served by an investor-owned electric utility, under default vs. 1.0 vs. 2.0 heat pump electric delivery rates. Assumes default gas delivery rates, 2024 supply rates for gas and electricity, and 100% participation in low-income programs by eligible households"

bill_type <- 'annual_bill_change_energy'
electric_utility_option <- 'Statewide EDC'
baseline_heating_type_option <- 'All Fuels'

building_type_group <- "2-4 Units"

annual_change_2_4_units <- annual_change_bldg_type |>
  filter(electric_utility != "municipal") |>
  filter(!is.na(annual_bill_change) & is.finite(annual_bill_change),
        bill_type == !!bill_type,
        if (electric_utility_option != "Statewide EDC") electric_utility == !!electric_utility_option else TRUE,
        if (baseline_heating_type_option != "All Fuels") baseline_heating_type == !!baseline_heating_type_option else TRUE,
        year == supply_year,
        building_type_group == !!building_type_group
        )


result_2_4_units <- plot_bill_change_histograms(
  annual_change_table = annual_change_2_4_units,
  bill_type = bill_type,
  baseline_heating_type_option = baseline_heating_type_option,
  electric_utility_option = electric_utility_option,
  supply_year = supply_year,
  second_subtitle = "2-4 Unit Building",
  x_limits = c(-2500, 2500),
  y_limits = c(0, 100000)/242.13,
  binwidth = 100
)

# First return item is the plot
plot_2_4_units <- result_2_4_units[[1]]


# results[2-4] are the category percentages for baseline, approved, and doer, respectively.
# Here we assigned them to workspace variables for easy access in the report.
rate_versions <- list(
  list(name = "baseline", result_index = 2, prefix = "baseline"),
  list(name = "approved", result_index = 3, prefix = "approved"),
  list(name = "doer", result_index = 4, prefix = "doer")
)

# Loop through each rate version
for (version in rate_versions) {
  # Get the result for this version
  result <- result_2_4_units[[version$result_index]]

  # Extract large savings percentage
  assign(paste0("fig_hists_2_4_units_", version$prefix, "_pct_large_savings"),
         round(result |> filter(bill_category == "large_savings") |> pull(percentage), round_percent))

  # Extract small savings percentage
  assign(paste0("fig_hists_2_4_units_", version$prefix, "_pct_small_savings"),
         round(result |> filter(bill_category == "small_savings") |> pull(percentage), round_percent))

  # Calculate total savings percentage
  assign(paste0("fig_hists_2_4_units_", version$prefix, "_pct_that_save"),
         result |>
           filter(bill_category == "small_savings" | bill_category == "large_savings") |>
           pull(percentage) |>
           sum() |>
           round(round_percent))
}

print(plot_2_4_units)
```

### fig-hists-5-plus-units
```{r}
#| label: fig-hists-5-plus-units
#| fig-cap: "Change in total winter energy bills after switching to heat pumps, for every MA home in a 5+ unit building served by an investor-owned electric utility, under default vs. 1.0 vs. 2.0 heat pump electric delivery rates. Assumes default gas delivery rates, 2024 supply rates for gas and electricity, and 100% participation in low-income programs by eligible households"

bill_type <- 'annual_bill_change_energy'
electric_utility_option <- 'Statewide EDC'
baseline_heating_type_option <- 'All Fuels'

building_type_group <- "5+ Units"

annual_change_5_plus_units <- annual_change_bldg_type |>
  filter(electric_utility != "municipal") |>
  filter(!is.na(annual_bill_change) & is.finite(annual_bill_change),
        bill_type == !!bill_type,
        if (electric_utility_option != "Statewide EDC") electric_utility == !!electric_utility_option else TRUE,
        if (baseline_heating_type_option != "All Fuels") baseline_heating_type == !!baseline_heating_type_option else TRUE,
        year == supply_year,
        building_type_group == !!building_type_group
        )


result_5_plus_units <- plot_bill_change_histograms(
  annual_change_table = annual_change_5_plus_units,
  bill_type = bill_type,
  baseline_heating_type_option = baseline_heating_type_option,
  electric_utility_option = electric_utility_option,
  supply_year = supply_year,
  second_subtitle = "5+ Unit Building",
  x_limits = c(-2500, 2500),
  y_limits = c(0, 100000)/242.13,
  binwidth = 100
)

# First return item is the plot
plot_5_plus_units <- result_5_plus_units[[1]]


# results[2-4] are the category percentages for baseline, approved, and doer, respectively.
# Here we assigned them to workspace variables for easy access in the report.
rate_versions <- list(
  list(name = "baseline", result_index = 2, prefix = "baseline"),
  list(name = "approved", result_index = 3, prefix = "approved"),
  list(name = "doer", result_index = 4, prefix = "doer")
)

# Loop through each rate version
for (version in rate_versions) {
  # Get the result for this version
  result <- result_5_plus_units[[version$result_index]]

  # Extract large savings percentage
  assign(paste0("fig_hists_5_plus_units_", version$prefix, "_pct_large_savings"),
         round(result |> filter(bill_category == "large_savings") |> pull(percentage), round_percent))

  # Extract small savings percentage
  assign(paste0("fig_hists_5_plus_units_", version$prefix, "_pct_small_savings"),
         round(result |> filter(bill_category == "small_savings") |> pull(percentage), round_percent))

  # Calculate total savings percentage
  assign(paste0("fig_hists_5_plus_units_", version$prefix, "_pct_that_save"),
         result |>
           filter(bill_category == "small_savings" | bill_category == "large_savings") |>
           pull(percentage) |>
           sum() |>
           round(round_percent))
}

print(plot_5_plus_units)
```

### fig-box-building-type
```{r data-box-building-type}
# Create filtered dataset for ridge plots - remove top/bottom 2.5% of values for each group
data_ridges_trimmed_building_type <- data_box_building_type |>
  group_by(building_type_group, version_elec) |>
  mutate(
    lower_bound = quantile(annual_bill_change, 0.05, na.rm = TRUE),
    upper_bound = quantile(annual_bill_change, 0.95, na.rm = TRUE)
  ) |>
  filter(annual_bill_change >= lower_bound & annual_bill_change <= upper_bound) |>
  ungroup()
```

### fig-box-building-type-default-only
```{r}
pct_save_5plus_baseline <- round((box_building_type_stats |> filter(building_type_group == "5+ Units", version_elec == "baseline") |> pull(pct_that_save)*100))

pct_save_5plus_approved <- round((box_building_type_stats |> filter(building_type_group == "5+ Units", version_elec == "hpr_approved") |> pull(pct_that_save)*100))

pct_save_5plus_doer <- round((box_building_type_stats |> filter(building_type_group == "5+ Units", version_elec == "hpr_doer") |> pull(pct_that_save)*100))


pct_save_2to4_baseline <- round((box_building_type_stats |> filter(building_type_group == "2-4 Units", version_elec == "baseline") |> pull(pct_that_save))*100)

pct_save_2to4_approved <- round((box_building_type_stats |> filter(building_type_group == "2-4 Units", version_elec == "hpr_approved") |> pull(pct_that_save))*100)

pct_save_2to4_doer <- round((box_building_type_stats |> filter(building_type_group == "2-4 Units", version_elec == "hpr_doer") |> pull(pct_that_save))*100)


pct_save_singlefam_baseline <- round((box_building_type_stats |> filter(building_type_group == "Single-Family", version_elec == "baseline") |> pull(pct_that_save))*100)

pct_save_singlefam_approved <- round((box_building_type_stats |> filter(building_type_group == "Single-Family", version_elec == "hpr_approved") |> pull(pct_that_save))*100)

pct_save_singlefam_doer <- round((box_building_type_stats |> filter(building_type_group == "Single-Family", version_elec == "hpr_doer") |> pull(pct_that_save))*100)

# Define median_savings variables
median_savings_singlefam_doer <- round(box_building_type_stats |>
  filter(building_type_group == "Single-Family", version_elec == "hpr_doer") |> pull(median_savings))

median_savings_singlefam_approved <- round(box_building_type_stats |>
  filter(building_type_group == "Single-Family", version_elec == "hpr_approved") |> pull(median_savings))

median_savings_singlefam_baseline <- round(box_building_type_stats |>
  filter(building_type_group == "Single-Family", version_elec == "baseline") |> pull(median_savings))

median_savings_2to4_doer <- round(box_building_type_stats |>
  filter(building_type_group == "2-4 Units", version_elec == "hpr_doer") |> pull(median_savings))

median_savings_5plus_doer <- round(box_building_type_stats |>
  filter(building_type_group == "5+ Units", version_elec == "hpr_doer") |> pull(median_savings))


```


## ---------- How do heat pump rates affect winter energy burdens? ----------
### data-impact-of-low-income-discounts
```{r}
#| label: data-impact-of-low-income-discounts

incomes_threshold <- 50000
binwidth <- 250

# Energy Burdens today
li_baseline_rates_current_hvac_0pct_participation <- annual_bills_undiscounted |>
  filter(bill_type == "annual_bill_raw_energy" & hvac == "current" & version_elec == "baseline" & assigned_income <= incomes_threshold)

li_baseline_rates_current_hvac_100pct_participation <- annual_bills |>
  filter(bill_type == "annual_bill_energy" & hvac == "current" & version_elec == "baseline" & assigned_income <= incomes_threshold)

# After switching to heat pumps, but with current rates
li_baseline_undiscounted <- annual_bills_undiscounted |>
  filter(bill_type == "annual_bill_raw_energy" & hvac == hvac_for_utility_bills & version_elec == "baseline" & assigned_income <= incomes_threshold)
li_baseline_no_elec_discount <- annual_bills_no_elec_discount |>
  filter(bill_type == "annual_bill_energy" & hvac == hvac_for_utility_bills & version_elec == "baseline" & assigned_income <= incomes_threshold)
li_baseline_discounted <- annual_bills |>
  filter(bill_type == "annual_bill_energy" & hvac == hvac_for_utility_bills & version_elec == "baseline" & assigned_income <= incomes_threshold)

# Heat Pump Rates
li_hpr_approved_discounted <- annual_bills |>
  filter(bill_type == "annual_bill_energy" & hvac == hvac_for_utility_bills & version_elec == "hpr_approved" & assigned_income <= incomes_threshold)
li_hpr_doer_discounted <- annual_bills |>
  filter(bill_type == "annual_bill_energy" & hvac == hvac_for_utility_bills & version_elec == "hpr_doer" & assigned_income <= incomes_threshold)

```

### fig-hists-impact-of-low-income-discounts
```{r}
#| label: fig-hists-impact-of-low-income-discounts
#|

plot_li_discount_histograms <- function(
  datasets,
  hvac_for_utility_bills,
  incomes_threshold,
  x_limits = c(0, 0.2),
  y_limits = c(0, 150000/242.13),
  binwidth = 0.01
) {
  # # Define dataset list with names for easy processing
  # datasets <- list(
  #   "Baseline Rates, Current HVAC, 100% Participation" = li_baseline_rates_current_hvac_100pct_participation,
  #   "Baseline Rates, Current HVAC, 0% Participation" = li_baseline_rates_current_hvac_0pct_participation,
  #   "Baseline Rates, Mid HP, Undiscounted" = li_baseline_undiscounted,
  #   "Baseline Rates, Mid HP, Discounted" = li_baseline_discounted,
  #   "Approved Rates, Mid HP, Discounted" = li_hpr_approved_discounted,
  #   "DOER Rates, Mid HP, Discounted" = li_hpr_doer_discounted
  # )

  # Define color palette for each dataset
  colors <- c(
    "Baseline Rates, Current HVAC, 100% Participation" = "magenta",
    "Baseline Rates, Current HVAC, 0% Participation" = "red",
    "Baseline Rates, Mid HP, Undiscounted" = "grey50",
    "Baseline Rates, Mid HP, Discounted" = "#023047",
    "Approved Rates, Mid HP, Discounted" = "#FC9706",
    "DOER Rates, Mid HP, Discounted" = "#68BED8"
  )

  # Calculate y_max across all datasets if not provided
  if (is.null(y_limits)) {
    y_max <- 10
    for (dataset_name in names(datasets)) {
      hist_plot <- ggplot(datasets[[dataset_name]], aes(x = energy_burden)) +
        geom_histogram(binwidth = binwidth)

      current_max <- max(ggplot_build(hist_plot)$data[[1]]$count, na.rm = TRUE)
      y_max <- max(y_max, current_max)
    }

    y_max <- ceiling(y_max * 1.1)
    y_limits <- c(0, y_max)
  }

  #Function to create a single histogram for a dataset
  create_single_histogram <- function(data, name, is_first = FALSE, is_last = FALSE) {
    low_burden_threshold <- 0.04
    moderate_burden_threshold <- 0.06
    high_burden_threshold <- 0.12


    # Calculate bill categories
    data <- data |>
      mutate(bill_category = case_when(
        energy_burden < low_burden_threshold ~ "low_energy_burden",
        energy_burden < moderate_burden_threshold ~ "moderate_energy_burden",
        energy_burden < high_burden_threshold ~ "high_energy_burden",
        TRUE ~ "very_high_energy_burden"
      ))

    # Calculate percentages for each category
    category_percentages <- data |>
      group_by(bill_category) |>
      summarise(count = n(), .groups = "drop") |>
      mutate(percentage = count / sum(count) * 100)

    # Get percentages for annotation
    pct_low_energy_burden <- category_percentages |>
      filter(bill_category == "low_energy_burden") |>
      pull(percentage) |>
      round(1)

    pct_moderate_energy_burden <- category_percentages |>
      filter(bill_category == "moderate_energy_burden") |>
      pull(percentage) |>
      round(1)

    pct_high_energy_burden <- category_percentages |>
      filter(bill_category == "high_energy_burden") |>
      pull(percentage) |>
      round(1)

    pct_very_high_energy_burden <- category_percentages |>
      filter(bill_category == "very_high_energy_burden") |>
      pull(percentage) |>
      round(1)

    # Calculate bin breaks and midpoints
    breaks <- seq(floor(x_limits[1]/binwidth)*binwidth, ceiling(x_limits[2]/binwidth)*binwidth, by = binwidth)

    # Manually bin the data
    binned_data <- data |>
      mutate(bin = cut(energy_burden,
                      breaks = breaks,
                      include.lowest = TRUE,
                      right = FALSE)) |>
      group_by(bin, bill_category) |>
      summarise(count = n(), .groups = "drop") |>
      # Add bin midpoint for plotting
      mutate(bin_mid = as.numeric(as.character(
        sapply(bin, function(b) {
          # Extract bin boundaries and calculate midpoint
          vals <- as.numeric(gsub("\\(|\\]|\\[|\\)", "", strsplit(as.character(b), ",")[[1]]))
          return(mean(vals))
        })
      )))

    # More balanced color palette
    dark_red <- "#db8b87"        # dark red
    light_red <- "#eaada9"       # light red
    light_green <- "#8ebd85"     # light green
    dark_green <- "#5e8a5e"      # dark green

    # Position the bars at the top and text just below
    bar_y_pos <- y_limits[2]
    text_y_pos <- y_limits[2] * 0.90

    # Calculate x positions for percentage labels
    x_low_energy_burden <- mean(c(0, low_burden_threshold-0.002))
    x_moderate_energy_burden <- mean(c(low_burden_threshold, moderate_burden_threshold))
    x_high_energy_burden <- mean(c(moderate_burden_threshold, high_burden_threshold))
    x_very_high_energy_burden <- mean(c(high_burden_threshold, x_limits[2]))

    # Create the plot
    plot <- ggplot(binned_data, aes(x = bin_mid, y = count, fill = bill_category)) +
      # Add vertical dotted lines at the category boundaries
      geom_vline(xintercept = c(low_burden_threshold, moderate_burden_threshold, high_burden_threshold), linetype = "dotted", color = "gray50", size = 0.5) +

      # Add the bars
      geom_col(position = "stack", width = binwidth * 0.9) +

      # Set the fill colors
      scale_fill_manual(
        values = c(
          "low_energy_burden" = dark_green,
          "moderate_energy_burden" = light_green,
          "high_energy_burden" = light_red,
          "very_high_energy_burden" = dark_red
        ),
        guide = "none"
      ) +

      # Add horizontal colored lines at the top
      annotate("segment",
                x = x_limits[1], xend = low_burden_threshold,
                y = bar_y_pos, yend = bar_y_pos,
                color = dark_green, size = 1) +

      annotate("segment",
                x = low_burden_threshold, xend = moderate_burden_threshold,
                y = bar_y_pos, yend = bar_y_pos,
                color = light_green, size = 1) +

      annotate("segment",
                x = moderate_burden_threshold, xend = high_burden_threshold,
                y = bar_y_pos, yend = bar_y_pos,
                color = light_red, size = 1) +

      annotate("segment",
                x = high_burden_threshold, xend = x_limits[2],
                y = bar_y_pos, yend = bar_y_pos,
                color = dark_red, size = 1) +

      # Add percentage labels
      annotate("text",
                x = x_low_energy_burden,
                y = text_y_pos * 0.95,
                label = paste0(pct_low_energy_burden, "%"),
                size = 3,
                fontface = "bold") +

      annotate("text",
                x = x_moderate_energy_burden,
                y = text_y_pos * 0.95,
                label = paste0(pct_moderate_energy_burden, "%"),
                size = 3,
                fontface = "bold") +

      annotate("text",
                x = x_high_energy_burden,
                y = text_y_pos * 0.95,
                label = paste0(pct_high_energy_burden, "%"),
                size = 3,
                fontface = "bold") +

      annotate("text",
                x = x_very_high_energy_burden,
                y = text_y_pos * 0.95,
                label = paste0(pct_very_high_energy_burden, "%"),
                size = 3,
                fontface = "bold") +

      # Add dataset name on the right
      annotate("text",
                x = x_limits[2] * 0.95,
                y = y_limits[2] * 0.7,
                label = name,
                hjust = 1,
                fontface = "bold",
                size = 3,
                color = colors[name]) +

      # Format axes
      scale_y_continuous(
        labels = function(x) paste0(round(x * 242.13 * 0.001), "k"),
        name = "# of Homes",
        limits = y_limits,
        breaks = c(0, 50000, 100000, 150000)/242.13
      ) +
      scale_x_continuous(
        labels = scales::percent_format(accuracy = 1),
        name = "Energy Burden",
        limits = x_limits,
        breaks = c(low_burden_threshold, moderate_burden_threshold, high_burden_threshold)
      ) +

      # Apply theme
      theme_minimal() +
      theme(
        panel.grid.major = element_line(linewidth = 0.2),
        panel.grid.minor = element_blank(),
        aspect.ratio = 0.15
      )

    # Configure axes based on position
    if (!is_last) {
      plot <- plot +
        theme(
          axis.title.x = element_blank(),
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank()
        )
    }
    return(list(plot, category_percentages))
  }


  # Create each histogram
  dataset_names <- names(datasets)
  plots <- list()
  list_of_category_percentages <- list()
  for (i in 1:length(datasets)) {
    name <- dataset_names[i]
    is_first <- (i == 1)
    is_last <- (i == length(datasets))

    result_single_histogram <- create_single_histogram(
      data = datasets[[name]],
      name = name,
      is_first = is_first,
      is_last = is_last
    )

    plots[[i]] <- result_single_histogram[[1]]
    list_of_category_percentages[[i]] <- result_single_histogram[[2]]
  }

  # Combine plots with patchwork
  combined_plot <- plots[[1]]
  for (i in 2:length(plots)) {
    combined_plot <- combined_plot / plots[[i]]
  }

  # Add overall title and formatting
  final_plot <- combined_plot +
    plot_layout(
      heights = rep(1, length(plots)),
      guides = "collect"
    ) +
    plot_annotation(
      title = "Winter Energy Burdens in Massachusetts, by Scenario",
      subtitle = paste0("Among households below $", incomes_threshold),
      theme = theme(
        plot.title = element_text(hjust = 0.5, size = 9, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 8),
        plot.margin = margin(5, 5, 5, 5, "pt")
      )
    ) &
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_line(linewidth = 0.2, color = "gray90"),
      panel.grid.major.y = element_line(linewidth = 0.2, color = "gray90"),
      plot.margin = margin(1, 5, 1, 5, "pt")
    )

  return(list(final_plot, list_of_category_percentages))
}

# ---------
# Function Inputs

datasets <- list(
  "Current HVAC, 100% Discounts, Default Rates" = li_baseline_rates_current_hvac_100pct_participation,
  "Current HVAC, 0% Discounts, Default Rates" = li_baseline_rates_current_hvac_0pct_participation,
  "ccASHP, 0% Discounts, Default Rates" = li_baseline_undiscounted,
  "ccASHP, 100% Discounts, Default Rates" = li_baseline_discounted,
  "ccASHP, 100% Discounts, 1.0 Rates" = li_hpr_approved_discounted,
  "ccASHP, 100% Discounts, 2.0 Rates" = li_hpr_doer_discounted
)





# ---------
# Actually use the function
result_li_discount_histograms <- plot_li_discount_histograms(
  datasets,
  incomes_threshold = incomes_threshold,
  x_limits = c(0, 0.2),
  y_limits = c(0, 150000/242.13),
  binwidth = 0.01
)

# Extract results with different variable names
combined_plot <- result_li_discount_histograms[[1]]
list_of_category_percentages <- result_li_discount_histograms[[2]]

# Display the plot
print(combined_plot)

# Fix the rate_versions to match your actual datasets
rate_versions <- list(
  list(name = "baseline_undiscounted", result_index = 3, prefix = "baseline_undiscounted"),
  list(name = "baseline_discounted", result_index = 4, prefix = "baseline_discounted"),
  list(name = "approved", result_index = 5, prefix = "approved"),
  list(name = "doer", result_index = 6, prefix = "doer")
)

# Fix the category name and loop structure
for (i in seq_along(datasets)) {
  dataset_name <- names(datasets)[i]
  dataset_name <- tolower(gsub("[,]", "", dataset_name))
  dataset_name <- tolower(gsub("[%]", "pct", dataset_name))
  dataset_name <- tolower(gsub("__", "_", dataset_name))
  # Get the result for this dataset
  result <- list_of_category_percentages[[i]]

  # Use the correct category names from your function
  for (category in c("low_energy_burden", "moderate_energy_burden", "high_energy_burden", "very_high_energy_burden")) {
    pct_value <- result |>
      filter(bill_category == category) |>
      pull(percentage)

    if (length(pct_value) > 0) {
      assign(paste0("fig_hists_li_discount_histograms_", gsub(" ", "_", dataset_name), "_", category),
             round(pct_value, round_percent))
    }
  }
}

```

### data-pct_below_6_li_discount_variables
```{r pct_below_6_li_discount_variables}
pct_below_6_baseline_rates_current_hvac_100pct_participation <- fig_hists_li_discount_histograms_current_hvac_100pct_discounts_default_rates_low_energy_burden + fig_hists_li_discount_histograms_current_hvac_100pct_discounts_default_rates_moderate_energy_burden

pct_below_6_baseline_rates_current_hvac_0pct_participation <- fig_hists_li_discount_histograms_current_hvac_0pct_discounts_default_rates_low_energy_burden + fig_hists_li_discount_histograms_current_hvac_0pct_discounts_default_rates_moderate_energy_burden

pct_below_6_ccashp_100pct_undiscounted <- fig_hists_li_discount_histograms_ccashp_0pct_discounts_default_rates_low_energy_burden + fig_hists_li_discount_histograms_ccashp_0pct_discounts_default_rates_moderate_energy_burden

pct_below_6_ccashp_100pct_discounted <- fig_hists_li_discount_histograms_ccashp_100pct_discounts_default_rates_low_energy_burden + fig_hists_li_discount_histograms_ccashp_100pct_discounts_default_rates_moderate_energy_burden

pct_below_6_approved_rates_mid_hp_discounted <- fig_hists_li_discount_histograms_ccashp_100pct_discounts_1.0_rates_low_energy_burden + fig_hists_li_discount_histograms_ccashp_100pct_discounts_1.0_rates_moderate_energy_burden

pct_below_6_doer_rates_mid_hp_discounted <- fig_hists_li_discount_histograms_ccashp_100pct_discounts_2.0_rates_low_energy_burden + fig_hists_li_discount_histograms_ccashp_100pct_discounts_2.0_rates_moderate_energy_burden

```

### create_single_histogram_standalone()
```{r}
# Create helper function with better aspect ratio
create_single_histogram_standalone <- function(data, name, colors, x_limits = c(0, 0.2), y_limits = c(0, 150000/242.13), binwidth = 0.01, show_x_axis = TRUE, show_y_axis = TRUE) {
  low_burden_threshold <- 0.04
  moderate_burden_threshold <- 0.06
  high_burden_threshold <- 0.12

  # Calculate bill categories
  data <- data |>
    mutate(bill_category = case_when(
      energy_burden < low_burden_threshold ~ "low_energy_burden",
      energy_burden < moderate_burden_threshold ~ "moderate_energy_burden",
      energy_burden < high_burden_threshold ~ "high_energy_burden",
      TRUE ~ "very_high_energy_burden"
    ))

  # Calculate percentages for each category
  category_percentages <- data |>
    group_by(bill_category) |>
    summarise(count = n(), .groups = "drop") |>
    mutate(percentage = count / sum(count) * 100)

  # Get percentages for annotation
  pct_low_energy_burden <- category_percentages |>
    filter(bill_category == "low_energy_burden") |>
    pull(percentage) |>
    round(1)

  pct_moderate_energy_burden <- category_percentages |>
    filter(bill_category == "moderate_energy_burden") |>
    pull(percentage) |>
    round(1)

  pct_high_energy_burden <- category_percentages |>
    filter(bill_category == "high_energy_burden") |>
    pull(percentage) |>
    round(1)

  pct_very_high_energy_burden <- category_percentages |>
    filter(bill_category == "very_high_energy_burden") |>
    pull(percentage) |>
    round(1)

  # Calculate bin breaks and midpoints
  breaks <- seq(floor(x_limits[1]/binwidth)*binwidth, ceiling(x_limits[2]/binwidth)*binwidth, by = binwidth)

  # Manually bin the data
  binned_data <- data |>
    mutate(bin = cut(energy_burden,
                    breaks = breaks,
                    include.lowest = TRUE,
                    right = FALSE)) |>
    group_by(bin, bill_category) |>
    summarise(count = n(), .groups = "drop") |>
    mutate(bin_mid = as.numeric(as.character(
      sapply(bin, function(b) {
        vals <- as.numeric(gsub("\\(|\\]|\\[|\\)", "", strsplit(as.character(b), ",")[[1]]))
        return(mean(vals))
      })
    )))

  # More balanced color palette
  dark_red <- "#db8b87"
  light_red <- "#eaada9"
  light_green <- "#8ebd85"
  dark_green <- "#5e8a5e"

  # Position the bars at the top and text just below
  bar_y_pos <- y_limits[2]
  text_y_pos <- y_limits[2] * 0.90

  # Calculate x positions for percentage labels
  x_low_energy_burden <- mean(c(0, low_burden_threshold-0.002))
  x_moderate_energy_burden <- mean(c(low_burden_threshold, moderate_burden_threshold))
  x_high_energy_burden <- mean(c(moderate_burden_threshold, high_burden_threshold))
  x_very_high_energy_burden <- mean(c(high_burden_threshold, x_limits[2]))

  # Create the plot
  plot <- ggplot(binned_data, aes(x = bin_mid, y = count, fill = bill_category)) +
    # Add vertical dotted lines at the category boundaries
    geom_vline(xintercept = c(low_burden_threshold, moderate_burden_threshold, high_burden_threshold),
               linetype = "dotted", color = "gray50", size = 0.5) +
    # Add bars
    geom_col(position = "stack", width = binwidth * 0.9) +

    # Set the fill colors for the bars
    scale_fill_manual(
      values = c(
        "low_energy_burden" = dark_green,
        "moderate_energy_burden" = light_green,
        "high_energy_burden" = light_red,
        "very_high_energy_burden" = dark_red
      ),
      guide = "none"
    ) +
    # Add vertical lines at the category boundaries
    annotate("segment",
              x = x_limits[1], xend = low_burden_threshold,
              y = bar_y_pos, yend = bar_y_pos,
              color = dark_green, size = 1) +
    annotate("segment",
              x = low_burden_threshold, xend = moderate_burden_threshold,
              y = bar_y_pos, yend = bar_y_pos,
              color = light_green, size = 1) +
    annotate("segment",
              x = moderate_burden_threshold, xend = high_burden_threshold,
              y = bar_y_pos, yend = bar_y_pos,
              color = light_red, size = 1) +
    annotate("segment",
              x = high_burden_threshold, xend = x_limits[2],
              y = bar_y_pos, yend = bar_y_pos,
              color = dark_red, size = 1) +
    annotate("text",
              x = x_low_energy_burden,
              y = text_y_pos * 0.95,
              label = paste0(pct_low_energy_burden, "%"),
              size = 3,
              fontface = "bold") +
    annotate("text",
              x = x_moderate_energy_burden,
              y = text_y_pos * 0.95,
              label = paste0(pct_moderate_energy_burden, "%"),
              size = 3,
              fontface = "bold") +
    annotate("text",
              x = x_high_energy_burden,
              y = text_y_pos * 0.95,
              label = paste0(pct_high_energy_burden, "%"),
              size = 3,
              fontface = "bold") +
    annotate("text",
              x = x_very_high_energy_burden,
              y = text_y_pos * 0.95,
              label = paste0(pct_very_high_energy_burden, "%"),
              size = 3,
              fontface = "bold") +
    annotate("text",
              x = x_limits[2] * 0.95,
              y = y_limits[2] * 0.7,
              label = name,
              hjust = 1,
              fontface = "bold",
              size = 3,
              color = colors[name]) +
    scale_y_continuous(
      labels = function(x) paste0(round(x * 242.13 * 0.001), "k"),
      #name = "# of Homes",
      limits = y_limits,
      breaks = c(0, 50000, 100000, 150000)/242.13
    ) +
    scale_x_continuous(
      labels = scales::percent_format(accuracy = 1),
      name = "Winter Energy Burden",
      limits = x_limits,
      breaks = c(low_burden_threshold, moderate_burden_threshold, high_burden_threshold),
    ) +
    theme_minimal() +
    theme(
      panel.grid.major = element_line(linewidth = 0.2),
      panel.grid.minor = element_blank(),
      aspect.ratio = 0.18,  # CHANGED: Increased from 0.15 to match bill change histograms
      axis.title.x = element_text(size = 8)
    )

  # Remove x-axis elements if requested
  if (!show_x_axis) {
    plot <- plot +
      theme(
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()
      )
  }

  # Remove y-axis elements if requested
  if (!show_y_axis) {
    plot <- plot +
      theme(
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()
      )
  }

  return(plot)
}

```

### fig-li-burdens-current-hvac (fig13)
```{r}
#| label: fig-li-burdens-current-hvac
#| fig-cap: "Winter energy burdens for every MA household with annual incomes less than $50,000, under current HVAC systems, with and without 100% participation in low-income gas and electric discount programs. Assumes default delivery rates and 2024 supply rates for gas and electricity."


# Define colors
colors <- c(
  "Current HVAC, 100% Discounts, Default Rates" = "magenta",
  "Current HVAC, 0% Discounts, Default Rates" = "red",
  "ccASHP, 0% Discounts, Default Rates" = "grey50",
  "ccASHP, 100% Discounts, Default Rates" = "#023047",
  "ccASHP, 100% Discounts, 1.0 Rates" = "#FC9706",
  "ccASHP, 100% Discounts, 2.0 Rates" = "#68BED8"
)

# PLOT 1: Histograms 1 and 2 with tighter spacing
plot1_hist1 <- create_single_histogram_standalone(
  data = li_baseline_rates_current_hvac_100pct_participation,
  name = "Current HVAC, 100% Discounts, Default Rates",
  colors = colors,
  show_x_axis = TRUE,
  show_y_axis = FALSE
) + theme(plot.margin = margin(0, 5, 0, 5))  # Even tighter margins

plot1_hist2 <- create_single_histogram_standalone(
  data = li_baseline_rates_current_hvac_0pct_participation,
  name = "Current HVAC, 0% Discounts, Default Rates",
  colors = colors,
  show_x_axis = TRUE,
  show_y_axis = FALSE
) + theme(plot.margin = margin(0, 5, 0, 5))  # Even tighter margins

plot1_combined <- plot1_hist1 / plot1_hist2 +
  plot_layout(
    heights = c(1, 1)
  ) +
  plot_annotation(
    title = "Winter Energy Burdens in Massachusetts: Current HVAC",
    subtitle = paste0("Among households below $", incomes_threshold),
    theme = theme(
      plot.title = element_text(hjust = 0.5, size = 9, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 8),
      plot.margin = margin(10, 10, 10, 10)  # Much smaller overall margins
    )
  )


# Wrap the plot to add a shared y-axis label on the left
plot1_combined <- wrap_elements(plot1_combined) +
  labs(tag = "# of Homes") +
  theme(
    plot.tag = element_text(angle = 90, vjust = 0.5, hjust = 0.5, face = "plain", size = 9),
    plot.tag.position = "left"
  )

print(plot1_combined)
```

### fig-li-burdens-grid-comparison (fig14)
```{r}
#| label: fig-li-burdens-grid-comparison
#| fig-cap: "Winter energy burdens for every MA household with annual income less than $50,000, current HVAC vs. ccASHPS, with and without 100% participation in low-income gas and electric discount programs. Assumes default delivery rates and 2024 supply rates for gas and electricity."

# Define colors
colors <- c(
  "Current HVAC, 100% Discounts" = "magenta",
  "Current HVAC, 0% Discounts" = "red",
  "ccASHP, 0% Discounts" = "grey50",
  "ccASHP, 100% Discounts" = "#023047",
  "ccASHP, 100% Discounts" = "#FC9706",
  "ccASHP, 100% Discounts" = "#68BED8"
)


# PLOT 3: 2x2 Grid with better aspect ratio and tighter spacing
plot3_hist2 <- create_single_histogram_standalone(
  data = li_baseline_rates_current_hvac_0pct_participation,
  name = "Current HVAC, 0% Discounts",
  colors = colors,
  show_x_axis = FALSE,
  show_y_axis = TRUE  # Keep y-axis for left column
) + theme(
  aspect.ratio = 0.5,
  plot.margin = margin(2, 1, 1, 2),
  axis.title.y = element_blank()  # Remove individual y-axis title
)

plot3_hist3 <- create_single_histogram_standalone(
  data = li_baseline_undiscounted,
  name = "ccASHP, 0% Discounts",
  colors = colors,
  show_x_axis = TRUE,
  show_y_axis = TRUE  # Keep y-axis for left column
) + theme(
  aspect.ratio = 0.5,
  plot.margin = margin(1, 1, 2, 2),
  axis.title.y = element_blank()  # Remove individual y-axis title
)

plot3_hist1 <- create_single_histogram_standalone(
  data = li_baseline_rates_current_hvac_100pct_participation,
  name = "Current HVAC, 100% Discounts",
  colors = colors,
  show_x_axis = FALSE,
  show_y_axis = FALSE  # Remove y-axis for right column
) + theme(aspect.ratio = 0.5, plot.margin = margin(2, 2, 1, 1))

plot3_hist4 <- create_single_histogram_standalone(
  data = li_baseline_discounted,
  name = "ccASHP, 100% Discounts",
  colors = colors,
  show_x_axis = TRUE,
  show_y_axis = FALSE  # Remove y-axis for right column
) + theme(aspect.ratio = 0.5, plot.margin = margin(1, 2, 2, 1))

# Combine with patchwork
plot3_combined <- (plot3_hist2 | plot3_hist1) / (plot3_hist3 | plot3_hist4) +
  plot_layout(heights = c(1, 1), widths = c(1, 1)) +
  plot_annotation(
    title = "Winter Energy Burdens: Comparison of HVAC and Discount Scenarios",
    subtitle = paste0("Among households below $", incomes_threshold, "| Default Rates"),
    theme = theme(
      plot.title = element_text(hjust = 0.5, size = 9, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 8),
      plot.margin = margin(10, 10, 10, 30)  # Extra left margin for y-axis label
    )
  )

# Wrap the plot to add a shared y-axis label on the left
plot3_combined <- wrap_elements(plot3_combined) +
  labs(tag = "# of Homes") +
  theme(
    plot.tag = element_text(angle = 90, vjust = 0.5, hjust = 0.5, face = "plain", size = 9),
    plot.tag.position = "left"
  )

print(plot3_combined)
```

### fig-li-burdens-hp-rates (fig15)
```{r}
#| label: fig-li-burdens-hp-rates
#| fig-cap: "Winter energy burdens after switching to ccASHPS, for every MA household with annual income less than $50,000, under default vs. 1.0 vs. 2.0 heat pump electric delivery rates combined with 100% participation in low-income gas and electric discount programs. Assumes default gas delivery rates and 2024 supply rates for gas and electricity."

# PLOT 15: Histograms 4, 5, 6 with tighter spacing
plot2_hist4 <- create_single_histogram_standalone(
  data = li_baseline_discounted,
  name = "ccASHP, 100% Discounts, Default Rates",
  colors = colors,
  show_x_axis = FALSE,
  show_y_axis = FALSE
) + theme(plot.margin = margin(5, 5, 0, 5))  # Reduced margins

plot2_hist5 <- create_single_histogram_standalone(
  data = li_hpr_approved_discounted,
  name = "ccASHP, 100% Discounts, 1.0 Rates",
  colors = colors,
  show_x_axis = FALSE,
  show_y_axis = FALSE
) + theme(plot.margin = margin(0, 5, 0, 5))  # Reduced margins

plot2_hist6 <- create_single_histogram_standalone(
  data = li_hpr_doer_discounted,
  name = "ccASHP, 100% Discounts, 2.0 Rates",
  colors = colors,
  show_x_axis = TRUE,
  show_y_axis = FALSE
) + theme(plot.margin = margin(0, 5, 5, 5))  # Reduced margins

plot2_combined <- plot2_hist4 / plot2_hist5 / plot2_hist6 +
  plot_layout(heights = c(1, 1, 1)) +
  plot_annotation(
    title = "Winter Energy Burdens: Heat Pump Scenarios with Different Rates",
    subtitle = paste0("Among households below $", incomes_threshold, " with 100% discount participation"),
    theme = theme(
      plot.title = element_text(hjust = 0.5, size = 9, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 8),
      plot.margin = margin(10, 10, 10, 10)  # Reduced overall margins
    )
  )

# Wrap the plot to add a shared y-axis label on the left
plot2_combined <- wrap_elements(plot2_combined) +
  labs(tag = "# of Homes") +
  theme(
    plot.tag = element_text(angle = 90, vjust = 0.5, hjust = 0.5, face = "plain", size = 9),
    plot.tag.position = "left"
  )

print(plot2_combined)
```


## ---------- How do savings vary by utility? ----------
### fig-stacked-bars-utility-rate-scenarios (fig16)
```{r}
#| label: fig-stacked-bars-utility-rate-scenarios
#| fig-cap: "Proportion of MA homes served by each investor-owned electric utility that would see a large or small bill savings or increase in total winter energy bills when switching to heat pumps, under default vs. 1.0 vs. 2.0 heat pump electric delivery rates. Assumes default gas delivery rates, 2024 supply rates for gas and electricity, and 100% participation in low-income programs by eligible households."

# Create a function to categorize bill changes
categorize_bill_changes_by_utility <- function(data) {
  data %>%
    filter(bill_type == "annual_bill_change_energy",
           hvac == hvac_for_utility_bills,
           version_elec %in% c("baseline", "hpr_approved", "hpr_doer")) %>%
    mutate(category = case_when(
      annual_bill_change <= -1000 ~ "large_savings",
      annual_bill_change < 0 ~ "small_savings",
      annual_bill_change < 1000 ~ "small_increase",
      TRUE ~ "large_increase"
    )) %>%
    group_by(electric_utility, version_elec, category) %>%
    summarise(count = n(), .groups = "drop") %>%
    group_by(electric_utility, version_elec) %>%
    mutate(
      total = sum(count),
      percentage = count / total * 100,
      utility_scenario = paste(electric_utility, version_elec, sep = "_")
    ) %>%
    ungroup()
}

# Apply the function to the data
utility_scenario_counts <- categorize_bill_changes_by_utility(annual_change_with_municipal)

# Set factor levels for correct ordering
utility_scenario_counts <- utility_scenario_counts %>%
  mutate(
    # FIXED: Order for stacking - this controls the visual stacking order
    # For negative values, the order is visually reversed, so we put large_savings before small_savings
    category = factor(
      category,
      levels = c("large_increase", "small_increase", "large_savings", "small_savings")
    ),
    electric_utility = factor(
      electric_utility,
      levels = c("eversource", "nationalgrid", "unitil", "municipal")
    ),
    version_elec = factor(
      version_elec,
      levels = c("baseline", "hpr_approved", "hpr_doer")
    ),
    utility_scenario = factor(
      utility_scenario,
      levels = paste(rep(c("eversource", "nationalgrid", "unitil", "municipal"), each = 3),
                    rep(c("baseline", "hpr_approved", "hpr_doer"), 4),
                    sep = "_")
    )
  )

# Create the stacked bar chart
ggplot(utility_scenario_counts, aes(x = utility_scenario, y = percentage, fill = category)) +
  # Reference line at 0 (boundary between savings and increases)
  geom_hline(yintercept = 0, linetype = "solid", color = "gray40", linewidth = 0.5) +
  # For savings categories (plotted downward)
  geom_col(
    data = utility_scenario_counts %>% filter(category %in% c("large_savings", "small_savings")),
    aes(y = -percentage),
    position = "stack"
  ) +
  # For increase categories (plotted upward)
  geom_col(
    data = utility_scenario_counts %>% filter(category %in% c("small_increase", "large_increase")),
    position = "stack"
  ) +
  # Add percentage labels for savings (negative values)
  geom_text(
    data = utility_scenario_counts %>% filter(category %in% c("large_savings", "small_savings")),
    aes(y = -percentage, label = paste0(round(percentage, 1), "%")),
    position = position_stack(vjust = 0.5),
    color = "white",
    fontface = "bold",
    size = 2.5
  ) +
  # Add percentage labels for increases (positive values)
  geom_text(
    data = utility_scenario_counts %>% filter(category %in% c("small_increase", "large_increase")),
    aes(label = paste0(round(percentage, 1), "%")),
    position = position_stack(vjust = 0.5),
    color = "white",
    fontface = "bold",
    size = 2.5
  ) +
  # FIXED: Swapped the colors for increase categories
  scale_fill_manual(
    values = c(
      "large_savings" = "#5e8a5e",      # dark green
      "small_savings" = "#8ebd85",      # light green
      "small_increase" = "#eaada9",     # FIXED: now light red (was dark red)
      "large_increase" = "#db8b87"      # FIXED: now dark red (was light red)
    ),
    labels = c(
      "large_savings" = ">$1,000 savings",
      "small_savings" = "$0-$1,000 savings",
      "small_increase" = "$0-$1,000 increase",
      "large_increase" = ">$1,000 increase"
    ),
    breaks = c("large_increase", "small_increase", "small_savings", "large_savings")
  ) +
  # Add utility grouping
  scale_y_continuous(
    labels = NULL,
  ) +
  scale_x_discrete(
    labels = rep(c("Default", "1.0", "2.0 rates"), 4),
    guide = guide_axis(n.dodge = 1)
  ) +
  # Add utility labels with faceting
  facet_wrap(
    ~ factor(electric_utility,
            levels = c("eversource", "nationalgrid", "unitil", "municipal"),
            labels = c("Eversource", "National Grid", "Unitil", "Municipal")),
    nrow = 1,
    scales = "free_x"
  ) +
  labs(
    title = "Change in Total Winter Energy Bills when Switching to Heat Pump, by Utility",
    subtitle = glue::glue("{supply_year} Supply Prices"),
    x = NULL,
    y = "",
    fill = ""
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 10, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 8),
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_blank(),
    strip.text = element_text(face = "bold")
  )

# Create and export report variables for each utility, scenario, and category combination
for (utility in c("eversource", "nationalgrid", "unitil", "municipal")) {
  for (scenario in c("baseline", "hpr_approved", "hpr_doer")) {
    for (category in c("large_savings", "small_savings", "large_increase", "small_increase")) {
      # Get the percentage for this specific combination
      pct_value <- utility_scenario_counts |>
        filter(electric_utility == utility & version_elec == scenario & category == !!category) |>
        pull(percentage)

      # Handle cases where the category might not exist (assign 0)
      if (length(pct_value) == 0) {
        pct_value <- 0
      }

      # Create variable name like: utility_scenario_counts_eversource_baseline_large_savings
      var_name <- paste("utility_change_", utility, "_", scenario, "_", category, sep = "")
      assign(var_name, round(pct_value, round_percent))
    }
  }
}

```

### fig-boxplot-utility-bills (fig17)
```{r}
#| label: fig-boxplot-utility-bills
#| fig-cap: "Distribution of total winter energy bills, for all MA homes served by an investor-owned electric utility, before and after switching to heat pumps, under default vs. 1.0 vs. 2.0 heat pump electric delivery rates, by utility. Assumes default gas delivery rates, 2024 supply rates for gas and electricity, and 100% participation in low-income programs by eligible households."

# Prepare the data for boxplots (exclude municipal)
utility_bills_boxplot <- annual_bills_with_municipal |>
  filter(
    bill_type == "annual_bill_energy",
    raw_or_discounted == "discounted",
    electric_utility %in% c("eversource", "nationalgrid", "unitil")  # Remove municipal
  ) |>
  mutate(scenario = case_when(
      hvac == 'current' & version_elec == "baseline" ~ "baseline_rates_current_hvac",
      hvac == hvac_for_utility_bills & version_elec == "baseline" ~ "baseline",
      hvac == hvac_for_utility_bills & version_elec == "hpr_approved" ~ "hpr_approved",
      hvac == hvac_for_utility_bills & version_elec == "hpr_doer" ~ "hpr_doer",
      TRUE ~ NA_character_
  )) |>
  filter(!is.na(scenario)) |>
  mutate(
    scenario = factor(
      scenario,
      levels = c(
        "hpr_doer",
        "hpr_approved",
        "baseline",
        "baseline_rates_current_hvac"
      )
    ),
    electric_utility = factor(electric_utility,
                            levels = c("eversource", "nationalgrid", "unitil"))
  )

# Loop through versions to calculate statistics
for (eu in c("eversource", "nationalgrid", "unitil")) {
  for (s in c("baseline_rates_current_hvac", "baseline", "hpr_approved", "hpr_doer")) {
    # Get the stats for this version
    stats <- utility_bills_boxplot |>
      filter(electric_utility == eu) |>
      filter(scenario == s) |>
      summarise(
        p25 = quantile(annual_bill, 0.25),
        p50 = quantile(annual_bill, 0.50),
        p75 = quantile(annual_bill, 0.75)
      )
    assign(paste0("utility_bills_", eu, "_", s, "_p25"),
           round(stats |> pull(p25), round_dollars))
    assign(paste0("utility_bills_", eu, "_", s, "_p50"),
           round(stats |> pull(p50), round_dollars))
    assign(paste0("utility_bills_", eu, "_", s, "_p75"),
           round(stats |> pull(p75), round_dollars))
  }
}

# Create the boxplot with flipped axes and no outliers
ggplot(utility_bills_boxplot, aes(y = electric_utility, x = annual_bill, fill = scenario)) +
  geom_boxplot(
    position = position_dodge(width = 0.85),
    alpha = 0.8,
    outlier.shape = NA  # Remove outliers but keep whiskers
  ) +
  scale_fill_manual(
    values = c(
      "baseline_rates_current_hvac" = "grey50",
      "baseline" = "#023047",
      "hpr_approved" = "#FC9706",
      "hpr_doer" = "#68BED8"
    ),
    labels = c(
      "baseline_rates_current_hvac" = "Baseline Rates, Current HVAC",
      "baseline" = "Baseline Rates, Mid HP",
      "hpr_approved" = "Heat Pump, 1.0 HP Rates",
      "hpr_doer" = "Heat Pump, 2.0 HP Rates"
    )
  ) +
  guides(fill = guide_legend(reverse = TRUE)) +
  scale_y_discrete(
    labels = c(
      "eversource" = "Eversource",
      "nationalgrid" = "National Grid",
      "unitil" = "Unitil"
    ),
    expand = expansion(mult = c(0.3, 0.3))  # Add space at top for labels
  ) +
  scale_x_continuous(
    labels = scales::dollar_format(),
    expand = expansion(mult = c(0.0, 0.05)),
    limits = c(0, 8000)
  ) +
  labs(
    title = "Distribution of Winter Energy Bills by Utility and Scenario",
    subtitle = "Vertical lines represent 25th, 50th, and 75th percentiles; outliers removed",
    y = NULL,
    x = "Annual Energy Bill",
    fill = "Scenario"
  ) +
    # Add colored annotations instead of legend
  annotate("text", x = 5600, y = 3.39, label = "Current HVAC, Default Rates",
           color = "grey50", fontface = "bold", size = 3.5, hjust = 0) +
  annotate("text", x = 5500, y = 3.19, label = "Heat Pump, Default Rates",
           color = "#023047", fontface = "bold", size = 3.5, hjust = 0) +
  annotate("text", x = 5400, y = 2.99, label = "Heat Pump, 1.0 HP Rates",
           color = "#FC9706", fontface = "bold", size = 3.5, hjust = 0) +
  annotate("text", x = 5300, y = 2.78, label = "Heat Pump, 2.0 HP Rates",
           color = "#68BED8", fontface = "bold", size = 3.5, hjust = 0) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 10, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 8, color = "gray60"),
    panel.grid.major.y = element_blank(),
    aspect.ratio = 0.5,
    plot.margin = margin(t = 0.05, r = 0.1, b = 0.05, l = 0.1),
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    axis.text.x = element_text(size = 8),
    axis.text.y = element_text(size = 8),
    legend.position = "none"  # Remove the legend
  )

```

### fig-boxplot-utility-bills-munis (fig18)
```{r}
#| label: fig-boxplot-utility-bills-munis
#| fig-cap: "Distribution of total winter energy bills after switching to heat pumps, for all MA homes served by an investor-owned electric utility and municipal light plants, under default vs. 2.0 heat pump electric delivery rates, by IOUs vs. MLPS. Municipal supply and delivery rates are averaged across all MLPs. Assumes default IOU gas delivery rates, 2024 IOU supply rates for gas and electricity, and 100% participation in IOU low-income programs by eligible households."
#| fig-width: 6
#| fig-height: 4
#|
# Prepare the data for boxplots (exclude municipal)
utility_bills_boxplot <- annual_bills_with_municipal |>
  filter(
    bill_type == "annual_bill_energy",
    raw_or_discounted == "discounted",
    electric_utility %in% c("municipal", "eversource", "nationalgrid", "unitil")
  ) |>
  mutate(scenario = case_when(
      hvac == 'current' & version_elec == "baseline" ~ "baseline_rates_current_hvac",
      hvac == hvac_for_utility_bills & version_elec == "baseline" ~ "baseline",
      hvac == hvac_for_utility_bills & version_elec == "hpr_approved" ~ "hpr_approved",
      hvac == hvac_for_utility_bills & version_elec == "hpr_doer" ~ "hpr_doer",
      TRUE ~ NA_character_
  )) |>
  filter(!is.na(scenario)) |>
  filter(!(scenario %in% c("baseline_rates_current_hvac", "hpr_approved"))) |>
  mutate(
    annual_bill = ifelse(electric_utility == "municipal" & scenario == "hpr_doer", 0, annual_bill),
    scenario = factor(
      scenario,
      levels = c(
        "hpr_doer",
        "baseline"
      )
    ),
    electric_utility = factor(electric_utility,
                            levels = c("municipal", "eversource", "nationalgrid", "unitil"))
  )

# Create the boxplot with flipped axes and no outliers
ggplot(utility_bills_boxplot, aes(y = electric_utility, x = annual_bill, fill = scenario)) +
  geom_boxplot(
    position = position_dodge(width = 0.85),
    alpha = 0.8,
    outlier.shape = NA  # Remove outliers but keep whiskers
  ) +
  scale_fill_manual(
    values = c(
      "baseline" = "#023047",
      "hpr_doer" = "#68BED8"
    ),
    labels = c(
      "baseline" = "Heat Pump, Default Rates",
      "hpr_doer" = "Heat Pump, 2.0 HP Rates"
    )
  ) +
  scale_y_discrete(
    labels = c(
      "municipal" = "Municipal Light Plants",
      "eversource" = "Eversource",
      "nationalgrid" = "National Grid",
      "unitil" = "Unitil"
    ),
    expand = expansion(mult = c(0.2, 0.3))  # Add space at top for labels
  ) +
  scale_x_continuous(
    labels = scales::dollar_format(),
    expand = expansion(mult = c(0.0, 0.05)),
    limits = c(0, 8000)
  ) +
  labs(
    title = "Distribution of Winter Energy Bills, by Utility ",
    subtitle = "Vertical lines represent 25th, 50th, and 75th percentiles; outliers removed",
    y = NULL,
    x = "Annual Energy Bill",
    fill = "Scenario"
  ) +
  # Add colored annotations instead of legend
  annotate("text", x = 5300, y = 4.06, label = "Heat Pump, Default Rates",
           color = "#023047", fontface = "bold", size = 3.5, hjust = 0) +
  annotate("text", x = 4200, y = 3.66, label = "Heat Pump, 2.0 HP Rates",
           color = "#68BED8", fontface = "bold", size = 3.5, hjust = 0) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 10, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 8, color = "gray60"),
    panel.grid.major.y = element_blank(),
    aspect.ratio = 0.5,
    plot.margin = margin(t = 0.05, r = 0.1, b = 0.05, l = 0.1),
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    axis.text.x = element_text(size = 8),
    axis.text.y = element_text(size = 8),
    legend.position = "none"  # Remove the legend
  )
```


### fig-bldg-type-by-utility
```{r}
#| label: fig-bldg-type-by-utility
#| fig-cap: "Building Type Distribution by Utility"

annual_change_bldg_type <- annual_change |>
  mutate(
    building_type_group = case_when(
      building_type %in% c("Single-Family Detached", "Mobile Home", "Single-Family Attached") ~ "Single-Family",
      building_type %in% c("2 Unit", "3 or 4 Unit") ~ "2-4 Units",
      building_type %in% c("5 to 9 Unit", "10 to 19 Unit", "20 to 49 Unit", "50 or more Unit") ~ "5+ Units",
      TRUE ~ "Other"
    ),
  )

# Create box plots by building type and version_elec
building_type_group_order = c("Single-Family", "2-4 Units", "5+ Units")

building_type_group_names = c("Single-Family", "2-4 Units", "5+ Units")



# Calculate proportions of building types by utility
building_type_by_utility <- annual_change_bldg_type |>
  filter(
    building_type_group %in% building_type_group_order,
    electric_utility != "municipal"
  ) |>
  group_by(electric_utility, building_type_group) |>
  summarise(n = n_distinct(bldg_id)) |>
  group_by(electric_utility) |>
  mutate(
    pct = n / sum(n),
    building_type_group = factor(building_type_group, levels = building_type_group_order)
  ) |>
  ungroup()

# Create stacked bar plot
ggplot(building_type_by_utility, aes(x = electric_utility, y = pct, fill = building_type_group)) +
  geom_col(position = "stack", width = 0.7) +
  geom_text(
    aes(label = scales::percent(pct, accuracy = 1)),
    position = position_stack(vjust = 0.5),
    color = "white",
    fontface = "bold"
  ) +
  scale_y_continuous(
    labels = scales::percent,
    expand = expansion(mult = c(0, 0.05))
  ) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Building Type Distribution by Utility",
    x = NULL,
    y = "Percentage of Buildings",
    fill = "Building Type"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 0),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )

```

### fig-heating-type-by-utility
```{r}
#| label: fig-heating-type-by-utility
#| fig-cap: "Heating Type Distribution by Utility"

# Create box plots by building type and version_elec
baseline_heating_type_order = c("None", "Other Fuel", "Heat Pump", "Propane", "Resistance", "Fuel Oil", "Natural Gas")

baseline_heating_type_names = c("None", "Other Fuel", "Heat Pump", "Propane", "Electric Resistance", "Fuel Oil", "Natural Gas")


# Choose a color palette for building types
library(RColorBrewer)
heating_type_palette <- rev(brewer.pal(n = length(baseline_heating_type_order), name = "Dark2"))


# Calculate proportions of heating types by utility
heating_type_by_utility <- annual_change_bldg_type |>
  filter(electric_utility != "municipal") |>
  filter(baseline_heating_type != "None") |>
  group_by(electric_utility, baseline_heating_type) |>
  summarise(n = n_distinct(bldg_id)) |>
  group_by(electric_utility) |>
  mutate(
    pct = n / sum(n),
    baseline_heating_type = factor(baseline_heating_type, levels = (baseline_heating_type_order))
  ) |>
  ungroup()

# Get Unitil data for label positioning
unitil_data <- heating_type_by_utility |>
  filter(electric_utility == "unitil") |>
  arrange(baseline_heating_type) |>
  mutate(
    cumsum_pct = cumsum(pct),
    label_y = 1- (cumsum_pct - pct/2)
  )


# Create stacked bar plot
ggplot(heating_type_by_utility, aes(x = electric_utility, y = pct, fill = baseline_heating_type)) +
  geom_col(position = "stack", width = 0.7) +
  geom_text(
    aes(label = scales::percent(pct, accuracy = 1)),
    position = position_stack(vjust = 0.5),
    color = "white",
    fontface = "bold"
  ) +
  # Add text annotations aligned with Unitil bars
  geom_text(
    data = unitil_data,
    aes(
      x = 'unitil',
      y = label_y,
      label = baseline_heating_type,
      color = baseline_heating_type
    ),
    hjust = 0,
    position = position_nudge(x = 0.39)  # Move labels right by 0.5 * column width (0.7)
  ) +
  scale_y_continuous(
    labels = scales::percent,
    expand = expansion(mult = c(0, 0.05))
  ) +
  scale_fill_manual(values = heating_type_palette) +
  scale_color_manual(values = heating_type_palette) +
  labs(
    title = "Heating Type Distribution by Utility",
    x = NULL,
    y = "Percentage of Buildings"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text(angle = 0),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.margin = margin(10, 120, 10, 10)
  ) +
  coord_cartesian(clip = "off")



```




# ------------------------------------------------------------------------------
# Appendx
# ------------------------------------------------------------------------------


## - - - -  MLP Subsection - - - -
### Load MLP Delivery Tariffs
```{r}
#| label: load-mlp-delivery-tariffs
path_mlp_delivery_tariffs <- "/workspaces/reports/data/utility_rates/electric/madpu/ma_mlp_delivery_tariffs.rds"
mlp_delivery_tariffs <- readRDS(path_mlp_delivery_tariffs)
```

### fig-mlp-charges
```{r}
#| label: fig-mlp-charges
#| fig-cap: "Distribution of customer and delivery charges for MA's 41 MLPs. The average customer charge increased 15% from 2019 to 2024, less than inflation in the same time period. The average delivery charge increase 26% from 2019 to 2024, on pace with inflation."

# Customer Charge
mlp_customer_charge_median_2019 <- mlp_delivery_tariffs |>
  filter(Year == 2019, charge_type == "customer_charge") |>
  summarise(median = median(Value, na.rm = TRUE)) |>
  pull(median)

mlp_customer_charge_median_2024 <- mlp_delivery_tariffs |>
  filter(Year == 2024, charge_type == "customer_charge") |>
  summarise(median = median(Value, na.rm = TRUE)) |>
  pull(median)

mlp_customer_charge_increase_pct <- (mlp_customer_charge_median_2024 - mlp_customer_charge_median_2019) / mlp_customer_charge_median_2019

customer_charge_by_year <- mlp_delivery_tariffs |>
  filter(charge_type == "customer_charge") |>
  filter(Year %in% c(2019, 2024)) |>
  mutate(Year = as.factor(Year))

# Delivery Charge
mlp_delivery_charge_median_2019 <- mlp_delivery_tariffs |>
  filter(Year == 2019, charge_type == "delivery_rate") |>
  summarise(median = median(Value, na.rm = TRUE)) |>
  pull(median)

mlp_delivery_charge_median_2024 <- mlp_delivery_tariffs |>
  filter(Year == 2024, charge_type == "delivery_rate") |>
  summarise(median = median(Value, na.rm = TRUE)) |>
  pull(median)

mlp_delivery_charge_increase_pct <- (mlp_delivery_charge_median_2024 - mlp_delivery_charge_median_2019) / mlp_delivery_charge_median_2019

delivery_charge_by_year <- mlp_delivery_tariffs |>
  filter(charge_type == "delivery_rate") |>
  filter(Year %in% c(2019, 2024)) |>
  mutate(Year = as.factor(Year))


# Make sure patchwork is loaded for plot arrangement
library(patchwork)
library(ggridges)

# Create the customer charge plot (remove legend)
colors <- c("2019" = "#FFC729", "2024" = "#A0AF12")
customer_plot <- customer_charge_by_year |>
  ggplot(aes(x = Value, y = fct_rev(Year), fill = Year)) +
  geom_density_ridges(
    alpha = 0.7,
    rel_min_height = 0.01,
    scale = 1,
    color = NA
  ) +
  stat_density_ridges(
    quantile_lines = TRUE,
    quantiles = 2,
    alpha = 0,
    color = "black",
    linetype = "dashed",
    scale = 1
  ) +
  geom_text(
    data = customer_charge_by_year |>
      group_by(Year) |>
      summarise(
        median = median(Value, na.rm = TRUE),
        y = as.numeric(fct_rev(Year)) + 0.03
      ),
    aes(
      x = median+0.2,
      y = y,
      label = paste0(Year, ": $", round(median, 2))
    ),
    vjust = 0,
    hjust = 0,
    size = 2.5
  ) +
  scale_x_continuous(
    limits = c(0, 15),
    labels = scales::label_dollar()
  ) +
  scale_y_discrete(expand = expansion(mult = c(0, 0.1))) +
  scale_fill_manual(values = colors) +
  labs(
    title = NULL,
    x = "Customer Charge ($)",
    y = "Year"
  ) +
  theme_minimal() +
  theme(
    aspect.ratio = 0.85,  # Set aspect ratio for the individual plot
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, size = 10),
    plot.margin = margin(10, 10, 10, 10),  #
    axis.text.x = element_text(size = 8)

  )  # Remove legend from this plot

# Create the delivery charge plot (keep legend)
delivery_plot <- delivery_charge_by_year |>
  ggplot(aes(x = Value, y = fct_rev(Year), fill = Year)) +
  geom_density_ridges(
    alpha = 0.7,
    rel_min_height = 0.01,
    scale = 1,
    color = NA
  ) +
  stat_density_ridges(
    quantile_lines = TRUE,
    quantiles = 2,
    alpha = 0,
    color = "black",
    linetype = "dashed",
    scale = 1
  ) +
  geom_text(
    data = delivery_charge_by_year |>
      group_by(Year) |>
      summarise(
        median = median(Value, na.rm = TRUE),
        y = as.numeric(fct_rev(Year)) + 0.03
      ),
    aes(
      x = median+0.005,
      y = y,
      label = paste0(Year, ": $", round(median, 3))
    ),
    vjust = 0,
    hjust = 0,
    size = 2.5
  ) +
  scale_x_continuous(
    limits = c(0.05, 0.25),
    labels = scales::label_dollar(),
    expand = expansion(mult = c(0, 0.05))
  ) +
  scale_y_discrete(expand = expansion(mult = c(0, 0.1))) +
  scale_fill_manual(values = colors) +
  labs(
    title = NULL,
    x = "Delivery Charge ($)",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    aspect.ratio = 0.85,  # Set aspect ratio for the individual plot
    plot.title = element_text(hjust = 0.5, size = 10),
    plot.margin = margin(10, 10, 10, 25),  #
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.text.x = element_text(size = 8)
  )

# Create a 2x1 grid with the plots and shared legend at the bottom
combined_plot <- (customer_plot + delivery_plot) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title = "Distribution of MLP Customer and Delivery Charges, by Year",
    theme = theme(plot.title = element_text(hjust = 0.5, size = 12))
  ) +
  theme(
    legend.position = "none",
   plot.margin = margin(2, 2, 2, 2)
  )

# Display the combined plot
combined_plot
```

### fig-hists-edc-mlp-burden
```{r}
#| label: fig-hists-edc-mlp-burden
#| fig-cap: "fig-hists-edc-mlp-burden"

plot_edc_mlp_burden_histograms <- function(
  datasets,
  hvac_for_utility_bills,
  x_limits = c(0, 0.2),
  y_limits = NULL,
  binwidth = 0.01
) {
  # Define color palette for each dataset
  colors <- c(
    "EDC Baseline" = "grey50",
    "EDC Approved" = "#d13b40",
    "EDC DOER" = "#023047",
    "MLP Baseline" = "#FC9706"
  )

  # # Calculate y_max across all datasets if not provided
  # if (is.null(y_limits)) {
  #   y_max <- 10
  #   for (dataset_name in names(datasets)) {
  #     hist_plot <- ggplot(datasets[[dataset_name]], aes(x = energy_burden)) +
  #       geom_histogram(binwidth = binwidth)

  #     current_max <- max(ggplot_build(hist_plot)$data[[1]]$count, na.rm = TRUE)
  #     y_max <- max(y_max, current_max)
  #   }

  #   y_max <- ceiling(y_max * 1.1)
  #
  # }



  # Function to create a single histogram for a dataset
  create_single_histogram <- function(data, name, is_first = FALSE, is_last = FALSE) {
    low_burden_threshold <- 0.04
    moderate_burden_threshold <- 0.06
    high_burden_threshold <- 0.12

    # Calculate bill categories based on energy burden
    data <- data |>
      mutate(bill_category = case_when(
        energy_burden < low_burden_threshold ~ "low_energy_burden",
        energy_burden < moderate_burden_threshold ~ "moderate_energy_burden",
        energy_burden < high_burden_threshold ~ "high_energy_burden",
        TRUE ~ "very_high_energy_burden"
      ))

    # Calculate percentages for each category
    category_percentages <- data |>
      group_by(bill_category) |>
      summarise(count = n(), .groups = "drop") |>
      mutate(percentage = count / sum(count) * 100)

    # Get percentages for annotation
    pct_low_energy_burden <- category_percentages |>
      filter(bill_category == "low_energy_burden") |>
      pull(percentage) |>
      round(1)

    pct_moderate_energy_burden <- category_percentages |>
      filter(bill_category == "moderate_energy_burden") |>
      pull(percentage) |>
      round(1)

    pct_high_energy_burden <- category_percentages |>
      filter(bill_category == "high_energy_burden") |>
      pull(percentage) |>
      round(1)

    pct_very_high_energy_burden <- category_percentages |>
      filter(bill_category == "very_high_energy_burden") |>
      pull(percentage) |>
      round(1)

    # Calculate bin breaks and midpoints
    breaks <- seq(floor(x_limits[1]/binwidth)*binwidth, ceiling(x_limits[2]/binwidth)*binwidth, by = binwidth)

    # Manually bin the data
    binned_data <- data |>
      mutate(bin = cut(energy_burden,
                      breaks = breaks,
                      include.lowest = TRUE,
                      right = FALSE)) |>
      group_by(bin, bill_category) |>
      summarise(count = n(), .groups = "drop") |>
      # Add bin midpoint for plotting
      mutate(bin_mid = as.numeric(as.character(
        sapply(bin, function(b) {
          # Extract bin boundaries and calculate midpoint
          vals <- as.numeric(gsub("\\(|\\]|\\[|\\)", "", strsplit(as.character(b), ",")[[1]]))
          return(mean(vals))
        })
      )))

    # Color palette for energy burden categories
    dark_red <- "#db8b87"        # dark red
    light_red <- "#eaada9"       # light red
    light_green <- "#8ebd85"     # light green
    dark_green <- "#5e8a5e"      # dark green

    # Position the bars at the top and text just below
    bar_y_pos <- y_limits[2]
    text_y_pos <- y_limits[2] * 0.90

    # Calculate x positions for percentage labels
    x_low_energy_burden <- mean(c(0, low_burden_threshold-0.002))
    x_moderate_energy_burden <- mean(c(low_burden_threshold, moderate_burden_threshold))
    x_high_energy_burden <- mean(c(moderate_burden_threshold, high_burden_threshold))
    x_very_high_energy_burden <- mean(c(high_burden_threshold, x_limits[2]))


    # # Calculate y_max dynamically based on highest bin count
    # y_max <- binned_data |>
    #   group_by(bin_mid) |>
    #   summarise(total_count = sum(count)) |>
    #   pull(total_count) |>
    #   max() |>
    #   {function(x) ceiling(x/(25000/242.13))*(25000/242.13)}()

    # # Update y_limits with new y_max
    # y_limits <- c(0, y_max)

    # Set y_max based on plot number
    if (name == "MLP Baseline") {
      y_max <- 120000/242.13
      y_tick_step <- 50000
    } else {
      y_max <- 850000/242.13
      y_tick_step <- 250000
    }

    y_limits <- c(0, y_max)

    # Recalculate bar and text positions with new y_max
    bar_y_pos <- y_max
    text_y_pos <- y_max * 0.90

    # Create the plot
    plot <- ggplot(binned_data, aes(x = bin_mid, y = count, fill = bill_category)) +
      # Add vertical dotted lines at the category boundaries
      geom_vline(xintercept = c(low_burden_threshold, moderate_burden_threshold, high_burden_threshold), linetype = "dotted", color = "gray50", size = 0.5) +

      # Add the bars
      geom_col(position = "stack", width = binwidth * 0.9) +

      # Set the fill colors
      scale_fill_manual(
        values = c(
          "low_energy_burden" = dark_green,
          "moderate_energy_burden" = light_green,
          "high_energy_burden" = light_red,
          "very_high_energy_burden" = dark_red
        ),
        guide = "none"
      ) +

      # Add horizontal colored lines at the top
      annotate("segment",
                x = x_limits[1], xend = low_burden_threshold,
                y = bar_y_pos, yend = bar_y_pos,
                color = dark_green, size = 1) +

      annotate("segment",
                x = low_burden_threshold, xend = moderate_burden_threshold,
                y = bar_y_pos, yend = bar_y_pos,
                color = light_green, size = 1) +

      annotate("segment",
                x = moderate_burden_threshold, xend = high_burden_threshold,
                y = bar_y_pos, yend = bar_y_pos,
                color = light_red, size = 1) +

      annotate("segment",
                x = high_burden_threshold, xend = x_limits[2],
                y = bar_y_pos, yend = bar_y_pos,
                color = dark_red, size = 1) +

      # Add percentage labels
      annotate("text",
                x = x_low_energy_burden,
                y = text_y_pos,
                label = paste0(pct_low_energy_burden, "%"),
                size = 3,
                fontface = "bold") +

      annotate("text",
                x = x_moderate_energy_burden,
                y = text_y_pos,
                label = paste0(pct_moderate_energy_burden, "%"),
                size = 3,
                fontface = "bold") +

      annotate("text",
                x = x_high_energy_burden,
                y = text_y_pos,
                label = paste0(pct_high_energy_burden, "%"),
                size = 3,
                fontface = "bold") +

      annotate("text",
                x = x_very_high_energy_burden,
                y = text_y_pos,
                label = paste0(pct_very_high_energy_burden, "%"),
                size = 3,
                fontface = "bold") +

      # Add dataset name on the right
      annotate("text",
                x = x_limits[2] * 0.95,
                y = y_limits[2] * 0.7,
                label = name,
                hjust = 1,
                fontface = "bold",
                size = 3,
                color = colors[name]) +

      # Format axes
      scale_y_continuous(
        labels = function(x) paste0(round(x * 242.13 * 0.001), "k"),
        name = "# of Homes",
        limits = y_limits,
        breaks = seq(0, y_max*242.13, by = y_tick_step)/242.13
      ) +
      scale_x_continuous(
        labels = scales::percent_format(accuracy = 1),
        name = "Energy Burden",
        limits = x_limits,
        breaks = c(low_burden_threshold, moderate_burden_threshold, high_burden_threshold)
      ) +

      # Apply theme
      theme_minimal() +
      theme(
        panel.grid.major = element_line(linewidth = 0.2),
        panel.grid.minor = element_blank(),
        aspect.ratio = 0.15
      )

    # Configure axes based on position
    if (!is_last) {
      plot <- plot +
        theme(
          axis.title.x = element_blank(),
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank()
        )
    }
    return(list(plot, category_percentages))
  }

  # Create each histogram
  dataset_names <- names(datasets)
  plots <- list()
  list_of_category_percentages <- list()

  for (i in 1:length(datasets)) {
    name <- dataset_names[i]
    is_first <- (i == 1)
    is_last <- (i == length(datasets))

    result_single_histogram <- create_single_histogram(
      data = datasets[[name]],
      name = name,
      is_first = is_first,
      is_last = is_last
    )

    plots[[i]] <- result_single_histogram[[1]]
    list_of_category_percentages[[i]] <- result_single_histogram[[2]]
  }

  # Combine plots with patchwork
  combined_plot <- plots[[1]]
  for (i in 2:length(plots)) {
    combined_plot <- combined_plot / plots[[i]]
  }

  # Add overall title and formatting
  final_plot <- combined_plot +
    plot_layout(heights = rep(1, length(plots))) +
    plot_annotation(
      title = "Energy Burden Comparison: EDC vs MLP",
      subtitle = glue::glue("Medium HP Efficiency | {supply_year} | 100% LI Discount Participation"),
      theme = theme(
        plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 12)
      )
    ) &
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_line(linewidth = 0.2, color = "gray90"),
      panel.grid.major.y = element_line(linewidth = 0.2, color = "gray90")
    )

  return(list(final_plot, list_of_category_percentages))
}


annual_bills_with_municipal_x <- annual_bills_with_municipal |> filter(bldg_id == bldg_id_for_debugging)


annual_bills_edc_baseline <- annual_bills_with_municipal |>
  filter(bill_type == "annual_bill_energy" &
          electric_utility %in% c("nationalgrid", "eversource", "unitil") &
          hvac == hvac_for_utility_bills &
          version_elec == "baseline")
annual_bills_edc_hpr_approved <- annual_bills_with_municipal |>
  filter(bill_type == "annual_bill_energy" &
          electric_utility %in% c("nationalgrid", "eversource", "unitil") &
          hvac == hvac_for_utility_bills &
          version_elec == "hpr_approved")
annual_bills_edc_hpr_doer <- annual_bills_with_municipal |>
  filter(bill_type == "annual_bill_energy" &
          electric_utility %in% c("nationalgrid", "eversource", "unitil") &
          hvac == hvac_for_utility_bills &
          version_elec == "hpr_doer")
annual_bills_mlp_baseline <- annual_bills_with_municipal |>
  filter(bill_type == "annual_bill_energy" &
          electric_utility == "municipal" &
          hvac == hvac_for_utility_bills &
          version_elec == "baseline")


dataset = list(
  "EDC Baseline" = annual_bills_edc_baseline,
  "EDC Approved" = annual_bills_edc_hpr_approved,
  "EDC DOER" = annual_bills_edc_hpr_doer,
  "MLP Baseline" = annual_bills_mlp_baseline
)


# Actually use the function
result_edc_mlp_burden_histograms <- plot_edc_mlp_burden_histograms(
  datasets = dataset,
  hvac_for_utility_bills = hvac_for_utility_bills
)

# Extract results
edc_mlp_combined_plot <- result_edc_mlp_burden_histograms[[1]]
edc_mlp_list_of_category_percentages <- result_edc_mlp_burden_histograms[[2]]

# Display the plot
print(edc_mlp_combined_plot)

# Create workspace variables for each dataset and category
for (i in seq_along(dataset)) {
  dataset_name <- names(dataset)[i]
  dataset_name <- tolower(gsub(" ", "_", dataset_name))

  # Get the result for this dataset
  result <- edc_mlp_list_of_category_percentages[[i]]

  # Use the correct category names from your function
  for (category in c("low_energy_burden", "moderate_energy_burden", "high_energy_burden", "very_high_energy_burden")) {
    pct_value <- result |>
      filter(bill_category == category) |>
      pull(percentage)

    if (length(pct_value) > 0) {
      assign(paste0("fig_edc_mlp_burden_histograms_", dataset_name, "_", category),
             round(pct_value, round_percent))
    }
  }
}

```



## Historical Supply Rates
### data-supply-rates
```{r}

# Load inflation factors
#inflation_factors <- readRDS("/workspaces/reports/data/fred/inflation_factors.rds")


# Load supply rates
con <- DBI::dbConnect(duckdb::duckdb(), path_rs_db)
inflation_factors <- DBI::dbReadTable(con, "inflation_factors")
supply_rates_elec <- DBI::dbReadTable(con, "supply_rates_elec")
supply_rates_gas <- DBI::dbReadTable(con, "supply_rates_gas")
supply_rates_fueloil <- DBI::dbReadTable(con, "supply_rates_fueloil")
supply_rates_propane <- DBI::dbReadTable(con, "supply_rates_propane")
DBI::dbDisconnect(con)

# --------------------------
# Electric
electric_utilities = c("eversource", "nationalgrid", "unitil")
weights = c(0.455, 0.431, 0.008)
weights = weights / sum(weights)

# Create tibble of utility weights
utility_weights <- tibble(
  electric_utility = electric_utilities,
  weight_edc = weights
)


# Create time series plot of supply rates by utility
supply_rates_elec_weighted_inflated <- supply_rates_elec |>
  filter(electric_utility %in% electric_utilities) |>
  left_join(utility_weights, by = "electric_utility") |>
  left_join(inflation_factors, by = "year") |>
  mutate(
    date = lubridate::ymd(paste0(year, "-", month, "-01")),
    supply_rate_weighted_inflated = supply_rate * weight_edc * inflation_factor
  ) |>
  group_by(date) |>
  summarise(supply_rate = sum(supply_rate_weighted_inflated), inflation_factor = mean(inflation_factor), weight_edc = mean(weight_edc))


# --------------------------
# Gas
gas_utilities = c("nationalgrid", "eversource", "liberty", "berkshire", "unitil")
weights_gas = c(0.268, 0.221, 0.002, 0.015, 0.005)
weights_gas = weights_gas / sum(weights_gas)

# Create tibble of utility weights
utility_weights_gas <- tibble(
  gas_utility = gas_utilities,
  weight_edc = weights_gas
)

supply_rates_gas_weighted_inflated <- supply_rates_gas |>
  filter(gas_utility %in% gas_utilities) |>
  left_join(utility_weights_gas, by = "gas_utility") |>
  left_join(inflation_factors, by = "year") |>
  mutate(
    date = lubridate::ymd(paste0(year, "-", month, "-01")),
    supply_rate_weighted_inflated = supply_rate * weight_edc * inflation_factor
  ) |>
  group_by(date) |>
  summarise(supply_rate = sum(supply_rate_weighted_inflated), inflation_factor = mean(inflation_factor), weight_edc = mean(weight_edc))


# --------------------------
# Fuel Oil
# Create tibble of utility weights

supply_rates_fueloil_weighted_inflated <- supply_rates_fueloil |>
  left_join(inflation_factors, by = "year") |>
  mutate(
    date = lubridate::ymd(paste0(year, "-", month, "-01")),
    supply_rate_weighted_inflated = fuel_oil_dollars_per_kwh * inflation_factor
  ) |>
  group_by(date) |>
  summarise(supply_rate = sum(supply_rate_weighted_inflated), inflation_factor = mean(inflation_factor))


```

### fig-historical-supply-rates
```{r}
#| label: fig-historical-supply-rates
#| fig-cap: "Historical supply rates for electric, gas, and fuel oil, inflation adjusted to 2024 dollars. Electricity and gas prices are weighted by utility share."

final_value_elec = supply_rates_elec_weighted_inflated |> filter(date == as.Date("2024-12-01")) |> pull(supply_rate)
final_value_gas = supply_rates_gas_weighted_inflated |> filter(date == as.Date("2024-12-01")) |> pull(supply_rate)
final_value_fueloil = supply_rates_fueloil_weighted_inflated |> filter(date == as.Date("2024-12-01")) |> pull(supply_rate)


# --------------------------
# Create supply rate plots
first_year = 2010
first_date = as.Date(paste0(first_year, "-01-01"))

last_year = 2024
last_date = as.Date(paste0(last_year, "-01-01"))


supply_plot <- ggplot() +
  # Electric supply rates
  geom_line(data = supply_rates_elec_weighted_inflated |> filter(year(date) >= first_year),
            aes(x = date, y = supply_rate, color = "Electricity"),
            linewidth = 1) +
  # Gas supply rates
  geom_line(data = supply_rates_gas_weighted_inflated |> filter(year(date) >= first_year),
            aes(x = date, y = supply_rate, color = "Gas"),
            linewidth = 1) +
  # Fuel Oil supply rates
  geom_line(data = supply_rates_fueloil_weighted_inflated |> filter(year(date) >= first_year),
            aes(x = date, y = supply_rate, color = "Fuel Oil"),
            linewidth = 1) +
  # Add grey rectangle for 2024
  annotate("rect",
           xmin = as.Date("2024-01-01"),
           xmax = as.Date("2024-12-31"),
           ymin = -Inf,
           ymax = Inf,
           alpha = 0.2,
           fill = "grey70") +
  # Add colored annotations instead of legend
  annotate("text", x = as.Date("2024-12-31"), y = final_value_elec, label = "Electricity",
           color = "#E69F00", fontface = "bold", size = 4, hjust = 0) +
  annotate("text", x = as.Date("2024-12-31"), y = final_value_gas, label = "Gas",
           color = "#56B4E9", fontface = "bold", size = 4, hjust = 0) +
  annotate("text", x = as.Date("2024-12-31"), y = final_value_fueloil, label = "Fuel Oil",
           color = "#009E73", fontface = "bold", size = 4, hjust = 0) +
  scale_y_continuous(labels = scales::label_dollar(size = 8)) +
  scale_x_date(
    limits = c(first_date, as.Date(paste0(last_year, "-12-31"))),
    breaks = c(as.Date("2010-01-01"), as.Date("2015-01-01"), as.Date("2020-01-01"), as.Date("2025-01-01")),  # Major breaks at 2010, 2015, 2020, 2025
    minor_breaks = seq(as.Date("2010-01-01"), as.Date("2025-01-01"), by = "1 year"),  # Minor breaks every year
    date_labels = "%Y",  # Show only year labels
    expand = expansion(mult = c(0, 0.22))  # Add 15% padding on right side only
  ) +
  scale_color_manual(values = c("Electricity" = "#E69F00", "Gas" = "#56B4E9", "Fuel Oil" = "#009E73")) +
  labs(
    title = "Historical Supply Rates - Electric, Gas, and Fuel Oil",
    subtitle = "Weighted by Utility Share | Inflation Adjusted to 2024",
    x = NULL,
    y = "Supply Rate [$/kWh]",
    color = "Fuel Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    panel.grid.major.x = element_line(color = "gray80", linetype = "solid"),
    panel.grid.minor.x = element_line(color = "gray50", linetype = "dotted"),
    legend.position = "none"  # Remove the legend
  )

supply_plot

```

### fig-historical-spark-gap
```{r}
#| label: fig-historical-spark-gap
#| fig-cap: "The 'Spark Gap' is the ratio of electric energy to natural gas energy, and is a key driver of the bill impacts of switching to heat pumps."

# Trim to years > 2010-01-01 and join datasets
supply_rates_elec_trimmed <- supply_rates_elec_weighted_inflated |>
  filter(date > as.Date("2010-01-01"))

supply_rates_gas_trimmed <- supply_rates_gas_weighted_inflated |>
  filter(date > as.Date("2010-01-01"))

supply_rates_fueloil_trimmed <- supply_rates_fueloil_weighted_inflated |>
  filter(date > as.Date("2010-01-01"))

# Join on date and create ratio column
supply_rates_combined <- supply_rates_elec_trimmed |>
  rename(supply_rate_elec = supply_rate) |>
  left_join(
    supply_rates_gas_trimmed |> rename(supply_rate_gas = supply_rate),
    by = "date"
  ) |>
  left_join(
    supply_rates_fueloil_trimmed |> rename(supply_rate_fueloil = supply_rate),
    by = "date"
  ) |>
  mutate(
    elec_gas_ratio = supply_rate_elec / supply_rate_gas,
    elec_fueloil_ratio = supply_rate_elec / supply_rate_fueloil
  )

spark_gap_plot <- ggplot(supply_rates_combined, aes(x = date)) +
  geom_line(aes(y = elec_gas_ratio, color = "Electric vs Gas"), linewidth = 1) +
  #geom_line(aes(y = elec_fueloil_ratio, color = "Electric vs Fuel Oil"), linewidth = 1) +
  scale_y_continuous(labels = scales::label_number(accuracy = 0.1), limits = c(0, 6)) +
  scale_color_manual(values = c("Electric vs Gas" = "#E69F00", "Electric vs Fuel Oil" = "#009E73")) +
  labs(
    title = "Spark Gap - Electric vs Gas",
    subtitle = "Weighted by Utility Share | Inflation Adjusted to 2024",
    x = NULL,
    y = "Spark Gap",
    color = "Comparison"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "none"
  )

spark_gap_plot
```


## Supply Year Sensitivity
```{r}
#| label: fig-supply-year-sensitivity
#| fig-cap: "The relative price of electric energy to natural gas drives the bill impacts of switching to heat pumps."

path_annual_change_db = "/workspaces/reports/data/ma_hp_rates/annual_change.db"

con <- DBI::dbConnect(duckdb::duckdb(), path_annual_change_db)
annual_change_with_municipal <- DBI::dbReadTable(con, "annual_change_with_municipal")
DBI::dbDisconnect(con)

categorize_bill_changes_by_utility <- function(data) {
  data |>
    filter(bill_type == "annual_bill_change_energy",
           hvac == hvac_for_utility_bills,
           version_elec %in% c("baseline", "hpr_approved", "hpr_doer")) |>
    mutate(category = case_when(
      annual_bill_change <= -1000 ~ "large_savings",
      annual_bill_change < 0 ~ "small_savings",
      annual_bill_change < 1000 ~ "small_increase",
      TRUE ~ "large_increase"
    )) |>
    group_by(version_elec, category, year) |>
    summarise(count = n(), .groups = "drop") |>
    group_by(version_elec, year) |>
    mutate(
      total = sum(count),
      percentage = count / total * 100,
      utility_scenario = paste(version_elec, year, sep = "_")
    ) |>
    ungroup()
}


# Apply the function to the data
supply_year_scenario_bins <- categorize_bill_changes_by_utility(annual_change_with_municipal)

# Set factor levels for correct ordering
supply_year_scenario_bins <- supply_year_scenario_bins |>
  mutate(
    # FIXED: Order for stacking - this controls the visual stacking order
    # For negative values, the order is visually reversed, so we put large_savings before small_savings
    category = factor(
      category,
      levels = c("large_increase", "small_increase", "large_savings", "small_savings")
    ),
    version_elec = factor(
      version_elec,
      levels = c("baseline", "hpr_approved", "hpr_doer")
    ),
    year = factor(
      year,
      levels = c("2019", "2020", "2021", "2022", "2023", "2024")
    )
  )

col_width = 0.8

# Create a single faceted plot showing all three scenarios
faceted_plot <- ggplot(supply_year_scenario_bins, aes(x = year, y = percentage, fill = category)) +
  # Reference line at 0 (boundary between savings and increases)
  geom_hline(yintercept = 0, linetype = "solid", color = "gray40", linewidth = 0.5) +
  # Stacked bars for increases (positive values)
  geom_col(
    data = supply_year_scenario_bins |> filter(category %in% c("small_increase", "large_increase")),
    aes(y = percentage),
    position = "stack",
    width = col_width
    ) +
  # Stacked bars for savings (negative values)
  geom_col(
    data = supply_year_scenario_bins |> filter(category %in% c("small_savings", "large_savings")),
    aes(y = -percentage),
    position = "stack",
    width = col_width
    ) +
  # Add percentage labels for savings (negative values)
  geom_text(
    data = supply_year_scenario_bins |> filter(category %in% c("large_savings", "small_savings")),
    aes(y = -percentage, label = paste0(round(percentage, 1), "%")),
    position = position_stack(vjust = 0.5),
    color = "white",
    fontface = "bold",
    size = 1.5
  ) +
  # Add percentage labels for increases (positive values)
  geom_text(
    data = supply_year_scenario_bins |> filter(category %in% c("small_increase", "large_increase")),
    aes(label = paste0(round(percentage, 1), "%")),
    position = position_stack(vjust = 0.5),
    color = "white",
    fontface = "bold",
    size = 1.5
  ) +
  # Facet by version_elec to create separate panels
  facet_wrap(~ version_elec,
             labeller = labeller(version_elec = c(
               "baseline" = "Baseline Rates",
               "hpr_approved" = "HPR Approved Rates",
               "hpr_doer" = "HPR DOER Rates"
             ))) +
  # Color scheme for categories
  scale_fill_manual(
    values = c(
      "large_increase" = "#db8b87",     # dark red (top)
      "small_increase" = "#eaada9",     # light red
      "small_savings" = "#8ebd85",      # light green
      "large_savings" = "#5e8a5e"       # dark green (bottom)
    ),
    labels = c(
      "large_increase" = ">$1,000 increase",
      "small_increase" = "$0-$1,000 increase",
      "small_savings" = "$0-$1,000 savings",
      "large_savings" = ">$1,000 savings"
    ),
    breaks = c("large_increase", "small_increase", "small_savings", "large_savings")
  ) +
  scale_y_continuous(
    labels = function(x) paste0(abs(x), "%")
  ) +
  scale_x_discrete(
    #name = "Supply Year",
    expand = expansion(mult = c(0.2, 0.2))
  ) +
  labs(
    title = "Heating Bill Changes by Supply Year and Rate Scenario",
    x = NULL,
    y = NULL,
    fill = "Bill Change Category"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_blank(),
    strip.text = element_text(face = "bold"),
    panel.spacing.x = unit(1.3, "lines")  # Reduce spacing between facet panels
  )

faceted_plot
```

## Revenue Impacts
### fig-stacked-bar-components-supply-on-top
```{r}
#| label: fig-stacked-bar-components-supply-on-top
#| fig-cap: "Median electric bill for Massachusetts homes, with current HVAC vs. mid-efficiency heat pump, under default, 1.0 and 2.0 heat pump rates."

bill_component_colors = c(
  "annual_gas_bill" = "grey50", # grey
  "annual_elec_supply_charge" = "#A0AF12",
  "annual_elec_delivery_charge" = "#023047",
  "annual_elec_customer_charge" = "#68BED8",
  "pre_hp_delivery_total" = "#FC9706"
)


stack_order_supply_on_top = c(
  "annual_gas_bill", # tippy top
  "annual_elec_supply_charge", # top
  "annual_elec_delivery_charge", # middle
  "annual_elec_customer_charge" # bottom
)

components <- factor(stack_order_supply_on_top, levels = stack_order_supply_on_top)

annual_bills_by_component_plot <- annual_bills_by_component |>
  arrange(match(charge_type, components))

y_current_baseline_customer_charge <- annual_bills_by_component_plot |> filter(version == "baseline", hvac == "current", charge_type == "annual_elec_customer_charge") |> pull(amount)
y_current_baseline_delivery_charge <- annual_bills_by_component_plot |> filter(version == "baseline", hvac == "current", charge_type == "annual_elec_delivery_charge") |> pull(amount)
y_current_baseline_supply_charge <- annual_bills_by_component_plot |> filter(version == "baseline", hvac == "current", charge_type == "annual_elec_supply_charge") |> pull(amount)

y_doer_customer_charge <- annual_bills_by_component_plot |> filter(version == "hpr_doer", hvac == "mid_hp", charge_type == "annual_elec_customer_charge") |> pull(amount)
y_doer_delivery_charge <- annual_bills_by_component_plot |> filter(version == "hpr_doer", hvac == "mid_hp", charge_type == "annual_elec_delivery_charge") |> pull(amount)
y_doer_supply_charge <- annual_bills_by_component_plot |> filter(version == "hpr_doer", hvac == "mid_hp", charge_type == "annual_elec_supply_charge") |> pull(amount)

y_annotation_offset <- 0
x_label_offset <- 4.5

fig_bill_stacked_bar_supply_on_top <- ggplot(annual_bills_by_component_plot |>
  mutate(charge_type = factor(charge_type, levels = components)),
       aes(x = interaction(hvac, version), y = amount, fill = charge_type)) +
  geom_bar(stat = "identity") +
  labs(title = "Components of Annual Electricity Bill",
       x = "",
       y = "Annual Cost ($)",
       fill = "Charge Type") +
  scale_fill_manual(values = bill_component_colors) +
  scale_y_continuous(labels = scales::dollar_format(), limits = c(0, 4500), breaks = seq(0, 5000, by = 1000), expand = c(0,0)) +
  scale_x_discrete(labels = c("Current HVAC\nCurrent Rates",
                           "Mid-Efficiency HP\nCurrent Rates",
                           "Mid-Efficiency HP\n1.0 Rates",
                           "Mid-Efficiency HP\n2.0 Rates")) +
  coord_cartesian(clip = "off") +
  theme(
    text = element_text(family = "IBM-Plex-Sans"),
    axis.title.x = element_text(margin = margin(t = 3)),
    axis.title.y = element_text(margin = margin(r = 3)),
    legend.position = "none",
    plot.margin = margin(t = 3, r = 200, b = 0, l = 10, unit = "pt")
  ) +

  # horizonta line at pre-hp delivery total
  # geom_hline(
  #   yintercept = y_current_baseline_customer_charge + y_current_baseline_delivery_charge,
  #   linetype = "dashed",
  #   color = bill_component_colors["pre_hp_delivery_total"],
  #   linewidth = 3
  # ) +

  # instead of a legend, write annotation for the 3 categories to the right of the right-most bar
  annotate(
    "text",
    x = x_label_offset,
    y = (y_doer_customer_charge/1.6) + y_annotation_offset,
    label = "Customer Charge",
    hjust = 0,
    family = "IBM-Plex-Sans",
    color = bill_component_colors["annual_elec_customer_charge"],
    size = 6,
    fontface = "bold"
  ) +
  annotate(
    "text",
    x = x_label_offset,
    y = (y_doer_delivery_charge/2) + y_doer_customer_charge + y_annotation_offset,
    label = "Delivery Charge",
    hjust = 0,
    family = "IBM-Plex-Sans",
    color = bill_component_colors["annual_elec_delivery_charge"],
    size = 6,
    fontface = "bold"
  ) +
  annotate(
    "text",
    x = x_label_offset,
    y = (y_doer_supply_charge/2) + y_doer_delivery_charge + y_doer_customer_charge + y_annotation_offset,
    label = "Supply Charge",
    hjust = 0,
    family = "IBM-Plex-Sans",
    color = bill_component_colors["annual_elec_supply_charge"],
    size = 6,
    fontface = "bold"
  ) +

  # No legend
  theme(legend.position = "none")

fig_bill_stacked_bar_supply_on_top

```


## Revenue Impact estimate -just volumetric
## prototype for fig-hists-revenue-impact
```{r}
#| label: fig-hists-revenue-impact-prototype
#| fig-cap: "The revenue impact of the HPR approved rates and HPR DOER rates."

hist_for_single_rate_version <- function(
  annual_change_table,
  bill_type,
  baseline_heating_type_option,
  hvac,
  appliances,
  shell,
  version_elec,
  x_limits = c(-3000, 3000),
  y_limits = c(0, 100),
  y_step = 10,
  binwidth = 100,
  show_category_labels = TRUE
) {

  # --------------------------------------
  # Plot parameters
  x_min <- x_limits[1]
  x_max <- x_limits[2]

  # Define the breaks for binning
  breaks <- seq(floor(x_min/binwidth)*binwidth, ceiling(x_max/binwidth)*binwidth, by = binwidth)

  # More balanced color palette with less contrast
  dark_red <- "#db8b87"        # dark red
  light_red <- "#eaada9"       # light red
  light_green <- "#8ebd85"     # light green
  dark_green <- "#5e8a5e"      # dark green
  # --------------------------------------

  # Filter data for mid_hp
  plot_data <- annual_change_table |>
    filter(hvac == !!hvac, appliances == !!appliances, shell == !!shell, version_elec == !!version_elec) |>
    filter(bill_type == !!bill_type) |>
    filter(if (baseline_heating_type_option != "All Fuels") baseline_heating_type == !!baseline_heating_type_option else TRUE) |>
    # Add category column before binning with new cutoffs
    mutate(bill_category = case_when(
      annual_bill_change <= -1000 ~ "large_savings",
      annual_bill_change < 0 ~ "small_savings",
      annual_bill_change < 1000 ~ "small_increase",
      TRUE ~ "large_increase"
    ))

  # Calculate percentages for each category
  category_percentages <- plot_data |>
    group_by(bill_category) |>
    summarise(count = n(), .groups = "drop") |>
    mutate(percentage = count / sum(count) * 100)

  # Get percentages for annotation
  pct_large_savings <- category_percentages |>
    filter(bill_category == "large_savings") |>
    pull(percentage) |>
    round(1)
  if(length(pct_large_savings) == 0) pct_large_savings <- 0

  pct_small_savings <- category_percentages |>
    filter(bill_category == "small_savings") |>
    pull(percentage) |>
    round(1)
  if(length(pct_small_savings) == 0) pct_small_savings <- 0

  pct_that_save <- pct_large_savings + pct_small_savings

  pct_small_increase <- category_percentages |>
    filter(bill_category == "small_increase") |>
    pull(percentage) |>
    round(1)
  if(length(pct_small_increase) == 0) pct_small_increase <- 0

  pct_large_increase <- category_percentages |>
    filter(bill_category == "large_increase") |>
    pull(percentage) |>
    round(1)
  if(length(pct_large_increase) == 0) pct_large_increase <- 0

  pct_that_lose <- pct_small_increase + pct_large_increase

  # Manually bin the data and calculate counts - this preserves the bill_category
  binned_data <- plot_data |>
    mutate(bin = cut(annual_bill_change,
                     breaks = breaks,
                     include.lowest = TRUE,
                     right = FALSE)) |>
    group_by(bin, bill_category) |>
    summarise(count = n(), .groups = "drop") |>
    # Add bin midpoint for plotting
    mutate(bin_mid = as.numeric(as.character(
      sapply(bin, function(b) {
        # Extract bin boundaries and calculate midpoint
        vals <- as.numeric(gsub("\\(|\\]|\\[|\\)", "", strsplit(as.character(b), ",")[[1]]))
        return(mean(vals))
      })
    )))

  # Position the bars at the very top and text just below
  bar_y_pos <- y_limits[2]  # Exactly at the top
  text_y_pos <- y_limits[2] * 0.95  # Just below the colored line

  # Calculate x positions for percentage labels (center of each category)
  x_large_savings <- mean(c(x_min, -1000))
  x_small_savings <- mean(c(-1000, 0))
  x_that_save <- x_min * 0.7
  x_small_increase <- mean(c(0, 1000))
  x_large_increase <- mean(c(1000, x_max))
  x_that_lose <- x_max * 0.7

  # Create the plot using geom_col() with binned data
  px_hist_by_rate_version <- ggplot(binned_data, aes(x = bin_mid, y = count, fill = bill_category)) +
    # Add vertical dotted lines at the category boundaries
    geom_vline(xintercept = c(-1000, 0, 1000), linetype = "dotted", color = "gray50", size = 0.5) +

    # Add the bars
    geom_col(position = "stack", width = binwidth * 0.9) +

    # Set the fill colors manually with expanded categories
    scale_fill_manual(
      values = c(
        "large_savings" = "#5e8a5e",      # dark green
        "small_savings" = "#8ebd85",      # light green
        "small_increase" = "#eaada9",     # light red
        "large_increase" = "#db8b87"      # dark red
      ),
      guide = "none"
    ) +

    # Add horizontal lines at the very top of the plot
    # With colors matching their respective categories
    annotate("segment",
             x = x_min, xend = -1000,
             y = bar_y_pos, yend = bar_y_pos,
             color = dark_green, size = 1) +

    annotate("segment",
             x = -1000, xend = 0,
             y = bar_y_pos, yend = bar_y_pos,
             color = light_green, size = 1) +

    annotate("segment",
             x = 0, xend = 1000,
             y = bar_y_pos, yend = bar_y_pos,
             color = light_red, size = 1) +

    annotate("segment",
             x = 1000, xend = x_max,
             y = bar_y_pos, yend = bar_y_pos,
             color = dark_red, size = 1) +

    # Add percentage labels for each category directly below the lines
    annotate("text",
             x = x_large_savings,
             y = text_y_pos,
             label = paste0(pct_large_savings, "%"),
             size = 3,
             fontface = "bold") +

    annotate("text",
             x = x_small_savings,
             y = text_y_pos,
             label = paste0(pct_small_savings, "%"),
             size = 3,
             fontface = "bold") +

    annotate("text",
             x = x_small_increase,
             y = text_y_pos,
             label = paste0(pct_small_increase, "%"),
             size = 3,
             fontface = "bold") +

    annotate("text",
             x = x_large_increase,
             y = text_y_pos,
             label = paste0(pct_large_increase, "%"),
             size = 3,
             fontface = "bold") +

    # Add the percentage labels for the "that save" and "that lose" categories
    annotate("text",
             x = x_that_save,
             y = text_y_pos*0.8,
             label = paste0(pct_that_save, "% pay lower electric bills"),
             size = 3,
             fontface = "bold") +

    annotate("text",
             x = x_that_lose,
             y = text_y_pos*0.8,
             label = paste0(pct_that_lose, "% pay higher electric bills"),
             size = 3,
             fontface = "bold") +

    # Rest of the plot styling
    labs(
      title = glue::glue("Change in Total Winter Energy Bills after Switching to Heat Pumps"),
      x = "Annual Bill Change ($)",
      y = "# of Homes"
    ) +
    scale_y_continuous(
      labels = function(x) paste0(round(x * 1), ""),
      name = "# of Homes",
      limits = y_limits,
      breaks = seq(0, y_limits[2], by=y_step),
      expand = expansion(mult = c(0, 0))
    ) +
    scale_x_continuous(
      labels = scales::dollar_format(),
      limits = x_limits,
      breaks = seq(x_min, x_max, by = 500),
      expand = expansion(mult = c(0, 0))
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 10, face = "bold"),
      panel.grid.major = element_line(linewidth = 0.2),
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "none",
      aspect.ratio = 0.6,
      margin = margin(0, 0, 0, 0),
      panel.background = element_rect(fill = "#f2f0eb", color = NA),
      #plot.background = element_rect(fill = "#FAFAF5", color = NA)
    ) +
    {
      if (show_category_labels) {
        list(
          annotate("text", x = x_large_savings,   y = y_limits[2]*0.97,
                   label = "savings > $1k",  size = 3, fontface = "bold"),
          annotate("text", x = x_small_savings,   y = y_limits[2]*0.97,
                   label = "savings $0–1k", size = 3, fontface = "bold"),  # make bold
          annotate("text", x = x_small_increase,  y = y_limits[2]*0.97,
                   label = "losses  $0–1k", size = 3, fontface = "bold"),  # make bold
          annotate("text", x = x_large_increase,  y = y_limits[2]*0.97,
                   label = "losses  > $1k", size = 3, fontface = "bold")
        )
      }
    } +
    coord_cartesian(clip = "off")

  return(list(px_hist_by_rate_version, category_percentages))
}



plot_elec_bill_change_histograms <- function(
  annual_change_table,
  bill_type,
  baseline_heating_type_option,
  hvac,
  appliances,
  shell,
  electric_utility_option,
  supply_year,
  second_subtitle = NULL,
  x_limits = c(-2500, 2500),
  y_limits = c(0, 100),
  y_step = 10,
  binwidth = 100
) {


  # Create nicer version labels
  version_labels <- c(
    "baseline" = "     Default\n     Rates",
    "hpr_approved" = "     1.0\n     Rates",
    "hpr_doer" = "     2.0\n     Rates"
  )

  # Create three histograms without individual titles
  result_baseline <- hist_for_single_rate_version(
    annual_change_table = annual_change_table,
    bill_type = bill_type,
    baseline_heating_type_option = baseline_heating_type_option,
    hvac = hvac,
    appliances = appliances,
    shell = shell,
    version_elec = "baseline",
    x_limits = x_limits,
    y_limits = y_limits,
    y_step = y_step,
    binwidth = binwidth,
    show_category_labels = FALSE
  )

  p1_baseline <- result_baseline[[1]] +
    labs(title = NULL) +  # Remove title
    ylab(NULL) +                 # Remove y-axis label
    theme(
      axis.title.x = element_blank(),  # Remove x axis title from top two plots
      axis.text.x = element_blank(),   # Remove x tick labels from top two plots
      axis.ticks.x = element_blank(),  # Remove x ticks from top two plots
      plot.margin = margin(5, 5, 0, 5),  # Consistent margins (top, right, bottom, left)
      axis.title.y = element_blank()     # Remove y title from top plot
    ) +
    # Add elegant version label on right side
    annotate("text",
             x = Inf,  # Use Inf to position at the right edge of the plot
             y = y_limits[2] * 0.7,
             label = version_labels["baseline"],
             hjust = 0,  # Align to the right edge
             fontface = "bold",
             size = 4,
             color = "black",
             vjust = 0.5)  # Center text vertically

  result_approved <- hist_for_single_rate_version(
    annual_change_table = annual_change_table,
    bill_type = bill_type,
    baseline_heating_type_option = baseline_heating_type_option,
    hvac = hvac,
    appliances = appliances,
    shell = shell,
    version_elec = "hpr_approved",
    x_limits = x_limits,
    y_limits = y_limits,
    y_step = y_step,
    binwidth = binwidth,
    show_category_labels = FALSE
  )

  p2_approved <- result_approved[[1]] +
    labs(title = NULL) +  # Remove title
    ylab(NULL) +
    theme(
      axis.title.x = element_blank(),  # Remove x axis title from top two plots
      axis.text.x = element_blank(),   # Remove x tick labels from top two plots
      axis.ticks.x = element_blank(),  # Remove x ticks from top two plots
      axis.title.y = element_blank(),   # Remove y title from middle plot
      plot.margin = margin(0, 5, 0, 5)  # Consistent margins (no top/bottom margin)
    ) +
    # Add elegant version label on right side
    annotate("text",
             x = Inf,  # Use Inf to position at the right edge of the plot
             y = y_limits[2] * 0.7,
             label = version_labels["hpr_approved"],
             hjust = 0,  # Align to the right edge
             fontface = "bold",
             size = 4,
             color = "black",
             vjust = 0.5)  # Center text vertically

  result_doer <- hist_for_single_rate_version(
    annual_change_table = annual_change_table,
    bill_type = bill_type,
    baseline_heating_type_option = baseline_heating_type_option,
    hvac = hvac,
    appliances = appliances,
    shell = shell,
    version_elec = "hpr_doer",
    x_limits = x_limits,
    y_limits = y_limits,
    y_step = y_step,
    binwidth = binwidth,
    show_category_labels = FALSE
  )

  p3_doer <- result_doer[[1]] +
    labs(title = NULL) +  # Remove title
    ylab(NULL) +
    theme(
      axis.title.y = element_blank(),  # Remove y title from bottom plot
      plot.margin = margin(0, 5, 5, 5),  # Consistent margins
      axis.title.x = element_text(size = 9),
    ) +
    # Add elegant version label on right side
    annotate("text",
             x = Inf,  # Use Inf to position at the right edge of the plot
             y = y_limits[2] * 0.7,
             label = version_labels["hpr_doer"],
             hjust = 0,  # Align to the right edge
             fontface = "bold",
             size = 4,
             color = "black",
             vjust = 0.5)  # Center text vertically

  # Combine plots with shared x and y axes
  combined_plot <- p1_baseline / p2_approved / p3_doer +
    plot_layout(heights = c(1, 1, 1)) +  # Equal heights for all plots
    plot_annotation(
      title = case_when(
        bill_type == "annual_bill_change_elec" ~ glue::glue("Switching from {baseline_heating_type_option} to Heat Pump: Change in Annual Electric Bill"),
        bill_type == "annual_bill_change_energy" ~ glue::glue("Change in Total Winter Energy Bills after Switching to Heat Pump from {baseline_heating_type_option}"),
        bill_type == "annual_bill_change_gas" ~ glue::glue("Switching from {baseline_heating_type_option} to Heat Pump: Change in Annual Gas Bill"),
        TRUE ~ "Heat Pump Rate Bill Changes" # Default case
      ),
      subtitle = second_subtitle,
      theme = theme(
        plot.margin = margin(1,1,5,1),  # top, right, bottom, left
        plot.title = element_text(hjust = 0.5, size = 10, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 8),
      )
    ) &
    # Apply consistent theming to all plots
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_line(linewidth = 0.2, color = "gray90"),
      panel.grid.major.y = element_line(linewidth = 0.2, color = "gray90"),
      aspect.ratio = 0.18,
      plot.margin = margin(1,25,8,1),  # top, right, bottom, left
    )

  # Wrap the combined plot to add a shared y-axis label
  combined_plot <- patchwork::wrap_elements(combined_plot) +
    labs(tag = "# of Homes") +
    theme(
      plot.tag = element_text(angle = 90, vjust = 0.5, hjust = 0.5, face = "plain", size = 9),
      plot.tag.position = "left"
    )

  return(list(combined_plot, result_baseline[[2]], result_approved[[2]], result_doer[[2]]))
}
```


## Revenue Impact estimate - just volumetric
```{r}
#| label: data-hist-delivery-revenue-volumetric-estimate
#| fig-cap: "The electric delivery revenue impact of the HPR approved rates and HPR DOER rates."

default_delivery_rate_eversource = 0.19122
hpr1_delivery_rate_eversource = 0.12094
hpr2_delivery_rate_eversource = 0.0702

default_delivery_rate_nationalgrid = 0.18091
hpr1_delivery_rate_nationalgrid = 0.12669
hpr2_delivery_rate_nationalgrid = 0.06652

default_delivery_rate_unitil = 0.26493
hpr1_delivery_rate_unitil = 0.20307
hpr2_delivery_rate_unitil = 0.09652



elec_revenue_currrent_hvac <- winter_energy |>
  filter(hvac == "current") |>
  filter(fuel == "electricity") |>
  select(bldg_id, hvac, total_winter_kwh, electric_utility) |>
  mutate(
    elec_revenue_default_current =  case_when(
      electric_utility == "unitil" ~ total_winter_kwh * default_delivery_rate_unitil,
      electric_utility == "nationalgrid" ~ total_winter_kwh * default_delivery_rate_nationalgrid,
      electric_utility == "eversource" ~ total_winter_kwh * default_delivery_rate_eversource,
      TRUE ~ 0
    )
  ) |>
  pivot_longer(
    cols = starts_with("elec_revenue_"),
    names_to = "version_elec",
    values_to = "elec_revenue"
  ) |>
  mutate(
    version_elec = case_when(
      version_elec == "elec_revenue_default_current" ~ "baseline"
    )
  ) |>
  select(bldg_id, hvac, total_winter_kwh, version_elec, elec_revenue, electric_utility)


elec_revenue_mid_hp <- winter_energy |>
  filter(hvac == "mid_hp") |>
  filter(fuel == "electricity") |>
  select(bldg_id, hvac, total_winter_kwh, electric_utility) |>
  mutate(
    elec_revenue_default_mid_hp = case_when(
      electric_utility == "unitil" ~ total_winter_kwh * default_delivery_rate_unitil,
      electric_utility == "nationalgrid" ~ total_winter_kwh * default_delivery_rate_nationalgrid,
      electric_utility == "eversource" ~ total_winter_kwh * default_delivery_rate_eversource,
      TRUE ~ 0
    ),
    elec_revenue_hpr1_mid_hp = case_when(
      electric_utility == "unitil" ~ total_winter_kwh * hpr1_delivery_rate_unitil,
      electric_utility == "nationalgrid" ~ total_winter_kwh * hpr1_delivery_rate_nationalgrid,
      electric_utility == "eversource" ~ total_winter_kwh * hpr1_delivery_rate_eversource,
      TRUE ~ 0
    ),
    elec_revenue_hpr2_mid_hp = case_when(
      electric_utility == "unitil" ~ total_winter_kwh * hpr2_delivery_rate_unitil,
      electric_utility == "nationalgrid" ~ total_winter_kwh * hpr2_delivery_rate_nationalgrid,
      electric_utility == "eversource" ~ total_winter_kwh * hpr2_delivery_rate_eversource,
      TRUE ~ 0
    )
  ) |>
  pivot_longer(
    cols = starts_with("elec_revenue_"),
    names_to = "version_elec",
    values_to = "elec_revenue"
  ) |>
  mutate(
    version_elec = case_when(
      version_elec == "elec_revenue_default_mid_hp" ~ "baseline",
      version_elec == "elec_revenue_hpr1_mid_hp" ~ "hpr_approved",
      version_elec == "elec_revenue_hpr2_mid_hp" ~ "hpr_doer"
    )
  ) |>
  select(bldg_id, hvac, total_winter_kwh, version_elec, elec_revenue, electric_utility)

elec_revenue <- elec_revenue_mid_hp |>
  bind_rows(elec_revenue_currrent_hvac) |>
  select(bldg_id, hvac, version_elec, total_winter_kwh, elec_revenue, electric_utility) |>
  left_join(housing_units |> select(bldg_id, in.sqft, electric_utility, gas_utility, assigned_income, building_type_group, baseline_heating_type), by = c("bldg_id","electric_utility")) |>
  mutate(
    shell = "current",
    appliances = "current",
    bill_type = "annual_delivery_revenue_elec"
  )

rm(elec_revenue_currrent_hvac, elec_revenue_mid_hp)

elec_revenue_change <- elec_revenue |>
  # Get the baseline revenue for each building
  left_join(
    elec_revenue |>
      filter(hvac == "current", version_elec == "baseline") |>
      select(bldg_id, baseline_revenue = elec_revenue, baseline_total_winter_kwh = total_winter_kwh, electric_utility),
    by = c("bldg_id", "electric_utility")
  ) |>
  # Calculate the change in revenue
  mutate(
    bill_type = "annual_delivery_revenue_change_elec",
    annual_kwh_change = total_winter_kwh - baseline_total_winter_kwh,
    annual_bill_change = elec_revenue - baseline_revenue
  ) |>
  select(-baseline_revenue)

```


```{r}
#| label: stats-annual-elec-revenue-change
# Stats
calc_elec_revenue_stats_by_rate_version <- function(annual_change_table, hp_eff_for_stats, print_table = FALSE) {
  scenario_stats <- annual_change_table |>
    filter(hvac == hp_eff_for_stats) |>
    filter(!is.na(annual_bill_change)) |>  # Remove NA values
    group_by(version_elec) |>
    summarise(
      # kWh change
      median_kwh_change = median(annual_kwh_change),

      # Bill change
      median_bill_change = median(annual_bill_change),
      pct_that_save = mean(annual_bill_change < 0) * 100,
      pct_that_save_big = mean(annual_bill_change < -1000) * 100,
      pct_that_lose = mean(annual_bill_change > 0) * 100,
      pct_that_lose_big = mean(annual_bill_change > 1000) * 100,
      median_bill_savings = median(annual_bill_change[annual_bill_change < 0]),
      median_bill_loss = median(annual_bill_change[annual_bill_change > 0])
    )

  if (print_table) {
    print(knitr::kable(scenario_stats,
               caption = glue::glue("Summary Statistics by Rate Version (HP Efficiency = {hp_eff_for_stats})"),
               format = "pipe",
               digits = 1,
               col.names = c("Rate Version", "Median Bill Change ($)", "% That Save", "Median Savings ($)", "Median Loss ($)")))
  }

  return(scenario_stats)
}

scenario_stats_elec_types_mid_hp <- calc_elec_revenue_stats_by_rate_version(elec_revenue_change, hp_eff_for_stats = "mid_hp", print_table = FALSE)

# Some stats used in the report
# Baseline
fig_hists_delivery_revenue_baseline_median_bill_change <- scenario_stats_elec_types_mid_hp |> filter(version_elec == "baseline") |> pull(median_bill_change)
fig_hists_delivery_revenue_baseline_pct_that_save <- scenario_stats_elec_types_mid_hp |> filter(version_elec == "baseline") |> pull(pct_that_save)
fig_hists_delivery_revenue_baseline_pct_that_lose <- scenario_stats_elec_types_mid_hp |> filter(version_elec == "baseline") |> pull(pct_that_lose)

# HPR 2.0
fig_hists_delivery_revenue_hpr2_median_bill_change <- scenario_stats_elec_types_mid_hp |> filter(version_elec == "hpr_doer") |> pull(median_bill_change)
fig_hists_delivery_revenue_hpr2_median_kwh_change <- scenario_stats_elec_types_mid_hp |> filter(version_elec == "hpr_doer") |> pull(median_kwh_change)
fig_hists_delivery_revenue_hpr2_pct_that_save <- scenario_stats_elec_types_mid_hp |> filter(version_elec == "hpr_doer") |> pull(pct_that_save)
fig_hists_delivery_revenue_hpr2_pct_that_lose <- scenario_stats_elec_types_mid_hp |> filter(version_elec == "hpr_doer") |> pull(pct_that_lose)

## Sanity check: median_kwh_change/median_bill_change ~ = $0.20/kWh (rough average of the delivery rates)
#crude_marginal_delivery_rate_baseline = round(fig_hists_delivery_revenue_baseline_median_bill_change / fig_hists_delivery_revenue_hpr2_median_kwh_change, 3)

#Sanity check: median_kwh_change/median_bill_change[hpr2] ~  ZERO (zero would mean that the total electric revenue doesn't change for this new heat pump household
#crude_marginal_delivery_rate_hp2 = round(fig_hists_delivery_revenue_hpr2_median_bill_change / fig_hists_delivery_revenue_hpr2_median_kwh_change, 3)

```

### fig-hist-delivery-revenue-volumetric-estimate
```{r}
#| label: fig-hists-delivery-revenue-volumetric-estimate
#| fig-cap: "The distribution of the change in winter electric delivery bills for heat pump households."

elec_revenue_change_plot <- elec_revenue_change |>
  filter(hvac == "mid_hp")

revenue_hists <- plot_elec_bill_change_histograms(
  annual_change_table = elec_revenue_change_plot,
  bill_type = "annual_delivery_revenue_change_elec",
  baseline_heating_type_option = "All Fuels",
  hvac = "mid_hp",
  appliances = "current",
  shell = "current",
  electric_utility_option = "all",
  supply_year = "2024",
  second_subtitle = "Change in Winter Electric Delivery Bills (Volumetric Component Only)",
  x_limits = c(-3000, 3000),
  y_limits = c(0, 3000),
  y_step = 500,
  binwidth = 250
)

print(revenue_hists[[1]])
```



## Spark Gap
### fig-scatter-elec-gas-kwh-changes
```{r}
#| label: fig-scatter-elec-gas-kwh-changes
#| fig-cap: "The relationship between the change in annual gas kwh and the change in annual electric kwh."

# rate_colors = c(
#   "current" = "#023047",  # sb midnight
#   "low_hp" = "#68BED8", # sb sky
#   "mid_hp" = "#FC9706", # sb carrot
#   "high_hp" = "#009E73" # sb pistachio
# )


# # First calculate the linear model for Natural Gas heated buildings
# # Calculate linear models for each version
# # winter_energy_delta_wide <- winter_energy_delta |>
# #   filter(baseline_heating_type == "Natural Gas") |>
# #   pivot_wider(
# #     id_cols = c(bldg_id, hvac),
# #     names_from = fuel,
# #     values_from = delta_winter_kwh,
# #     names_prefix = "delta_winter_kwh_"
# #   ) |>
# #   mutate(
# #     ratio_electricity_to_gas = delta_winter_kwh_electricity / -delta_winter_kwh_gas
# #   )


# lm_models <- winter_energy_delta |>
#   filter(baseline_heating_type == "Natural Gas") |>
#   group_by(hvac) |>
#   pivot_wider(
#     id_cols = c(bldg_id),
#     names_from = fuel,
#     values_from = delta_winter_kwh,
#     names_prefix = "delta_winter_kwh_"
#   ) |>
#   # summarise(
#   #   model = list(lm(delta_winter_kwh_electricity ~ delta_winter_kwh_gas)),
#   #   r_squared = summary(model[[1]])$r.squared,
#   #   slope = coef(model[[1]])[2],
#   #   intercept = coef(model[[1]])[1]
#   # ) |>
#   ungroup()



# # # Extract coefficients into separate variables for each version
# # current_slope <- lm_models |> filter(fuel == "electricity") |> pull(slope)
# # current_intercept <- lm_models |> filter(hvac == "current") |> pull(intercept)

# low_hp_slope <- lm_models |> filter(hvac == "low_hp") |> pull(slope)
# low_hp_intercept <- lm_models |> filter(hvac == "low_hp") |> pull(intercept)

# mid_hp_slope <- lm_models |> filter(hvac == "mid_hp") |> pull(slope)
# mid_hp_intercept <- lm_models |> filter(hvac == "mid_hp") |> pull(intercept)

# high_hp_slope <- lm_models |> filter(hvac == "high_hp") |> pull(slope)
# high_hp_intercept <- lm_models |> filter(hvac == "high_hp") |> pull(intercept)

# # Create the plot
# xmin = 0
# xmax = 50000
# xstep = 5000
# ymin = 0
# ymax = 30000
# ystep = 5000


# ggplot(winter_energy_delta, aes(x = delta_winter_kwh_gas,
#                               y = delta_winter_kwh_electricity,
#                               color = hvac)) +
#   scale_color_manual(values = rate_colors) +
#   geom_point(alpha = 0.3, size = 1) +

#   geom_abline(slope = current_slope, intercept = current_intercept, color = rate_colors["current"]) +
#   geom_abline(slope = low_hp_slope, intercept = low_hp_intercept, color = rate_colors["low_hp"]) +
#   geom_abline(slope = mid_hp_slope, intercept = mid_hp_intercept, color = rate_colors["mid_hp"]) +
#   geom_abline(slope = high_hp_slope, intercept = high_hp_intercept, color = rate_colors["high_hp"]) +

#   annotate("text", x = xmax, y = (current_slope * xmax) + current_intercept,
#            label = sprintf(" Current: y = %.2fx + %.2f", current_slope, current_intercept),
#            hjust = 0, size = 3.5, color = rate_colors["current"]) +
#   annotate("text", x = xmax, y = (low_hp_slope * xmax) + low_hp_intercept,
#            label = sprintf("Low HP: y = %.2fx + %.2f", low_hp_slope, low_hp_intercept),
#            hjust = 0, size = 3.5, color = rate_colors["low_hp"]) +
#   annotate("text", x = xmax, y = (mid_hp_slope * xmax) + mid_hp_intercept,
#            label = sprintf("Mid HP: y = %.2fx + %.2f", mid_hp_slope, mid_hp_intercept),
#            hjust = 0, size = 3.5, color = rate_colors["mid_hp"]) +
#   annotate("text", x = xmax, y = (high_hp_slope * xmax) + high_hp_intercept,
#            label = sprintf("High HP: y = %.2fx + %.2f", high_hp_slope, high_hp_intercept),
#            hjust = 0, size = 3.5, color = rate_colors["high_hp"]) +
#   labs(x = "Annual Natural Gas Energy (kWh) BEFORE Electrification",
#        y = "Annual Electric Energy (kWh) AFTER Electrification",
#        title = "Annual Energy Consumption by Heating System") +
#   scale_x_continuous(labels = scales::comma_format(),
#                      limits = c(xmin, xmax),
#                      breaks = seq(xmin, xmax, xstep)) +
#   scale_y_continuous(labels = scales::comma_format(),
#                      limits = c(ymin, ymax),
#                      breaks = seq(ymin, ymax, ystep)) +
#   coord_cartesian(xlim = c(xmin, xmax+(2*xstep)),
#                  ylim = c(ymin, ymax)) +
#   theme_minimal() +
#   theme(
#     plot.title = element_text(hjust = 0.5),
#     axis.text.x = element_text(vjust = 0),
#     axis.ticks.x = element_line(),
#     legend.position = "none",
#     plot.margin = margin(t = 10, r = 100, b = 10, l = 10)
#   )


```

### data-effective-spark-gap
```{r}
#| label: data-effective-spark-gap

# con <- DBI::dbConnect(duckdb::duckdb(), path_rs_db)

# # Query annual energy consumption by building and HVAC scenario
# annual_energy <- tbl(con, "ma_monthly") |>
#   group_by(bldg_id, hvac) |>
#   summarise(
#     total_elec_kwh = sum(total_elec_kwh),
#     total_gas_kwh = sum(total_gas_kwh),
#     total_propane_kwh = sum(total_propane_kwh),
#     total_fuel_oil_kwh = sum(total_fuel_oil_kwh)
#   ) |>
#   collect()

# DBI::dbDisconnect(con)

# annual_energy_long <- annual_energy |>
#   pivot_longer(cols = c(total_elec_kwh, total_gas_kwh, total_propane_kwh, total_fuel_oil_kwh),
#                names_to = "bill_type",
#                values_to = "annual_kwh") |>
#   mutate(
#     bill_type = case_when(
#       bill_type == "total_elec_kwh" ~ "annual_bill_elec",
#       bill_type == "total_gas_kwh" ~ "annual_bill_gas",
#       bill_type == "total_propane_kwh" ~ "annual_bill_propane",
#       bill_type == "total_fuel_oil_kwh" ~ "annual_bill_fueloil"
#     )
#   )


# annual_bills_with_kwh <- annual_bills |>
#   left_join(annual_energy_long, by = c("bldg_id", "hvac", "bill_type")) |>
#   mutate(
#     energy_rate = ifelse(is.na(annual_kwh) | annual_kwh == 0, NA_real_, annual_bill / annual_kwh)
#   ) |>
#   select(bldg_id, hvac, bill_type, annual_bill, annual_kwh, energy_rate, version_elec, baseline_heating_type)

```


### fig-scatter-effective-spark-gap
```{r}
#| label: fig-scatter-effective-spark-gap
#| fig-cap: "The effective spark gap is the ratio of electric energy to natural gas energy, and is a key driver of the bill impacts of switching to heat pumps."


# # --------------------------------------
# # Create scatter plot of gas vs electric rates
# # --------------------------------------
# plot_these_rates <- c("baseline", "hpr_approved", "hpr_doer")
# annual_bills_with_kwh_plot <- annual_bills_with_kwh |>
#   filter(
#     hvac == "mid_hp",
#     baseline_heating_type == "Natural Gas",
#     version_elec %in% plot_these_rates
#   ) |>

#   # Pivot wider to get gas and electric rates in separate columns
#   select(bldg_id, bill_type, energy_rate, version_elec) |>
#   filter(energy_rate > 0.01) |>
#   filter(energy_rate < 0.5) |>
#   pivot_wider(
#     id_cols = c(bldg_id, version_elec),
#     names_from = bill_type,
#     values_from = energy_rate
#   ) |>
#   mutate(
#     energy_rate_elec = annual_bill_elec,
#     energy_rate_gas = annual_bill_gas,
#     energy_rate_fuel_oil = annual_bill_fueloil,
#     energy_rate_propane = annual_bill_propane
#   ) |>
#   mutate(
#     spark_gap_gas = energy_rate_elec / energy_rate_gas,
#   )

# #View(annual_bills_with_kwh_plot |> filter(bldg_id == 21))


# rate_colors = c(
#   "baseline" = "#023047",  # sb midnight
#   "hpr_approved" = "#FC9706", # sb carrot
#   "hpr_doer" = "#009E73" # sb pistachio
# )


# # First calculate the linear model for Natural Gas heated buildings
# # Calculate linear models for each version
# lm_models <- annual_bills_with_kwh_plot |>
#   group_by(version_elec) |>
#   summarise(
#     model = list(lm(energy_rate_elec ~ energy_rate_gas)),
#     slope_linear = coef(model[[1]])[2],
#     intercept = coef(model[[1]])[1]
#   ) |>
#   ungroup()

# # Extract coefficients into separate variables for each version
# lm_slope_baseline <- lm_models |> filter(version_elec == "baseline") |> pull(slope_linear)
# lm_intercept_baseline <- lm_models |> filter(version_elec == "baseline") |> pull(intercept)

# lm_slope_hpr_approved <- lm_models |> filter(version_elec == "hpr_approved") |> pull(slope_linear)
# lm_intercept_hpr_approved <- lm_models |> filter(version_elec == "hpr_approved") |> pull(intercept)

# lm_slope_hpr_doer <- lm_models |> filter(version_elec == "hpr_doer") |> pull(slope_linear)
# lm_intercept_hpr_doer <- lm_models |> filter(version_elec == "hpr_doer") |> pull(intercept)

# # Create scatter plot of gas vs electric rates
# xmin = 0
# xmax = 0.5
# ymin = 0
# ymax = 0.5

# annual_bills_with_kwh_plot |>
#   # Create scatter plot
#   ggplot(aes(x = annual_bill_gas, y = annual_bill_elec, color = version_elec)) +
#   scale_color_manual(values = rate_colors) +
#   geom_point(alpha = 0.5, size = 1) +

#   #
#   geom_abline(slope = lm_slope_baseline, intercept = lm_intercept_baseline, color = rate_colors["baseline"]) +
#   geom_abline(slope = lm_slope_hpr_approved, intercept = lm_intercept_hpr_approved, color = rate_colors["hpr_approved"]) +
#   geom_abline(slope = lm_slope_hpr_doer, intercept = lm_intercept_hpr_doer, color = rate_colors["hpr_doer"]) +

#   annotate("text", x = .35, y = (lm_slope_baseline * .35) + lm_intercept_baseline,
#            label = sprintf("Baseline: y = %.2fx + %.2f", lm_slope_baseline, lm_intercept_baseline),
#            hjust = 0, size = 3.5, color = rate_colors["baseline"]) +
#   annotate("text", x = .35, y = (lm_slope_hpr_approved * .35) + lm_intercept_hpr_approved,
#            label = sprintf("HPR Approved: y = %.2fx + %.2f", lm_slope_hpr_approved, lm_intercept_hpr_approved),
#            hjust = 0, size = 3.5, color = rate_colors["hpr_approved"]) +
#   annotate("text", x = .35, y = (lm_slope_hpr_doer * .35) + lm_intercept_hpr_doer,
#            label = sprintf("HPR DOER: y = %.2fx + %.2f", lm_slope_hpr_doer, lm_intercept_hpr_doer),
#            hjust = 0, size = 3.5, color = rate_colors["hpr_doer"]) +

#   # Add reference line where electric rate equals gas rate (slope=1, intercept=0)
#   geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "darkgrey") +

#   scale_x_continuous(labels = scales::dollar_format(),
#                      limits = c(xmin, xmax),
#                      breaks = seq(xmin, xmax, 0.05)) +
#   scale_y_continuous(labels = scales::dollar_format(),
#                      limits = c(ymin, ymax),
#                      breaks = seq(ymin, ymax, 0.05)) +
#   coord_cartesian(xlim = c(xmin, xmax+0.05),
#                  ylim = c(ymin, ymax)) +
#   labs(
#     x = "Natural Gas Rate ($/kWh)",
#     y = "Electricity Rate ($/kWh)",
#     color = "Rate Version",
#     title = "Effective Spark Gap: Electricity vs. Natural Gas Rates"
#   ) +
#   theme_minimal() +
#   theme(
#     legend.position = "bottom",
#     plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
#     panel.grid.minor = element_blank(),
#     margin = margin(t = 10, r = 100, b = 10, l = 10)
#   )

```


# ------------------------------------------------------------------------------
# Export Data
# ------------------------------------------------------------------------------
```{r}
#| label: save-image
 save.image(file = "/workspaces/reports/src/ma_hp_rates/cache/report_variables.RData")
```
