---
title: analysis - MA HP Rates

toc: true
reference-location: margin
fig-cap-location: bottom

appendix-style: default
citation-location: document
citation:
  container-title: Switchbox

format:
  html:
    page-layout: full
---

```{r}
library(lubridate)
library(patchwork)
library(tidyverse)

# Switchbox Libraries
source("lib/ggplot/switchbox_theme.R")
source("lib/rates_analysis/heat_pump_rate_funcs.R")

path_rs_db <- "data/ResStock/2022_resstock_amy2018_release_1.1/rs_ma_hp_targeted.db"

# Choose county
county = 'Suffolk'
electric_utility = 'national_grid'
gas_utility = 'nationalgrid'

county_code_rs <- case_when(
  county == 'Suffolk' ~ "G2500250", # Winthrop
  county == 'Worcester' ~ "G2500270" # Leominster
)
```


### --------------------------------------------------------------------------
# County Metadata
### --------------------------------------------------------------------------

## Houshold Heating Types
```{r}
# Make a copy of the housing_units table
con <- DBI::dbConnect(duckdb::duckdb(), path_rs_db)
DBI::dbExecute(con, "CREATE TABLE IF NOT EXISTS housing_units_original AS SELECT * FROM housing_units")
DBI::dbDisconnect(con)


# 1. trim to our counties of interest
#counties_of_interest <- c("G2500250", "G2500270")
con <- DBI::dbConnect(duckdb::duckdb(), path_rs_db)
DBI::dbExecute(con, sprintf("
  CREATE OR REPLACE TABLE housing_units AS
  SELECT *
  FROM housing_units_original
  WHERE \"in.county\" IN ('%s')
", paste(county_code_rs, collapse = "','")))
DBI::dbDisconnect(con)

# 2. Force all gas utilities to be nationalgrid
con <- DBI::dbConnect(duckdb::duckdb(), path_rs_db)
DBI::dbExecute(con, "
  UPDATE housing_units
  SET gas_utility = 'nationalgrid',
      electric_utility = 'national_grid'
")
DBI::dbDisconnect(con)


# 3. add a column to the housing_units table that is the baseline heating type (like in.heating_fuel but splits electricity into resistance and heat pump)
add_baseline_heating_type(path_rs_db)

# 4. add a column to the housing_units table that is the income in dollars
add_income_dollars(path_rs_db)

```

```{r}
# --------------------------------------------------------------------------
# Get some summary counts
# Heating Type Counts
heating_type_counts <- get_housing_units_column_counts("baseline_heating_type", county_code_rs)
print("Heating Type Counts")
print(heating_type_counts)

# Gas Utility Counts
gas_utility_counts <- get_housing_units_column_counts("gas_utility", county_code_rs)
print("Gas Utility Counts")
print(gas_utility_counts)

# Electric Utility Counts
electric_utility_counts <- get_housing_units_column_counts("electric_utility", county_code_rs)
print("Electric Utility Counts")
print(electric_utility_counts)

# City Counts
city_counts <- get_housing_units_column_counts("\"in.city\"", county_code_rs)
print("City Counts")
print(city_counts)

# County Counts
county_counts <- get_housing_units_column_counts("\"in.county\"", county_code_rs)
print("County Counts")
print(county_counts)

# Housing Types Counts
housing_types_counts <- get_housing_units_column_counts("\"in.geometry_building_type_acs\"", county_code_rs)
print("Housing Types Counts")
print(housing_types_counts)
```

## Heating Type Bar Chart
```{r}
con <- DBI::dbConnect(duckdb::duckdb(), path_rs_db)
housing_units <- DBI::dbReadTable(con, "housing_units")
DBI::dbDisconnect(con)

heating_fuel_counts <- housing_units |>
  group_by(baseline_heating_type) |>
  summarize(count = n()) |>
  mutate(percentage = count / sum(count) * 100) |>
  arrange(desc(count)) |>
  # Convert to factor with levels in reverse order for bottom-to-top stacking
  mutate(baseline_heating_type = factor(baseline_heating_type, levels = rev(baseline_heating_type)))

# Then create the plot
ggplot(heating_fuel_counts, aes(x = fct_rev(baseline_heating_type), y = count)) +
  geom_bar(stat = "identity", fill = sb_colors["midnight"], width = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", percentage), y = count + max(count)*0.05),
            size = 10) +
  labs(
    title = "Distribution of Primary Heating Types",
    subtitle = glue::glue("{county} County"),
    x = "Heating Type",
    y = "Number of Buildings"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 14),
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 16),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )

ggsave(
  "/workspaces/reports/src/ma_hp_rates/img/heating_fuel_counts.png",
  width = 3,
  height = 3,
  dpi = 300,
  bg = "white"
)
```

## Income Brackets Bar Chart
```{r}
# con <- DBI::dbConnect(duckdb::duckdb(), path_rs_db)
# housing_units <- DBI::dbReadTable(con, "housing_units")
# DBI::dbDisconnect(con)

# # Create two dataframes filtered by occupants
# income_small_household <- housing_units |>
#   filter(in.occupants <= 3) |>
#   group_by(income_dollars) |>
#   summarize(count = n()) |>
#   mutate(percentage = count / sum(count) * 100) |>
#   filter(!is.na(income_dollars)) |>
#   arrange(desc(income_dollars)) |>
#   mutate(income_dollars = factor(income_dollars, levels = income_dollars))

# income_large_household <- housing_units |>
#   filter(in.occupants >= 4) |>
#   group_by(income_dollars) |>
#   summarize(count = n()) |>
#   mutate(percentage = count / sum(count) * 100) |>
#   filter(!is.na(income_dollars)) |>
#   arrange(desc(income_dollars)) |>
#   mutate(income_dollars = factor(income_dollars, levels = income_dollars))

# # Create the two plots
# p1 <- ggplot(income_small_household, aes(x = fct_rev(income_dollars), y = count)) +
#   geom_bar(stat = "identity", fill = sb_colors["midnight"], width = 0.7) +
#   geom_text(aes(label = sprintf("%.1f%%", percentage), y = count + max(count)*0.05),
#             size = 10) +
#   labs(
#     title = "Income Distribution (1-3 Occupants)",
#     subtitle = glue::glue("{county} County"),
#     x = "Income ($)",
#     y = "Number of Buildings"
#   ) +
#   scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
#   theme(
#     plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
#     plot.subtitle = element_text(hjust = 0.5, size = 14),
#     axis.text.x = element_text(angle = 45, hjust = 1, size = 16),
#     panel.grid.major.x = element_blank(),
#     panel.grid.minor = element_blank()
#   )

# p2 <- ggplot(income_large_household, aes(x = fct_rev(income_dollars), y = count)) +
#   geom_bar(stat = "identity", fill = sb_colors["midnight"], width = 0.7) +
#   geom_text(aes(label = sprintf("%.1f%%", percentage), y = count + max(count)*0.05),
#             size = 10) +
#   labs(
#     title = "Income Distribution (4+ Occupants)",
#     subtitle = glue::glue("{county} County"),
#     x = "Income ($)",
#     y = "Number of Buildings"
#   ) +
#   scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
#   theme(
#     plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
#     plot.subtitle = element_text(hjust = 0.5, size = 14),
#     axis.text.x = element_text(angle = 45, hjust = 1, size = 16),
#     panel.grid.major.x = element_blank(),
#     panel.grid.minor = element_blank()
#   )

# # Combine plots using patchwork
# combined_plot <- p1 / p2

# # Save the combined plot
# ggsave(
#   "/workspaces/reports/src/ma_hp_rates/img/income_by_occupants.png",
#   combined_plot,
#   width = 12,
#   height = 8,
#   dpi = 300,
#   bg = "white"
# )

# # Display the combined plot
# combined_plot





```

## Housing Types Bar Chart
```{r}
con <- DBI::dbConnect(duckdb::duckdb(), path_rs_db)
housing_units <- DBI::dbReadTable(con, "housing_units")
DBI::dbDisconnect(con)

single_family <- c("Single-Family Detached", "Single-Family Attached")
multi_family <- c("2 Unit", "3 or 4 Unit", "5 to 9 Unit", "10 to 19 Unit", "20 to 49 Unit", "50 or more Unit")
order = c(single_family, multi_family)

housing_types_counts <- housing_units |>
  filter(`in.geometry_building_type_acs` %in% order) |>
  group_by(`in.geometry_building_type_acs`) |>
  summarize(count = n()) |>
  arrange(order) |>
  mutate(percentage = count / sum(count) * 100) |>
  mutate(`in.geometry_building_type_acs` = factor(`in.geometry_building_type_acs`))


# First create a new dataframe with housing category
housing_types_grouped <- housing_types_counts |>
  mutate(
    housing_category = case_when(
      `in.geometry_building_type_acs` %in% single_family ~ "Single Family",
      `in.geometry_building_type_acs` %in% multi_family ~ "Multi Family"
    )
  )

ggplot(housing_types_grouped,
       aes(x = housing_category,
           y = count,
           fill = factor(`in.geometry_building_type_acs`, levels = rev(order)))) +
  geom_bar(stat = "identity", position = "stack", width = 0.7) +
  # Add text for each building type segment
  geom_text(data = housing_types_grouped,
            aes(label = `in.geometry_building_type_acs`,
                y = count),
            position = position_stack(vjust = 0.5),  # Center text in each segment
            color = "white",  # Make text white for better visibility
            size = 3) +  # Smaller text size for building types
  # Keep existing percentage labels at the top
  geom_text(data = housing_types_grouped |>
              group_by(housing_category) |>
              summarize(total = sum(count)) |>
              mutate(pct = total/sum(total)*100),
            aes(x = housing_category,
                y = total + max(total)*0.05,
                label = sprintf("%.1f%%", pct)),
            inherit.aes = FALSE,
            size = 10) +
  labs(
    title = "Distribution of Housing Types",
    subtitle = glue::glue("{county} County"),
    x = "Housing Category",
    y = "Number of Buildings",
    fill = "Building Type"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  scale_fill_manual(values = colorRampPalette(c(sb_colors["midnight"], sb_colors["sky"]))(length(order))) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 14),
    axis.text.x = element_text(angle = 0, hjust = 0.5, size = 16),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "right"
  )

ggsave(
  "/workspaces/reports/src/ma_hp_rates/img/housing_types_counts.png",
  width = 3,
  height = 3,
)
```


### --------------------------------------------------------------------------
# Calculate Bills
### --------------------------------------------------------------------------

# Electric Bills
## Electric Delivery Tariffs
```{r}
con <- DBI::dbConnect(duckdb::duckdb(), path_rs_db)
delivery_tariffs_elec <- DBI::dbReadTable(con, "delivery_tariffs_elec")
DBI::dbDisconnect(con)

```

## Electric Supply Rates
```{r}
source("lib/rates_analysis/heat_pump_rate_funcs.R")
# path_to_supply_rates <- "/workspaces/reports/data/utility_rates/electric/ConEd/coned_supply_rates.csv"
# supply_rates <- read_csv(path_to_supply_rates)

con <- DBI::dbConnect(duckdb::duckdb(), path_rs_db)
supply_rates_elec <- DBI::dbReadTable(con, "supply_rates_elec")
DBI::dbDisconnect(con)

# Monthly Supply Rates are used parts of the bill calculation
supply_monthly_2024 <- get_supply_rates_monthly(supply_rates_elec, target_year = 2024, type = "all_hours", zone = "everywhere")

```

### Electric Supply Plots
```{r}
# plot each year as a separate line against 12 months
spaghetti_plot <- plot_supply_rates_12_months(supply_rates_monthly_long = supply_rates_elec,
                  electric_utility = "national_grid",
                  highlight_years = c("2020", "2024"))
print(spaghetti_plot)
```


## Compute Monthly Electric Bills
```{r}
source("lib/rates_analysis/heat_pump_rate_funcs.R")
#monthly_bills_elec_2014 <- get_monthly_bills_elec(path_rs_db, 2014)
#monthly_bills_elec_2018 <- get_monthly_bills_elec(path_rs_db, 2018)
monthly_bills_elec_2024 <- get_monthly_bills_elec(path_rs_db, 2024)

#monthly_bills_elec <- bind_rows(monthly_bills_elec_2014, monthly_bills_elec_2018, monthly_bills_elec_2024)
monthly_bills_elec <- monthly_bills_elec_2024

head(monthly_bills_elec)

annual_elec_bills <- calc_annual_bills_from_monthly(monthly_bills_elec, "electricity")
```


# Gas Bills
## Gas Delivery Tariffs
```{r}
con <- DBI::dbConnect(duckdb::duckdb(), path_rs_db)
delivery_tariffs_gas <- DBI::dbReadTable(con, "delivery_tariffs_gas")
DBI::dbDisconnect(con)
```

## Gas Supply Rates
```{r}
con <- DBI::dbConnect(duckdb::duckdb(), path_rs_db)
supply_rates_gas <- DBI::dbReadTable(con, "supply_rates_gas")
DBI::dbDisconnect(con)

# Print unique rate classes in supply_rates_gas
unique_rate_classes <- unique(supply_rates_gas$rate_class)
print("Unique rate classes in supply_rates_gas:")
print(unique_rate_classes)
```

## Compute Gas Bills
```{r}
source("lib/rates_analysis/heat_pump_rate_funcs.R")

monthly_bills_gas_2014 <- get_monthly_bills_gas(path_rs_db, year = 2014)
monthly_bills_gas_2018 <- get_monthly_bills_gas(path_rs_db, year = 2018)
monthly_bills_gas_2024 <- get_monthly_bills_gas(path_rs_db, year = 2024)

#monthly_bills_gas <- bind_rows(monthly_bills_gas_2014, monthly_bills_gas_2018, monthly_bills_gas_2024)
monthly_bills_gas <- monthly_bills_gas_2024

annual_gas_bills <- calc_annual_bills_from_monthly(monthly_bills_gas, "gas")
```



# Fuel Oil Bills
```{r}
source("lib/rates_analysis/heat_pump_rate_funcs.R")
monthly_bills_fueloil_2024 <- get_monthly_bills_fueloil(path_rs_db, year = 2024)
monthly_bills_fueloil <- monthly_bills_fueloil_2024

annual_fueloil_bills <- calc_annual_bills_from_monthly(monthly_bills_fueloil, "fueloil")
```

# Propane Bills
```{r}
source("lib/rates_analysis/heat_pump_rate_funcs.R")
monthly_bills_propane_2024 <- get_monthly_bills_propane(path_rs_db, year = 2024)
monthly_bills_propane <- monthly_bills_propane_2024

annual_propane_bills <- calc_annual_bills_from_monthly(monthly_bills_propane, "propane")
```

# Annual Bills (All Fuels)
```{r}
source("lib/rates_analysis/heat_pump_rate_funcs.R")
annual_bills_all <- annual_elec_bills |>
  left_join(annual_gas_bills, by = c("bldg_id", "year", "upgrade"), suffix = c("_elec", "_gas"))|>
  left_join(annual_fueloil_bills, by = c("bldg_id", "year", "upgrade"), suffix = c("_gas", "_fueloil"))|>
  left_join(annual_propane_bills, by = c("bldg_id", "year", "upgrade"), suffix = c("_fueloil", "_propane"))|>
  mutate(annual_bill_energy = annual_bill_gas + annual_bill_elec + annual_bill_fueloil + annual_bill_propane) |>
  pivot_longer(cols = c(annual_bill_elec, annual_bill_gas, annual_bill_fueloil, annual_bill_propane, annual_bill_energy), names_to = "bill_type", values_to = "annual_bill")


  # filter down to just 4 cases
  # baseline: upgrade==0, version_elec=="baseline"
  # upgrade with baseline rates: upgrade==4, version_elec=="baseline"
  # upgrade with approved rates: upgrade==4, version_elec=="hpr_approved"
  # upgrade with doer rates: upgrade==4, version_elec=="hpr_doer"
annual_bills <- annual_bills_all |> filter(

    # Baseline Cases
    # Case 0: baseline pre-HP
    (upgrade == 0 & version_elec == "baseline") |
    # Case 7: Whole-home electrification, min-efficiency
    (upgrade == 7 & version_elec == "baseline") |
    # Case 8: Whole-home electrification, high efficiency
    (upgrade == 8 & version_elec == "baseline") |
    # Case 9: Whole-home electrification, high efficiency ​+ basic enclosure package
    (upgrade == 9 & version_elec == "baseline") |

    # Approached HPR Cases
    (upgrade == 0 & version_elec == "hpr_approved") |
    # Case 7: Whole-home electrification, min-efficiency
    (upgrade == 7 & version_elec == "hpr_approved") |
    # Case 8: Whole-home electrification, high efficiency
    (upgrade == 8 & version_elec == "hpr_approved") |
    # Case 9: Whole-home electrification, high efficiency ​+ basic enclosure package
    (upgrade == 9 & version_elec == "hpr_approved")
  ) |>
  filter(bill_type %in% c("annual_bill_elec", "annual_bill_gas", "annual_bill_energy"))

annual_bills <- annual_bills |> filter(year == 2024)

annual_bills_x = annual_bills |> filter(bldg_id == 182)
```

```{r}

# Calculate savings by comparing to baseline version
annual_change <- annual_bills |>
  # First get baseline values to compare against
  filter(upgrade == 0 & version_elec == "baseline") |>
  select(bldg_id, year, electric_utility, bill_type, annual_bill) |>
  rename(baseline_bill = annual_bill) |>
  # Join back to full dataset
  right_join(annual_bills |> filter(upgrade %in% c(4, 7, 8, 9, 10)),
            by = c("bldg_id", "year", "electric_utility", "bill_type")) |>
  # Calculate savings
  mutate(
    annual_bill_change = annual_bill - baseline_bill,
    bill_type = case_when(
      bill_type == "annual_bill_elec" ~ "annual_bill_change_elec",
      bill_type == "annual_bill_gas" ~ "annual_bill_change_gas",
      bill_type == "annual_bill_energy" ~ "annual_bill_change_energy"
    )
  ) |>
  select(-baseline_bill, -annual_bill)
```



# Annual Change Statistics
```{r}
source("lib/rates_analysis/heat_pump_rate_funcs.R")
# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Parameters
year_to_plot <- 2024
baseline_heating_fuel <- "Natural Gas"
electric_utility_option <- "national_grid"
bill_type <- "annual_bill_change_energy"
# ~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

# Get bldg_ids that start with Natural Gas
bldgs_by_heating_fuel <- get_bldgs_by_heating_fuel(path_rs_db, baseline_heating_fuel)

building_types <- c("Single-Family Attached", "Single-Family Detached", "2 Unit", "3 or 4 Unit")
bldgs_by_building_type <- get_bldgs_by_building_type(path_rs_db, building_types)

# Filter to just bldg_ids that start with Natural Gas
annual_change_to_plot <- annual_change |>
  filter(
    !is.na(annual_bill_change) & is.finite(annual_bill_change),
    year == year_to_plot,
    bldg_id %in% bldgs_by_heating_fuel$bldg_id,
    bldg_id %in% bldgs_by_building_type$bldg_id
  )


# --------------------------------------------------------------------------
# Transformations and Summary Statistics
# --------------------------------------------------------------------------
get_scenario_stats <- function(annual_change_to_plot, upgrade, electric_utility_option, bill_type, version_elec) {
  # Filter data for scenario
  annual_change_scenario <- annual_change_to_plot |>
    filter(
      upgrade == !!upgrade,
      if (electric_utility_option != "statewide") electric_utility == !!electric_utility_option else TRUE,
      bill_type == !!bill_type,
      version_elec == !!version_elec
    )

  # Calculate summary statistics
  scenario_stats <- annual_change_scenario |>
    summarise(
      median_change = median(annual_bill_change),
      pct_that_save = mean(annual_bill_change < 0) * 100,
      median_savings = median(annual_bill_change[annual_bill_change < 0]),
      median_loss = median(annual_bill_change[annual_bill_change > 0])
    )

  return(list(
    annual_change_scenario = annual_change_scenario,
    scenario_stats = scenario_stats
  ))
}

# Baseline
# Upgrade 7 Change
upgrade7_baseline_results <- get_scenario_stats(annual_change_to_plot, 7, electric_utility_option, bill_type, 'baseline')
annual_change_upgrade7_baseline <- upgrade7_baseline_results$annual_change_scenario
upgrade7_baseline_stats <- upgrade7_baseline_results$scenario_stats

# Upgrade 8 Change
upgrade8_baseline_results <- get_scenario_stats(annual_change_to_plot, 8, electric_utility_option, bill_type, 'baseline')
annual_change_upgrade8_baseline <- upgrade8_baseline_results$annual_change_scenario
upgrade8_baseline_stats <- upgrade8_baseline_results$scenario_stats

# Upgrade 9 Change
upgrade9_baseline_results <- get_scenario_stats(annual_change_to_plot, 9, electric_utility_option, bill_type, 'baseline')
annual_change_upgrade9_baseline <- upgrade9_baseline_results$annual_change_scenario
upgrade9_baseline_stats <- upgrade9_baseline_results$scenario_stats

# HPR Approved
# Upgrade 7 Change
upgrade7_approved_results <- get_scenario_stats(annual_change_to_plot, 7, electric_utility_option, bill_type, 'hpr_approved')
annual_change_upgrade7_approved <- upgrade7_approved_results$annual_change_scenario
upgrade7_approved_stats <- upgrade7_approved_results$scenario_stats

# Upgrade 8 Change
upgrade8_approved_results <- get_scenario_stats(annual_change_to_plot, 8, electric_utility_option, bill_type, 'hpr_approved')
annual_change_upgrade8_approved <- upgrade8_approved_results$annual_change_scenario
upgrade8_approved_stats <- upgrade8_approved_results$scenario_stats

# Upgrade 9 Change
upgrade9_approved_results <- get_scenario_stats(annual_change_to_plot, 9, electric_utility_option, bill_type, 'hpr_approved')
annual_change_upgrade9_approved <- upgrade9_approved_results$annual_change_scenario
upgrade9_approved_stats <- upgrade9_approved_results$scenario_stats


# Combine stats into single table
stats_combined <- bind_rows(
  upgrade7_baseline_stats |> mutate(upgrade = "Upgrade 7 - Baseline"),
  upgrade8_baseline_stats |> mutate(upgrade = "Upgrade 8 - Baseline"),
  upgrade9_baseline_stats |> mutate(upgrade = "Upgrade 9 - Baseline"),
  upgrade7_approved_stats |> mutate(upgrade = "Upgrade 7 - Approved"),
  upgrade8_approved_stats |> mutate(upgrade = "Upgrade 8 - Approved"),
  upgrade9_approved_stats |> mutate(upgrade = "Upgrade 9 - Approved")
) |>
  pivot_longer(
    cols = c(median_change, pct_that_save, median_savings, median_loss),
    names_to = "metric"
  ) |>
  pivot_wider(
    names_from = upgrade,
    values_from = value
  )
# Round all numeric columns to 2 decimal places
stats_combined <- stats_combined |>
  mutate(across(where(is.numeric), ~round(., 2)))

# Print table
print(stats_combined)

# Save stats_combined to CSV
write.csv(stats_combined, file = glue::glue("/workspaces/reports/src/ma_hp_rates/notebooks/{county}_stats_combined.csv"), row.names = FALSE)
```

## Bar chart by income
```{r}
con <- DBI::dbConnect(duckdb::duckdb(), path_rs_db)
housing_units <- DBI::dbReadTable(con, "housing_units")
DBI::dbDisconnect(con)


# Get bldg_ids where annual_bill_change > 0
upgrade <- 8
version_elec <- "baseline"
bill_type <- "annual_bill_change_energy"

high_bill_bldgs <- annual_change_to_plot |>
  filter(upgrade == !!upgrade,
         version_elec == !!version_elec,
         bill_type == !!bill_type,
         annual_bill_change > 10) |>
  pull(bldg_id)

housing_units_high_bill <- housing_units |>
  filter(bldg_id %in% high_bill_bldgs)

# Create histogram of income distribution for high bill buildings
ggplot(housing_units_high_bill, aes(x = as.numeric(income_dollars))) +
  geom_histogram(binwidth = 10000, fill = sb_colors["midnight"]) +
  labs(
    title = "Income Distribution of Buildings with Bill Increases",
    subtitle = glue::glue("Upgrade {upgrade} | {version_elec} tariffs | {county} County"),
    x = "Annual Income ($)",
    y = "Number of Buildings"
  ) +
  scale_x_continuous(labels = scales::dollar_format()) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

## Bar chart by building type
- stacks by upgrade_that_saves
```{r}
#upgrade <- 9
version_elec <- "hpr_approved"
bill_type <- "annual_bill_change_energy"

bill_changes_with_data <- annual_change_to_plot |>
  filter(bill_type == !!bill_type) |>
  left_join(housing_units, by = "bldg_id") |>
  mutate(income_dollars = as.numeric(income_dollars)) |>
  mutate(package_that_saves = ifelse(annual_bill_change < 0, upgrade, NA))


# Calculate counts for each building type and bill change category
bill_change_counts <- bill_changes_with_data |>
  group_by(bldg_id, version_elec, `in.geometry_building_type_acs`, income_dollars) |>
  summarize(
    first_saving_package = min(package_that_saves, na.rm = TRUE)
  )  |>
  mutate(package_description = case_when(
    first_saving_package == 7 ~ "Minimum Efficiency Appliances",
    first_saving_package == 8 ~ "High Efficiency Appliances",
    first_saving_package == 9 ~ "Weatherization",
    is.na(first_saving_package) ~ "No Savings"
  ))|>
  filter(version_elec == "hpr_approved")


# --------------------------------------------------------------------------
# Print some summary stats
# Calculate summary statistics by group
summary_stats <- bill_changes_with_data |>
  group_by(upgrade, `in.geometry_building_type_acs`, version_elec) |>
  summarize(
    n_buildings = n(),
    pct_with_savings = mean(annual_bill_change < 0, na.rm = TRUE) * 100,
    median_savings = median(annual_bill_change[annual_bill_change < 0], na.rm = TRUE),
    median_increase = median(annual_bill_change[annual_bill_change > 0], na.rm = TRUE)
  ) |>
  ungroup()

# Print formatted results
print("Summary Statistics by Group:")
print("----------------------------")
summary_stats |>
  mutate(
    pct_with_savings = round(pct_with_savings, 1),
    median_savings = round(median_savings, 2),
    median_increase = round(median_increase, 2)
  ) |>
  arrange(upgrade, `in.geometry_building_type_acs`, version_elec) |>
  print(n = Inf)






# --------------------------------------------------------------------------
# Plot
# --------------------------------------------------------------------------
single_family <- c("Single-Family Attached", "Single-Family Detached")
multi_family <- c("2 Unit", "3 or 4 Unit", "5 to 9 Unit", "10 to 19 Unit", "20 to 49 Unit", "50 or more Unit")
order = c(single_family, multi_family)

tariff_aliases <- c(
  "baseline" = "Default (R-1)",
  "hpr_approved" = "Approved (64% Distribution Charge Discount)",
  "hpr_doer" = "DOER Proposal"
)

# Then modify the plot to match exactly
ggplot(bill_change_counts,
       aes(x = factor(`in.geometry_building_type_acs`, levels = order),
           fill = factor(package_description,
                        levels = c("No Savings",  "Weatherization", "High Efficiency Appliances", "Minimum Efficiency Appliances")))) +
  geom_bar(position = "stack") +
  scale_fill_manual(
    values = c("No Savings" = "#023047",
               "Weatherization" = "#87CEEB",
               "High Efficiency Appliances" = "#FF8C00",
               "Minimum Efficiency Appliances" = "#A0AF12"
               ),
    name = "First Package with Savings"
  ) +
  labs(
    title = "Bill Savings by Building Type",
    subtitle = glue::glue("Tariff: {tariff_aliases[version_elec]} | {county} County"),
    x = "Building Type",
    y = "Number of Buildings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 14),
    axis.text.x = element_text(angle = 0, hjust = 0.5)
  )

ggsave(glue::glue("/workspaces/reports/src/ma_hp_rates/img/{county}_bill_savings_by_building_type.png"), width = 6, height = 3, dpi = 300)


```

## Box-and-whisker: Bill change by housing unit type
```{r}
bill_changes_with_data <- annual_change_upgrade9_approved |>
  left_join(housing_units, by = "bldg_id") |>
  mutate(income_dollars = as.numeric(income_dollars))

# Create box plot of bill changes by building type
ggplot(bill_changes_with_data, aes(x = `in.geometry_building_type_acs`, y = annual_bill_change)) +
  geom_boxplot(fill = sb_colors["midnight"], alpha = 0.6) +
  labs(
    title = "Annual Bill Change Distribution by Building Type",
    subtitle = glue::glue("{county} County"),
    x = "Building Type",
    y = "Annual Bill Change ($)"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 14),
    axis.text.x = element_text(angle = 90, hjust = 1)
  ) +
  scale_y_continuous(labels = scales::dollar_format())



```


## Scatter sqft vs bill change
```{r}

# Join annual bill changes with housing unit data
bill_changes_with_data <- annual_change_upgrade8_baseline |>
  left_join(housing_units, by = "bldg_id") |>
  mutate(income_dollars = as.numeric(income_dollars))

# Create scatter plot
ggplot(bill_changes_with_data, aes(x = `in.sqft`, y = annual_bill_change, color = income_dollars)) +
  geom_point(alpha = 0.6) +
  scale_color_viridis_c(labels = scales::dollar_format()) +
  labs(
    title = "Annual Bill Change vs Building Square Footage",
    subtitle = glue::glue("{county} County"),
    x = "Building Square Footage",
    y = "Annual Bill Change ($)",
    color = "Annual Income"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 14),
    legend.position = "right"
  )




```


# Plots
## 3 Hists = Baseline
```{r}
# --------------------------------------------------------------------------
# Plotting
# --------------------------------------------------------------------------
# Set up common parameters
max_change <- 3000
min_change <- -max_change
binwidth <- 100

# --------------------------------------
# y_max calculation

# Create the histograms first
p7 <- annual_change_upgrade7 |>
  ggplot(aes(x = annual_bill_change)) +
  geom_histogram(binwidth = binwidth)

p8 <- annual_change_upgrade8 |>
  ggplot(aes(x = annual_bill_change)) +
  geom_histogram(binwidth = binwidth)

p9 <- annual_change_upgrade9 |>
  ggplot(aes(x = annual_bill_change)) +
  geom_histogram(binwidth = binwidth)

# Get the maximum bin height across all three histograms
y_max <- ceiling(max(
  ggplot_build(p7)$data[[1]]$count,
  ggplot_build(p8)$data[[1]]$count,
  ggplot_build(p9)$data[[1]]$count,
  ggplot_build(p10)$data[[1]]$count
) / 25) * 25

y_limits <- c(0, y_max)
# --------------------------------------


# Create the three plots
p7 <- annual_change_upgrade7 |>
  ggplot(aes(x = annual_bill_change)) +
  geom_histogram(
    binwidth = binwidth, position = "identity",
    fill = sb_colors["midnight"]
  ) +
  geom_vline(
    data = . %>% summarise(median = median(annual_bill_change)),
    aes(xintercept = upgrade7_stats$median_savings),
    color = sb_colors["midnight"],
    linetype = "dashed",
    linewidth = .75,
    alpha = 0.8
  ) +
  # geom_vline(xintercept = 0, color = "grey40", linetype = "solid") +
  labs(title = "Bill Change with Package 7 (Minimum Efficiency All-Electric Appliances)", x = NULL, y = "Homes Count") +
  scale_y_continuous(labels = scales::comma, limits = y_limits) +
  scale_x_continuous(
    labels = scales::dollar_format(),
    limits = c(min_change, max_change),
    breaks = seq(min_change, max_change, by = 1000)
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid.major = element_line(linewidth = 0.2),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

p8 <- annual_change_upgrade8 |>
  ggplot(aes(x = annual_bill_change)) +
  geom_histogram(
    binwidth = binwidth, position = "identity",
    fill = sb_colors["carrot"]
  ) +
  geom_vline(
    data = . %>% summarise(median = median(annual_bill_change)),
    aes(xintercept = upgrade8_stats$median_savings),
    color = sb_colors["carrot"],
    linetype = "dashed",
    linewidth = .75,
    alpha = 0.8
  ) +
  # geom_vline(xintercept = 0, color = "grey40", linetype = "solid") +
  labs(title = "Bill Change with Package 8 (High Efficiency All-Electric Appliances)", x = NULL, y = "Homes Count") +
  scale_y_continuous(labels = scales::comma, limits = y_limits) +
  scale_x_continuous(
    labels = scales::dollar_format(),
    limits = c(min_change, max_change),
    breaks = seq(min_change, max_change, by = 1000)
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid.major = element_line(linewidth = 0.2),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

p9 <- annual_change_upgrade9 |>
  ggplot(aes(x = annual_bill_change)) +
  geom_histogram(
    binwidth = binwidth, position = "identity",
    fill = sb_colors["sky"]
  ) +
  geom_vline(
    data = . %>% summarise(median = median(annual_bill_change)),
    aes(xintercept = upgrade9_stats$median_savings),
    color = sb_colors["sky"],
    linetype = "dashed",
    linewidth = .75,
    alpha = 0.8
  ) +
  # geom_vline(xintercept = 0, color = "grey40", linetype = "solid") +
  labs(title = "Bill Change with Package 9 (High Efficiency Appliances+ Basic Weatherization)", x = "Annual Bill Change ($)", y = "Homes Count") +
  scale_y_continuous(labels = scales::comma, limits = y_limits) +
  scale_x_continuous(
    labels = scales::dollar_format(),
    limits = c(min_change, max_change),
    breaks = seq(min_change, max_change, by = 1000)
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid.major = element_line(linewidth = 0.2),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

p10 <- annual_change_upgrade10 |>
  ggplot(aes(x = annual_bill_change)) +
  geom_histogram(
    binwidth = binwidth, position = "identity",
    fill = sb_colors["pistachio"]
  ) +
  geom_vline(
    data = . %>% summarise(median = median(annual_bill_change)),
    aes(xintercept = upgrade10_stats$median_savings),
    color = sb_colors["pistachio"],
    linetype = "dashed",
    linewidth = .75,
    alpha = 0.8
  ) +
  # geom_vline(xintercept = 0, color = "grey40", linetype = "solid") +
  labs(title = "Bill Change with Package 10 (High Efficiency Appliances + Enhanced Weatherization)", x = "Annual Bill Change ($)", y = "Homes Count") +
  scale_y_continuous(labels = scales::comma, limits = y_limits) +
  scale_x_continuous(
    labels = scales::dollar_format(),
    limits = c(min_change, max_change),
    breaks = seq(min_change, max_change, by = 1000)
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid.major = element_line(linewidth = 0.2),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

# Combine plots vertically using patchwork
p9 / p8 / p7 +
  plot_layout(heights = c(1, 1, 1)) +
  plot_annotation(
    title = case_when(
      bill_type == "annual_bill_change_elec" ~ glue::glue("Switching from {baseline_heating_fuel} to Heat Pump: Change in Annual Electric Bill"),
      bill_type == "annual_bill_change_energy" ~ glue::glue("Switching from {baseline_heating_fuel} to Heat Pump: Change in Annual Total Energy Bill"),
      bill_type == "annual_bill_change_gas" ~ glue::glue("Switching from {baseline_heating_fuel} to Heat Pump: Change in Annual Gas Bill"),
      TRUE ~ "Heat Pump Rate Bill Changes" # Default case
    ),
    subtitle = glue::glue("{county} County  | Electric Utility: {electric_utility_option} | Supply Year: {year_to_plot}"),
    theme = theme(
      plot.margin = margin(3, 3, 3, 3),
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 12)
    )
  )

# Save the plot
ggsave(
  "/workspaces/reports/src/ma_hp_rates/img/annual_change_3hists_baseline.png",
  width = 4,
  height = 5,
  dpi = 300,
  bg = "white"
)
```

## 3 Hists x2 = Baseline + HPR Approved
```{r}
# --------------------------------------------------------------------------
# Plotting
# --------------------------------------------------------------------------
# Set up common parameters
max_change <- 3000
min_change <- -max_change
binwidth <- 100

# --------------------------------------
# y_max calculation

# Create the histograms first
p7<- annual_change_upgrade7_baseline |>
  ggplot(aes(x = annual_bill_change)) +
  geom_histogram(binwidth = binwidth)

p8 <- annual_change_upgrade8_baseline |>
  ggplot(aes(x = annual_bill_change)) +
  geom_histogram(binwidth = binwidth)

p9 <- annual_change_upgrade9_baseline |>
  ggplot(aes(x = annual_bill_change)) +
  geom_histogram(binwidth = binwidth)

# Get the maximum bin height across all three histograms
y_max <- ceiling(max(
  ggplot_build(p7)$data[[1]]$count,
  ggplot_build(p8)$data[[1]]$count,
  ggplot_build(p9)$data[[1]]$count
) / 25) * 25

y_limits <- c(0, y_max)
# --------------------------------------


# Create the three plots
p7 <- annual_change_upgrade7_baseline |>
  ggplot(aes(x = annual_bill_change)) +
  geom_histogram(binwidth = binwidth, position = "identity",
                fill = sb_colors["midnight"], alpha = 0.3) +
  geom_histogram(data = annual_change_upgrade7_approved,
                binwidth = binwidth, position = "identity",
                fill = sb_colors["pistachio"], alpha = 0.7) +
  geom_vline(data = . %>% summarise(median = median(annual_bill_change)),
             aes(xintercept = upgrade7_baseline_stats$median_savings),
             color = sb_colors["pistachio"],
             linetype = "dashed",
             linewidth = .75,
             alpha = 0.8) +
  #geom_vline(xintercept = 0, color = "grey40", linetype = "solid") +

  labs(title = "Bill Change with Package 9 (High Efficiency Appliances+ Basic Weatherization)", x = "Annual Bill Change ($)", y = "Homes Count") +
  scale_y_continuous(labels = scales::comma, limits = y_limits) +
  scale_x_continuous(
    labels = scales::dollar_format(),
    limits = c(min_change, max_change)
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid.major = element_line(linewidth = 0.2),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

p8 <- annual_change_upgrade8_baseline |>
  ggplot(aes(x = annual_bill_change)) +

  geom_histogram(binwidth = binwidth, position = "identity",
                fill = sb_colors["midnight"], alpha = 0.3) +
  geom_histogram(data = annual_change_upgrade8_approved,
                binwidth = binwidth, position = "identity",
                fill = sb_colors["carrot"], alpha = 0.7) +
  geom_vline(data = . %>% summarise(median = median(annual_bill_change)),
             aes(xintercept = upgrade8_baseline_stats$median_savings),
             color = sb_colors["carrot"],
             linetype = "dashed",
             linewidth = .75,
             alpha = 0.8) +
  #geom_vline(xintercept = 0, color = "grey40", linetype = "solid") +
  labs(title = "Bill Change with Package 8 (High Efficiency All-Electric Appliances)", x = NULL, y = "Homes Count") +

  scale_y_continuous(labels = scales::comma, limits = y_limits) +
  scale_x_continuous(
    labels = scales::dollar_format(),
    limits = c(min_change, max_change)
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid.major = element_line(linewidth = 0.2),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

p9 <- annual_change_upgrade9_baseline |>
  ggplot(aes(x = annual_bill_change)) +

  geom_histogram(binwidth = binwidth, position = "identity",
                fill = sb_colors["midnight"], alpha = 0.3) +
  geom_histogram(data = annual_change_upgrade9_approved,
                binwidth = binwidth, position = "identity",
                fill = sb_colors["sky"], alpha = 0.7) +
  geom_vline(data = . %>% summarise(median = median(annual_bill_change)),
             aes(xintercept = upgrade9_baseline_stats$median_savings),
             color = sb_colors["sky"],
             linetype = "dashed",
             linewidth = .75,
             alpha = 0.8) +
  #geom_vline(xintercept = 0, color = "grey40", linetype = "solid") +
  labs(title = "Bill Change with Package 7 (Minimum Efficiency All-Electric Appliances)", x = NULL, y = "Homes Count") +
  scale_y_continuous(labels = scales::comma, limits = y_limits) +
  scale_x_continuous(
    labels = scales::dollar_format(),
    limits = c(min_change, max_change)
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid.major = element_line(linewidth = 0.2),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

# Combine plots vertically using patchwork
p9 / p8 / p7 +
  plot_layout(heights = c(1, 1, 1)) +
  plot_annotation(
    title = case_when(
      bill_type == "annual_bill_change_elec" ~ glue::glue("Switching from {baseline_heating_fuel} to Heat Pump: Change in Annual Electric Bill"),
      bill_type == "annual_bill_change_energy" ~ glue::glue("Switching from {baseline_heating_fuel} to Heat Pump: Change in Annual Total Energy Bill"),
      bill_type == "annual_bill_change_gas" ~ glue::glue("Switching from {baseline_heating_fuel} to Heat Pump: Change in Annual Gas Bill"),
      TRUE ~ "Heat Pump Rate Bill Changes" # Default case
    ),
    subtitle = glue::glue("{county} County  | Electric Utility: {electric_utility_option} | Supply Year: {year_to_plot}"),
    theme = theme(
      plot.margin = margin(3, 3, 3, 3),
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 12)
    )
  )

#Save the plot
ggsave(
  "/workspaces/reports/src/ma_hp_rates/img/annual_change_3hists_baseline.png",
  width = 4,
  height = 5,
  dpi = 300,
  bg = "white"
)# HPR Approved
```


### --------------------------------------------------------------------------
# Sanity Checks
```{r}
#| label: sanity_check_fitchburg_data
# data from https://www.documentcloud.org/documents/25555039-electric-rate-case-impacts-july-1-2024/
rd1 <- data.frame(
  label = paste0("rd1-",c("10%", "20%", "30%", "40%", "50%", "60%", "70%", "80%", "90%", "100%")),
  res_average_monthly_kwh = c(70.1, 162.9, 262.3, 335.3, 410.3, 493.6, 593.3, 715.5, 928.9, 1650.9),
  current_rates_total_rev = c(36.48, 83.91, 117.28, 147.97, 179.50, 214.45, 256.43, 309.60, 397.53, 701.06)
)
rd2 <- data.frame(
  label = paste0("rd2-",c("10%", "20%", "30%", "40%", "50%", "60%", "70%", "80%", "90%", "100%")),
  res_average_monthly_kwh = c(70.1, 182.9, 262.3, 335.3, 410.3, 493.5, 593.3, 724.6, 928.9, 1650.9),
  current_rates_total_rev = c(23.89, 54.95, 76.81, 96.91, 117.57, 140.46, 167.82, 204.91, 260.37, 459.19)
)

fitchburg_data <- bind_rows(rd1, rd2)

```

## Plot Monthly Bills vs Monthly Usage
```{r}
monthly_bills_elec_october <- monthly_bills_elec |>
  filter(electric_utility == "national_grid") |>
  filter(year == 2024) |>
  filter(month %in% c(10))


# Fit linear model
# From "Exhibit HP-3 - Heat Pump Rate Design and Customer Bill Impact - Final", slope should be $0.34/kWh and intercept should be $10
lm_fit <- lm(total_bill ~ elec_consumption, data = monthly_bills_elec_october)
slope <- round(coef(lm_fit)[2], 3)
intercept <- round(coef(lm_fit)[1], 0)

# given points
# Create data frame with reference points
# added reference point from malkani_NearTermRateDesign_2024, pg 9
reference_points <- data.frame(
  x = c(841, 2269, 841, 2269, 1127, 3341, 1127, 3341, 600),
  y = c(296.94, 784.18, 261.55, 688.68, 394.54, 1149.96, 347.10, 1009.34, 225),
  label = c(
    "october_3ton_gas", "january_3ton_gas",
    "october_3ton_hp", "january_3ton_hp",
    "october_5ton_gas", "january_5ton_gas",
    "october_5ton_hp", "january_5ton_hp", "malkani_monthly_avg"
  )
)

# Plot with reference points
monthly_bills_elec_october |>
  ggplot(aes(x = elec_consumption, y = total_bill, color = factor(month))) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  geom_point(data = reference_points, aes(x = x, y = y), color = "red", size = 3) +
  geom_point(data = fitchburg_data, aes(x = res_average_monthly_kwh, y = current_rates_total_rev), color = "blue", size = 3) +
  geom_text(
    data = reference_points,
    aes(x = x, y = y, label = label),
    hjust = -0.1,
    vjust = -0.1,
    color = "black",
    size = 3
  ) +
  geom_text(
    data = fitchburg_data,
    aes(x = res_average_monthly_kwh, y = current_rates_total_rev, label = label),
    hjust = -0.1,
    vjust = -0.1,
    color = "black",
    size = 3
  ) +
  annotate("text",
    x = min(monthly_bills_elec_october$elec_consumption),
    y = max(monthly_bills_elec_october$total_bill),
    label = sprintf("y = %.3fx + %.0f", slope, intercept),
    hjust = 0,
    vjust = 1
  ) +
  labs(
    title = "Monthly Bills vs Monthly Usage",
    x = "Electricity Consumption (kWh)",
    y = "Total Bill ($)"
  )

```
