---
title: analysis - MA HP Rates

toc: true
reference-location: margin
fig-cap-location: bottom

appendix-style: default
citation-location: document
citation:
  container-title: Switchbox

format:
  html:
    page-layout: full
    fig-dpi: 600
execute:
  #fig-width: 9 # Expand default width
  fig-dpi: 600

knitr:
  opts_chunk:
    dev: "ragg_png"
---
# Setup
# Setup
```{r}
#| label: setup
library(lubridate)
library(patchwork)
library(tidyverse)
library(ggrepel)

# Switchbox Libraries
source("/workspaces/reports/lib/ggplot/switchbox_theme.R")
source("/workspaces/reports/lib/rates_analysis/heat_pump_rate_funcs.R")


# Global parameters
exclude_municipal <- TRUE
supply_year = 2024
path_rs_db <- "/workspaces/reports/data/ResStock/2022_resstock_amy2018_release_1.1/rs_ma_hp_rates_2.db"

hp_eff_for_stats <- "mid_hp"
hvac_for_utility_bills <- "mid_hp"
round_dollars <- 0
round_percent <- 1

scenario_colors = c(
  "baseline_rates_current_hvac" = "grey50",

  "baseline" = "#023047",
  "baseline_rates_low_hp" = "#023047",
  "baseline_rates_mid_hp" = "#023047",
  "baseline_rates_high_hp" = "#023047",

  "hpr_approved" = "#FC9706",
  "approved_low_hp" = "#FC9706",
  "approved_mid_hp" = "#FC9706",
  "approved_high_hp" = "#FC9706",

  "hpr_doer" = "#68BED8",
  "doer_low_hp" = "#68BED8",
  "doer_mid_hp" = "#68BED8",
  "doer_high_hp" = "#68BED8"
)

bldg_id_for_debugging = 134260

# Option to pull from cached data
# ---------------------------------
cached_or_recompute = "recompute"
# ---------------------------------

# if (cached_or_recompute == "cached") {
#   load("/workspaces/reports/src/ma_hp_rates/cache/full_workspace.RData")
# }

con <- DBI::dbConnect(duckdb::duckdb(), path_rs_db)
housing_units <- DBI::dbReadTable(con, "housing_units")
DBI::dbDisconnect(con)



```

## Add Columns to Housing Units Table
```{r}
#| label: data-add-housing-units-columns
source("/workspaces/reports/lib/rates_analysis/heat_pump_rate_funcs.R")

# 1. add a column to the housing_units table that is the baseline heating type (basically 'in.heating_fuel' but splits electricity into resistance and heat pump)
# add_baseline_heating_type(path_rs_db)

# 2. add a column to the housing_units table that is the income in dollars
#assign_precise_income_dollars(path_rs_db)

# 3. Add building_type and building_type_group columns to housing_units table
#add_building_type_group(path_rs_db)
```

## Housing Units Summary Counts
```{r}
#| label: data-get-housing-units-counts
# --------------------------------------------------------------------------

# Get some summary counts
county_code_rs <- 'statewide'
display_stats <- TRUE

if (display_stats) {

  con <- DBI::dbConnect(duckdb::duckdb(), path_rs_db)
  housing_units <- DBI::dbReadTable(con, "housing_units")
  DBI::dbDisconnect(con)

  # Heating Type Counts
  heating_type_counts <- get_housing_units_column_counts("baseline_heating_type", county_code_rs)
  print("Heating Type Counts")
  print(heating_type_counts)

  # Gas Utility Counts
  gas_utility_counts <- get_housing_units_column_counts("gas_utility", county_code_rs)
  print("Gas Utility Counts")
    print(gas_utility_counts)

  # Electric Utility Counts
  electric_utility_counts <- get_housing_units_column_counts("electric_utility", county_code_rs)
  print("Electric Utility Counts")
  print(electric_utility_counts)

  # City Counts
  city_counts <- get_housing_units_column_counts("\"in.city\"", county_code_rs)
  print("City Counts")
  print(city_counts)

  # County Counts
  county_counts <- get_housing_units_column_counts("\"in.county\"", county_code_rs)
  print("County Counts")
  print(county_counts)

  # Housing Types Counts
  housing_types_counts <- get_housing_units_column_counts("\"in.geometry_building_type_acs\"", county_code_rs)
  print("Housing Types Counts")
  print(housing_types_counts)
}
```


# Monthly Bills
## Electric Bills
```{r}
#| label: data-electric-bills

source("/workspaces/reports/lib/ggplot/switchbox_theme.R")
source("/workspaces/reports/lib/rates_analysis/heat_pump_rate_funcs.R")

tables <- c("ma_monthly", "housing_units", "delivery_tariffs_elec", "supply_rates_elec")

con <- DBI::dbConnect(duckdb::duckdb(), path_rs_db)
for (table in tables) {
  print(paste0("\nColumns in ", table, ":"))
  print(DBI::dbListFields(con, table))
}
DBI::dbDisconnect(con)


if (cached_or_recompute == "recompute") {
  con <- DBI::dbConnect(duckdb::duckdb(), path_rs_db)
  delivery_tariffs_elec <- DBI::dbReadTable(con, "delivery_tariffs_elec")
  supply_rates_elec <- DBI::dbReadTable(con, "supply_rates_elec")
  DBI::dbDisconnect(con)

  # Get the monthly bills
  monthly_bills_elec <- get_monthly_bills_elec(path_rs_db, supply_year)

  # apply low-income discounts
  url_sheet_low_income_thresholds <- "https://docs.google.com/spreadsheets/d/1QSBYVgaxIVmyT23Xlz_jYe153jaA6XUK2WgPTVphCEk/edit?gid=1688752002#gid=1688752002"
  sheet_name = "low_income_thresholds_elec"

  monthly_bills_elec_discounted <- apply_low_income_discounts(
    path_rs_db,
    url_sheet_low_income_thresholds,
    sheet_name,
    monthly_bills_elec,
    "electricity"
  )

  # ------------------------------------------------------------
  # Annual bills
  full_year_bills_elec <- calc_annual_bills_from_monthly(monthly_bills_elec_discounted, "electricity")

  # ------------------------------------------------------------
  # Heating Season ("Winter") bills
  # add the season column (Winter/Summer); note this is different for each utility
  monthly_bills_elec_discounted <- add_season_column(monthly_bills_elec_discounted)
  winter_bills_elec <- calc_annual_bills_from_monthly(monthly_bills_elec_discounted |> filter(season == "winter"), "electricity")

}

monthly_bills_elec_x <- monthly_bills_elec_discounted |> slice_sample(n = 200)
full_year_bills_elec_x <- full_year_bills_elec |> slice_sample(n = 200)
winter_bills_elec_x <- winter_bills_elec |> slice_sample(n = 200)

```


## Gas Bills
```{r}
#| label: data-gas-bills
source("/workspaces/reports/lib/rates_analysis/heat_pump_rate_funcs.R")

if (cached_or_recompute == "recompute") {
  con <- DBI::dbConnect(duckdb::duckdb(), path_rs_db)
  supply_rates_gas <- DBI::dbReadTable(con, "supply_rates_gas")
  delivery_tariffs_gas <- DBI::dbReadTable(con, "delivery_tariffs_gas")
  DBI::dbDisconnect(con)

  monthly_bills_gas <- get_monthly_bills_gas(path_rs_db, year = supply_year)

  # apply low-income discounts`
  url_sheet_low_income_thresholds <- "https://docs.google.com/spreadsheets/d/1QSBYVgaxIVmyT23Xlz_jYe153jaA6XUK2WgPTVphCEk/edit?gid=1688752002#gid=1688752002"
  sheet_name = "low_income_thresholds_gas"

  monthly_bills_gas_discounted <- apply_low_income_discounts(
    path_rs_db,
    url_sheet_low_income_thresholds,
    sheet_name,
    monthly_bills_gas,
    "gas"
  )


  # ------------------------------------------------------------
  # Annual bills
  full_year_bills_gas <- calc_annual_bills_from_monthly(monthly_bills_gas_discounted, "gas")

  # ------------------------------------------------------------
  # Heating Season ("Winter") bills
  # add the season column (Winter/Summer); note this is different for each utility
  monthly_bills_gas_discounted <- add_season_column(monthly_bills_gas_discounted)
  winter_bills_gas <- calc_annual_bills_from_monthly(monthly_bills_gas_discounted |> filter(season == "winter"), "gas")


}

monthly_bills_gas_x <- monthly_bills_gas_discounted |> slice_sample(n = 200)
monthly_bills_gas_21 <- monthly_bills_gas_discounted |> filter(bldg_id == 21)
full_year_bills_gas_x <- full_year_bills_gas |> slice_sample(n = 200)
winter_bills_gas_x <- winter_bills_gas |> slice_sample(n = 200)
```

## Fuel Oil Bills
```{r}
#| label: data-fuel-oil-bills
source("/workspaces/reports/lib/rates_analysis/heat_pump_rate_funcs.R")

if (cached_or_recompute == "recompute") {
  monthly_bills_fueloil <- get_monthly_bills_fueloil(path_rs_db, year = supply_year)

  # apply low-income discounts
  url_sheet_low_income_thresholds <- "https://docs.google.com/spreadsheets/d/1QSBYVgaxIVmyT23Xlz_jYe153jaA6XUK2WgPTVphCEk/edit?gid=1688752002#gid=1688752002"
  sheet_name = "low_income_thresholds_fueloil"

  monthly_bills_fueloil_discounted <- apply_low_income_discounts(
    path_rs_db,
    url_sheet_low_income_thresholds,
    sheet_name,
    monthly_bills_fueloil,
    "fueloil"
  )

  # ------------------------------------------------------------
  # Annual bills
  full_year_bills_fueloil <- calc_annual_bills_from_monthly(monthly_bills_fueloil_discounted, "fueloil")

  # ------------------------------------------------------------
  # Heating Season ("Winter") bills
  # add the season column (Winter/Summer); note this is different for each utility
  monthly_bills_fueloil_discounted <- add_season_column(monthly_bills_fueloil_discounted)
  winter_bills_fueloil <- calc_annual_bills_from_monthly(monthly_bills_fueloil_discounted |> filter(season == "winter"), "fueloil")

}

monthly_bills_fueloil_x <- monthly_bills_fueloil_discounted |> slice_sample(n = 200)
full_year_bills_fueloil_x <- full_year_bills_fueloil |> slice_sample(n = 200)
winter_bills_fueloil_x <- winter_bills_fueloil |> slice_sample(n = 200)
```

## Propane Bills
```{r}
#| label: data-propane-bills

source("/workspaces/reports/lib/rates_analysis/heat_pump_rate_funcs.R")


if (cached_or_recompute == "recompute") {
  monthly_bills_propane <- get_monthly_bills_propane(path_rs_db, year = supply_year)

  # apply low-income discounts
  url_sheet_low_income_thresholds <- "https://docs.google.com/spreadsheets/d/1QSBYVgaxIVmyT23Xlz_jYe153jaA6XUK2WgPTVphCEk/edit?gid=1688752002#gid=1688752002"
  sheet_name = "low_income_thresholds_propane"

  monthly_bills_propane_discounted <- apply_low_income_discounts(
    path_rs_db,
    url_sheet_low_income_thresholds,
    sheet_name,
    monthly_bills_propane,
    "propane"
  )

  # ------------------------------------------------------------
  # full_year bills
  full_year_bills_propane <- calc_annual_bills_from_monthly(monthly_bills_propane_discounted, "propane")

  # ------------------------------------------------------------
  # Heating Season ("Winter") bills
  # add the season column (Winter/Summer); note this is different for each utility
  monthly_bills_propane_discounted <- add_season_column(monthly_bills_propane_discounted)
  winter_bills_propane <- calc_annual_bills_from_monthly(monthly_bills_propane_discounted |> filter(season == "winter"), "propane")
}

monthly_bills_propane_x <- monthly_bills_propane_discounted |> slice_sample(n = 200)
full_year_bills_propane_x <- full_year_bills_propane |> slice_sample(n = 200)
winter_bills_propane_x <- winter_bills_propane |> slice_sample(n = 200)

```

# Annual Bills by Component
## data-annual-bills-by-component
```{r}
annual_bills_elec_components <- monthly_bills_elec |>
  filter(!(electric_utility == 'municipal')) |>
  group_by(bldg_id, electric_utility, version, hvac) |>
  summarise(annual_elec_supply_charge = sum(supply_charge),
            annual_elec_delivery_charge = sum(delivery_charge),
            annual_elec_customer_charge = sum(customer_charge),
            .groups = "drop") |>
  ungroup()

annual_bills_gas_totals <- monthly_bills_gas_discounted |>
  filter(!is.na(gas_utility)) |>
  select(bldg_id, hvac, monthly_bill) |>
  group_by(bldg_id, hvac) |>
  summarise(annual_gas_bill = sum(monthly_bill)) |>
  ungroup()

annual_bills_by_component_wide <- annual_bills_elec_components |>
  left_join(annual_bills_gas_totals, by = c("bldg_id", "hvac"))


annual_bills_by_component <- annual_bills_by_component_wide |>
  filter(!is.na(annual_gas_bill)) |>
  group_by(version, hvac) |>
  summarise(annual_elec_supply_charge = median(annual_elec_supply_charge),
            annual_elec_delivery_charge = median(annual_elec_delivery_charge),
            annual_elec_customer_charge = median(annual_elec_customer_charge),
            annual_gas_bill = median(annual_gas_bill),
            .groups = "drop") |>
  pivot_longer(cols = starts_with("annual_"),
               names_to = "charge_type",
               values_to = "amount") |>
  mutate(
    hvac = factor(hvac, levels = c("current", "low_hp", "mid_hp", "high_hp")),
    version = factor(version, levels = c("baseline", "hpr_approved", "hpr_doer")),
    #charge_type = factor(charge_type, levels = c("supply_charge", "delivery_charge", "customer_charge"))
  ) |>
  # filter down to four cases using a single filter with OR conditions:
  filter(
    (hvac == "current" & version == "baseline") |
    (hvac == "mid_hp" & version == "baseline") |
    (hvac == "mid_hp" & version == "hpr_approved") |
    (hvac == "mid_hp" & version == "hpr_doer")
  )

# for now, remove the gas bills
annual_bills_by_component <- annual_bills_by_component |>
  filter(charge_type != "annual_gas_bill")

# remove intermediate tables
rm(annual_bills_elec_components, annual_bills_gas_totals, annual_bills_by_component_wide)

```

# Annual
## Annual Bills
```{r}
#| label: data-annual-bills-combined
source("/workspaces/reports/lib/rates_analysis/heat_pump_rate_funcs.R")

winter_or_full_year <- "winter"

if (winter_or_full_year == "winter") {
  bills_elec <- winter_bills_elec
  bills_gas <- winter_bills_gas
  bills_fueloil <- winter_bills_fueloil
  bills_propane <- winter_bills_propane
} else {
  bills_elec <- full_year_bills_elec
  bills_gas <- full_year_bills_gas
  bills_fueloil <- full_year_bills_fueloil
  bills_propane <- full_year_bills_propane
}

if (cached_or_recompute == "recompute") {
  # --------------------------
  annual_bills_all <- bills_elec |>
    left_join(bills_gas, by = c("bldg_id", "year", "hvac", "shell", "appliances", "county", "building_type", "occupants"), suffix = c("_elec", "_gas"))|>
    left_join(bills_fueloil, by = c("bldg_id", "year","hvac", "shell", "appliances", "county", "building_type", "occupants"), suffix = c("_gas", "_fueloil"))|>
    left_join(bills_propane, by = c("bldg_id", "year","hvac", "shell", "appliances", "county", "building_type", "occupants"), suffix = c("_fueloil", "_propane"))|>
    mutate(annual_bill_energy = annual_bill_gas + annual_bill_elec + annual_bill_fueloil + annual_bill_propane,
          annual_bill_raw_energy = annual_bill_raw_gas + annual_bill_raw_elec + annual_bill_raw_fueloil + annual_bill_raw_propane
    ) |>
    pivot_longer(cols = c(annual_bill_elec, annual_bill_gas, annual_bill_fueloil, annual_bill_propane, annual_bill_energy, annual_bill_raw_gas, annual_bill_raw_elec, annual_bill_raw_fueloil, annual_bill_raw_propane, annual_bill_raw_energy), names_to = "bill_type", values_to = "annual_bill") |>
    mutate(raw_or_discounted = ifelse(grepl("raw", bill_type), "raw", "discounted")) |>

    # filter down to just 4 cases
    # ---------------------------
    # baseline: upgrade==0, version_elec=="baseline"
    # upgrade with baseline rates: upgrade==4, version_elec=="baseline"
    # upgrade with approved rates: upgrade==4, version_elec=="hpr_approved"
    # upgrade with doer rates: upgrade==4, version_elec=="hpr_doer"
    filter(
        # Case 1: baseline pre-HP
        (hvac == 'current' & version_elec == "baseline") |

        # Case 2: post-HP with baseline rates
        (hvac == 'low_hp' & version_elec == "baseline") |
        # Case 3: post-HP with approved rates
        (hvac == 'low_hp' & version_elec == "hpr_approved") |
        # Case 4: post-HP with DOER rates
        (hvac == 'low_hp' & version_elec == "hpr_doer") |

        # Case 5: post-HP with baseline rates
        (hvac == 'mid_hp' & version_elec == "baseline") |
        # Case 6: post-HP with approved rates
        (hvac == 'mid_hp' & version_elec == "hpr_approved") |
        # Case 7: post-HP with DOER rates
        (hvac == 'mid_hp' & version_elec == "hpr_doer") |

        # Case 8: post-HP with baseline rates
        (hvac == 'high_hp' & version_elec == "baseline") |
        # Case 9: post-HP with approved rates
        (hvac == 'high_hp' & version_elec == "hpr_approved") |
        # Case 10: post-HP with DOER rates
        (hvac == 'high_hp' & version_elec == "hpr_doer")
      ) |>
      filter(bill_type %in% c("annual_bill_elec", "annual_bill_gas", "annual_bill_fueloil", "annual_bill_propane", "annual_bill_energy", "annual_bill_raw_elec", "annual_bill_raw_gas", "annual_bill_raw_fueloil", "annual_bill_raw_propane", "annual_bill_raw_energy"))


  # --------------------------



  # Energy Burden calculation
  # Add income data, calculate energy burden
  annual_bills_all <- annual_bills_all |>
    left_join(housing_units |> select(bldg_id, assigned_income), by = "bldg_id") |>
    mutate(energy_burden = annual_bill / assigned_income)

  # Add baseline_heating_type
  annual_bills_all <- annual_bills_all |>
    left_join(housing_units |> select(bldg_id, baseline_heating_type), by = "bldg_id")

  # Filter to just supply_year
  annual_bills_20YY <- annual_bills_all |> filter(year == supply_year)

  annual_bills_20YY_x = annual_bills_20YY |> sample_n(100)

  # ---------------------------
  # First table = annual bill WITH LI discounts
  annual_bills <- annual_bills_20YY |> filter(bill_type %in% c("annual_bill_elec", "annual_bill_gas", "annual_bill_fueloil", "annual_bill_propane", "annual_bill_energy"))

  # ---------------------------
  # Second table = annual bill WITHOUT LI discounts
  annual_bills_undiscounted <- annual_bills_20YY |> filter(bill_type %in% c("annual_bill_raw_elec", "annual_bill_raw_gas", "annual_bill_raw_fueloil", "annual_bill_raw_propane", "annual_bill_raw_energy"))

  annual_bills_no_elec_discount <- annual_bills_20YY %>%
  # Filter out municipal utilities if needed
  filter(electric_utility != "municipal") %>%
  # Keep only the relevant bill types
  filter(bill_type %in% c("annual_bill_raw_elec", "annual_bill_gas", "annual_bill_fueloil", "annual_bill_propane")) %>%
  # Group by all identifier columns
  group_by(bldg_id, year, hvac, shell, appliances, county, building_type, occupants,
           electric_utility, version_elec, tariff_id_elec, gas_utility, version_gas,
           tariff_id_gas, assigned_income, baseline_heating_type) %>%
  # Calculate total energy bill
  summarize(
    annual_bill_energy = sum(annual_bill, na.rm = TRUE),
    # Maintain these individual bills if needed
    annual_bill_raw_elec = sum(annual_bill[bill_type == "annual_bill_raw_elec"], na.rm = TRUE),
    annual_bill_gas = sum(annual_bill[bill_type == "annual_bill_gas"], na.rm = TRUE),
    annual_bill_fueloil = sum(annual_bill[bill_type == "annual_bill_fueloil"], na.rm = TRUE),
    annual_bill_propane = sum(annual_bill[bill_type == "annual_bill_propane"], na.rm = TRUE),
    # Keep raw_or_discounted if it's consistent within groups
    raw_or_discounted = "mixed",
    .groups = "drop"
  ) %>%
  # Convert back to long format if needed
  pivot_longer(
    cols = c(annual_bill_energy, annual_bill_raw_elec, annual_bill_gas,
             annual_bill_fueloil, annual_bill_propane),
    names_to = "bill_type",
    values_to = "annual_bill"
  ) %>%
  # Recalculate energy burden if needed
  mutate(energy_burden = case_when(
    !is.na(assigned_income) & assigned_income > 0 ~ annual_bill / assigned_income,
    TRUE ~ NA_real_
  ))

  # --------------------------
  # Cleanup and save
  if (exclude_municipal) {
    annual_bills_with_municipal <- annual_bills
    annual_bills_with_municipal_undiscounted <- annual_bills_undiscounted
    annual_bills_with_municipal_no_elec_discount <- annual_bills_no_elec_discount


    annual_bills <- annual_bills |> filter(electric_utility != "municipal")
    annual_bills_undiscounted <- annual_bills_undiscounted |> filter(electric_utility != "municipal")
    annual_bills_no_elec_discount <- annual_bills_no_elec_discount |> filter(electric_utility != "municipal")
  }

  con <- DBI::dbConnect(duckdb::duckdb(), path_rs_db)
  DBI::dbWriteTable(con, "annual_bills", annual_bills, overwrite = TRUE)
  DBI::dbWriteTable(con, "annual_bills_undiscounted", annual_bills_undiscounted, overwrite = TRUE)
  DBI::dbWriteTable(con, "annual_bills_no_elec_discount", annual_bills_no_elec_discount, overwrite = TRUE)
  DBI::dbDisconnect(con)

}
annual_bills_x = annual_bills |> filter(bldg_id == bldg_id_for_debugging)
annual_bills_undiscounted_x = annual_bills_undiscounted |> filter(bldg_id == bldg_id_for_debugging)
annual_bills_no_elec_discount_x = annual_bills_no_elec_discount |> filter(bldg_id == bldg_id_for_debugging)
```

## Annual Bill Change
```{r}
#| label: data-annual-bill-change
source("/workspaces/reports/lib/rates_analysis/heat_pump_rate_funcs.R")

# Calculate savings by comparing to baseline version
if (cached_or_recompute == "recompute") {
  # --------------------------
  annual_change <- annual_bills |>
    # First get baseline values to compare against
    filter(hvac == 'current' & version_elec == "baseline") |>
    select(bldg_id, year, electric_utility, county, building_type, occupants, bill_type, annual_bill, raw_or_discounted, assigned_income) |>
    rename(baseline_bill = annual_bill) |>
    # Join back to full dataset
    right_join(annual_bills |> filter(hvac %in% c('low_hp', 'mid_hp', 'high_hp')),
              by = c("bldg_id", "year", "electric_utility", "county", "building_type", "occupants", "bill_type", "raw_or_discounted", "assigned_income")) |>
    # Calculate savings
    mutate(
      annual_bill_change = annual_bill - baseline_bill,
      bill_type = case_when(
        bill_type == "annual_bill_elec" ~ "annual_bill_change_elec",
        bill_type == "annual_bill_gas" ~ "annual_bill_change_gas",
        bill_type == "annual_bill_fueloil" ~ "annual_bill_change_fueloil",
        bill_type == "annual_bill_propane" ~ "annual_bill_change_propane",
        bill_type == "annual_bill_energy" ~ "annual_bill_change_energy"
      )
    ) |>
    select(-baseline_bill, -annual_bill)

  con <- DBI::dbConnect(duckdb::duckdb(), path_rs_db)
  DBI::dbWriteTable(con, "annual_change", annual_change, overwrite = TRUE)
  DBI::dbDisconnect(con)

  # --------------------------
  # UNDISCOUNTED annual bill change
  # Calculate savings by comparing to baseline version
  annual_change_undiscounted <- annual_bills_undiscounted |>
    # First get baseline values to compare against
    filter(hvac == 'current' & version_elec == "baseline") |>
    select(bldg_id, year, electric_utility, county, building_type, occupants, bill_type, annual_bill, raw_or_discounted, assigned_income) |>
    rename(baseline_bill = annual_bill) |>
    # Join back to full dataset
    right_join(annual_bills_undiscounted |> filter(hvac %in% c('low_hp', 'mid_hp', 'high_hp')),
              by = c("bldg_id", "year", "electric_utility", "county", "building_type", "occupants", "bill_type", "raw_or_discounted", "assigned_income")) |>
    # Calculate savings
    mutate(
      annual_bill_change = annual_bill - baseline_bill,
      bill_type = case_when(
        bill_type == "annual_bill_raw_elec" ~ "annual_bill_change_raw_elec",
        bill_type == "annual_bill_raw_gas" ~ "annual_bill_change_raw_gas",
        bill_type == "annual_bill_raw_fueloil" ~ "annual_bill_change_raw_fueloil",
        bill_type == "annual_bill_raw_propane" ~ "annual_bill_change_raw_propane",
        bill_type == "annual_bill_raw_energy" ~ "annual_bill_change_raw_energy"
      )
    ) |>
    # only cases where upgrade==4 are interesting...these are the post-HP install cases
    #filter(upgrade == 4) |>
    select(-baseline_bill, -annual_bill)

  con <- DBI::dbConnect(duckdb::duckdb(), path_rs_db)
  DBI::dbWriteTable(con, "annual_change_undiscounted", annual_change_undiscounted, overwrite = TRUE)
  DBI::dbDisconnect(con)

  # --------------------------
  # NO ELECTRIC DISCOUNT annual bill change
  # Calculate savings by comparing to baseline version
  annual_change_no_elec_discount <- annual_bills_no_elec_discount |>
    # First get baseline values to compare against
    filter(hvac == 'current' & version_elec == "baseline") |>
    select(bldg_id, year, electric_utility, county, building_type, occupants, bill_type, annual_bill, raw_or_discounted, assigned_income) |>
    rename(baseline_bill = annual_bill) |>
    # Join back to full dataset
    right_join(annual_bills_no_elec_discount |> filter(hvac %in% c('low_hp', 'mid_hp', 'high_hp')),
              by = c("bldg_id", "year", "electric_utility", "county", "building_type", "occupants", "bill_type", "raw_or_discounted", "assigned_income")) |>
    # Calculate savings
    mutate(
      annual_bill_change = annual_bill - baseline_bill,
      bill_type = case_when(
        bill_type == "annual_bill_elec" ~ "annual_bill_change_elec",
        bill_type == "annual_bill_gas" ~ "annual_bill_change_gas",
        bill_type == "annual_bill_fueloil" ~ "annual_bill_change_fueloil",
        bill_type == "annual_bill_propane" ~ "annual_bill_change_propane",
        bill_type == "annual_bill_energy" ~ "annual_bill_change_energy"
      )
    ) |>
    # only cases where upgrade==4 are interesting...these are the post-HP install cases
    #filter(upgrade == 4) |>
    select(-baseline_bill, -annual_bill)


  # --------------------------
  if (exclude_municipal) {
    annual_change_with_municipal <- annual_change
    annual_change_with_municipal_undiscounted <- annual_change_undiscounted
    annual_change_with_municipal_no_elec_discount <- annual_change_no_elec_discount

    annual_change <- annual_change |> filter(electric_utility != "municipal")
    annual_change_undiscounted <- annual_change_undiscounted |> filter(electric_utility != "municipal")
    annual_change_no_elec_discount <- annual_change_no_elec_discount |> filter(electric_utility != "municipal")
  }

  con <- DBI::dbConnect(duckdb::duckdb(), path_rs_db)
  DBI::dbWriteTable(con, "annual_change_no_elec_discount", annual_change_no_elec_discount, overwrite = TRUE)
  DBI::dbDisconnect(con)
}


annual_change_x = annual_change |> filter(bldg_id == bldg_id_for_debugging)
annual_change_undiscounted_x = annual_change_undiscounted |> filter(bldg_id == bldg_id_for_debugging)
annual_change_no_elec_discount_x = annual_change_no_elec_discount |> filter(bldg_id == bldg_id_for_debugging)
```


## Winter Energy (kWh)
```{r}
#| label: data-annual-energy
#| echo: false
#| output: false


con <- DBI::dbConnect(duckdb::duckdb(), path_rs_db)

# Query annual energy consumption by building and HVAC scenario
winter_months_6 = c(11, 12, 1, 2, 3, 4)
winter_energy_6_months <- tbl(con, "ma_monthly") |>
  filter(month %in% winter_months_6) |>
  group_by(bldg_id, hvac) |>
  summarise(
    total_elec_kwh = sum(total_elec_kwh),
    total_gas_kwh = sum(total_gas_kwh),
    total_propane_kwh = sum(total_propane_kwh),
    total_fuel_oil_kwh = sum(total_fuel_oil_kwh)
  ) |>
  collect()

DBI::dbDisconnect(con)


winter_energy_6_months <- winter_energy_6_months |>
  left_join(housing_units |> select(bldg_id, baseline_heating_type, electric_utility), by = "bldg_id") |>
  filter(electric_utility %in% c("unitil", "nationalgrid"))


winter_months_8 = c(10, 11, 12, 1, 2, 3, 4, 5)
con <- DBI::dbConnect(duckdb::duckdb(), path_rs_db)

winter_energy_8_months <- tbl(con, "ma_monthly") |>
  filter(month %in% winter_months_8) |>
  group_by(bldg_id, hvac) |>
  summarise(
    total_elec_kwh = sum(total_elec_kwh),
    total_gas_kwh = sum(total_gas_kwh),
    total_propane_kwh = sum(total_propane_kwh),
    total_fuel_oil_kwh = sum(total_fuel_oil_kwh)
  ) |>
  collect()

DBI::dbDisconnect(con)

winter_energy_8_months <- winter_energy_8_months |>
  left_join(housing_units |> select(bldg_id, baseline_heating_type, electric_utility), by = "bldg_id") |>
  filter(electric_utility == "eversource")


winter_energy <- bind_rows(winter_energy_6_months, winter_energy_8_months)
rm(winter_energy_6_months, winter_energy_8_months)

# throw out any rows whereh total elec kwh < 0 (net producers)
winter_energy <- winter_energy |>
  filter(total_elec_kwh > 0) |>
  pivot_longer(cols = c(total_elec_kwh, total_gas_kwh, total_propane_kwh, total_fuel_oil_kwh), names_to = "fuel", values_to = "total_winter_kwh") |>
  mutate(
    fuel = case_when(
      fuel == "total_elec_kwh" ~ "electricity",
      fuel == "total_gas_kwh" ~ "gas",
      fuel == "total_propane_kwh" ~ "propane",
      fuel == "total_fuel_oil_kwh" ~ "fuel_oil",
      TRUE ~ "other"
    )
  )

winter_energy_delta <- winter_energy |>
  # Pivot wider to get each hvac scenario in separate columns
  pivot_wider(
    id_cols = c(bldg_id, fuel),
    names_from = hvac,
    values_from = total_winter_kwh
  ) |>
  # Calculate deltas compared to current
  mutate(
    delta_low_hp = low_hp - current,
    delta_mid_hp = mid_hp - current,
    delta_high_hp = high_hp - current
  ) |>
  # Select only relevant columns
  select(bldg_id, fuel, starts_with("delta_")) |>
  # Pivot longer to get back to tidy format
  pivot_longer(
    cols = starts_with("delta_"),
    names_to = "hvac",
    values_to = "delta_winter_kwh"
  ) |>
  # Clean up hvac names by removing "delta_" prefix
  mutate(
    hvac = str_remove(hvac, "delta_")
  ) |>
  # Join back with housing_units to get baseline heating type and utility
  left_join(
    housing_units |>
    select(bldg_id, baseline_heating_type, electric_utility),
    by = "bldg_id"
  )

```
# ------------------------------------------------------------------------------
# Export Data
# ------------------------------------------------------------------------------

## Export just "annual_change_with_municipal"
```{r}
#| label: data-export-annual-change-with-municipal
path_annual_change_db = "/workspaces/reports/data/ma_hp_rates/annual_change.db"

export_annual_change <- FALSE
if (export_annual_change) {
  # Create directory if it doesn't exist
  dir.create(dirname(path_annual_change_db), recursive = TRUE, showWarnings = FALSE)

  # Connect to the database and add the table
  con <- DBI::dbConnect(duckdb::duckdb(), path_annual_change_db, create = TRUE)

  # Write the annual_change_with_municipal table to the database
  DBI::dbWriteTable(con, "annual_change_with_municipal", annual_change_with_municipal, append = TRUE)

  # Disconnect from the database
  DBI::dbDisconnect(con)

  cat("Successfully exported annual_change_with_municipal table to database at:", path_annual_change_db, "\n")
}

```



## Export of "full workspace" and "output_for_analysis"
```{r}
#| label: data-export

save_full_workspace <- FALSE

# Create cache directory if it doesn't exist
dir.create("/workspaces/reports/src/ma_hp_rates/cache/", recursive = TRUE, showWarnings = FALSE)


if (cached_or_recompute == "recompute") {


  if (save_full_workspace) {
    save.image(file = "/workspaces/reports/src/ma_hp_rates/cache/full_calcs_workspace.RData")
  }

  rm(monthly_bills_elec)
  rm(monthly_bills_elec_discounted)
  rm(monthly_bills_gas)
  rm(monthly_bills_gas_discounted)
  rm(monthly_bills_fueloil)
  rm(monthly_bills_fueloil_discounted)
  rm(monthly_bills_propane)
  rm(monthly_bills_propane_discounted)

  rm(bills_elec)
  rm(bills_fueloil)
  rm(bills_gas)
  rm(bills_propane)
  rm(bills_resistance)

  rm(annual_bills_all)
  rm(annual_bills_20YY)
  rm(annual_bills_x)
  rm(annual_bills_heating_oil)
  rm(annual_bills_natural_gas)
  rm(annual_bills_propane)
  rm(annual_bills_raw)

  rm(annual_change_all_fuels)
  rm(annual_change_heating_oil)
  rm(annual_change_natural_gas)
  rm(annual_change_propane)
  rm(annual_change_resistance)

  rm(full_year_bills_elec)
  rm(full_year_bills_fueloil)
  rm(full_year_bills_gas)
  rm(full_year_bills_propane)
  rm(full_year_bills_resistance)

  rm(winter_bills_elec)
  rm(winter_bills_fueloil)
  rm(winter_bills_gas)
  rm(winter_bills_propane)
  rm(winter_bills_resistance)

  save.image(file = "/workspaces/reports/src/ma_hp_rates/cache/output_for_analysis.RData")
}
```
