---
title: eda_muni_rates

toc: true
reference-location: margin
fig-cap-location: bottom

appendix-style: default
citation-location: document
citation:
  container-title: Switchbox

format:
  html:
    page-layout: full
---

# Context

Massachusetts has many Municipal Light Plants (MLPs), which are quasi-public utilities that own and operate local electricity distribution systems.

Unlike for the investor-owned utilities (IOUs) in the state, we only have data on their supply and deliery rates through the end of 2023, not the 2024 rates used in the analysis.

So we need to extrapolate the 2023 rates to 2024, which we do in this notebook by inflating the 2023 rates by the inflation rate.

# Setup

First, we load the libraries and define the paths.
```{r load-libraries}
# Load all required libraries
library(tidyverse)
library(scales)
library(gt)
library(gtExtras)
library(readxl)
library(DBI)
library(duckdb)
library(ggridges)
library(fredr)
library(tidycensus)

# Source additional functions
source("/workspaces/reports/lib/ggplot/switchbox_theme.R")
source("/workspaces/reports/lib/inflation.R")
```

Here, we define the paths to the dataset we'll be using—the E3's MA Residential Electricity Rates Database, and the ResStock database.
```{r define-paths}
# E3's MA Residential Electricity Rates Database - big excel sheet developed by E3, owned by massgov
# https://www.mass.gov/doc/massachusetts-residential-electricity-rates-database
path_ma_rates_xlsx <- "/workspaces/reports/data/utility_rates/electric/madpu/residential_electricity_rates_database_20250520.xlsx"

# Path to ResStock database
path_rs_db <- "/workspaces/reports/data/ResStock/2022_resstock_amy2018_release_1.1/rs_ma_hp_rates_2.db"

```

It's best practice to define all output paths at the top of the notebook.
```{r define-output-path}
# Path for saving MLP delivery tariffs
path_mlp_tariffs_output <- "/workspaces/reports/data/utility_rates/electric/madpu/ma_mlp_delivery_tariffs.rds"

# Path to FRED API key
path_fred_key <- "/workspaces/reports/.secrets/fred.key"
```

# Inflation Adjustment

Next, we fetch the inflation time series data from FRED.
```{r fetch-inflation-data}
# Load the FRED API key
get_eci_inflation_factor <- function(input_year, target_year, api_key) {
  # Set FRED API key
  fredr_set_key(api_key)

  # Employment Cost Index: Wages & salaries for Private industry workers in Middle Atlantic
  ECI <- "CIU2020000000212I"

  # Get inflation index for date range
  start_date <- paste0(input_year, "-01-01")
  end_date <- paste0(target_year, "-01-01")
  wage_index <- get_inflation_index(series = ECI,
                                  start = start_date,
                                  end = end_date,
                                  api_key = api_key)

  # Calculate and return inflation factor
  inflation_adj_factor <- get_inflation_factor(wage_index,
                                             input_year = input_year,
                                             target_year = target_year)
  print(paste0("Inflation factor for ", input_year, " to ", target_year, ": ", inflation_adj_factor))
  return(inflation_adj_factor)
}

# Get inflation factor for 2023 to 2024
FRED_KEY <- file(path_fred_key) |> readLines() # Needs API key

inf_2023_2024 <- get_eci_inflation_factor(2006, 2024, FRED_KEY)

years = seq(2006, 2023)
result_tibble <- tibble(
  year = years,
  inflation_factor = map_dbl(years, ~get_eci_inflation_factor(.x, 2024, FRED_KEY))
)

# Add row for 2024 (inflation factor = 1.0)
result_tibble <- bind_rows(
  result_tibble,
  tibble(year = 2024, inflation_factor = 1.0),
  tibble(year = 2025, inflation_factor = 1.0)
)

file_path <- "/workspaces/reports/data/fred/inflation_factors.rds"
dir.create(dirname(file_path), recursive = TRUE, showWarnings = FALSE)

# Save the file
saveRDS(result_tibble, file_path)

#
inflation_adj_2023_2024 <- get_inflation_factor(wage_index, input_year = 2023, target_year = 2024)
```


```{r}
con <- DBI::dbConnect(duckdb::duckdb(), path_rs_db)
DBI::dbWriteTable(con, "inflation_factors", result_tibble, overwrite = TRUE)
DBI::dbDisconnect(con)
```


# MLP Customer Charge

In this section, we inflate the MLP customer charge from 2023 to 2024.
```{r inflate-mlp-customer-charge}
# Read the E3 rates data
mlp_customer_charge <- readxl::read_xlsx(
  path = path_ma_rates_xlsx,
  sheet = "Summary",
  range = "C259:E454"
)

# Get 2023 values and inflate to 2024
inflation_adj_2023_2024 <- get_inflation_factor(wage_index, input_year = 2023, target_year = 2024)
mlp_customer_charge_2024 <- mlp_customer_charge |>
  filter(Year == 2023) |>
  mutate(
    Value = Value * inflation_adj_2023_2024,
    Year = 2024
  )

# Add the 2024 values back to the original data
mlp_customer_charge <- bind_rows(
  mlp_customer_charge,
  mlp_customer_charge_2024
)

mlp_customer_charge <- mlp_customer_charge |>
  mutate(charge_type = "customer_charge")
```

Let's inspect some summary statistics of the inflated 2024 customer charge we've constructed.
```{r summary-stats-mlp-customer-charge}
# Summary statistics
mlp_customer_charge |>
  filter(Year == 2024) |>
  summarise(
    min = min(`Value`, na.rm = TRUE),
    max = max(`Value`, na.rm = TRUE),
    p10 = quantile(`Value`, probs = 0.10, na.rm = TRUE),
    p25 = quantile(`Value`, probs = 0.25, na.rm = TRUE),
    p50 = quantile(`Value`, probs = 0.50, na.rm = TRUE),
    p75 = quantile(`Value`, probs = 0.75, na.rm = TRUE),
    p90 = quantile(`Value`, probs = 0.90, na.rm = TRUE)
  )
```

Now, let's plot the distribution of the customer charge by year, including both the historical and inflated 2024 values, so we sanity check our work.

```{r plot-mlp-customer-charge}
# Create histogram of delivery charges by year
# Filter to recent years
customer_charge_by_year <- mlp_customer_charge |>
  filter(Year %in% c(2019, 2024)) |>
  mutate(Year = as.factor(Year))

# Create the plot
customer_charge_plot <- customer_charge_by_year |>
  ggplot(aes(x = Value, y = fct_rev(Year), fill = Year)) +
  geom_density_ridges(
    alpha = 0.7,
    rel_min_height = 0.01,
    scale = 1,
    color = NA # Remove the outline
  ) +
  stat_density_ridges(
    quantile_lines = TRUE,
    quantiles = 2,
    alpha = 0,
    color = "black",
    linetype = "dashed",
    scale = 1
  ) +
  geom_text(
    data = customer_charge_by_year |>
      group_by(Year) |>
      summarise(
        median = median(Value, na.rm = TRUE),
        y = as.numeric(fct_rev(Year)) + 0.2
      ),
    aes(
      x = median+0.6,
      y = y,
      label = paste0(Year, ": $", round(median, 2))
    ),
    vjust = -0.5,
    size = 3
  ) +
  scale_x_continuous(
    limits = c(0, max(mlp_customer_charge |> filter(Year == 2024) |> pull(Value))),
    labels = scales::label_dollar()
  ) +
  scale_y_discrete(expand = expansion(mult = c(0, 0.1))) +
  scale_fill_manual(values = c("steelblue", "darkred")) +
  labs(
    title = "Distribution of MLP Customer Charges by Year",
    x = "Customer Charge ($)",
    y = "Year"
  ) +
  theme_minimal()

# Display the plot
customer_charge_plot
```

# MLP Delivery Charge

In this section, we repeat what we did for the MLP customer charge, but for the MLP delivery charge.
```{r inflate-mlp-delivery-charge}
# Read the E3 rates data
mlp_delivery_rate <- readxl::read_xlsx(
  path = path_ma_rates_xlsx,
  sheet = "Summary",
  range = "C459:E653",
  col_names = FALSE,
  col_types = c("text", "numeric", "numeric")
)

names(mlp_delivery_rate) <- c("Utility","Year", "Value")

# Get 2023 values and inflate to 2024
mlp_delivery_rate_2024 <- mlp_delivery_rate |>
  filter(Year == 2023) |>
  mutate(
    Value = Value * inflation_adj_2023_2024,
    Year = 2024
  )

# Add the 2024 values back to the original data
mlp_delivery_rate <- bind_rows(
  mlp_delivery_rate,
  mlp_delivery_rate_2024
)

mlp_delivery_rate <- mlp_delivery_rate |>
  mutate(charge_type = "delivery_rate")
```

Let's inspect some summary statistics of the inflated 2024 delivery charge we've constructed.

```{r summary-stats-mlp-delivery-charge}
stats_mlp_delivery_rate_2019 <- mlp_delivery_rate |>
  filter(Year == 2019) |>
  summarise(
    p10 = quantile(`Value`, probs = 0.10, na.rm = TRUE),
    p25 = quantile(`Value`, probs = 0.25, na.rm = TRUE),
    p50 = quantile(`Value`, probs = 0.50, na.rm = TRUE),
    p75 = quantile(`Value`, probs = 0.75, na.rm = TRUE),
    p90 = quantile(`Value`, probs = 0.90, na.rm = TRUE)
  )

stats_mlp_delivery_rate_2024 <- mlp_delivery_rate |>
  filter(Year == 2024) |>
  summarise(
    p10 = quantile(`Value`, probs = 0.10, na.rm = TRUE),
    p25 = quantile(`Value`, probs = 0.25, na.rm = TRUE),
    p50 = quantile(`Value`, probs = 0.50, na.rm = TRUE),
    p75 = quantile(`Value`, probs = 0.75, na.rm = TRUE),
    p90 = quantile(`Value`, probs = 0.90, na.rm = TRUE)
  )

stats_mlp_delivery_rate_2019
stats_mlp_delivery_rate_2024
```

Here again, let's plot sanity check our work by plotting the distribution of the delivery charge by year, including both the historical and inflated 2024 values.

```{r plot-mlp-delivery-rate}
# Create histogram of delivery charges by year
# Filter to recent years
delivery_by_year <- mlp_delivery_rate |>
  filter(Year %in% c(2019, 2024)) |>
  mutate(Year = as.factor(Year))

# Create the plot
delivery_rate_plot <- delivery_by_year |>
  ggplot(aes(x = Value, y = fct_rev(Year), fill = Year)) +
  geom_density_ridges(
    alpha = 0.7,
    rel_min_height = 0.01,
    scale = 1,
    color = NA # Remove the outline
  ) +
  stat_density_ridges(
    quantile_lines = TRUE,
    quantiles = 2,
    alpha = 0,
    color = "black",
    linetype = "dashed",
    scale = 1
  ) +
  geom_text(
    data = delivery_by_year |>
      group_by(Year) |>
      summarise(
        median = median(Value, na.rm = TRUE),
        y = as.numeric(fct_rev(Year)) + 0.02
      ),
    aes(
      x = median+0.01,
      y = y,
      label = paste0(Year, ": $", round(median, 3))
    ),
    vjust = -0.5,
    size = 3
  ) +
  scale_x_continuous(
    limits = c(0.075, max(mlp_delivery_rate |> filter(Year == 2024) |> pull(Value))),
    labels = scales::label_dollar()
  ) +
  scale_y_discrete(expand = expansion(mult = c(0, 0.01))) +
  scale_fill_manual(values = c("steelblue", "darkred")) +
  labs(
    title = "Distribution of MLP Delivery Charges by Year",
    x = "Delivery Charge ($/kWh)",
    y = "Year"
  ) +
  theme_minimal()

# Display the plot
delivery_rate_plot
```

# Save MLP delivery tariffs

Finally, we combine the delivery and customer charges and save them as an RDS file, for use in our main analysis.

```{r save-mlp-delivery-tariffs}
mlp_delivery_tariffs <- bind_rows(
  mlp_delivery_rate,
  mlp_customer_charge
)

# Save MLP delivery tariffs as RDS file
saveRDS(mlp_delivery_tariffs, path_mlp_tariffs_output)
```
